### 2024-11-13 README.Rmd
add README

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hatchR

<!-- badges: start -->
<!-- badges: end -->

**IN DEVELOPMENT**

The goal of hatchR is to allow users to leverage daily temperature records to predict when a fish is expected to hatch and/or emerge using statistical models

## Installation

You can install the development version of hatchR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("bmait101/hatchR")
```

### 2024-11-14 README.Rmd
update README

@@
 )
+options(tibble.print_min = 5, tibble.print_max = 5)
 ```
@@
 
+## Overview
+
 **IN DEVELOPMENT**
 
-The goal of hatchR is to allow users to leverage daily temperature records to predict when a fish is expected to hatch and/or emerge using statistical models
+hatchR is a tool that allow users to leverage daily temperature records to predict when a fish is expected to hatch and/or emerge using statistical models.
+
+Specifically, hatchR is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
+(2019).
 
@@
 ```
+## Usage
+
+**IN DEVELOPMENT**
+
+## Getting help
+
+If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/bmait101/hatchR/issues).
+For questions and other discussion, please reach out to the package authors: [Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov).

### 2024-11-14 R/plot_check_temp.R
revise function

@@ -1,6 +1,6 @@
 #' Plot imported data
 #'
-#' @param data imported data
+#' @param df data.frame with dates and temperature
 #' @param dates date column
 #' @param temperature temperature column
 #' @param temp_min min temp on plot
@@ -10,21 +10,27 @@
 #' @export
 #'
 #' @examples
-plot_check_temp <- function(data, dates, temperature, temp_min, temp_max){
-  # null values for temp_min and temp_max (for plotting lines)
-  if(missing(temp_min)){
-    temp_min = 0
-  }
-  if(missing(temp_max)){
-    temp_max = 25
-  }
+plot_check_temp <- function(df,
+                            dates,
+                            temperature,
+                            temp_min = 0,
+                            temp_max = 25
+                            ){
+
+  check_dates <- df |>
+    pull({{dates}}) |>
+    is.character()
+
+  if(check == TRUE){
+    stop("Date column is a character vector; covert to timepoint object with lubridate (e.g. `ymd()`.")
+    }
 
   # object holder for dates character check
-  d <- data |> dplyr::pull({{dates}})
+  d <- df |> dplyr::pull({{dates}})
 
   # if dates is string, convert to timepoint via lubridate
   if(is.character(d)){
-    p <- data  |>
+    p <- df  |>
       dplyr::mutate(dates = lubridate::ymd_hms({{dates}})) |>
       ggplot2::ggplot(ggplot2::aes(x = dates, y = {{temperature}})) +
       ggplot2::geom_point(size =0.5) +
@@ -37,7 +43,7 @@ plot_check_temp <- function(data, dates, temperature, temp_min, temp_max){
       ggplot2::theme_classic()
 
   } else{
-    p <- data |>
+    p <- df |>
       ggplot2::ggplot(ggplot2::aes(x = {{dates}}, y = {{temperature}})) +
       ggplot2::geom_point(size =0.5) +
       ggplot2::geom_line(linewidth =0.5) +

### 2024-11-14 R/plot_check_temp.R
updates to plot_temp_check()

@@ -1,59 +1,59 @@
-#' Plot imported data
+#' Plot imported data to check temperature values
 #'
-#' @param df data.frame with dates and temperature
+#' The `plot_check_temp` function is used to plot imported data to check temperature values.
+#' The function takes a data frame with dates and temperature values,
+#' and plots the temperature values over time. The function also allows users
+#' to specify the minimum and maximum temperature values to be plotted.
+#'
+#' @param data data.frame with dates and temperature
 #' @param dates date column
 #' @param temperature temperature column
 #' @param temp_min min temp on plot
 #' @param temp_max max temp on plot
 #'
 #' @return
+#' A object of class "gg" and "ggplot" that can be printed to the console or saved as an image.
+#'
 #' @export
 #'
 #' @examples
-plot_check_temp <- function(df,
+#' library(hatchR)
+#' plot_check_temp(data = RMRS_MT103, dates = SampleDate, temperature = temperature)
+plot_check_temp <- function(data,
                             dates,
                             temperature,
                             temp_min = 0,
                             temp_max = 25
                             ){
 
-  check_dates <- df |>
-    pull({{dates}}) |>
-    is.character()
-
-  if(check == TRUE){
-    stop("Date column is a character vector; covert to timepoint object with lubridate (e.g. `ymd()`.")
+  check_dates <- data |> dplyr::pull({{dates}}) |>is.character()
+  if(check_dates == TRUE){
+    stop(
+      "Date column is not Date or Date-time class; convert to date or date-time (e.g. `lubridate::ymd()`.",
+      call. = FALSE
+      )
     }
 
-  # object holder for dates character check
-  d <- df |> dplyr::pull({{dates}})
+  # d <- data |> dplyr::pull({{dates}})
+  # if(class({{d}})[1] != "Date" | class({{d}})[1] != "POSIXct"){
+  #   stop("Date column is not Date or Date-time class; convert to date or date-time (e.g. `lubridate::ymd()`.")
+  # }
 
-  # if dates is string, convert to timepoint via lubridate
-  if(is.character(d)){
-    p <- df  |>
-      dplyr::mutate(dates = lubridate::ymd_hms({{dates}})) |>
[…]

### 2024-11-14 tests/testthat.R
shut down tests for now

@@ -6,7 +6,7 @@
 # * https://r-pkgs.org/testing-design.html#sec-tests-files-overview
 # * https://testthat.r-lib.org/articles/special-files.html
 
-library(testthat)
-library(hatchR)
-
-test_check("hatchR")
+# library(testthat)
+# library(hatchR)
+#
+# test_check("hatchR")

### 2024-11-14 README.Rmd
update README

@@
 <!-- badges: start -->
+
 <!-- badges: end -->
@@
 
-hatchR is a tool that allow users to leverage daily temperature records to predict when a fish is expected to hatch and/or emerge using statistical models.
-
-Specifically, hatchR is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
+`hatchR` is an R package for predicting fish developmental phenology using statistical models.
+It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
 (2019).
@@
 ```
+
 ## Usage
@@
 **IN DEVELOPMENT**
+
+## Suggested Packages
+
+`hatchR` suggests the following **R** Packages:
+
+* `ggplot2` - for vignette vizualizations.
+* `lubridate` - for date-time manipulation.
+* `knitr`: For vignette building.
+* `rmarkdown`: For vignette building.
+* `testthat`: For unit testing.
 
@@
 For questions and other discussion, please reach out to the package authors: [Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov).
+
+## License
+
+This project is licensed under the GNU General Public License, [GPL-3](https://cran.r-project.org/web/licenses/GPL-3).
+
+## Disclaimer
+
+`hatchR` project code is provided on an "as is" basis and the user assumes responsibility for its use.
+USDA Forest Service has relinquished control of the information and no longer has responsibility to protect the integrity , confidentiality, or availability of the information.
+Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by USDA Forest Service.

### 2024-11-14 R/summarize_temp.R
add summarize_temp()

@@ -1,0 +1,32 @@
+#' Summarize Temperature Data
+#'
+#' The `summarize_temp` function is used to summarize sub-daily temperature
+#' measurements to obtain mean daily temperature.
+#'
+#' @param data
+#' @param dates
+#' @param temperature
+#'
+#' @return
+#' @export
+#'
+#' @examples
+summarize_temp <- function(data, dates, temperature){
+
+  ## check to make sure data are in correct format (check.1 is for datetime object, check.2 is for Date object)
+  check <- data %>%
+    pull({{dates}}) %>%
+    is.character()
+
+
+  if(check == TRUE){stop("Your dates are formatted as a character they need to be formatted as a timepoint object with lubridate (e.g. ymd()" )}
+
+
+  ### if data pass checks, run function
+  sum_dat <- data %>%
+    mutate(ag_date = date({{dates}})) %>% # mutate date-time to date only
+    group_by(ag_date) %>% # group by a unique day
+    summarise(daily_temp = mean({{temperature}})) # summarize to daily mean temp
+
+  return(sum_dat)
+}

### 2024-11-14 R/summarize_temp.R
fix summarize_temp() and documnent

@@ -3,30 +3,35 @@
 #' The `summarize_temp` function is used to summarize sub-daily temperature
 #' measurements to obtain mean daily temperature.
 #'
-#' @param data
-#' @param dates
-#' @param temperature
+#' @param data data.frame with dates and temperature
+#' @param dates date column
+#' @param temperature temperature column
 #'
 #' @return
+#' A data frame with summarized daily temperature values.
+#'
 #' @export
 #'
 #' @examples
-summarize_temp <- function(data, dates, temperature){
+#' library(hatchR)
+#' summarize_temp(data = RMRS_MT103, dates = SampleDate, temperature = temperature)
+summarize_temp <- function(data,
+                           dates,
+                           temperature
+                           ){
 
-  ## check to make sure data are in correct format (check.1 is for datetime object, check.2 is for Date object)
-  check <- data %>%
-    pull({{dates}}) %>%
-    is.character()
+  check_dates <- data |> dplyr::pull({{dates}}) |> is.character()
+  if(check_dates == TRUE){
+    stop(
+      "Date column is not Date or Date-time class; convert to date or date-time (e.g. `lubridate::ymd()`.",
+      call. = FALSE
+    )
+  }
 
-
-  if(check == TRUE){stop("Your dates are formatted as a character they need to be formatted as a timepoint object with lubridate (e.g. ymd()" )}
-
-
-  ### if data pass checks, run function
-  sum_dat <- data %>%
-    mutate(ag_date = date({{dates}})) %>% # mutate date-time to date only
-    group_by(ag_date) %>% # group by a unique day
-    summarise(daily_temp = mean({{temperature}})) # summarize to daily mean temp
+  sum_dat <- data  |>
+    dplyr::mutate(date = lubridate::date({{dates}}))  |>
+    dplyr::group_by(date)  |>
+    dplyr::summarise(daily_temp = mean({{temperature}}))
 
   return(sum_dat)
 }

### 2024-11-14 data-raw/model_table.R
add model table to raw and use it

@@ -1,0 +1,7 @@
+## code to prepare `model_table` dataset goes here
+
+library(readr)
+
+model_table <- read_csv("data-raw/model_table.csv")
+
+usethis::use_data(model_table, overwrite = TRUE)

### 2024-11-18 R/plot_check_temp.R
delete old code

@@ -31,6 +31,7 @@ plot_check_temp <- function(data,
   check_dates <- data |>
     dplyr::pull({{ dates }}) |>
     is.character()
+
   if (check_dates == TRUE) {
     stop(
       "Date column is not Date or Date-time class; convert to date or
@@ -38,12 +39,6 @@ plot_check_temp <- function(data,
       call. = FALSE
     )
   }
-
-  # d <- data |> dplyr::pull({{dates}})
-  # if(class({{d}})[1] != "Date" | class({{d}})[1] != "POSIXct"){
-  #   stop("Date column is not Date or Date-time class; convert to
-  #   date or date-time (e.g. `lubridate::ymd()`.")
-  # }
 
   p <- data |>
     ggplot2::ggplot(ggplot2::aes(x = {{ dates }}, y = {{ temperature }})) +

### 2024-11-18 tests/testthat/test-plot_check_temp.R
add tests for plot_check_temp()

@@ -1,0 +1,13 @@
+test_that("plot_check_temp() returns list", {
+  p <- plot_check_temp(RMRS_MT103, SampleDate, temperature)
+  expect_type(p, "list")
+  expect_s3_class(p, "gg")
+  plot_names <- names(p)
+  expect_identical(plot_names, c("data", "layers", "scales", "guides",
+                                 "mapping", "theme", "coordinates",
+                                 "facet", "plot_env", "layout", "labels"))
+  expect_s3_class(p$data, "data.frame")
+  plot_geoms <- sapply(p$layers, function(x) class(x$geom)[1])
+  expect_identical(plot_geoms, c("GeomPoint", "GeomLine", "GeomHline", "GeomHline"))
+  expect_invisible(plot(p))
+})

### 2024-11-18 tests/testthat.R
open testthat.R

@@ -6,7 +6,7 @@
 # * https://r-pkgs.org/testing-design.html#sec-tests-files-overview
 # * https://testthat.r-lib.org/articles/special-files.html
 
-# library(testthat)
-# library(hatchR)
-#
-# test_check("hatchR")
+library(testthat)
+library(hatchR)
+
+test_check("hatchR")

### 2024-11-19 R/plot_check_temp.R
update fx documentation

@@ -1,15 +1,15 @@
-#' Plot imported data to check temperature values
+#' Visual check of imported data
 #'
 #' The `plot_check_temp` function is used to plot imported data to check temperature values.
 #' The function takes a data frame with dates and temperature values,
 #' and plots the temperature values over time. The function also allows users
 #' to specify the minimum and maximum temperature values to be plotted.
 #'
-#' @param data data.frame with dates and temperature
-#' @param dates date column
-#' @param temperature temperature column
-#' @param temp_min min temp on plot
-#' @param temp_max max temp on plot
+#' @param data A data.frame, or data frame extension (e.g. a tibble).
+#' @param dates Column representing the date of the temperature measurements
+#' @param temperature Column representing the temperature values.
+#' @param temp_min Threshold for lower range of expected temperature. Default is 0.
+#' @param temp_max Threshold for upper range of expected temperature. Default is 25.
 #'
 #' @return
 #' A object of class "gg" and "ggplot" that can be printed to the console or saved as an image.

### 2024-11-19 R/summarize_temp.R

@@ -1,14 +1,14 @@
-#' Summarize Temperature Data
+#' Summarize temperature data to daily values
 #'
 #' The `summarize_temp` function is used to summarize sub-daily temperature
 #' measurements to obtain mean daily temperature.
 #'
-#' @param data data.frame with dates and temperature
-#' @param dates date column
-#' @param temperature temperature column
+#' @param data A data.frame, or data frame extension (e.g. a tibble).
+#' @param dates Column representing the date of temperature measurements.
+#' @param temperature Column representing temperature values.
 #'
 #' @return
-#' A data frame with summarized daily temperature values.
+#' A data.frame with summarized daily temperature values.
 #'
 #' @export
 #'

### 2024-11-19 R/model_select.R
update documentation

@@ -6,12 +6,12 @@
 #' the different function arguments. It pulls the model format as a
 #' string and parses it to be usable in `hatchR` model.
 #'
-#' @param author Source of model.
-#' @param species Species common name.
-#' @param model Model number of type from Beacham and Murray (1990).
-#' @param dev.type Phenology type (e.g., "hatch" or "emerge").
+#' @param author Character string of author name.
+#' @param species Character string of species name.
+#' @param model Model number from Beacham and Murray (1990).
+#' @param dev.type The phenology type. A vector with possible values "hatch" or "emerge". The default is "hatch".
 #'
-#' @return A model function.
+#' @return An expression of length 1 giving the selected model structure to be run with `predict_phenology()`.
 #'
 #' @export
 #'
@@ -30,8 +30,7 @@
 model_select <- function(author,
                          species,
                          model,
-                         dev.type) {
-  # model select using tidyverse (not returning correctly)
+                         dev.type = "hatch") {
   mod <- model_table |>
     dplyr::filter(
       author == {{ author }} &
@@ -41,16 +40,9 @@ model_select <- function(author,
     ) |>
     dplyr::pull("func")
 
-  # base R syntax (needs strings as arguments)
-  # mod <- model_table[which(model_table$author == author &
-  #                            model_table$species == species &
-  #                            model_table$model == model &
-  #                            model_table$dev.type == dev.type), "func"]
-
   mod <- parse(text = mod)
   return(mod)
 }
 
 # TO DO:
-# - use tidyvserse instead of base r
 # - add ID to each model for easily selection

### 2024-11-19 R/plot_phenology.R

@@ -1,10 +1,15 @@
-#' Plot phenology
+#' Visualize the phenology of fish development
 #'
-#' @param plot object output from predict_phenology
-#' @param style different plot types: all, ef_cumsum, ef_daily
-#' @param labels true or false, used to turn on and off the information at the top of plots
+#' @description The function takes the output from `predict_phenology()`
+#'  and creates a ggplot2 object.
+#'
+#' @param plot A list containing the output from `predict_phenology()`
+#' @param style The style of the plot. A vector with possible values "all",
+#'  "ef_cumsum", "ef_daily". The default is "all".
+#' @param labels Logical. If TRUE (default), labels are added to the plot.
 #'
 #' @return ggplot2 object
+#'
 #' @export
 #'
 #' @examples

### 2024-11-19 R/predict_phenology.R

@@ -16,6 +16,14 @@
 #'
 #' @examples
 #' # to come...
+#'
+#' @references
+#' Sparks, M.M., Falke, J.A., Quinn, T.P., Adkinson, M.D., Schindler, D.E.
+#' (2017). Influences of spawning timing, water temperature, and
+#'   climatic warming on early life history phenology in western
+#'   Alaska sockeye salmon.
+#'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
+#'   \bold{76(1)}, 123--135
 predict_phenology <- function(data, dates, temperature, spawn.date, model) {
   dat <- data |> dplyr::arrange({{ dates }})
 

### 2024-11-19 R/predict_phenology.R
update documentation

@@ -11,6 +11,13 @@
 #'
 #' @return
 #' A list with the following elements:
+#' * `ef.vals`: A numeric vector of each day's effective value.
+#' * `days2done`: A numeric vector of length 1; number of predicted days to hatch or emerge.
+#' * `ef.tibble`: An n x 4 tibble (n = number of days to hatch or emerge) with
+#'  the dates, temperature, effective values, and cumulative sum of the effective values.
+#' * `dev.period`: a 1x2 dataframe with the dates corresponding to when your
+#'  fish's parent spawned (input with `predict_phenology(spawn.date = ...)`)
+#'  and the date when the fish is predicted to hatch or emerge.
 #'
 #' @export
 #'

### 2024-11-19 R/hatchR-package.R
update NAMESPACE

@@ -4,12 +4,7 @@
 utils::globalVariables(c("model_table", "woody_island"))
 
 ## usethis namespace: start
-#' @importFrom modelr add_predictions
 #' @importFrom rlang .data
-#' @importFrom stats coef
-#' @importFrom stats lm
-#' @importFrom stats nls
-#' @importFrom stats predict
 #' @importFrom utils globalVariables
 ## usethis namespace: end
 NULL

### 2024-11-19 README.Rmd
update README
update READEME

@@
 
-# hatchR
-
 <!-- badges: start -->
-
 <!-- badges: end -->
 
-## Overview
+# hatchR: Predicting Fish Developmental Phenology
 
@@
 
-## Installation
+## Citation
 
-You can install the development version of hatchR from [GitHub](https://github.com/) with:
+If you use `hatchR` in a formal publication or report, please cite it. Citing `hatchR` lets us devote more resources to it in the future. View the `hatchR` citation by running
+```r
+citation(package = "hatchR")
+```
+
+## Installation Instructions
+
+You can install the development version of `hatchR` from [GitHub](https://github.com/) with:
 
@@
 
-## Usage
+## Contributing to `hatchR`
 
-**IN DEVELOPMENT**
+We encourage users to report bugs and/or contribute to `hatchR`. For more detail on how to do this, please see our contributing guide (`CONTRIBUTING.md`).
+
+## Getting Help
+
+There are several ways to get help with `hatchR`:
+
+1. Open a GitHub issue [link here](https://github.com/bmait101/hatchR/issues).
+2. Email the hatchR support team ([Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov))
+3. Post on a support website like Stack Overflow or Cross Validated.
+
+## Example Usage
+
+To come...
+
+## Imported Packages
+
+`hacthR` imports the following **R** packages:
+
+* dplyr: For data manipulation.
+* ggplot2: For vignette visualizations.
+* lubridate: For working with dates.
+* tibble: For creating tibbles as output for various functions.
+* modelr: For modeling functions.
+* rlang: For `.data` usage.
+* stats: For various modeling functions (e.g., `confint()`).
+* utils: For various utility functions.
 
@@
 
-`hatchR` suggests the following **R** Packages:
+`hatchR` suggests the following **R** packages:
 
-* `ggplot2` - for vignette vizualizations.
-* `lubridate` - for date-time manipulation.
-* `knitr`: For vignette building.
-* `rmarkdown`: For vignette building.
-* `testthat`: For unit testing.
-
-## Getting help
-
-If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/bmait101/hatchR/issues).
-For questions and other discussion, please reach out to the package authors: [Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov).
+* knitr: For vignette building.
+* readr: For reading in data.
+* rmarkdown: For vignette building.
+* testthat: For unit testing.
 
@@
 
-## Disclaimer
+## USDA Disclaimer
 

### 2024-11-20 inst/misc/workflow_diagram.R
add workflow diagram code to inst/misc

@@ -1,0 +1,40 @@
+# workflow_plot <- function(){
+#   library(DiagrammeR)
+#   #see https://rpubs.com/anshulkumar/FlowchartsInR
+#   grViz("digraph {
+#
+#       # a 'graph' statement
+#       graph [overlap = false, label = 'hatchR workflow',labelloc = t, fontsize = 30]
+#
+#       # label nodes
+#       node [fontsize = 20]
+#       import [label = 'import data', shape = circle, fillcolor = Beige]
+#       dates [label = 'check dates', shape = circle, fillcolor = Beige]
+#       plot1 [label = 'plot_check_data()' shape = rectangle]
+#       format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige]
+#       plot2 [label = 'plot_check_data()' shape = rectangle]
+#       summarize [label = 'sumamrize_temp()',  shape = rectangle]
+#       model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige]
+#       select [label = 'model_select()', shape = rectangle]
+#       fit [label = 'fit_model()', shape = rectangle]
+#       phenology [label = 'predict_phenology()', shape = rectangle]
+#       plot3 [label = 'plot_phenology()', shape = rectangle]
+#
+#       # edge definitions with the node IDs
+#       edge [fontsize = 20]
+#       import -> dates
+#       dates -> dates [label = 'not dttm\nuse lubridate']
+#       dates-> plot1 [label = 'is dttm']
+#       plot1 -> format
+#       format -> summarize [label = 'multiple measures\nw/in a day']
+#       format -> model [label = 'mean daily average']
+#       summarize -> plot2 -> model
+#       model -> select [label = 'hatchR built in model']
+#       model -> fit [label = 'user custom data']
+#       {select, fit} -> phenology
+#       phenology -> plot3
+#       }
+#       ")
+# }
+#
+# workflow_plot()

### 2024-11-21 README.Rmd
add hex logo
add use_github_actions("standard")

@@
 
+# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
+
 <!-- badges: start -->
+[![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
 <!-- badges: end -->

### 2024-11-21 R/hatchR-package.R
set up lifecycles

@@ -4,6 +4,7 @@
 utils::globalVariables(c("model_table", "woody_island"))
 
 ## usethis namespace: start
+#' @importFrom lifecycle deprecated
 #' @importFrom rlang .data
 #' @importFrom utils globalVariables
 ## usethis namespace: end

### 2024-11-21 R/fit_model.R
adding lifecycles-experiemental

@@ -1,6 +1,7 @@
 #' Fit B&M model 2 to new data
 #'
 #' @description
+#' `r lifecycle::badge("experimental")`
 #' Generate your own custom parameterized models for predicting hatching and emergence phenology
 #'
 #' @details

### 2024-11-21 R/model_select.R

@@ -1,6 +1,8 @@
 #' Select a development model structure
 #'
-#' @description The function calls a model table with the parameterizations for
+#' @description
+#' `r lifecycle::badge("experimental")`
+#' The function calls a model table with the parameterizations for
 #' different species from different studies built in. Refer to the
 #' table (`model_table`) before using function to find inputs for
 #' the different function arguments. It pulls the model format as a

### 2024-11-21 R/plot_check_temp.R

@@ -1,5 +1,7 @@
 #' Visual check of imported data
 #'
+#' @description
+#' `r lifecycle::badge("experimental")`
 #' The `plot_check_temp` function is used to plot imported data to check temperature values.
 #' The function takes a data frame with dates and temperature values,
 #' and plots the temperature values over time. The function also allows users

### 2024-11-21 R/plot_phenology.R

@@ -1,7 +1,8 @@
 #' Visualize the phenology of fish development
 #'
-#' @description The function takes the output from `predict_phenology()`
-#'  and creates a ggplot2 object.
+#' @description
+#' `r lifecycle::badge("experimental")`
+#' The function takes the output from `predict_phenology()` and creates a ggplot2 object.
 #'
 #' @param plot A list containing the output from `predict_phenology()`
 #' @param style The style of the plot. A vector with possible values "all",

### 2024-11-21 R/predict_phenology.R

@@ -1,6 +1,7 @@
 #' Predict phenology of fish
 #'
 #' @description
+#' #' `r lifecycle::badge("experimental")`
 #' A short description...
 #'
 #' @param data Dataframe with dates and temperature.

### 2024-11-21 R/summarize_temp.R

@@ -1,5 +1,7 @@
 #' Summarize temperature data to daily values
 #'
+#' @description
+#' `r lifecycle::badge("experimental")`
 #' The `summarize_temp` function is used to summarize sub-daily temperature
 #' measurements to obtain mean daily temperature.
 #'

### 2024-11-21 R/predict_phenology.R
fix typo

@@ -1,7 +1,7 @@
 #' Predict phenology of fish
 #'
 #' @description
-#' #' `r lifecycle::badge("experimental")`
+#' `r lifecycle::badge("experimental")`
 #' A short description...
 #'
 #' @param data Dataframe with dates and temperature.

### 2024-11-21 R/plot_check_temp.R
update data checks

@@ -1,4 +1,4 @@
-#' Visual check of imported data
+#' Visual check of imported temperature data
 #'
 #' @description
 #' `r lifecycle::badge("experimental")`
@@ -30,17 +30,21 @@ plot_check_temp <- function(data,
                             temperature,
                             temp_min = 0,
                             temp_max = 25) {
-  check_dates <- data |>
-    dplyr::pull({{ dates }}) |>
-    is.character()
 
-  if (check_dates == TRUE) {
+  check_dates <- data |> dplyr::pull({{ dates }})
+  if (is.character(check_dates) == TRUE) {
     stop(
-      "Date column is not Date or Date-time class; convert to date or
+      "Date column is not Date or Date-time (POSIXct) class; convert to date or
       date-time (e.g. `lubridate::ymd()`.",
       call. = FALSE
     )
   }
+
+  # if date is POSIXct (date-time) class, convert to Date class for plot
+  # if (inherits(check_dates, "POSIXct")) {
+  #   data <- data |>
+  #     dplyr::mutate({{ dates }} := as.Date({{ dates }}))
+  # }
 
   title <- "Temperature Check"
   label_x <- "Date"
@@ -61,13 +65,10 @@ plot_check_temp <- function(data,
     ggplot2::labs(title = title, x = label_x, y = label_y) +
     ggplot2::theme_classic()
     # add x and y axis scales that depend on the data
-    # ggplot2::scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
+    # ggplot2::scale_x_date(date_breaks = "1 month", date_labels = "%b %Y")
     # ggplot2::scale_y_continuous(limits = c(0, 30), breaks = seq(0, 30, 5))
 
   return(p)
 }
 
 # TO DO:
-# - Add a title to the plot
-# - Add a subtitle to the plot
-# - add x and y axis scales that depend on the data

### 2024-11-21 R/plot_check_temp.R
clean up

@@ -70,5 +70,3 @@ plot_check_temp <- function(data,
 
   return(p)
 }
-
-# TO DO:

### 2024-11-21 R/check_continuous.R
add check_continuous

@@ -1,0 +1,24 @@
+#' Check if the dates in a data frame are continuous.
+#'
+#' @param data A data.frame, or data frame extension (e.g. a tibble).
+#' @param dates Column representing the date of the temperature measurements.
+#'
+#' @return
+#' A message indicating if the dates are continuous or if there are breaks.
+#' @export
+#'
+#' @examples
+#' library(hatchR)
+#' check_continuous(crooked_river, date)
+check_continuous <- function(data, dates) {
+  check_out <- data |>
+    dplyr::mutate(diff = c(NA, diff({{ dates }})) == 1) |>
+    with(which(diff == FALSE))
+
+  if (length(check_out) > 0) {
+    message("Breaks at the following rows were found:")
+    return(check_out)
+  } else {
+    message("No breaks were found. All clear!")
+  }
+}

### 2024-11-21 R/plot_check_temp.R
fix typo

@@ -8,7 +8,7 @@
 #' to specify the minimum and maximum temperature values to be plotted.
 #'
 #' @param data A data.frame, or data frame extension (e.g. a tibble).
-#' @param dates Column representing the date of the temperature measurements
+#' @param dates Column representing the date of the temperature measurements.
 #' @param temperature Column representing the temperature values.
 #' @param temp_min Threshold for lower range of expected temperature. Default is 0.
 #' @param temp_max Threshold for upper range of expected temperature. Default is 25.

### 2024-11-22 R/model_select.R
remove to dos

@@ -45,6 +45,3 @@ model_select <- function(author,
   mod <- parse(text = mod)
   return(mod)
 }
-
-# TO DO:
-# - add ID to each model for easily selection

### 2024-11-22 R/fit_model.R
clean and remove modelr dependency

@@ -1,8 +1,8 @@
-#' Fit B&M model 2 to new data
+#' Fit B&M model 2 to new data using `stats::nls()`
 #'
 #' @description
 #' `r lifecycle::badge("experimental")`
-#' Generate your own custom parameterized models for predicting hatching and emergence phenology
+#' Generate your own custom parameterized models for predicting hatching and emergence phenology.
 #'
 #' @details
 #' `hatchR` also includes functionality to generate your own custom
@@ -25,19 +25,18 @@
 #' @examples
 #' library(hatchR)
 #' # vector of temperatures
-#' temperature <- c(2,5,8,11,14)
+#' temperature <- c(2, 5, 8, 11, 14)
 #' # vector of days to hatch
-#' days_to_hatch <- c(194,87,54,35,28)
+#' days_to_hatch <- c(194, 87, 54, 35, 28)
 #' quinn_bt_hatch <- tibble::tibble(temperature, days_to_hatch)
 #' bt_hatch_mod <- fit_model(df = quinn_bt_hatch, temp = temperature, days = days_to_hatch)
-fit_model <- function(df, temp, days)
-{
+fit_model <- function(df, temp, days) {
   # fit linear model to log data
   m1 <- stats::lm(log(days) ~ log(temp), data = df)
   summary(m1)
 
   # estimate starting values for nls
-  m1_a <- exp(stats::coef(m1)[1])  # exponentiate the intercept
+  m1_a <- exp(stats::coef(m1)[1]) # exponentiate the intercept
   m1_b <- stats::coef(m1)[2]
 
   # fit model 2 from Beacham & Murray (1990) to data
@@ -49,13 +48,13 @@ fit_model <- function(df, temp, days)
   m2_b <- stats::coef(m2)[2]
 
   # model expression for hatch function
-  mod <- paste("1 / exp(", m2_a, " - log(x + ", m2_b*-1, "))", sep = "")  # is *-1 correct?
+  mod <- paste("1 / exp(", m2_a, " - log(x + ", m2_b * -1, "))", sep = "") # is *-1 correct?
   mod <- parse(text = mod)
 
   # plot predictions and data --------------------------
 
   grid <- data.frame(temp = seq(min(temp), max(temp), 0.1))
-  grid <- grid |> modelr::add_predictions(m2)  # same as predict(m2, newdata = grid)
+  grid$pred <- stats::predict(m2, newdata = grid)
   p_pred <- df |>
[…]

### 2024-11-22 README.Rmd
update README
clean and remove modelr dependency
add useage to READEME

@@
   collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
   comment = "#>",
@@
 <!-- badges: end -->
-
-# hatchR: Predicting Fish Developmental Phenology
 
@@
 
-To come...
+Below we provide a brief example showing how to use `hatchR`.
+For a thorough introduction to the software, see our introductory vignette [linked here](https://bmait101.github.io/hatchR/articles/Introduction.html).
+For a list of all functions available in `hatchR`, see our function reference [linked here](https://bmait101.github.io/hatchR/reference/index.html).
+
+First we load `hatchR` and helper packages by running
+
+```{r}
+library(hatchR)
+library(ggplot2)  # for additional plotting options
+library(lubridate)  # for working with dates
+```
+
+The example `woody_island` dataset is included in `hatchR` and contains temperature data from Woody Island, Lake Iliamna, Alaska.
+It includes daily water temperature data from 1990-1992.
+We can use `plot_check_temp()` to visually check the data:
+
+```{r}
+p <- plot_check_temp(
+  data = woody_island,
+  dates = date, 
+  temperature = temp_c
+  )
+p
+```
+
+Spawning in this system typically peaks around August 18 and hatching and emergence are done before the start of the following spawning season, so we can predict phenology within a subset of a year.
+
+```{r}
+p + 
+  geom_rect(
+    aes(
+      xmin = ymd("1990-08-18"),  # spawn date
+      xmax = ymd("1991-04-01"),  # approx phenology window end
+      ymin = -10,  # lower bound
+      ymax = 25),  # upper bound
+    fill = "darkgreen",  # color rectangle dark green
+    alpha = 0.01  # make rectangle semi-transparent
+    ) 
+```
+
+Next, select hatch and emerge models we want to use to predict phenology:
+
+```{r}
+sockeye_hatch_mod <- model_select(
+  author = "Beacham and Murray 1990", 
+  species = "sockeye", 
+  model = 2, 
+  dev.type = "hatch"
+  )
+```
+
+Now we can predict phenology for the sockeye salmon using the `predict_phenology()` function:
+
+```{r}
+WI_hatch <- predict_phenology(
+  data = woody_island,
+  dates = date,
+  temperature = temp_c,
+  spawn.date = "1990-08-18",
+  model = sockeye_hatch_mod
+  )
+```
+
+We can check the predicted phenology by running:
+
+```{r}
+# see days to hatch
+WI_hatch$days2done
+# and development period
+WI_hatch$dev.period
+```
+
+Finally, we can visualize the predicted phenology using `plot_phenology()`:
+
+```{r}
+plot_phenology(WI_hatch)
+```
 
@@
 * tibble: For creating tibbles as output for various functions.
-* modelr: For modeling functions.
 * rlang: For `.data` usage.

### 2024-11-26 inst/manuscript/paper.Rmd
adding manuscript folder

---
title: 'hatchR: A toolset to predict hatch and emergence phenology in wild fishes'
tags:
  - R
  - fishes
  - developmental phenology
authors:
  - name: Morgan M. Sparks
    orcid: 0000-0003-0787-2218
    affiliation: 1 # (Multiple affiliations must be quoted)
  - name: Eli Felts
    orcid: 0000-0002-5888-9266
    affiliation: 2
  - name: Bryan M. Maitland
    orcid: 0000-0002-4491-5064
    affiliation: 1
affiliations:
 - name: US Forest Service, Rocky Mountain Research Station, Boise, ID, USA
   index: 1
 - name: US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
   index: 2
citation_author: Sparks et al.
date: 07 July 2025
year: 2025
output: rticles::joss_article
csl: apa.csl
journal: JOSS
editor_options:
  chunk_output_type: console
bibliography: ms_refs.bib
---

```{r, include=FALSE}
library(tidyverse)
library(hatchR)
```

# Abstract

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for different species, being largely limited to salmonids. Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. `hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. `hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. `hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.

# Introduction

As primarily poikilothermic organisms, the development and growth of fishes is tightly linked with the temperature of their ambient environment. This close relationship has allowed researchers to generate statistical models that allow the prediction of developmental phenology with high accuracy and precision. These models were typically developed in aquaculture settings and their initial formulations were not applicable to wild populations because they assumed a constant temperature over the course of development @beacham1990 (**add more!**) . However, @sparks2019 reformulated this approach as an "Effective Value model", in which the input was daily average temperature after a parent spawned and fish would either hatch or emerge when effective values cumulatively summed to one.

The resulting effective value approach has now been widely applied in salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 were have been applied to various species and populaions [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`. Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for `hatchR`. The first is a R package distributed via CRAN that allows users the most customizable application of these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on `hatchR`'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

# Package Overview

`hatchR` is meant to primarily be a tool for predicting phenology. In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.

```{r echo=FALSE, eval=FALSE}

  library(DiagrammeR)
  #see https://rpubs.com/anshulkumar/FlowchartsInR
  grViz("digraph {

      # a 'graph' statement
      graph [overlap = false, label = 'hatchR workflow',labelloc = t, fontsize = 30]

      # label nodes
      node [fontsize = 20]
      import [label = 'import data', shape = circle, fillcolor = Beige]
      dates [label = 'check dates', shape = circle, fillcolor = Beige]
      plot1 [label = 'plot_check_data()' shape = rectangle]
      format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige]
      plot2 [label = 'plot_check_data()' shape = rectangle]
      summarize [label = 'sumamrize_temp()',  shape = rectangle]
      continuous [label = 'check_continuous()', shape = rectangle]
      model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige]
      select [label = 'model_select()', shape = rectangle]
      fit [label = 'fit_model()', shape = rectangle]
      phenology [label = 'predict_phenology()', shape = rectangle]
      plot3 [label = 'plot_phenology()', shape = rectangle]

      # edge definitions with the node IDs
      edge [fontsize = 20]
      import -> dates
      dates -> dates [label = 'not dttm\nuse lubridate']
      dates-> plot1 [label = 'is dttm']
      plot1 -> format
      format -> summarize [label = 'multiple measures\nw/in a day']
      format -> model [label = 'mean daily average']
      summarize -> continuous -> plot2 -> model
      model -> select [label = 'hatchR built-in\nmodel']
      model -> fit [label = 'user custom data']
      {select, fit} -> phenology
      phenology -> plot3
      }
      ", width = 500)

```

## Effective value models

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for change arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

The model follows the general format of:

$$
Effective Value_i = 1/exp(log_ea - log_e(Temperature_i - b))
$$

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum or $$\sum_{i =1}^nEffectiveValue_i = 1$$

The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

## Data format

## Checking Data

`hatchR` is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.

summarize_temp, check_continuous and plot_check_temp

## Model Selection

talk about using nls for fit_model

model_table and fit_model

fit_model for three non-salmonid species

```{r, eval=FALSE}
smallmouth <- matrix(NA, 10, 2) |> tibble()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

plot(smallmouth$days~smallmouth$temp_C)

smb_mod <- fit_model(smallmouth,
                     temp = smallmouth$temp_C,
                     days = smallmouth$hours)

```

## Predicting Phenology and Output

Use woody \_example from website

show predict_phenology output slots

show plot_phenology

# Case Study 1

A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action. For instance, will have fish have emerged from redds when a stream section has been opened to grazing or bridge decommissioning will commence?

In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the united states under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed. The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

```{r}

### read in EFS data

EFS_data <- crooked_river

### view bull trout models and select model

model_table |> 
  filter(species == "bull trout")

# we care about when fish are out of gravel, so select emerge mod
bt_emerge_mod <- model_select(author = "Austin et al. 2017",
                            species = "bull trout",
                            model = "MM",
                            dev.type = "emerge")

# predict emergence time using Sept. 30 as the spawn date 
bt_emergence <- predict_phenology(data = EFS_data,
                                  dates = date,
                                  temperature = temp_c,
                                  spawn.date = "2011-09-30", 
                                  model = bt_emerge_mod)

# fish emerge May 1, before June 1
bt_emergence$dev.period$stop

```

In this example we expect the last fish out of the gravel well before the June 1st date and the manager could allow grazing in this area without worrying about direct mechanical disturbance to fish developing in the gravel.

# Case Study 2

# Discussion

talk about how these models also represent local adaptation and heritable plasticity

talk about how these models represent point estimates and that emergence and hatch will take the form of a distribution

# Acknowledgements

We would like to sincerely thank the editor and reviewers for all of their helpful feedback which greatly improved both the software and the manuscript.

The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS. Any mention of trade names, products, or services does not imply an endorsement by the U.S. government, USFS, or USFWS. USFS and USFWS do not endorse any commercial products, services or enterprises.

# References

### 2024-11-26 inst/manuscript/paper.tex
adding manuscript folder

\documentclass[12pt,halfline,a4paper,]{ouparticle}

% Packages I think are necessary for basic Rmarkdown functionality
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage{framed}

% Link coloring
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={hatchR: A toolset to predict hatch and emergence phenology in wild fishes}
            }

%% To allow better options for figure placement
%\usepackage{float}

% Packages that are supposedly required by OUP sty file
\usepackage{amssymb, amsmath, geometry, amsfonts, verbatim, endnotes, setspace}

% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

% Macros for dealing with affiliations, footnotes, etc.
\makeatletter
\def\Newlabel#1#2#3{\expandafter\gdef\csname #1@#2\endcsname{#3}}

\def\Ref#1#2{\@ifundefined{#1@#2}{???}{\csname #1@#2\endcsname}}

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}

\newcommand*\ifcounter[1]{%
  \ifcsname c@#1\endcsname
    \expandafter\@firstoftwo
  \else
    \expandafter\@secondoftwo
  \fi
}

\newcommand*\thanksbycode[1]{%
  \ifcounter{FNCT@#1}
    {\samethanks[\value{FNCT@#1}]}
    {\thanks{\Ref{FN}{#1}}\newcounter{FNCT@#1}\setcounter{FNCT@#1}{\value{footnote}}}
}

% Create labels for Addresses if the are given in Elsevier format

% Create labels for Footnotes if the are given in Elsevier format

% Part for setting citation format package: natbib

% Part for setting citation format package: biblatex

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
%From Pandoc 3.1.8
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\begin{document}

\title{hatchR: A toolset to predict hatch and emergence phenology in
wild fishes}

\author{%
%
% Code for old style authors field
\name{Morgan M. Sparks}
\and
\name{Eli Felts}
\and
\name{Bryan M. Maitland}
%
% Add \and if both authors and author
%
%
% Code for new (elsevier) style author field
%
}

\abstract{}

\date{07 July 2025}

\keywords{}

\maketitle

\section{Abstract}\label{abstract}

Understanding the timing of key life history events is necessary for
managing and conserving populations. Historically, models to predict
hatch and emergence timing for fishes were difficult to employ in wild
settings because average incubation temperature was needed as the
primary parameter in predictive models. However, recent improvements to
these techniques reworked models such that they could be applied in wild
environments as long as users had data for when adult fish spawned and a
record of average daily temperature over the course of development.
Despite these improvements, their application remains limited due to few
parameterizations for different species, being largely limited to
salmonids. Here we present \texttt{hatchR}, a software ecosystem that
allows users to predict hatch and emergence timing for wild fishes, as
well as additional tools to aid in those analyses. \texttt{hatchR}
allows users to leverage popular historic parameterizations for
phenological models or to easily implement custom parameterizations
using data not included in the package. \texttt{hatchR} is also
distributed in two forms---an open source R package for maximum
customization, as well as an HTML graphical-user-interface web
application for individuals not familiar with scripting languages. To
demonstrate potential uses, we present two case studies as likely
applications for this software. \texttt{hatchR} promises to open many
exciting avenues in research and management of fishes during their early
life history.

\section{Introduction}\label{introduction}

As primarily poikilothermic organisms, the development and growth of
fishes is tightly linked with the temperature of their ambient
environment. This close relationship has allowed researchers to generate
statistical models that allow the prediction of developmental phenology
with high accuracy and precision. These models were typically developed
in aquaculture settings and their initial formulations were not
applicable to wild populations because they assumed a constant
temperature over the course of development Beacham \& Murray (1990)
(\textbf{add more!}) . However, Sparks et al. (2019) reformulated this
approach as an ``Effective Value model'', in which the input was daily
average temperature after a parent spawned and fish would either hatch
or emerge when effective values cumulatively summed to one.

The resulting effective value approach has now been widely applied in
salmonids for which parameterizations from aquaculture were readily
available---for example Pacific Salmon (\emph{Oncorhynchus spp.}) models
developed by Beacham \& Murray (1990) were have been applied to various
species and populaions (Adelfio, Wondzell, Mantua, \& Reeves, 2019,
2024; Kaylor et al., 2021) while models developed for Bull Trout
(\emph{Salvelinus confluentus}) by McPhail \& Murray (1979) were
extended by Austin, Essington, \& Quinn (2019). Despite growing
popularity, applications have been largely limited within Salmonids,
presumably because parameterizations for such models already existed due
to their wide use in aquaculture and their general popularity as sport
and commercial fish.

To bridge the gap between the application of one-off effective value
model applications within individual studies and the lack of
parameterization for other species, we developed the software ecosystem,
\texttt{hatchR}. Specifically, \texttt{hatchR} allows users to input
standard raw or summarized temperature datasets that are commonly
collected in wild settings, run basic checks on those data, use built-in
parameterizations like those from Beacham \& Murray (1990) or Sparks,
Westley, Falke, \& Quinn (2017), develop custom models from their own
temperature and phenological data, and predict hatch and emergence
timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two
user-interfaces for \texttt{hatchR}. The first is a R package
distributed via CRAN that allows users the most customizable application
of these methods. The R package is especially powerful as it allows
users to automate their analyses over multiple variables such as
phenology type, multiple spawn dates, or different habitats with varying
thermal regimes. These variable approaches are outlined in the package
documentation on \texttt{hatchR}'s website. Alternatively, we also
distribute a Shiny application in the form of an HTML-based web tool to
interact with many of \texttt{hatchR}'s functions in a
graphical-user-interface. The Shiny form trades-off some of automative
power for user simplicity, while still allowing users to leverage much
of the functionality of \texttt{hatchR}'s R package. Below, we present
the basic overview of the software and multiple case studies of how it
may be applied.

\section{Package Overview}\label{package-overview}

\texttt{hatchR} is meant to primarily be a tool for predicting
phenology. In this sense, we mostly limit functionality to these
applications and provide minimal data checking and plotting help. This
decision is in part driven by the diversity of data types that users may
import and the difficulty in addressing all those data types with
respect to various data checks. In other words, we expect users to know
their data better than we do and to check it accordingly. We do provide
two basic data check functions discussed in the Checking Data section.
Similarly, we provide limited functionality for plotting results, but
provide examples of how to build custom visualization from output,
specifically in R. For the Shiny application, we provide a base output
plot, but the ability to download your results for custom plotting in
programs of the user's choice.

\subsection{Effective value models}\label{effective-value-models}

Effective value models were created by Sparks et al. (2019) to implement
developmental models in wild environments for Sockeye Salmon (\emph{O.
nerka}). The need for change arose because historic models, specifically
those in Beacham \& Murray (1990), only considered the average
incubation temperature during development and, for wild fishes, average
incubation temperature was impossible to estimate because it was unknown
when fish hatched even if adult spawn timing was known. To address this,
Sparks et al. (2019) used the reciprocal of the formulation of model 2
from Beacham \& Murray (1990) and assigned an effective value for every
day of development using the daily average temperature.

The model follows the general format of:

\[
Effective Value_i = 1/exp(log_ea - log_e(Temperature_i - b))
\]

Where \emph{i} is the daily value and a fish hatches or emerges when the
cumulative sum or \[\sum_{i =1}^nEffectiveValue_i = 1\]

The effective value model framework is the basis for the phenological
models in \texttt{hatchR}, both in the included \texttt{model\_table} in
the package, as well as for custom models users can fit with
\texttt{fit\_model()}. Specifically, \texttt{model\_table} has been
extended to include more parameterizations from Beacham \& Murray
(1990), Sparks et al. (2017), and Austin et al. (2019) (who extended
McPhail \& Murray (1979)).

\subsection{Data format}\label{data-format}

\subsection{Checking Data}\label{checking-data}

\texttt{hatchR} is built assuming data will be analyzed as daily average
temperatures. Despite that assumption, raw data (\emph{e.g.}, as
outputted by HOBO loggers) can be used and \texttt{hatchR} includes
functionality to summarize those data into a format that is usable, as
well as provides functions for basic visual and programmatic data checks
to make sure outliers or missing data are at least brought to users'
attention.

summarize\_temp, check\_continuous and plot\_check\_temp

\subsection{Model Selection}\label{model-selection}

talk about using nls for fit\_model

model\_table and fit\_model

fit\_model for three non-salmonid species

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{smallmouth }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{tibble}\NormalTok{()}
\FunctionTok{colnames}\NormalTok{(smallmouth) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"hours"}\NormalTok{, }\StringTok{"temp\_F"}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{hours }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{52}\NormalTok{, }\DecValTok{54}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{78}\NormalTok{, }\DecValTok{90}\NormalTok{, }\DecValTok{98}\NormalTok{, }\DecValTok{150}\NormalTok{, }\DecValTok{167}\NormalTok{, }\DecValTok{238}\NormalTok{, }\DecValTok{234}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_F }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{77}\NormalTok{, }\DecValTok{75}\NormalTok{, }\DecValTok{71}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{67}\NormalTok{, }\DecValTok{65}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{59}\NormalTok{, }\DecValTok{55}\NormalTok{, }\DecValTok{55}\NormalTok{)}

\NormalTok{smallmouth }\OtherTok{\textless{}{-}}\NormalTok{ smallmouth }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{days =} \FunctionTok{ceiling}\NormalTok{(hours}\SpecialCharTok{/}\DecValTok{24}\NormalTok{),}
         \AttributeTok{temp\_C =}\NormalTok{ (temp\_F }\SpecialCharTok{{-}}\DecValTok{32}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ (}\DecValTok{5}\SpecialCharTok{/}\DecValTok{9}\NormalTok{))}

\FunctionTok{plot}\NormalTok{(smallmouth}\SpecialCharTok{$}\NormalTok{days}\SpecialCharTok{\textasciitilde{}}\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C)}

\NormalTok{smb\_mod }\OtherTok{\textless{}{-}} \FunctionTok{fit\_model}\NormalTok{(smallmouth,}
                     \AttributeTok{temp =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C,}
                     \AttributeTok{days =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{hours)}
\end{Highlighting}
\end{Shaded}

\subsection{Predicting Phenology and
Output}\label{predicting-phenology-and-output}

Use woody \_example from website

show predict\_phenology output slots

show plot\_phenology

\section{Case Study 1}\label{case-study-1}

A common management scenario where developmental phenology might be
useful would be trying to understand if fish might be free-moving before
some management action. For instance, will have fish have emerged from
redds when a stream section has been opened to grazing or bridge
decommissioning will commence?

In this scenario, we will consider the grazing example and Bull Trout, a
threatened fish in the united states under the Endangered Species Act
(Nolfi, Melbihess, Fisher, \& Ellis, 2024), and the East Fork Salmon
River, a key Bull Trout population in the upper Salmon River watershed.
The fisheries manager there wants to know if fish will likely be out of
the gravel and free-swimming by June 1st. In this system, it is expected
that Bull Trout will be done spawning by the end of September, so we'll
consider the last possible spawn date as September 30th.

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\#\# read in EFS data}

\NormalTok{EFS\_data }\OtherTok{\textless{}{-}}\NormalTok{ crooked\_river}

\DocumentationTok{\#\#\# view bull trout models and select model}

\NormalTok{model\_table }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{filter}\NormalTok{(species }\SpecialCharTok{==} \StringTok{"bull trout"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 2 x 5
##   author             species    model dev.type func                       
##   <chr>              <chr>      <chr> <chr>    <chr>                      
## 1 Austin et al. 2017 bull trout MM    hatch    1/exp(5.086 - (x * 0.131)) 
## 2 Austin et al. 2017 bull trout MM    emerge   1/exp(5.590 - (x  * 0.126))
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# we care about when fish are out of gravel, so select emerge mod}
\NormalTok{bt\_emerge\_mod }\OtherTok{\textless{}{-}} \FunctionTok{model\_select}\NormalTok{(}\AttributeTok{author =} \StringTok{"Austin et al. 2017"}\NormalTok{,}
                            \AttributeTok{species =} \StringTok{"bull trout"}\NormalTok{,}
                            \AttributeTok{model =} \StringTok{"MM"}\NormalTok{,}
                            \AttributeTok{dev.type =} \StringTok{"emerge"}\NormalTok{)}

\CommentTok{\# predict emergence time using Sept. 30 as the spawn date }
\NormalTok{bt\_emergence }\OtherTok{\textless{}{-}} \FunctionTok{predict\_phenology}\NormalTok{(}\AttributeTok{data =}\NormalTok{ EFS\_data,}
                                  \AttributeTok{dates =}\NormalTok{ date,}
                                  \AttributeTok{temperature =}\NormalTok{ temp\_c,}
                                  \AttributeTok{spawn.date =} \StringTok{"2011{-}09{-}30"}\NormalTok{, }
                                  \AttributeTok{model =}\NormalTok{ bt\_emerge\_mod)}

\CommentTok{\# fish emerge May 1, before June 1}
\NormalTok{bt\_emergence}\SpecialCharTok{$}\NormalTok{dev.period}\SpecialCharTok{$}\NormalTok{stop}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "2012-05-01 UTC"
\end{verbatim}

In this example we expect the last fish out of the gravel well before
the June 1st date and the manager could allow grazing in this area
without worrying about direct mechanical disturbance to fish developing
in the gravel.

\section{Case Study 2}\label{case-study-2}

\section{Discussion}\label{discussion}

talk about how these models also represent local adaptation and
heritable plasticity

talk about how these models represent point estimates and that emergence
and hatch will take the form of a distribution

\section{Acknowledgements}\label{acknowledgements}

We would like to sincerely thank the editor and reviewers for all of
their helpful feedback which greatly improved both the software and the
manuscript.

The views expressed in this manuscript are those of the authors and do
not necessarily represent the views or policies of USFS or USFWS. Any
mention of trade names, products, or services does not imply an
endorsement by the U.S. government, USFS, or USFWS. USFS and USFWS do
not endorse any commercial products, services or enterprises.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-adelfio2019}
Adelfio, L. A., Wondzell, S. M., Mantua, N. J., \& Reeves, G. H. (2019).
Warm winters reduce landscape-scale variability in the duration of egg
incubation for coho salmon (oncorhynchus kisutch) on the copper river
delta, alaska. \emph{Canadian Journal of Fisheries and Aquatic
Sciences}, \emph{76}(8), 1362--1375.
doi:\href{https://doi.org/10.1139/cjfas-2018-0152}{10.1139/cjfas-2018-0152}

\bibitem[\citeproctext]{ref-adelfio2024}
Adelfio, L. A., Wondzell, S. M., Mantua, N. J., \& Reeves, G. H. (2024).
Expanded, compressed, or equal? Interactions between spawning window and
stream thermal regime generate three responses in modeled juvenile
emergence for pacific salmon. \emph{Canadian Journal of Fisheries and
Aquatic Sciences}.
doi:\href{https://doi.org/10.1139/cjfas-2023-0238}{10.1139/cjfas-2023-0238}

\bibitem[\citeproctext]{ref-austin2019}
Austin, C. S., Essington, T. E., \& Quinn, T. P. (2019). Spawning and
emergence phenology of bull trout Salvelinus confluentus under differing
thermal regimes. \emph{Journal of Fish Biology}, \emph{94}(1), 191--195.
doi:\href{https://doi.org/10.1111/jfb.13864}{10.1111/jfb.13864}

\bibitem[\citeproctext]{ref-beacham1990}
Beacham, T. D., \& Murray, C. B. (1990). Temperature, egg size, and
development of embryos and alevins of five species of pacific salmon: A
comparative analysis. \emph{Transactions of the American Fisheries
Society}, \emph{119}(6), 927--945.
doi:\href{https://doi.org/10.1577/1548-8659(1990)119\%3C0927:TESADO\%3E2.3.CO;2}{10.1577/1548-8659(1990)119\textless0927:TESADO\textgreater2.3.CO;2}

\bibitem[\citeproctext]{ref-kaylor2021}
Kaylor, M. J., Justice, C., Armstrong, J. B., Staton, B. A., Burns, L.
A., Sedell, E., \& White, S. M. (2021). Temperature, emergence phenology
and consumption drive seasonal shifts in fish growth and production
across riverscapes. \emph{Journal of Animal Ecology}, \emph{90}(7),
1727--1741.
doi:\href{https://doi.org/10.1111/1365-2656.13491}{10.1111/1365-2656.13491}

\bibitem[\citeproctext]{ref-mcphail1979}
McPhail, J., \& Murray, C. (1979). \emph{The early life-history and
ecology of dolly varden (salvelimus malma) in the upper arrow lakes}.
Nelson, BC, Canada.

\bibitem[\citeproctext]{ref-nolfi2024}
Nolfi, D., Melbihess, T., Fisher, S., \& Ellis, L. (2024). 5-Year
Statuse Review Coterminous United States Population of Bull Trout
(Salvelinus confluentus).

\bibitem[\citeproctext]{ref-sparks2019}
Sparks, M. M., Falke, J. A., Quinn, T. P., Adkison, M. D., Schindler, D.
E., Bartz, K., Young, D., et al. (2019). Influences of spawning timing,
water temperature, and climatic warming on early life history phenology
in western alaska sockeye salmon. \emph{Canadian Journal of Fisheries
and Aquatic Sciences}, \emph{76}(1), 123--135.
doi:\href{https://doi.org/10.1139/cjfas-2017-0468}{10.1139/cjfas-2017-0468}

\bibitem[\citeproctext]{ref-sparks2017}
Sparks, M. M., Westley, P. A. H., Falke, J. A., \& Quinn, T. P. (2017).
Thermal adaptation and phenotypic plasticity in a warming world:
Insights from common garden experiments on Alaskan sockeye salmon.
\emph{Global Change Biology}, \emph{23}(12), 5203--5217.
doi:\href{https://doi.org/10.1111/gcb.13782}{10.1111/gcb.13782}

\end{CSLReferences}

\end{document}

### 2024-11-26 inst/manuscript/paper.Rmd
dealaing with diagram issues

@@
 ```{r echo=FALSE, eval=FALSE}
+## THIS WONT WORK WITH PDF OUTPUT
 
-  library(DiagrammeR)
-  #see https://rpubs.com/anshulkumar/FlowchartsInR
-  grViz("digraph {
-
+graph <- 
+  "digraph {
       # a 'graph' statement
       graph [overlap = false, label = 'hatchR workflow',labelloc = t, fontsize = 30]
-
       # label nodes
@@
       plot3 [label = 'plot_phenology()', shape = rectangle]
-
       # edge definitions with the node IDs
@@
     }
-      ", width = 500)
+"
+
+DiagrammeR::grViz(graph,
+    width = 500
+    )
 

### 2024-11-26 inst/manuscript/flowchart-tests.R

@@ -1,0 +1,42 @@
+
+# path <- here::here("manuscript", "flowchart.gv")
+#
+# DiagrammeR::grViz(g)
+#
+# g <- "digraph {
+#
+#       # a 'graph' statement
+#       graph [overlap = false, label = 'hatchR workflow', labelloc = t, fontsize = 30]
+#
+#       # label nodes
+#       node [fontsize = 20]
+#       import [label = 'import data', shape = circle, fillcolor = Beige, style=filled]
+#       dates [label = 'check dates', shape = circle, fillcolor = Beige, style=filled]
+#       plot1 [label = 'plot_check_data()' shape = rectangle]
+#       format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige, style=filled]
+#       plot2 [label = 'plot_check_data()' shape = rectangle]
+#       summarize [label = 'sumamrize_temp()',  shape = rectangle]
+#       continuous [label = 'check_continuous()', shape = rectangle]
+#       model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige, style=filled]
+#       select [label = 'model_select()', shape = rectangle]
+#       fit [label = 'fit_model()', shape = rectangle]
+#       phenology [label = 'predict_phenology()', shape = rectangle]
+#       plot3 [label = 'plot_phenology()', shape = rectangle]
+#
+#       # edge definitions with the node IDs
+#       edge [fontsize = 20]
+#       import -> dates
+#       dates -> dates [label = 'not dttm\nuse lubridate']
+#       dates -> plot1 [label = 'is dttm']
+#       plot1 -> format
+#       format -> summarize [label = 'multiple measures\nw/in a day']
+#       format -> model [label = 'mean daily average']
+#       summarize -> continuous -> plot2 -> model
+#       model -> select [label = 'hatchR built-in\nmodel']
+#       model -> fit [label = 'user custom data']
+#       {select, fit} -> phenology
+#       phenology -> plot3
+#
+#       }
+#
+# "

### 2024-11-26 inst/manuscript/paper.Rmd
workflow figure revisions

@@
 
-```{r echo=FALSE, eval=FALSE}
-## THIS WONT WORK WITH PDF OUTPUT
-
-graph <- 
-  "digraph {
-      # a 'graph' statement
-      graph [overlap = false, label = 'hatchR workflow',labelloc = t, fontsize = 30]
-      # label nodes
-      node [fontsize = 20]
-      import [label = 'import data', shape = circle, fillcolor = Beige]
-      dates [label = 'check dates', shape = circle, fillcolor = Beige]
-      plot1 [label = 'plot_check_data()' shape = rectangle]
-      format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige]
-      plot2 [label = 'plot_check_data()' shape = rectangle]
-      summarize [label = 'sumamrize_temp()',  shape = rectangle]
-      continuous [label = 'check_continuous()', shape = rectangle]
-      model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige]
-      select [label = 'model_select()', shape = rectangle]
-      fit [label = 'fit_model()', shape = rectangle]
-      phenology [label = 'predict_phenology()', shape = rectangle]
-      plot3 [label = 'plot_phenology()', shape = rectangle]
-      # edge definitions with the node IDs
-      edge [fontsize = 20]
-      import -> dates
-      dates -> dates [label = 'not dttm\nuse lubridate']
-      dates-> plot1 [label = 'is dttm']
-      plot1 -> format
-      format -> summarize [label = 'multiple measures\nw/in a day']
-      format -> model [label = 'mean daily average']
-      summarize -> continuous -> plot2 -> model
-      model -> select [label = 'hatchR built-in\nmodel']
-      model -> fit [label = 'user custom data']
-      {select, fit} -> phenology
-      phenology -> plot3
-    }
-"
-
-DiagrammeR::grViz(graph,
-    width = 500
-    )
-
+```{r echo = FALSE, out.height = '50%', fig.align = 'center'}
+knitr::include_graphics("flowchart.png")
 ```

### 2024-11-26 inst/manuscript/flowchart-render.R

@@ -1,18 +1,23 @@
-digraph {
+
+# path <- here::here("manuscript", "flowchart.gv")
+#
+# DiagrammeR::grViz(g)
+
+g <- "digraph {
 
       # a 'graph' statement
       graph [overlap = false, label = 'hatchR workflow', labelloc = t, fontsize = 30]
 
       # label nodes
       node [fontsize = 20]
-      import [label = 'import data', shape = circle, fillcolor = Green]
-      dates [label = 'check dates', shape = circle, fillcolor = Beige]
+      import [label = 'import data', shape = circle, fillcolor = Beige, style=filled]
+      dates [label = 'check dates', shape = circle, fillcolor = Beige, style=filled]
       plot1 [label = 'plot_check_data()' shape = rectangle]
-      format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige]
+      format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige, style=filled]
       plot2 [label = 'plot_check_data()' shape = rectangle]
       summarize [label = 'sumamrize_temp()',  shape = rectangle]
       continuous [label = 'check_continuous()', shape = rectangle]
-      model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige]
+      model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige, style=filled]
       select [label = 'model_select()', shape = rectangle]
       fit [label = 'fit_model()', shape = rectangle]
       phenology [label = 'predict_phenology()', shape = rectangle]
@@ -25,10 +30,20 @@ digraph {
       dates -> plot1 [label = 'is dttm']
       plot1 -> format
       format -> summarize [label = 'multiple measures\nw/in a day']
-      format -> model [label = 'mean daily average']
+      format -> model [label = '  mean daily average']
       summarize -> continuous -> plot2 -> model
-      model -> select [label = 'hatchR built-in\nmodel']
-      model -> fit [label = 'user custom data']
+      model -> select [label = 'hatchR built-\nin model']
+      model -> fit [label = ' user custom data']
       {select, fit} -> phenology
       phenology -> plot3
+
       }
+
+"
+DiagrammeR::grViz(g)
+
+DiagrammeR::grViz(g) |>
+  DiagrammeRsvg::export_svg() |>
+  charToRaw() |>
+  rsvg::rsvg()  |>
+  png::writePNG("inst/manuscript/flowchart.png")

### 2024-11-26 inst/manuscript/flowchart-tests.R

@@ -1,42 +1,0 @@
-
-# path <- here::here("manuscript", "flowchart.gv")
-#
-# DiagrammeR::grViz(g)
-#
-# g <- "digraph {
-#
-#       # a 'graph' statement
-#       graph [overlap = false, label = 'hatchR workflow', labelloc = t, fontsize = 30]
-#
-#       # label nodes
-#       node [fontsize = 20]
-#       import [label = 'import data', shape = circle, fillcolor = Beige, style=filled]
-#       dates [label = 'check dates', shape = circle, fillcolor = Beige, style=filled]
-#       plot1 [label = 'plot_check_data()' shape = rectangle]
-#       format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige, style=filled]
-#       plot2 [label = 'plot_check_data()' shape = rectangle]
-#       summarize [label = 'sumamrize_temp()',  shape = rectangle]
-#       continuous [label = 'check_continuous()', shape = rectangle]
-#       model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige, style=filled]
-#       select [label = 'model_select()', shape = rectangle]
-#       fit [label = 'fit_model()', shape = rectangle]
-#       phenology [label = 'predict_phenology()', shape = rectangle]
-#       plot3 [label = 'plot_phenology()', shape = rectangle]
-#
-#       # edge definitions with the node IDs
-#       edge [fontsize = 20]
-#       import -> dates
-#       dates -> dates [label = 'not dttm\nuse lubridate']
-#       dates -> plot1 [label = 'is dttm']
-#       plot1 -> format
-#       format -> summarize [label = 'multiple measures\nw/in a day']
-#       format -> model [label = 'mean daily average']
-#       summarize -> continuous -> plot2 -> model
-#       model -> select [label = 'hatchR built-in\nmodel']
-#       model -> fit [label = 'user custom data']
-#       {select, fit} -> phenology
-#       phenology -> plot3
-#
-#       }
-#
-# "

### 2024-11-26 inst/misc/workflow_diagram.R

@@ -1,40 +1,0 @@
-# workflow_plot <- function(){
-#   library(DiagrammeR)
-#   #see https://rpubs.com/anshulkumar/FlowchartsInR
-#   grViz("digraph {
-#
-#       # a 'graph' statement
-#       graph [overlap = false, label = 'hatchR workflow',labelloc = t, fontsize = 30]
-#
-#       # label nodes
-#       node [fontsize = 20]
-#       import [label = 'import data', shape = circle, fillcolor = Beige]
-#       dates [label = 'check dates', shape = circle, fillcolor = Beige]
-#       plot1 [label = 'plot_check_data()' shape = rectangle]
-#       format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige]
-#       plot2 [label = 'plot_check_data()' shape = rectangle]
-#       summarize [label = 'sumamrize_temp()',  shape = rectangle]
-#       model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige]
-#       select [label = 'model_select()', shape = rectangle]
-#       fit [label = 'fit_model()', shape = rectangle]
-#       phenology [label = 'predict_phenology()', shape = rectangle]
-#       plot3 [label = 'plot_phenology()', shape = rectangle]
-#
-#       # edge definitions with the node IDs
-#       edge [fontsize = 20]
-#       import -> dates
-#       dates -> dates [label = 'not dttm\nuse lubridate']
-#       dates-> plot1 [label = 'is dttm']
-#       plot1 -> format
-#       format -> summarize [label = 'multiple measures\nw/in a day']
-#       format -> model [label = 'mean daily average']
-#       summarize -> plot2 -> model
-#       model -> select [label = 'hatchR built in model']
-#       model -> fit [label = 'user custom data']
-#       {select, fit} -> phenology
-#       phenology -> plot3
-#       }
-#       ")
-# }
-#
-# workflow_plot()

### 2024-11-27 R/check_continuous.R
use cli messages

@@ -16,7 +16,7 @@ check_continuous <- function(data, dates) {
   check_dates <- data |> dplyr::pull({{ dates }})
   if (is.character(check_dates) == TRUE) {
     cli::cli_abort(c(
-      "You've supplied dates as a {.cls character} vector.",
+      "`dates` must be a vector of class {.cls date} or {.cls dttm}, not a {.cls character} vector.",
       "i" = "Use {.fn lubridate::ymd} to convert to {.cls date} or {.cls dttm} vector."
     ))
   }

### 2024-11-27 R/fit_model.R

@@ -38,12 +38,18 @@ fit_model <- function(df, temp, days, species = NULL, dev.type = NULL) {
 
   # check if species is NULL
   if (is.null(species)) {
-    stop("Please provide a species name.")
+    cli::cli_abort(c(
+            "`species` cannot be NULL.",
+      "i" = "Provide a species name using `species = 'your_species'`."
+      ))
   }
 
   # check if dev.type is NULL
   if (is.null(dev.type)) {
-    stop("Please provide a valid development type: 'hatch' or 'emerge'.")
+    cli::cli_abort(c(
+            "`dev.type` cannot be NULL",
+      "i" = "Provide a dev.type name using `dev.type = 'hatch' or `dev.type = 'emerge'`."
+    ))
   }
 
   # fit linear model to log data

### 2024-11-27 R/plot_check_temp.R

@@ -34,7 +34,7 @@ plot_check_temp <- function(data,
   check_dates <- data |> dplyr::pull({{ dates }})
   if (is.character(check_dates) == TRUE) {
     cli::cli_abort(c(
-            "You've supplied dates as a {.cls character} vector.",
+            "`dates` must be a vector of class {.cls date} or {.cls dttm}, not a {.cls character} vector.",
       "i" = "Use {.fn lubridate::ymd} to convert to {.cls date} or {.cls dttm} vector."
     ))
   }

### 2024-11-27 R/summarize_temp.R

@@ -24,13 +24,13 @@
 summarize_temp <- function(data,
                            dates,
                            temperature) {
-  # check if dates are a character vector
+
   check_dates <- data |> dplyr::pull({{ dates }})
   if (is.character(check_dates) == TRUE) {
-    stop(
-      "Date column is character vector; convert to date or date-time class.",
-      call. = FALSE
-    )
+    cli::cli_abort(c(
+      "`dates` must be a vector of class {.cls date} or {.cls dttm}, not a {.cls character} vector.",
+      "i" = "Use {.fn lubridate::ymd} to convert to {.cls date} or {.cls dttm} vector."
+    ))
   }
 
   sum_dat <- data |>

### 2024-11-27 inst/misc/interactive_checking.R
adding to do list and interactive checking script

@@ -1,0 +1,42 @@
+load_all()
+
+# get model parameterization
+sockeye_hatch_mod <- model_select(
+  author = "Beacham and Murray 1990",
+  species = "sockeye",
+  model = 2,
+  dev.type = "hatch"
+)
+# predict phenology
+sockeye_hatch <- predict_phenology(
+  data = woody_island,
+  dates = date,
+  temperature = temp_c,
+  spawn.date = "1990-08-18",
+  model = sockeye_hatch_mod
+)
+
+# vector of temperatures
+temp <- c(2, 5, 8, 11, 14)
+# vector of days to hatch
+days <- c(194, 87, 54, 35, 28)
+hatch <- tibble::tibble(temp, days)
+hatch <- fit_model(df = hatch, temp = temp, days = days, species = "fish")
+hatch$func
+
+crooked_river$date <- as.character(crooked_river$date)
+crooked_river$date <- as.Date(crooked_river$date)
+
+plot_check_temp(
+  data = crooked_river,
+  dates = date,
+  temperature = temp_c
+)
+
+# remove a 20 random rows from crooked_river
+tmp <- crooked_river[-sample(1:nrow(crooked_river), 20),]
+
+check_continuous(tmp, date)
+
+chk <- check_continuous(tmp, date)
+chk

### 2024-11-27 inst/misc/to-do-list.R

@@ -1,0 +1,4 @@
+# update lifecycle bages
+
+# add better error and warning messages
+

### 2024-11-27 R/predict_phenology.R
add dependency

@@ -111,7 +111,7 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     dev.period <- data.frame(matrix(NA, nrow = 1, ncol = 2))
     colnames(dev.period) <- c("start", "stop")
     dev.period$start <- min(ef.df$dates)
-    dev.period$stop <- as_date(NA)
+    dev.period$stop <- lubridate::as_date(NA)
 
     ef.results <- list(days2done=as.numeric(NA),
                        dev.period=dev.period,

### 2024-11-27 inst/manuscript/paper.tex
ms files

@@
-\documentclass[12pt,halfline,a4paper,]{ouparticle}
+\documentclass[10pt,a4paper,onecolumn]{article}
+\usepackage{marginnote}
+\usepackage{graphicx}
+\usepackage{xcolor}
+\usepackage{authblk,etoolbox}
+\usepackage{titlesec}
+\usepackage{calc}
+\usepackage{tikz}
+\usepackage{hyperref}
+\hypersetup{colorlinks,breaklinks,
+            urlcolor=[rgb]{0.0, 0.5, 1.0},
+            linkcolor=[rgb]{0.0, 0.5, 1.0}}
+\usepackage{caption}
+\usepackage{tcolorbox}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{seqsplit}
+\usepackage{fixltx2e} % provides \textsubscript
+\usepackage[
+  backend=biber,
+%  style=alphabetic,
+%  citestyle=numeric
+]{biblatex}
+\bibliography{ms\_refs.bib}
 
-% Packages I think are necessary for basic Rmarkdown functionality
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{listings}
-\usepackage{xcolor}
-\usepackage{fancyvrb}
-\usepackage{framed}
+% --- Page layout -------------------------------------------------------------
+\usepackage[top=3.5cm, bottom=3cm, right=1.5cm, left=1.0cm,
+            headheight=2.2cm, reversemp, includemp, marginparwidth=4.5cm]{geometry}
 
-% Link coloring
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={hatchR: A toolset to predict hatch and emergence phenology in wild fishes}
+% --- Default font ------------------------------------------------------------
+% \renewcommand\familydefault{\sfdefault}
+
+% --- Style -------------------------------------------------------------------
+\renewcommand{\bibfont}{\small \sffamily}
+\renewcommand{\captionfont}{\small\sffamily}
+\renewcommand{\captionlabelfont}{\bfseries}
+
+% --- Section/SubSection/SubSubSection ----------------------------------------
+\titleformat{\section}
+  {\normalfont\sffamily\Large\bfseries}
+  {}{0pt}{}
+\titleformat{\subsection}
+  {\normalfont\sffamily\large\bfseries}
+  {}{0pt}{}
+\titleformat{\subsubsection}
+  {\normalfont\sffamily\bfseries}
+  {}{0pt}{}
+\titleformat*{\paragraph}
+  {\sffamily\normalsize}
+
+% --- Header / Footer ---------------------------------------------------------
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+%\renewcommand{\headrulewidth}{0.50pt}
+\renewcommand{\headrulewidth}{0pt}
+\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{C:/Users/BryanMaitland/AppData/Local/R/win-library/4.4/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
+\fancyhead[C]{}
+\fancyhead[R]{}
+\renewcommand{\footrulewidth}{0.25pt}
+
+\fancyfoot[L]{\footnotesize{\sffamily Sparks et al., (2025). hatchR: A
+toolset to predict hatch and emergence phenology in wild
+fishes. \textit{Journal of Open Source Software}, (), . \href{https://doi.org/}{https://doi.org/}}}
+
+\fancyfoot[R]{\sffamily \thepage}
+\makeatletter
+\let\ps@plain\ps@fancy
+\fancyheadoffset[L]{4.5cm}
+\fancyfootoffset[L]{4.5cm}
+
+% --- Macros ---------
+
+\definecolor{linky}{rgb}{0.0, 0.5, 1.0}
+
+\newtcolorbox{repobox}
+   {colback=red, colframe=red!75!black,
+     boxrule=0.5pt, arc=2pt, left=6pt, right=6pt, top=3pt, bottom=3pt}
+
+\newcommand{\ExternalLink}{%
+   \tikz[x=1.2ex, y=1.2ex, baseline=-0.05ex]{%
+       \begin{scope}[x=1ex, y=1ex]
+           \clip (-0.1,-0.1)
+               --++ (-0, 1.2)
+               --++ (0.6, 0)
+               --++ (0, -0.6)
+               --++ (0.6, 0)
+               --++ (0, -1);
+           \path[draw,
+               line width = 0.5,
+               rounded corners=0.5]
+               (0,0) rectangle (1,1);
+       \end{scope}
+       \path[draw, line width = 0.5] (0.5, 0.5)
+           -- (1, 1);
+       \path[draw, line width = 0.5] (0.6, 1)
+           -- (1, 1) -- (1, 0.6);
+       }
    }
 
-%% To allow better options for figure placement
-%\usepackage{float}
+% --- Title / Authors ---------------------------------------------------------
+% patch \maketitle so that it doesn't center
+\patchcmd{\@maketitle}{center}{flushleft}{}{}
+\patchcmd{\@maketitle}{center}{flushleft}{}{}
+% patch \maketitle so that the font size for the title is normal
+\patchcmd{\@maketitle}{\LARGE}{\LARGE\sffamily}{}{}
+% patch the patch by authblk so that the author block is flush left
+\def\maketitle{{%
+  \renewenvironment{tabular}[2][]
+    {\begin{flushleft}}
+    {\end{flushleft}}
+  \AB@maketitle}}
+\makeatletter
+\renewcommand\AB@affilsepx{ \protect\Affilfont}
+%\renewcommand\AB@affilnote[1]{{\bfseries #1}\hspace{2pt}}
+\renewcommand\AB@affilnote[1]{{\bfseries #1}\hspace{3pt}}
+\makeatother
+\renewcommand\Authfont{\sffamily\bfseries}
+\renewcommand\Affilfont{\sffamily\small\mdseries}
+\setlength{\affilsep}{1em}
 
-% Packages that are supposedly required by OUP sty file
-\usepackage{amssymb, amsmath, geometry, amsfonts, verbatim, endnotes, setspace}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
 
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+
+\fi
 % use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
 
-% Macros for dealing with affiliations, footnotes, etc.
+\usepackage{hyperref}
+\hypersetup{unicode=true,
+            pdftitle={hatchR: A toolset to predict hatch and emergence phenology in wild fishes},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
 \makeatletter
-\def\Newlabel#1#2#3{\expandafter\gdef\csname #1@#2\endcsname{#3}}
-
-\def\Ref#1#2{\@ifundefined{#1@#2}{???}{\csname #1@#2\endcsname}}
-
-\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
-
-\newcommand*\ifcounter[1]{%
-  \ifcsname c@#1\endcsname
-    \expandafter\@firstoftwo
-  \else
-    \expandafter\@secondoftwo
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
-}
-
-\newcommand*\thanksbycode[1]{%
-  \ifcounter{FNCT@#1}
-    {\samethanks[\value{FNCT@#1}]}
-    {\thanks{\Ref{FN}{#1}}\newcounter{FNCT@#1}\setcounter{FNCT@#1}{\value{footnote}}}
-}
-
-% Create labels for Addresses if the are given in Elsevier format
-
-% Create labels for Footnotes if the are given in Elsevier format
-
-% Part for setting citation format package: natbib
-
-% Part for setting citation format package: biblatex
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 
-\begin{document}
-
 \title{hatchR: A toolset to predict hatch and emergence phenology in
@@
 
-\author{%
-%
-% Code for old style authors field
-\name{Morgan M. Sparks}
-\and
-\name{Eli Felts}
-\and
-\name{Bryan M. Maitland}
-%
-% Add \and if both authors and author
-%
-%
-% Code for new (elsevier) style author field
-%
+        \author[1]{Morgan M. Sparks}
+          \author[2]{Eli Felts}
+          \author[1]{Bryan M. Maitland}
+
+      \affil[1]{US Forest Service, Rocky Mountain Research Station,
+Boise, ID, USA}
+      \affil[2]{US Fish and Wildlife Service, Idaho Fish and Wildlife
+Conservation Office, Orofino, ID, USA}
+  \date{\vspace{-5ex}}
+
+\begin{document}
+\maketitle
+
+\marginpar{
+  %\hrule
+  \sffamily\small
+
+  {\bfseries DOI:} \href{https://doi.org/}{\color{linky}{}}
+
+  \vspace{2mm}
+
+  {\bfseries Software}
+  \begin{itemize}
+    \setlength\itemsep{0em}
+    \item \href{}{\color{linky}{Review}} \ExternalLink
+    \item \href{}{\color{linky}{Repository}} \ExternalLink
+    \item \href{}{\color{linky}{Archive}} \ExternalLink
+  \end{itemize}
+
+  \vspace{2mm}
+
+  {\bfseries Submitted:} \\
+  {\bfseries Published:}
+
+  \vspace{2mm}
+  {\bfseries License}\\
+  Authors of papers retain copyright and release the work under a Creative Commons Attribution 4.0 International License (\href{http://creativecommons.org/licenses/by/4.0/}{\color{linky}{CC-BY}}).
 }
-
-\abstract{}
-
-\date{07 July 2025}
-
-\keywords{}
-
-\maketitle
 
@@
 programs of the user's choice.
+
+\includegraphics{flowchart.png}
 

### 2024-12-03 R/fit_model.R
fix bugs

@@ -14,9 +14,8 @@
 #' specific discussion regarding model 2 and the development of the
 #' effective value approach.
 #'
-#' @param df Data.frame with new data
-#' @param temp Column with temperature measurement
-#' @param days Column with day of temperature measurement
+#' @param temp Numeric vector of temperatures
+#' @param days Numeric vector of days to hatch or emerge
 #' @param species Character string of species name (e.g., "sockeye")
 #' @param dev.type Character string of development type: "hatch" or "emerge"
 #'
@@ -31,10 +30,10 @@
 #' temperature <- c(2, 5, 8, 11, 14)
 #' # vector of days to hatch
 #' days_to_hatch <- c(194, 87, 54, 35, 28)
-#' quinn_bt_hatch <- tibble::tibble(temperature, days_to_hatch)
-#' bt_hatch_mod <- fit_model(df = quinn_bt_hatch, temp = temperature,
+#' bt_hatch_mod <- fit_model(temp = temperature,
 #' days = days_to_hatch, species = "sockeye", dev.type = "hatch")
-fit_model <- function(df, temp, days, species = NULL, dev.type = NULL) {
+fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
+  df <- tibble::tibble(x = temp, y = days)
 
   # check if species is NULL
   if (is.null(species)) {
@@ -52,19 +51,22 @@ fit_model <- function(df, temp, days, species = NULL, dev.type = NULL) {
     ))
   }
 
+
   # fit linear model to log data
-  m1 <- stats::lm(log(days) ~ log(temp), data = df)
-  summary(m1)
+  m1 <- stats::lm(log(y) ~ x, data = df)
+  # summary(m1)
 
   # estimate starting values for nls
-  m1_a <- exp(stats::coef(m1)[1]) # exponentiate the intercept
-  m1_b <- stats::coef(m1)[2]
+  st <- list(
+    a = exp(stats::coef(m1)[1]),
+    b = stats::coef(m1)[2]
+    )
 
   # fit model 2 from Beacham & Murray (1990) to data
-  m2 <- stats::nls(days ~ a / (temp - b), data = df, start = list(a = m1_a, b = m1_b))
-  summary(m2)
+  m2 <- stats::nls(y ~ a / (x - b), data = df, start = st)
+  # summary(m2)
 
-  # check coefficients
+  # get coefficients
   m2_a <- log(stats::coef(m2)[1])
[…]

### 2024-12-03 tests/testthat/test-fit_model.R
update for bug fix

@@ -2,7 +2,7 @@ test_that("fit_model works", {
   temp <- c(2, 5, 8, 11, 14)
   days <- c(194, 87, 54, 35, 28)
   df <- tibble::tibble(temp, days)
-  m <- fit_model(df = df, temp = temp, days = days, species = "sockeye", dev.type = "hatch")
+  m <- fit_model(temp = temp, days = days, species = "sockeye", dev.type = "hatch")
 
   expect_type(m, "list")
 

### 2024-12-04 data-raw/woody_island.R
fix data error

@@ -1,6 +1,9 @@
-library(readr)
-
-woody_island <- read_csv("data-raw/woody_island.csv")
-woody_island$date <- as.Date(woody_island$date, format = "%m/%d/%Y")
+woody_island <- readr::read_csv(
+  "data-raw/woody_island.csv",
+  col_types = readr::cols(
+    date = readr::col_date(format = "%m/%d/%Y"),
+    temp_c = readr::col_double()
+  )
+)
 
 usethis::use_data(woody_island, overwrite = TRUE)

### 2024-12-04 inst/manuscript/paper.Rmd
add catfish example

@@
 ```{r, eval=FALSE}
+#### temp regime
+set.seed(1234)
+temps <- sort(rnorm(n =30, mean = 20, sd = 1), decreasing = TRUE)
+
 smallmouth <- matrix(NA, 10, 2) |> tibble()
@@
 
-plot(smallmouth$days~smallmouth$temp_C)
+plot(smallmouth$hours~smallmouth$temp_C)
 
@@
                      temp = smallmouth$temp_C,
-                     days = smallmouth$hours)
+                     days = smallmouth$days,
+                     species = "smb",
+                     dev.type = "hatch")
+
+### catfish example from Small and Bates 2001
+### Effect of Low-Temperature Incubation of Channel Catfish Ictalarus punctatus
+### Eggs on Development, Survival, and Growth
+### Journal of the World Aquaculture Society
+catfish <- matrix(NA, 3, 2) |> tibble()
+colnames(catfish) <- c("days", "temp_C")
+catfish$days <- c(16,21,26)
+catfish$temp_C <- c(22,10,7)
+
+cat_mod <- fit_model(catfish,
+                     temp = catfish$temp_C,
+                     days = catfish$days,
+                     species = "catfish",
+                     dev.type = "hatch")
 

### 2024-12-04 inst/manuscript/paper.Rmd
add sturgeon model

--- 2024-11-26 inst/manuscript/paper.Rmd
@@
 ```{r, eval=FALSE}
+#### temp regime
+set.seed(1234)
+# create random temps and corresponding dates
+temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = TRUE)
+dates_sim <-  seq(from = ymd("2000-07-01"),
+             to = ymd("2000-07-31"), length.out = 30)
+
+data_sim <- cbind (dates_sim, temps_sim) |> data.frame()
+
+### smallmouth
 smallmouth <- matrix(NA, 10, 2) |> tibble()
@@
 
-plot(smallmouth$days~smallmouth$temp_C)
-
 smb_mod <- fit_model(smallmouth,
                      temp = smallmouth$temp_C,
-                     days = smallmouth$hours)
+                     days = smallmouth$days,
+                     species = "smb",
+                     dev.type = "hatch")
 
+### catfish example from Small and Bates 2001
+### Effect of Low-Temperature Incubation of Channel Catfish Ictalarus punctatus
+### Eggs on Development, Survival, and Growth
+### Journal of the World Aquaculture Society
+
+catfish <- matrix(NA, 3, 2) |> tibble()
+colnames(catfish) <- c("days", "temp_C")
+catfish$days <- c(16,21,26)
+catfish$temp_C <- c(22,10,7)
+
+cat_mod <- fit_model(catfish,
+                     temp = catfish$temp_C,
+                     days = catfish$days,
+                     species = "catfish",
+                     dev.type = "hatch")
+
+### lake sturgeon mod
+### Smith and King 2005
+### Dynamics and extent of larval lake sturgeon Acipenser fulvescens drift in the 
+### Upper Black River, Michigan
+### Journal of Applied Ecology
+
+sturgeon <-  matrix(NA, 7, 2) |> tibble()
+colnames(sturgeon) <- c("days", "CTU")
+sturgeon$days <- c(7,5,6,6,5,11,7)
+sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)
+
+sturgeon <- sturgeon |> 
+  mutate(temp_C = CTU/days)
+
+plot(sturgeon$days ~ sturgeon$temp_C)
+
+sturgeon_mod <- fit_model(df = sturgeon,
+                          days = sturgeon$days,
+                          temp = sturgeon$temp_C,
+                          species = "sturgeon",
+                          dev.type = "hatch")
+
+### predict_phenology
+
+catfish_res <- predict_phenology(data = data_sim,
+                  dates = dates_sim,
+                  temperature = temps_sim,
+                  spawn.date = "2000-07-01",
+                  model = cat_mod$func)
+
+# note that 16 C is pretty far out of range of temps for model fit, not best practice
+sturgeon_res <- predict_phenology(data = data_sim,
+                  dates = dates_sim,
+                  temperature = temps_sim,
+                  spawn.date = "2000-07-01",
+                  model = sturgeon_mod$func)
+
+catfish_res$days2done; sturgeon_res$days2done
 ```
@@
 
-Use woody \_example from website
+Use woody example from website
 

### 2024-12-04 R/predict_phenology.R
function overhaul

@@ -7,7 +7,8 @@
 #' @param data Dataframe with dates and temperature.
 #' @param dates Date of temperature measurements.
 #' @param temperature Temperature measurements.
-#' @param spawn.date Date of spawning, given as a character string (e.g., "1990-08-18")
+#' @param spawn.date Date of spawning, given as a character string
+#' (e.g., "1990-08-18")
 #' @param model A data.frame giving model specifications. This must have a
 #' column providing a model expression. Can be obtained using `model_select()`
 #' or using you own data to obtain a model expression (see `fit_model`).
@@ -17,7 +18,6 @@
 #'
 #' @return
 #' A list with the following elements:
-#' * `ef.vals`: A numeric vector of each day's effective value.
 #' * `days2done`: A numeric vector of length 1; number of predicted days to hatch or emerge.
 #' * `ef.tibble`: An n x 4 tibble (n = number of days to hatch or emerge) with
 #'  the dates, temperature, effective values, and cumulative sum of the effective values.
@@ -55,125 +55,167 @@
 #'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
 #'   \bold{76(1)}, 123--135
 predict_phenology <- function(data, dates, temperature, spawn.date, model) {
-  # arrange data by dates
-  dat <- data |> dplyr::arrange({{ dates }})
+  # assign data and arrange data by dates
+  dat <- data |>
+    dplyr::arrange({{ dates }}) |>
+    tibble::rownames_to_column(var = "index") |>
+    dplyr::mutate(index = as.numeric(.data$index))
 
   # check if dates are a character vector
   check_dates <- dat |> dplyr::pull({{ dates }})
   if (is.character(check_dates) == TRUE) {
-    stop(
-      "Date column is character vector; convert to date or date-time class.",
-      call. = FALSE
+    cli::cli_abort(c(
+      "`dates` must be a vector of class {.cls date} or {.cls dttm}, not a {.cls character} vector.",
+      "i" = "Use {.fn lubridate::ymd} to convert to {.cls date} or {.cls dttm} vector."
+    ))
+  }
+
+  # check if spawn.date is formatted as a date
+  if (lubridate::is.timepoint(spawn.date) == TRUE ||
[…]

### 2024-12-04 tests/testthat/test-predict_phenology.R
update for funciton overhaul

@@ -14,14 +14,13 @@ test_that("predict phenology works", {
     )
 
   expect_type(p, "list")
-  expect_length(p, 5)
+  expect_length(p, 4)
 
-  expect_type(p$days2done, "integer")
+  # expect_type(p$days2done, "integer")
 
-  expect_type(p$ef.vals, "double")
+  # expect_type(p$ef.vals, "double")
 
   expect_s3_class(p$ef.tibble, "data.frame")
-  expect_equal(ncol(p$ef.tibble), 4)
 
   expect_s3_class(p$dev.period, "data.frame")
   expect_equal(ncol(p$dev.period), 2)

### 2024-12-05 inst/manuscript/paper.Rmd
add species custom models and section on data checks and summary

--- 2024-11-26 inst/manuscript/paper.Rmd
@@
 
+```{r echo=FALSE}
+### 
+library(tidyverse); library(hatchR)
+```
+
 # Package Overview
@@
 
-summarize_temp, check_continuous and plot_check_temp
+We demonstrate the utility of the check summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year long dataset with temperature readings every thirty minutes.
+
+```{r}
+# create date object for a year with 30 min reading intervals
+dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
+             to = ymd_hms("2001-07-30 23:59:59"), length.out = 17568) 
+# create empty dataframe
+year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))
+
+# date column
+colnames(year_sim)[1] <- "date"
+
+# add dates vector to date column
+year_sim[1] <- dates
+
+#random seed
+set.seed(123)
+
+# take temps from a random normal dist with mean 10 sd 3 for every date time combo in dates and append to column (temp) in year_sim
+year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
+  abs()
+
+dim(year_sim)
+
+```
+
+First, we recommend checking import data for any outliers or strange inputs using `plot_check_temp()`
+
+```{r}
+plot_check_temp(data = year_sim,
+                dates = date,
+                temperature = temp,
+                temp_min = 0, # temp_min and max lines are user customizable
+                temp_max = 25)
+```
+
+It appears there are no outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`.
+We also recommend using `plot_check_temp()` again on the summarized data.
+
+```{r}
+# summarize
+year_sim_summ <- summarize_temp(data = year_sim,
+                                dates = date,
+                                temperature = temp)
+
+dim(year_sim_summ)
+
+# check continuous
+check_continuous(data = year_sim_summ,
+                 dates = date)
+
+# we can demonstrate an error by removing the Oct 8 (100th day)
+check_continuous(data = year_sim_summ[-100,],
+                 dates = date)
+
+# it is useful to plot again to check your summarized data
+plot_check_temp(data = year_sim_summ,
+                dates = date,
+                temperature = daily_temp,
+                temp_min = 0, # temp_min and max lines are user customizable
+                temp_max = 15)
+```
 
@@
 
-fit_model for three non-salmonid species
+fit_model for three non-salmonid specieslib
+
+Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
+Furthermore, we include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the original scope of the original effective value approach.
+We demonstrate parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+
+We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime.
 
 ```{r, eval=FALSE}
+###  make temp regime
+set.seed(1234)
+# create random temps and corresponding dates
+temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
+dates_sim <-  seq(from = ymd("2000-07-01"),
+             to = ymd("2000-07-31"), length.out = 30)
+
+data_sim <- cbind (dates_sim, temps_sim) |> data.frame()
+
+### smallmouth mod
 smallmouth <- matrix(NA, 10, 2) |> tibble()
@@
 
+# change F to C and hours to days
 smallmouth <- smallmouth |> 
@@
 
-plot(smallmouth$days~smallmouth$temp_C)
+smb_mod <- fit_model(temp = smallmouth$temp_C,
+                     days = smallmouth$days,
+                     species = "smb",
+                     dev.type = "hatch")
 
-smb_mod <- fit_model(smallmouth,
-                     temp = smallmouth$temp_C,
-                     days = smallmouth$hours)
+### catfish mod
+catfish <- matrix(NA, 3, 2) |> tibble()
+colnames(catfish) <- c("days", "temp_C")
+catfish$days <- c(16,21,26)
+catfish$temp_C <- c(22,10,7)
 
+cat_mod <- fit_model(temp = catfish$temp_C,
+                     days = catfish$days,
+                     species = "catfish",
+                     dev.type = "hatch")
+
+### lake sturgeon mod
+sturgeon <-  matrix(NA, 7, 2) |> tibble()
+colnames(sturgeon) <- c("days", "CTU")
+sturgeon$days <- c(7,5,6,6,5,11,7)
+sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)
+
+sturgeon <- sturgeon |> 
+  mutate(temp_C = CTU/days)
+
+plot(sturgeon$days ~ sturgeon$temp_C)
+
+sturgeon_mod <- fit_model(days = sturgeon$days,
+                          temp = sturgeon$temp_C,
+                          species = "sturgeon",
+                          dev.type = "hatch")
+
+### predict_phenology
+
+#smallmouth bass
+smb_res <- predict_phenology(data = data_sim,
+                  dates = dates_sim,
+                  temperature = temps_sim,
+                  spawn.date = "2000-07-01",
+                  model = smb_mod$func)
+
+# catfish
+catfish_res <- predict_phenology(data = data_sim,
+                  dates = dates_sim,
+                  temperature = temps_sim,
+                  spawn.date = "2000-07-01",
+                  model = cat_mod$func)
+
+# sturgeon
+# note that 16 C is pretty far out of range of temps for model fit, not best practice
+sturgeon_res <- predict_phenology(data = data_sim,
+                  dates = dates_sim,
+                  temperature = temps_sim,
+                  spawn.date = "2000-07-01",
+                  model = sturgeon_mod$func)
+
+```
+
+```{r, echo=FALSE}
+# summary for all species
+all_res <- data.frame(matrix(NA, 3, 2))
+colnames(all_res) <- c("start", "stop")
+all_res[1,] <- smb_res$dev.period
+all_res[2,] <- catfish_res$dev.period
+all_res[3,] <- sturgeon_res$dev.period
+
+all_res <- all_res |> 
+  mutate(days = stop-start)
+all_res$Species <- c("Smallmouth Bass", "Channel Catfish", "Lake Sturgeon")
 ```

### 2024-12-05 R/predict_phenology.R
redocument

@@ -18,13 +18,15 @@
 #'
 #' @return
 #' A list with the following elements:
-#' * `days2done`: A numeric vector of length 1; number of predicted days to hatch or emerge.
+#' * `days2done`: A numeric vector of length 1; number of predicted
+#'    days to hatch or emerge.
 #' * `ef.tibble`: An n x 4 tibble (n = number of days to hatch or emerge) with
-#'  the dates, temperature, effective values, and cumulative sum of the effective values.
+#'    the dates, temperature, effective values, and cumulative sum of the
+#'    effective values.
 #' * `dev.period`: a 1x2 dataframe with the dates corresponding to when your
-#'  fish's parent spawned (input with `predict_phenology(spawn.date = ...)`)
-#'  and the date when the fish is predicted to hatch or emerge.
-#'  * `model.specs`: A data.frame with the model specifications.
+#'    fish's parent spawned (input with `predict_phenology(spawn.date = ...)`)
+#'    and the date when the fish is predicted to hatch or emerge.
+#' * `model.specs`: A data.frame with the model specifications.
 #'
 #' @export
 #'

### 2024-12-05 data-raw/crooked_river.R
omit lib call

@@ -1,5 +1,3 @@
-library(readr)
-
-crooked_river <- read_csv("data-raw/crooked_river.csv")
+crooked_river <- readr::read_csv("data-raw/crooked_river.csv")
 
 usethis::use_data(crooked_river, overwrite = TRUE)

### 2024-12-05 R/check_continuous.R
fix warn

@@ -27,10 +27,10 @@ check_continuous <- function(data, dates) {
 
   if (length(check_out) > 0) {
     cli::cli_warn(c(
-      "!" = "Breaks at the following rows were found:",
+      "!" = "Data not continuous",
+      "i" = "Breaks found at rows:",
       "i" = paste(check_out, collapse = ", ")
       ))
-    check_out
   } else {
     cli::cli_inform(c("i" = "No breaks were found. All clear!"))
   }

### 2024-12-05 R/hatchR-package.R
import RDpack for roxy refs

@@ -5,6 +5,7 @@ utils::globalVariables(c("model_table", "woody_island"))
 
 ## usethis namespace: start
 #' @importFrom lifecycle deprecated
+#' @importFrom Rdpack reprompt
 #' @importFrom rlang .data
 #' @importFrom utils globalVariables
 ## usethis namespace: end

### 2024-12-05 R/model_table.R
adding refs

@@ -13,4 +13,11 @@
 #' }
 #' @examples
 #' model_table
+#' @references
+#' \insertRef{beacham1990}{hatchR}
+#'
+#' \insertRef{austin2019}{hatchR}
+#'
+#' \insertRef{sparks2019}{hatchR}
+
 "model_table"

### 2024-12-05 R/predict_phenology.R
update refs

@@ -50,12 +50,8 @@
 #' )
 #'
 #' @references
-#' Sparks, M.M., Falke, J.A., Quinn, T.P., Adkinson, M.D., Schindler, D.E.
-#' (2017). Influences of spawning timing, water temperature, and
-#'   climatic warming on early life history phenology in western
-#'   Alaska sockeye salmon.
-#'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
-#'   \bold{76(1)}, 123--135
+#' \insertRef{beacham1990}{hatchR}
+
 predict_phenology <- function(data, dates, temperature, spawn.date, model) {
   # assign data and arrange data by dates
   dat <- data |>

### 2024-12-05 R/hatchR-package.R
remove rdpack

@@ -5,7 +5,6 @@ utils::globalVariables(c("model_table", "woody_island"))
 
 ## usethis namespace: start
 #' @importFrom lifecycle deprecated
-#' @importFrom Rdpack reprompt
 #' @importFrom rlang .data
 #' @importFrom utils globalVariables
 ## usethis namespace: end

### 2024-12-05 R/model_table.R

@@ -14,10 +14,6 @@
 #' @examples
 #' model_table
 #' @references
-#' \insertRef{beacham1990}{hatchR}
-#'
-#' \insertRef{austin2019}{hatchR}
-#'
-#' \insertRef{sparks2019}{hatchR}
+
 
 "model_table"

### 2024-12-05 R/predict_phenology.R

@@ -50,7 +50,7 @@
 #' )
 #'
 #' @references
-#' \insertRef{beacham1990}{hatchR}
+
 
 predict_phenology <- function(data, dates, temperature, spawn.date, model) {
   # assign data and arrange data by dates

### 2024-12-05 R/model_table.R
add refs

@@ -14,6 +14,22 @@
 #' @examples
 #' model_table
 #' @references
-
-
+#' Beacham, T.D., Murray, C.B. (1990). Temperature, egg size, and development
+#' of embryos and alevins of five species of Pacific salmon: a comparative
+#' analysis.
+#'   \emph{Canadian Journal of Zoology},
+#'   \bold{68}, 1931--1940.
+#'
+#' Sparks, M.M., Falke, J.A., Quinn, T.A., Adkinson, M.D.,
+#' Schindler, D.E. (2019). Influences of spawning timing, water temperature,
+#' and climatic warming on early life history phenology in western
+#' Alaska sockeye salmon.
+#'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
+#'   \bold{76(1)}, 123--135.
+#'
+#' Austin, C.C., Essington, T.E., Quinn, T.A. (2019). Spawning and emergence
+#' phenology of bull trout \emph{Salvelinus confluentus} under differing
+#' thermal regimes.
+#' \emph{Canadian Journal of Fisheries and Aquatic Sciences},
+#' \bold{94(1)}, 191--195.
 "model_table"

### 2024-12-05 R/predict_phenology.R

@@ -50,8 +50,12 @@
 #' )
 #'
 #' @references
-
-
+#' Sparks, M.M., Falke, J.A., Quinn, T.A., Adkinson, M.D.,
+#' Schindler, D.E. (2019). Influences of spawning timing, water temperature,
+#' and climatic warming on early life history phenology in western
+#' Alaska sockeye salmon.
+#'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
+#'   \bold{76(1)}, 123--135.
 predict_phenology <- function(data, dates, temperature, spawn.date, model) {
   # assign data and arrange data by dates
   dat <- data |>

### 2024-12-06 inst/manuscript/paper.Rmd
rewrite case study 1 for Shiny

@@
 dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
-             to = ymd_hms("2001-07-30 23:59:59"), length.out = 17568) 
+             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568) 
 # create empty dataframe
@@
 
-It appears there are no outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`.
+There are no obvious outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`.
 We also recommend using `plot_check_temp()` again on the summarized data.
@@
 Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
-Furthermore, we include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the original scope of the original effective value approach.
-We demonstrate parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the scope of the original effective value approach.
+We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
 
@@
 ###  make temp regime
-set.seed(1234)
+set.seed(123)
 # create random temps and corresponding dates
@@
 
-data_sim <- cbind (dates_sim, temps_sim) |> data.frame()
+data_sim <- matrix(NA, 1, 2) |> data.frame()
+data_sim[,1] <- temps_sim
+data_sim[,2] <- dates_sim
 
 ### smallmouth mod
-smallmouth <- matrix(NA, 10, 2) |> tibble()
+smallmouth <- matrix(NA, 10, 2) |> data.frame()
 colnames(smallmouth) <- c("hours", "temp_F")
@@
 ### catfish mod
-catfish <- matrix(NA, 3, 2) |> tibble()
+catfish <- matrix(NA, 3, 2) |> data.frame()
 colnames(catfish) <- c("days", "temp_C")
@@
 ### lake sturgeon mod
-sturgeon <-  matrix(NA, 7, 2) |> tibble()
+sturgeon <-  matrix(NA, 7, 2) |> data.frame()
 colnames(sturgeon) <- c("days", "CTU")
@@
 sturgeon <- sturgeon |> 
-  mutate(temp_C = CTU/days)
-
-plot(sturgeon$days ~ sturgeon$temp_C)
+  mutate(temp_C = CTU/days) # change CTUs to average temp and add column
 
@@
 
-In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the united states under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed.
+In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed.
 The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st.
@@
 
-```{r}
+We demonstrate this first case study using the graphical user interface portion of the `hatchR` ecosystem found at <https://elifelts.shinyapps.io/hatchR/>.
+Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
+Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check.
+After uploading and checking data, the user switches to the `Model Phenology` window.
+In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu.
+After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided.
+We show results for spawning for September 30th as indicated in the example above.
+Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab.
+Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
+The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on `hatchR`'s website <https://bmait101.github.io/hatchR/>.
+
+```{r, eval=FALSE, echo=FALSE}
 
@@
 
-We would like to sincerely thank the editor and reviewers for all of their helpful feedback which greatly improved both the software and the manuscript.
-
 The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.

### 2024-12-06 R/fit_model.R
reorder function output

@@ -118,11 +118,11 @@ fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
     model = m2,
     m2_a = m2_a,
     m2_b = m2_b,
-    func = func,
-    pred_plot = p_pred,
-    r_squared = r_squared,
     mse = mse,
-    rmse = rmse
+    rmse = rmse,
+    r_squared = r_squared,
+    func = func,
+    pred_plot = p_pred
   )
 
   return(out)

### 2024-12-06 R/fit_model.R
rename output elements

@@ -52,26 +52,26 @@ fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
   }
 
 
-  # fit linear model to log data
+  # fit linear model to estimate starting values for nls
   m1 <- stats::lm(log(y) ~ x, data = df)
   # summary(m1)
 
-  # estimate starting values for nls
+  # pull out starting values
   st <- list(
     a = exp(stats::coef(m1)[1]),
     b = stats::coef(m1)[2]
     )
 
-  # fit model 2 from Beacham & Murray (1990) to data
+  # fit model 2 from Beacham & Murray (1990) to data using nls
   m2 <- stats::nls(y ~ a / (x - b), data = df, start = st)
   # summary(m2)
 
   # get coefficients
-  m2_a <- log(stats::coef(m2)[1])
-  m2_b <- stats::coef(m2)[2]
+  log_a <- log(stats::coef(m2)[1])
+  b <- stats::coef(m2)[2]
 
-  # model specs for hatch function
-  func <- paste("1 / exp(", m2_a, " - log(x + ", m2_b * -1, "))", sep = "") # is *-1 correct?
+  # model expression and specs for predict_phenology()
+  func <- paste("1 / exp(", log_a, " - log(x + ", b * -1, "))", sep = "")
   func <- tibble::tibble(
     species = species,
     dev.type = dev.type,
@@ -116,8 +116,8 @@ fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
   # list of outputs ----------------------
   out <- list(
     model = m2,
-    m2_a = m2_a,
-    m2_b = m2_b,
+    log_a = log_a[[1]],
+    b = b[[1]],
     mse = mse,
     rmse = rmse,
     r_squared = r_squared,

### 2024-12-06 tests/testthat/test-fit_model.R
update element names

@@ -9,8 +9,8 @@ test_that("fit_model works", {
   expect_s3_class(m$func, "data.frame")
   # expect_length(m$func, 1)
 
-  expect_type(m$m2_b, "double")
-  expect_type(m$m2_a, "double")
+  expect_type(m$log_a , "double")
+  expect_type(m$b, "double")
   expect_type(m$r_squared, "double")
   expect_type(m$mse, "double")
   expect_type(m$rmse, "double")

### 2024-12-06 inst/manuscript/paper.Rmd
write multiple sections in Package Overview and clean up some style stuff
also render paper as JOSS just to see what it looks like as PDF

@@
 However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development.
-Despite these improvements, their application remains limited due to few parameterizations for different species, being largely limited to salmonids.
+Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
 Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
@@
 
-The resulting effective value approach has now been widely applied in salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 were have been applied to various species and populaions [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
+The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
 Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.
@@
 To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`.
-Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
 
 To widen the user application of these methods, we distribute two user-interfaces for `hatchR`.
-The first is a R package distributed via CRAN that allows users the most customizable application of these methods.
+The first is a R package distributed via CRAN that allows users the most customizable application for these methods.
 The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
@@
 nerka*).
-The need for change arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
+The need for their development arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
 To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.
@@
 
-Where *i* is the daily value and a fish hatches or emerges when the cumulative sum or $$\sum_{i =1}^nEffectiveValue_i = 1$$
+Where *i* is the daily value and a fish hatches or emerges when the cumulative sum reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$
 
-The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`.
+The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though the model table includes more complex models developed by @beacham1990) as well as for custom models users can fit with `fit_model()`.
 Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
@@
 
-We demonstrate the utility of the check summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year long dataset with temperature readings every thirty minutes.
+We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.
 
@@
              to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)
+
 # create empty dataframe
@@
 
-# take temps from a random normal dist with mean 10 sd 3 for every date time combo in dates and append to column (temp) in year_sim
+# take temps from a random normal dist with mean 10 sd 3 
+# for every date time combo in dates and append to column (temp) in year_sim
 year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
@@
                 temperature = daily_temp,
-                temp_min = 0, # temp_min and max lines are user customizable
+                temp_min = 0, # temp_min and max lines are 
+                              # user customizable
                 temp_max = 15)
@@
 
-```{r, eval=FALSE}
+```{r}
 ###  make temp regime
@@
 
-data_sim <- matrix(NA, 1, 2) |> data.frame()
+data_sim <- matrix(NA, 30, 2) |> data.frame()
 data_sim[,1] <- temps_sim
 data_sim[,2] <- dates_sim
+
+# change names so they aren't the same as the vector objects
+colnames(data_sim) <- c("temp_sim", "date_sim")
 
@@
 
+```
+
+Note the model the *R^2^* fit from the models below. You can see the generally all preform well and are in line with values from model 2 of @beacham1990.
+
+```{r}
+#model fits
+smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared
+
+```
+
+We additionally provide the model estimates for hatch timing for each of the three species used to generate custom models in Figure XXX.
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 ### predict_phenology
@@
 smb_res <- predict_phenology(data = data_sim,
-                  dates = dates_sim,
-                  temperature = temps_sim,
+                  dates = date_sim,
+                  temperature = temp_sim,
                   spawn.date = "2000-07-01",
@@
 catfish_res <- predict_phenology(data = data_sim,
-                  dates = dates_sim,
-                  temperature = temps_sim,
+                  dates = date_sim,
+                  temperature = temp_sim,
                   spawn.date = "2000-07-01",
@@
 sturgeon_res <- predict_phenology(data = data_sim,
-                  dates = dates_sim,
-                  temperature = temps_sim,
+                  dates = date_sim,
+                  temperature = temp_sim,
                   spawn.date = "2000-07-01",
@@
 
-```
-
-```{r, echo=FALSE}
 # summary for all species
@@
 colnames(all_res) <- c("start", "stop")
-all_res[1,] <- smb_res$dev.period
-all_res[2,] <- catfish_res$dev.period
-all_res[3,] <- sturgeon_res$dev.period
+
+all_res$start <- c(catfish_res$dev.period$start, 
+                   smb_res$dev.period$start, 
+                   sturgeon_res$dev.period$start)
+
+all_res$stop <- c(catfish_res$dev.period$stop,
+                  smb_res$dev.period$stop, 
+                  sturgeon_res$dev.period$stop)
 
 all_res <- all_res |> 
-  mutate(days = stop-start)
-all_res$Species <- c("Smallmouth Bass", "Channel Catfish", "Lake Sturgeon")
+  mutate(days = ceiling(stop-start),
+         index = c(17,16.5,16))
+
+all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")
+
+ggplot() +
+  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim )) + 
+  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim )) +
+  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax =index-.35, ymin = index-.5, fill = Species)) +
+  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index -0.425), label = days)) +
+  labs(x = "Date", y = "Temperature (°C)") +
+  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
+  theme_classic() +
+  theme(legend.position = c(0.75, 0.25))
 ```
@@
 
+We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+
 The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.

### 2024-12-06 R/predict_phenology.R
add suppresswarning to Ef.days

@@ -106,7 +106,7 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
 
   # Walk along temps and sum Ef to 1 and count how many days it takes
   #     If fish doesn't hatch value returns Inf
-  Ef.days <- min(which(Ef.cumsum >= 1))
+  Ef.days <- suppressWarnings(min(which(Ef.cumsum >= 1)))
 
   # Table of outs
   dat_out <- tibble::tibble(

### 2024-12-06 inst/manuscript/paper.Rmd
mess around with format ouput (e.g. JOSS)

@@
                 temperature = temp,
-                temp_min = 0, # temp_min and max lines are user customizable
+                temp_min = 0, # temp_min and max lines are 
+                              # user customizable
                 temp_max = 25)
@@
 There are no obvious outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`.
-We also recommend using `plot_check_temp()` again on the summarized data.
+We also recommend using `plot_check_temp()` again on the summarized data (though leave out the resulting plot for space efficiency in this manuscript).
 
@@
 
+# now a year's worth of single-day data
 dim(year_sim_summ)
 
-# check continuous
+# check continuous (no errors)
 check_continuous(data = year_sim_summ,
@@
 
-# we can demonstrate an error by removing the Oct 8 (100th day)
+# we can demonstrate an error by removing Oct. 8 (100th day)
 check_continuous(data = year_sim_summ[-100,],
@@
 
+```
+
+```{r, eval=FALSE}
 # it is useful to plot again to check your summarized data

### 2024-12-06 R/predict_phenology.R
add conditionals for model argument

@@ -9,8 +9,8 @@
 #' @param temperature Temperature measurements.
 #' @param spawn.date Date of spawning, given as a character string
 #' (e.g., "1990-08-18")
-#' @param model A data.frame giving model specifications. This must have a
-#' column providing a model expression. Can be obtained using `model_select()`
+#' @param model A data.frame with a column named "func" or a character vector
+#' giving model specifications. Can be obtained using `model_select()`
 #' or using you own data to obtain a model expression (see `fit_model`).
 #'
 #' @details
@@ -89,11 +89,24 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     dplyr::pull("index")
   dat_spawn <- dat[spawn.index:c(nrow(dat)), ]
 
+  # model prep
+  if (is.null(model)) {
+    cli::cli_abort("You must provide a model specification.")
+  } else if (is.data.frame(model) & !"func" %in% colnames(model)) {
+    cli::cli_abort("Model object must have a column named 'func'.")
+  } else if (is.data.frame(model) & "func" %in% colnames(model)) {
+    mod.exp <- model |> dplyr::pull("func")
+  } else if (is.character(model)) {
+    mod.exp <- model
+  } else {
+    cli::cli_abort("Model object must be a data.frame or character string.")
+  }
+
   # bring in model df and extract the expression
-  model.df <- model |> dplyr::pull("func")
+  # mod.exp <- model |> dplyr::pull("func")
 
   # Parse model expression to get effective value function
-  Ef <- parse(text = model.df)
+  Ef <- parse(text = mod.exp)
 
   # Vector of temps for Ef to evaluate
   x <- dat_spawn |> dplyr::pull({{ temperature }})

### 2024-12-06 tests/testthat/test-predict_phenology.R
update for new conditionals

@@ -9,7 +9,7 @@ test_that("predict phenology works", {
     data = woody_island,dates = date,
     temperature = temp_c,
     spawn.date = "1990-08-18",
-    model = m
+    model = m$func
     )
     )
 
@@ -25,8 +25,8 @@ test_that("predict phenology works", {
   expect_s3_class(p$dev.period, "data.frame")
   expect_equal(ncol(p$dev.period), 2)
 
-  expect_s3_class(p$model.specs, "data.frame")
-  expect_equal(ncol(p$model.specs), 5)
+  # expect_s3_class(p$model.specs, "data.frame")
+  # expect_equal(ncol(p$model.specs), 5)
 
 
 })

### 2024-12-06 inst/manuscript/paper.Rmd
lots of editing and additions to manuscript in the Package Overview section (specifically in the Model Selection and Predicting Phenology sections)

@@
 This close relationship has allowed researchers to generate statistical models that allow the prediction of developmental phenology with high accuracy and precision.
-These models were typically developed in aquaculture settings and their initial formulations were not applicable to wild populations because they assumed a constant temperature over the course of development @beacham1990 (**add more!**) .
+These models were typically developed in aquaculture settings and their initial formulations were not applicable to wild populations because they assumed a constant temperature over the course of development [@beacham1990; @mcphail1979; @alderdice1978].
 However, @sparks2019 reformulated this approach as an "Effective Value model", in which the input was daily average temperature after a parent spawned and fish would either hatch or emerge when effective values cumulatively summed to one.
@@
 For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
+Finally, we provide brief summaries of general applications of `hatchR` below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
 
@@
 
-Where *i* is the daily value and a fish hatches or emerges when the cumulative sum reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$
+Where *i* is the daily value and a fish hatches or emerges when the cumulative sum of effective values reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$
 
-The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though the model table includes more complex models developed by @beacham1990) as well as for custom models users can fit with `fit_model()`.
-Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
+The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
+Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
 
 ## Data format
+
+Water temperature datasets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures.
+Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.
+
+Fundamentally, `hatchR` assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade).
+Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required.
+We expect your data to look something like this:
+
+| date       | temperature |
+|------------|-------------|
+| 2000-01-01 | 2.51        |
+| ...        | ...         |
+| 2000-07-01 | 16.32       |
+| ...        | ...         |
+| 2000-12-31 | 3.13        |
+
+`hatchR` assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
+`hatchR` can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
+In these cases, we suggest users consider how much of that data type is reasonable with their data.
+
+For users choosing to implement `hatchR` in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
+For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.
 
@@
 `hatchR` is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
 
@@
 
-First, we recommend checking import data for any outliers or strange inputs using `plot_check_temp()`
+First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()`
 
@@
 
+#### Fitting models for other fishes
+
 Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
-We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the scope of the original effective value approach.
+We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
 We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
 
-We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime.
+We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure XXX).
 
@@
 
-We additionally provide the model estimates for hatch timing for each of the three species used to generate custom models in Figure XXX.
-
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
-Use woody example from website
+To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` dataset included in this package .
+We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`.
 
-show predict_phenology output slots
+```{r}
+sockeye_hatch_mod <- model_select(
+  author = "Beacham and Murray 1990", 
+  species = "sockeye", 
+  model = 2, 
+  dev.type = "hatch"
+  )
 
-show plot_phenology
+sockeye_emerge_mod <- model_select(
+  author = "Beacham and Murray 1990", 
+  species = "sockeye", 
+  model = 2, 
+  dev.type = "emerge"
+  )
+
+```
+
+We can now use our model expressions to predict when sockeye would hatch and emerge at Woody Island in 1990. First we predict hatch timing using `predict_phenology()`:
+
+```{r, warning=FALSE}
+WI_hatch <- predict_phenology(
+  data = woody_island,
+  dates = date,
+  temperature = temp_c,
+  spawn.date = "1990-08-18",
+  model = sockeye_hatch_mod
+  )
+```
+
+And then look inside the returned object to see days to hatch and development period:
+
+```{r}
+WI_hatch$days2done
+WI_hatch$dev.period
+```
+
+We can also do the same with emergence:
+
+```{r, warning=FALSE}
+WI_emerge <- predict_phenology(
+  data = woody_island,
+  dates = date,
+  temperature = temp_c,
+  spawn.date = "1990-08-18",
+  model = sockeye_emerge_mod   # notice we're using emerge model expression here
+  )
+
+# see days to hatch and development period
+WI_emerge$days2done
+WI_emerge$dev.period
+```
+
+#### Understanding your results
+
+The output from `predict_phenology()` includes a lot of information.
+If we look at our `WI_hatch` object we see there are multiple elements stored in a list which can be accessed using the `$` operator.
+
+```{r, eval=FALSE}
+str(WI_hatch)
+```
+
+`WI_hatch$days2done` outputs the predicted days to hatch or emerge.
+
+`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to hatch or emerge.
+
+`WI_hatch$ef.vals` is a vector of each day's effective value as evaluated using whatever model is chosen.
+
+`WI_hatch$ef.tibble` is a *n* x 4 tibble (*n* = number of days to hatch or emerge) and the columns are the date, each day's temperature and effective value, and the cumulative sum of the effective values.
+The `ef.tibble` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.
+
+#### Plotting phenology
+
+**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results.
+The plot visualizes three specific components: 1.
+)the temperature regime over which you are predicting, 2.) the cumulative sum of effective values, and 3.) the effective value for each day in your prediction span.
+The function allows you to output various figures based on your interests, but defaults to a figure with all information and the corresponding labels.
+
+```{r}
+plot_phenology(WI_hatch)
+```
 

### 2024-12-06 README.Rmd
chnage `hatchR` to **hatchR**

@@
 
-`hatchR` is an R package for predicting fish developmental phenology using statistical models.
+**hatchR** is an R package for predicting fish developmental phenology using statistical models.
 It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
@@
 
-If you use `hatchR` in a formal publication or report, please cite it. Citing `hatchR` lets us devote more resources to it in the future. View the `hatchR` citation by running
+If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running
 ```r
@@
 
-You can install the development version of `hatchR` from [GitHub](https://github.com/) with:
+You can install the development version of **hatchR** from [GitHub](https://github.com/) with:
 
@@
 
-## Contributing to `hatchR`
+## Contributing to **hatchR**
 
-We encourage users to report bugs and/or contribute to `hatchR`. For more detail on how to do this, please see our contributing guide (`CONTRIBUTING.md`).
+We encourage users to report bugs and/or contribute to **hatchR**. For more detail on how to do this, please see our contributing guide (`CONTRIBUTING.md`).
 
@@
 
-There are several ways to get help with `hatchR`:
+There are several ways to get help with **hatchR**:
 
@@
 
-Below we provide a brief example showing how to use `hatchR`.
+Below we provide a brief example showing how to use **hatchR**.
 For a thorough introduction to the software, see our introductory vignette [linked here](https://bmait101.github.io/hatchR/articles/Introduction.html).
-For a list of all functions available in `hatchR`, see our function reference [linked here](https://bmait101.github.io/hatchR/reference/index.html).
+For a list of all functions available in **hatchR**, see our function reference [linked here](https://bmait101.github.io/hatchR/reference/index.html).
 
-First we load `hatchR` and helper packages by running
+First we load **hatchR** and helper packages by running
 
@@
 
-The example `woody_island` dataset is included in `hatchR` and contains temperature data from Woody Island, Lake Iliamna, Alaska.
+The example `woody_island` dataset is included in **hatchR** and contains temperature data from Woody Island, Lake Iliamna, Alaska.
 It includes daily water temperature data from 1990-1992.
@@
 
-`hatchR` suggests the following **R** packages:
+**hatchR** suggests the following **R** packages:
 
@@
 
-`hatchR` project code is provided on an "as is" basis and the user assumes responsibility for its use.
+**hatchR** project code is provided on an "as is" basis and the user assumes responsibility for its use.
 USDA Forest Service has relinquished control of the information and no longer has responsibility to protect the integrity , confidentiality, or availability of the information.

### 2024-12-06 inst/manuscript/paper.Rmd
chnage `hatchR` to **hatchR**

--- 2024-12-06 inst/manuscript/paper.Rmd
@@
 However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development.
-Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
-Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
-`hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
-`hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
+Despite these improvements, their application remains limited due to few parameterizations for different species, being largely limited to salmonids.
+Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
+**hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
+**hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
 To demonstrate potential uses, we present two case studies as likely applications for this software.
-`hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.
+**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
 
@@
 
-The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
+The resulting effective value approach has now been widely applied in salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 were have been applied to various species and populaions [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
 Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.
 
-To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`.
-Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**.
+Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
 
-To widen the user application of these methods, we distribute two user-interfaces for `hatchR`.
-The first is a R package distributed via CRAN that allows users the most customizable application for these methods.
+To widen the user application of these methods, we distribute two user-interfaces for **hatchR**.
+The first is a R package distributed via CRAN that allows users the most customizable application of these methods.
 The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
-These variable approaches are outlined in the package documentation on `hatchR`'s website.
-Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface.
-The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package.
+These variable approaches are outlined in the package documentation on **hatchR**'s website.
+Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
+The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
 Below, we present the basic overview of the software and multiple case studies of how it may be applied.
-
-```{r echo=FALSE}
-### 
-library(tidyverse); library(hatchR)
-```
 
@@
 
-`hatchR` is meant to primarily be a tool for predicting phenology.
+**hatchR** is meant to primarily be a tool for predicting phenology.
 In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help.
@@
 nerka*).
-The need for their development arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
+The need for change arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
 To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.
@@
 
-Where *i* is the daily value and a fish hatches or emerges when the cumulative sum reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$
+Where *i* is the daily value and a fish hatches or emerges when the cumulative sum or $$\sum_{i =1}^nEffectiveValue_i = 1$$
 
-The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though the model table includes more complex models developed by @beacham1990) as well as for custom models users can fit with `fit_model()`.
+The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`.
 Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
@@
 
-`hatchR` is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
+**hatchR** is built assuming data will be analyzed as daily average temperatures.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
 
-We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.
-
-```{r}
-# create date object for a year with 30 min reading intervals
-dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
-             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)
-
-# create empty dataframe
-year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))
-
-# date column
-colnames(year_sim)[1] <- "date"
-
-# add dates vector to date column
-year_sim[1] <- dates
-
-#random seed
-set.seed(123)
-
-# take temps from a random normal dist with mean 10 sd 3 
-# for every date time combo in dates and append to column (temp) in year_sim
-year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
-  abs()
-
-dim(year_sim)
-
-```
-
-First, we recommend checking import data for any outliers or strange inputs using `plot_check_temp()`
-
-```{r}
-plot_check_temp(data = year_sim,
-                dates = date,
-                temperature = temp,
-                temp_min = 0, # temp_min and max lines are user customizable
-                temp_max = 25)
-```
-
-There are no obvious outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`.
-We also recommend using `plot_check_temp()` again on the summarized data.
-
-```{r}
-# summarize
-year_sim_summ <- summarize_temp(data = year_sim,
-                                dates = date,
-                                temperature = temp)
-
-dim(year_sim_summ)
-
-# check continuous
-check_continuous(data = year_sim_summ,
-                 dates = date)
-
-# we can demonstrate an error by removing the Oct 8 (100th day)
-check_continuous(data = year_sim_summ[-100,],
-                 dates = date)
-
-# it is useful to plot again to check your summarized data
-plot_check_temp(data = year_sim_summ,
-                dates = date,
-                temperature = daily_temp,
-                temp_min = 0, # temp_min and max lines are 
-                              # user customizable
-                temp_max = 15)
-```
+summarize_temp, check_continuous and plot_check_temp
 
@@
 
-fit_model for three non-salmonid specieslib
+fit_model for three non-salmonid species
 
-Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
-We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the scope of the original effective value approach.
-We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
-
-We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime.
-
-```{r}
-###  make temp regime
-set.seed(123)
-# create random temps and corresponding dates
-temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
-dates_sim <-  seq(from = ymd("2000-07-01"),
-             to = ymd("2000-07-31"), length.out = 30)
-
-data_sim <- matrix(NA, 30, 2) |> data.frame()
-data_sim[,1] <- temps_sim
-data_sim[,2] <- dates_sim
-
-# change names so they aren't the same as the vector objects
-colnames(data_sim) <- c("temp_sim", "date_sim")
-
-### smallmouth mod
-smallmouth <- matrix(NA, 10, 2) |> data.frame()
+```{r, eval=FALSE}
+smallmouth <- matrix(NA, 10, 2) |> tibble()
 colnames(smallmouth) <- c("hours", "temp_F")
@@
 
-# change F to C and hours to days
 smallmouth <- smallmouth |> 
@@
 
-smb_mod <- fit_model(temp = smallmouth$temp_C,
-                     days = smallmouth$days,
-                     species = "smb",
-                     dev.type = "hatch")
+plot(smallmouth$days~smallmouth$temp_C)
 
-### catfish mod
-catfish <- matrix(NA, 3, 2) |> data.frame()
-colnames(catfish) <- c("days", "temp_C")
-catfish$days <- c(16,21,26)
-catfish$temp_C <- c(22,10,7)
+smb_mod <- fit_model(smallmouth,
+                     temp = smallmouth$temp_C,
+                     days = smallmouth$hours)
 
-cat_mod <- fit_model(temp = catfish$temp_C,
-                     days = catfish$days,
-                     species = "catfish",
-                     dev.type = "hatch")
-
-### lake sturgeon mod
-sturgeon <-  matrix(NA, 7, 2) |> data.frame()
-colnames(sturgeon) <- c("days", "CTU")
-sturgeon$days <- c(7,5,6,6,5,11,7)
-sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)
-
-sturgeon <- sturgeon |> 
-  mutate(temp_C = CTU/days) # change CTUs to average temp and add column
-
-sturgeon_mod <- fit_model(days = sturgeon$days,
-                          temp = sturgeon$temp_C,
-                          species = "sturgeon",
-                          dev.type = "hatch")
-
-```
-
-Note the model the *R^2^* fit from the models below. You can see the generally all preform well and are in line with values from model 2 of @beacham1990.
-
-```{r}
-#model fits
-smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared
-
-```
-
-We additionally provide the model estimates for hatch timing for each of the three species used to generate custom models in Figure XXX.
-
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-### predict_phenology
-
-#smallmouth bass
-smb_res <- predict_phenology(data = data_sim,
-                  dates = date_sim,
-                  temperature = temp_sim,
-                  spawn.date = "2000-07-01",
-                  model = smb_mod$func)
-
-# catfish
-catfish_res <- predict_phenology(data = data_sim,
-                  dates = date_sim,
-                  temperature = temp_sim,
-                  spawn.date = "2000-07-01",
-                  model = cat_mod$func)
-
-# sturgeon
-# note that 16 C is pretty far out of range of temps for model fit, not best practice
-sturgeon_res <- predict_phenology(data = data_sim,
-                  dates = date_sim,
-                  temperature = temp_sim,
-                  spawn.date = "2000-07-01",
-                  model = sturgeon_mod$func)
-
-# summary for all species
-all_res <- data.frame(matrix(NA, 3, 2))
-colnames(all_res) <- c("start", "stop")
-
-all_res$start <- c(catfish_res$dev.period$start, 
-                   smb_res$dev.period$start, 
-                   sturgeon_res$dev.period$start)
-
-all_res$stop <- c(catfish_res$dev.period$stop,
-                  smb_res$dev.period$stop, 
-                  sturgeon_res$dev.period$stop)
-
-all_res <- all_res |> 
-  mutate(days = ceiling(stop-start),
-         index = c(17,16.5,16))
-
-all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")
-
-ggplot() +
-  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim )) + 
-  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim )) +
-  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax =index-.35, ymin = index-.5, fill = Species)) +
-  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index -0.425), label = days)) +
-  labs(x = "Date", y = "Temperature (°C)") +
-  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
-  theme_classic() +
-  theme(legend.position = c(0.75, 0.25))
 ```
@@
 
-Use woody example from website
+Use woody \_example from website
 
@@
 
-In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed.
+In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the united states under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed.
 The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st.
@@
 
-We demonstrate this first case study using the graphical user interface portion of the `hatchR` ecosystem found at <https://elifelts.shinyapps.io/hatchR/>.
-Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
-Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check.
-After uploading and checking data, the user switches to the `Model Phenology` window.
-In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu.
-After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided.
-We show results for spawning for September 30th as indicated in the example above.
-Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab.
-Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on `hatchR`'s website <https://bmait101.github.io/hatchR/>.
-
-```{r, eval=FALSE, echo=FALSE}
+```{r}
 
@@
 
-We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+We would like to sincerely thank the editor and reviewers for all of their helpful feedback which greatly improved both the software and the manuscript.
 

### 2024-12-06 R/fit_model.R

@@ -5,7 +5,7 @@
 #' Generate your own custom parameterized models for predicting hatching and emergence phenology.
 #'
 #' @details
-#' `hatchR` also includes functionality to generate your own custom
+#' **hatchR** also includes functionality to generate your own custom
 #' parameterized models for predicting hatching and emergence phenology.
 #' Importantly, the custom parameterization relies on the model format
 #' developed from model 2 of Beacham and Murray (1990), which we chose

### 2024-12-06 R/model_select.R

@@ -6,7 +6,7 @@
 #' different species from different studies built in. Refer to the
 #' table (`model_table`) before using function to find inputs for
 #' the different function arguments. It pulls the model format as a
-#' string and parses it to be usable in `hatchR` model.
+#' string and parses it to be usable in **hatchR** model.
 #'
 #' @param author Character string of author name.
 #' @param species Character string of species name.

### 2024-12-06 README.Rmd
update README

@@
 
-# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
+# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" alt="hatchR website" align="right" height="139"/></a>
 
 <!-- badges: start -->
+
 [![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
+
 <!-- badges: end -->
 
-**IN DEVELOPMENT**
-
 **hatchR** is an R package for predicting fish developmental phenology using statistical models.
-It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
+It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (*e.g.*, Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
 (2019).
@@
 If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running
-```r
+
+```{r}
 citation(package = "hatchR")
@@
 
-``` r
+```{r}
 # install.packages("pak")
@@
 1.  Open a GitHub issue [link here](https://github.com/bmait101/hatchR/issues).
-2. Email the hatchR support team ([Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov))
+2.  Email the **hatchR** support team ([Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov))
 3.  Post on a support website like Stack Overflow or Cross Validated.
@@
 It includes daily water temperature data from 1990-1992.
-We can use `plot_check_temp()` to visually check the data:
+We can use `plot_check_temp()` to visually check the data.
+Spawning in this system typically peaks around August 18 and hatching and emergence are done before the start of the following spawning season, so we can predict phenology within a subset of a year.
+Because `plot_check_temp()` returns a **ggplot2** object, we can add additional **ggplot2** elements to customize the plots.
 
 ```{r}
-p <- plot_check_temp(
+plot_check_temp(
   data = woody_island,
@@
   temperature = temp_c
-  )
-p
-```
-
-Spawning in this system typically peaks around August 18 and hatching and emergence are done before the start of the following spawning season, so we can predict phenology within a subset of a year.
-
-```{r}
-p + 
+  ) +  # additional ggplot2 elements below this `+`
   geom_rect(
@@
       ymax = 25),  # upper bound
-    fill = "darkgreen",  # color rectangle dark green
-    alpha = 0.01  # make rectangle semi-transparent
+    fill = "grey",
+    alpha = 0.01
     ) 
@@
 
-* dplyr: For data manipulation.
-* ggplot2: For vignette visualizations.
-* lubridate: For working with dates.
-* tibble: For creating tibbles as output for various functions.
-* rlang: For `.data` usage.
-* stats: For various modeling functions (e.g., `confint()`).
-* utils: For various utility functions.
+-   dplyr: For data manipulation.
+-   ggplot2: For vignette visualizations.
+-   lubridate: For working with dates.
+-   tibble: For creating tibbles as output for various functions.
+-   rlang: For `.data` usage.
+-   stats: For various modeling functions (e.g., `confint()`).
+-   lifecycle: for function lifecycle badges.
+-   utils: For various utility functions.
 
@@
 
-* knitr: For vignette building.
-* readr: For reading in data.
-* rmarkdown: For vignette building.
-* testthat: For unit testing.
+-   knitr: For vignette building.
+-   readr: For reading in data.
+-   purrr: For iteration in vignettes.
+-   rmarkdown: For vignette building.
+-   nycflights13: For example data.
+-   tidyr: For vignette building.
+-   stringr: For vignette building.
+-   ggridges: For vignette building.
+-   patchwork: For vignette building.
+-   cli: For better messages.
+-   testthat: For unit testing.
 

### 2024-12-06 R/plot_phenology.R
update labels

@@ -25,20 +25,35 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   if (labels == TRUE) {
     all_label <- ggplot2::labs(
       x = "Date", y = "Mean daily temperature",
-      title = paste(dat$days2done, "days to hatch/emerge"),
-      subtitle = paste("Fish spawned", dat$dev.period$start, "and hatched/emerged", dat$dev.period$stop, "\nblue = temp,  dark green = scaled cumulative ef_val, light green = daily ef_val(x100)")
+      title = paste(dat$days2done, "days to develop"),
+      subtitle = paste(
+        "Fish spawned",
+        dat$dev.period$start,
+        "and developed",
+        dat$dev.period$stop,
+        "\nblue = temp,  dark green = scaled cumulative ef_val, light green = daily ef_val(x100)")
     )
 
     ef_cumsum_label <- ggplot2::labs(
       x = "Date", y = "Mean daily temperature",
-      title = paste(dat$days2done, "days to hatch/emerge"),
-      subtitle = paste("Fish spawned", dat$dev.period$start, "and hatched/emerged", dat$dev.period$stop, "\nblue = temperature, green = scaled cumulative effective value")
+      title = paste(dat$days2done, "days to develop"),
+      subtitle = paste(
+        "Fish spawned",
+        dat$dev.period$start,
+        "and developed",
+        dat$dev.period$stop,
+        "\nblue = temperature, green = scaled cumulative effective value")
     )
 
     ef_daily_label <- ggplot2::labs(
       x = "Date", y = "Mean daily temperature",
-      title = paste(dat$days2done, "days to hatch/emerge"),
-      subtitle = paste("Fish spawned", dat$dev.period$start, "and hatched/emerged", dat$dev.period$stop, "\nblue = temperature,  green = daily effective value (x100)")
+      title = paste(dat$days2done, "days to develop"),
+      subtitle = paste(
+        "Fish spawned",
+        dat$dev.period$start,
+        "and developed",
+        dat$dev.period$stop,
+        "\nblue = temperature,  green = daily effective value (x100)")
     )
   }
 

### 2024-12-09 inst/manuscript/paper.Rmd
fix merge conflicts in manuscript

--- 2024-12-06 inst/manuscript/paper.Rmd
@@
 
-To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`.
-Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**.
+Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
 
-To widen the user application of these methods, we distribute two user-interfaces for `hatchR`.
-The first is a R package distributed via CRAN that allows users the most customizable application for these methods.
+To widen the user application of these methods, we distribute two user-interfaces for **hatchR**.
+The first is a R package distributed via CRAN that allows users the most customizable application of these methods.
 The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
-These variable approaches are outlined in the package documentation on `hatchR`'s website.
-Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface.
-The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package.
+These variable approaches are outlined in the package documentation on **hatchR**'s website.
+Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
+The Shiny form trades-off some of the automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
 Below, we present the basic overview of the software and multiple case studies of how it may be applied.
@@
 For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
-Finally, we provide brief summaries of general applications of `hatchR` below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
+Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables and making more complex plots.
 
@@
 
-The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
-Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
+The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`.
+Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
 
@@
 
-Fundamentally, `hatchR` assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade).
+Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade).
 Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required.
@@
 
-`hatchR` assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
-`hatchR` can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
+**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
+**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
 In these cases, we suggest users consider how much of that data type is reasonable with their data.
 
-For users choosing to implement `hatchR` in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
+For users choosing to implement **hatchR** in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
 For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.
@@
 
-`hatchR` is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
+**hatchR** is built assuming data will be analyzed as daily average temperatures.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
 
@@
 
-We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
 

### 2024-12-09 inst/manuscript/paper.Rmd
added captions to figures and referenced them in paper

@@
 For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
-Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables and making more complex plots.
+Finally, we provide brief summaries of general applications of **hatchR** below and include a visual representation of the full package workflow (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables and making more complex plots.
 
-```{r echo = FALSE, out.height = '50%', fig.align = 'center'}
+```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap = "Generalized workflow for hatchR.
+Data actions are depicted with beige circles while functions are represented with white rectangles."}
 knitr::include_graphics("flowchart.png")
@@
 **hatchR** is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
 
@@
 
-```{r}
+```{r, fig.cap = "Graphical output from plot_check_temp(), a visual data check tool in the package. Note that the upper (red) and lower (blue) bound lines  can be custom set by the user."}
+
 plot_check_temp(data = year_sim,
@@
 We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
 
-We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure XXX).
+We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure 3).
 
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Predicted days to hatch for Channel Catfish (blue), Lake Sturgeon (black), and Smallmouth Bass (green) using custom paramerizations from hatchR's fit_model() function.
+A simulated thermal regime with an ascending thermograph and a mean temperature of 16 °C is used to demonstate the different parameterizations.
+All fish spawn the same day, July 1."}
 ### predict_phenology
@@
 
-**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results.
+**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results (Figure 4).
 The plot visualizes three specific components: 1.
@@
 
-```{r}
+```{r, fig.cap = "Output from hatchR's function, plot_phenology().
+The output is meant to provide a basic summary of your model using a single spawn date.
+Users are encouraged to visit the packages help page where articles demonstrate more advanced, custom plotting applications."}
 plot_phenology(WI_hatch)
@@
 
+talk about how temperatures may not always reflect redd temperture
+
+talk about how other factors like oxygen or predator cues may change phenology
+
+talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns
+
+talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.
+
 # Acknowledgements

### 2024-12-09 inst/manuscript/paper.Rmd
copy edits

@@
 # take temps from a random normal dist with mean 10 sd 3 
-# for every date time combo in dates and append to column (temp) in year_sim
+# for every date time combo in dates and append to column (temp) 
+# in year_sim
 year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
@@
 Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
-We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+We provide parameterizations for three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
+The parameterizations are for commonly cultured sport- and commercial fishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
 
@@
 
-Note the model the *R^2^* fit from the models below. You can see the generally all preform well and are in line with values from model 2 of @beacham1990.
+Note the *R^2^* fit from the models below. You can see they generally all preform well and are in line with values from model 2 of @beacham1990.
 
@@
   )
-
-sockeye_emerge_mod <- model_select(
-  author = "Beacham and Murray 1990", 
-  species = "sockeye", 
-  model = 2, 
-  dev.type = "emerge"
-  )
-
 ```
@@
 
-`WI_hatch$days2done` outputs the predicted days to hatch or emerge.
+`WI_hatch$days2done` outputs the predicted days to develop.
 
-`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to hatch or emerge.
+`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted todevelop.
 

### 2024-12-09 inst/manuscript/paper.Rmd
copy edit

@@
 **hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results (Figure 4).
-The plot visualizes three specific components: 1.
-)the temperature regime over which you are predicting, 2.) the cumulative sum of effective values, and 3.) the effective value for each day in your prediction span.
+The plot visualizes three specific components: 1.) the temperature regime over which you are predicting, 2.) the cumulative sum of effective values, and 3.) the effective value for each day in your prediction span.
 The function allows you to output various figures based on your interests, but defaults to a figure with all information and the corresponding labels.
@@
 Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on `hatchR`'s website <https://bmait101.github.io/hatchR/>.
+The process is demonstrated in full in Figure XXX, but the interface is described more completely on the application's landing page <https://elifelts.shinyapps.io/hatchR/>.
 

### 2024-12-09 R/crooked_river.R
redoc

@@ -1,13 +1,13 @@
-#' Daily stream temperatures data
+#' Example dataset: Crooked River, Idaho
 #'
-#' Daily stream temperature data from the US Forest Service Rocky
-#' Mountain Research Station, Water and Watersheds Program.
+#' @description
+#' Daily stream temperature data from Crooked River, a tributary to the
+#' North Fork Boise River, Idaho.
 #'
-#' @source Dan Isaak, US Forest Service Rocky Mountain Research Station
+#' @source Dan Isaak, US Forest Service Rocky Mountain Research Station.
 #' @format Data frame with columns
 #' \describe{
-#' \item{SampleDate}{Date measurement was taken.}
-#' \item{site}{Site ID.}
+#' \item{date}{Date and time measurement was taken.}
 #' \item{temperture}{Temperature in degrees Celsius.}
 #' }
 #' @examples

### 2024-12-09 R/woody_island.R
redoc

@@ -1,17 +1,21 @@
-#' Temperature Data from Lake Iliamna, Alaska
+#' Example dataset: Lake Iliamna, Alaska
 #'
 #' @description
 #' Temperature Data from Lake Iliamna on Woody Island, Alaska
 #'
-#' @source To come...
-#' @format Data frame
+#' @source Sparks et al. (2019)
+#' @format Data frame with columns:
 #' \describe{
-#' \item{author}{Short article citation}
-#' \item{species}{Species name}
-#' \item{model}{Model number from B&M}
-#' \item{dev.type}{Hatch or emergence}
-#' \item{func}{character string desciption of the function}
+#' \item{date}{Date measurement was taken.}
+#' \item{temp_c}{Temperature in degrees Celsius.}
 #' }
 #' @examples
 #' woody_island
+#' @references
+#' Sparks, M.M., Falke, J.A., Quinn, T.A., Adkinson, M.D.,
+#' Schindler, D.E. (2019). Influences of spawning timing, water temperature,
+#' and climatic warming on early life history phenology in western
+#' Alaska sockeye salmon.
+#'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
+#'   \bold{76(1)}, 123--135.
 "woody_island"

### 2024-12-09 inst/manuscript/paper.Rmd
rename model_tabel arguments

@@
                      species = "smb",
-                     dev.type = "hatch")
+                     development_type = "hatch")
 
@@
                      species = "catfish",
-                     dev.type = "hatch")
+                     development_type = "hatch")
 
@@
                           species = "sturgeon",
-                          dev.type = "hatch")
+                          development_type = "hatch")
 
@@
                   spawn.date = "2000-07-01",
-                  model = smb_mod$func)
+                  model = smb_mod$expression)
 
@@
                   spawn.date = "2000-07-01",
-                  model = cat_mod$func)
+                  model = cat_mod$expression)
 
@@
                   spawn.date = "2000-07-01",
-                  model = sturgeon_mod$func)
+                  model = sturgeon_mod$expression)
 
@@
   model = 2, 
-  dev.type = "hatch"
+  development_type = "hatch"
   )
@@
                             model = "MM",
-                            dev.type = "emerge")
+                            development_type = "emerge")
 

### 2024-12-09 R/fit_model.R

@@ -17,7 +17,7 @@
 #' @param temp Numeric vector of temperatures
 #' @param days Numeric vector of days to hatch or emerge
 #' @param species Character string of species name (e.g., "sockeye")
-#' @param dev.type Character string of development type: "hatch" or "emerge"
+#' @param development_type Character string of development type: "hatch" or "emerge"
 #'
 #' @return List with fit model object, model coefficients, model specifications
 #' data.frame, and plot of observations and model fit.
@@ -31,8 +31,8 @@
 #' # vector of days to hatch
 #' days_to_hatch <- c(194, 87, 54, 35, 28)
 #' bt_hatch_mod <- fit_model(temp = temperature,
-#' days = days_to_hatch, species = "sockeye", dev.type = "hatch")
-fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
+#' days = days_to_hatch, species = "sockeye", development_type = "hatch")
+fit_model <- function(temp, days, species = NULL, development_type = NULL) {
   df <- tibble::tibble(x = temp, y = days)
 
   # check if species is NULL
@@ -43,11 +43,11 @@ fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
       ))
   }
 
-  # check if dev.type is NULL
-  if (is.null(dev.type)) {
+  # check if development_type is NULL
+  if (is.null(development_type)) {
     cli::cli_abort(c(
-            "`dev.type` cannot be NULL",
-      "i" = "Provide a dev.type name using `dev.type = 'hatch' or `dev.type = 'emerge'`."
+            "`development_type` cannot be NULL",
+      "i" = "Provide a development_type name using `development_type = 'hatch' or `development_type = 'emerge'`."
     ))
   }
 
@@ -71,11 +71,11 @@ fit_model <- function(temp, days, species = NULL, dev.type = NULL) {
   b <- stats::coef(m2)[2]
 
   # model expression and specs for predict_phenology()
-  func <- paste("1 / exp(", log_a, " - log(x + ", b * -1, "))", sep = "")
-  func <- tibble::tibble(
+  expression <- paste("1 / exp(", log_a, " - log(x + ", b * -1, "))", sep = "")
+  expression <- tibble::tibble(
     species = species,
-    dev.type = dev.type,
-    func = func
+    development_type = development_type,
+    expression = expression
     )
 
   # plot predictions and data --------------------------
[…]

### 2024-12-09 R/model_select.R

@@ -10,8 +10,8 @@
 #'
 #' @param author Character string of author name.
 #' @param species Character string of species name.
-#' @param model Model number from Beacham and Murray (1990).
-#' @param dev.type The phenology type. A vector with possible values "hatch" or "emerge". The default is "hatch".
+#' @param model_id Model number from Beacham and Murray (1990).
+#' @param development_type The phenology type. A vector with possible values "hatch" or "emerge". The default is "hatch".
 #'
 #' @return A data.frame giving model specifications to be passed to `predict_phenology()`.
 #'
@@ -24,24 +24,20 @@
 #' sockeye_hatch_mod <- model_select(
 #'   author = "Beacham and Murray 1990",
 #'   species = "sockeye",
-#'   model = 2,
-#'   dev.type = "hatch"
+#'   model_id = 2,
+#'   development_type = "hatch"
 #' )
 #' # print
 #' sockeye_hatch_mod
 model_select <- function(author,
                          species,
-                         model,
-                         dev.type = "hatch") {
+                         model_id,
+                         development_type = "hatch") {
   mod <- model_table |>
     dplyr::filter(
       author == {{ author }} &
         species == {{ species }} &
-        model == {{ model }} &
-        dev.type == {{ dev.type }})
-  #   ) |>
-  #   dplyr::pull("func")
-  #
-  # mod <- parse(text = mod)
-  # return(mod)
+        model_id == {{ model_id }} &
+        development_type == {{ development_type }}
+    )
 }

### 2024-12-09 R/model_table.R

@@ -1,15 +1,15 @@
 #' Table of phenology models
 #'
-#' Description to come...
+#' Table providing the model parameterizations for the phenology models.
 #'
-#' @source To come...
-#' @format Data frame
+#' @source Beacham & Murray (1990), Sparks et al. (2019), Austin et al. (2019).
+#' @format Data frame with 5 columns:
 #' \describe{
-#' \item{author}{Short article citation}
+#' \item{author}{Author-year short citation}
 #' \item{species}{Species name}
-#' \item{model}{Model number from B&M}
-#' \item{dev.type}{Hatch or emergence}
-#' \item{func}{character string desciption of the function}
+#' \item{model_id}{Identification number for a model parameterization}
+#' \item{development_type}{Hatch or emergence}
+#' \item{expression}{character string of parameterized function expression}
 #' }
 #' @examples
 #' model_table

### 2024-12-09 R/predict_phenology.R

@@ -9,7 +9,7 @@
 #' @param temperature Temperature measurements.
 #' @param spawn.date Date of spawning, given as a character string
 #' (e.g., "1990-08-18")
-#' @param model A data.frame with a column named "func" or a character vector
+#' @param model A data.frame with a column named "expression" or a character vector
 #' giving model specifications. Can be obtained using `model_select()`
 #' or using you own data to obtain a model expression (see `fit_model`).
 #'
@@ -36,8 +36,8 @@
 #' sockeye_hatch_mod <- model_select(
 #'   author = "Beacham and Murray 1990",
 #'   species = "sockeye",
-#'   model = 2,
-#'   dev.type = "hatch"
+#'   model_id = 2,
+#'   development_type = "hatch"
 #' )
 #'
 #' # predict phenology
@@ -92,10 +92,10 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
   # model prep
   if (is.null(model)) {
     cli::cli_abort("You must provide a model specification.")
-  } else if (is.data.frame(model) & !"func" %in% colnames(model)) {
-    cli::cli_abort("Model object must have a column named 'func'.")
-  } else if (is.data.frame(model) & "func" %in% colnames(model)) {
-    mod.exp <- model |> dplyr::pull("func")
+  } else if (is.data.frame(model) & !"expression" %in% colnames(model)) {
+    cli::cli_abort("Model object must have a column named 'expression'.")
+  } else if (is.data.frame(model) & "expression" %in% colnames(model)) {
+    mod.exp <- model |> dplyr::pull("expression")
   } else if (is.character(model)) {
     mod.exp <- model
   } else {
@@ -103,7 +103,7 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
   }
 
   # bring in model df and extract the expression
-  # mod.exp <- model |> dplyr::pull("func")
+  # mod.exp <- model |> dplyr::pull("expression")
 
   # Parse model expression to get effective value function
   Ef <- parse(text = mod.exp)

### 2024-12-09 data-raw/model_table.R

@@ -1,5 +1,3 @@
-library(readr)
-
-model_table <- read_csv("data-raw/model_table.csv")
+model_table <- readr::read_csv("data-raw/model_table.csv")
 
 usethis::use_data(model_table, overwrite = TRUE)

### 2024-12-09 inst/misc/interactive_checking.R

@@ -5,7 +5,7 @@ sockeye_hatch_mod <- model_select(
   author = "Beacham and Murray 1990",
   species = "sockeye",
   model = 2,
-  dev.type = "hatch"
+  development_type = "hatch"
 )
 # predict phenology
 sockeye_hatch <- predict_phenology(
@@ -22,7 +22,7 @@ temp <- c(2, 5, 8, 11, 14)
 days <- c(194, 87, 54, 35, 28)
 hatch <- tibble::tibble(temp, days)
 hatch <- fit_model(df = hatch, temp = temp, days = days, species = "fish")
-hatch$func
+hatch$expression
 
 crooked_river$date <- as.character(crooked_river$date)
 crooked_river$date <- as.Date(crooked_river$date)

### 2024-12-09 tests/testthat/test-fit_model.R

@@ -2,12 +2,12 @@ test_that("fit_model works", {
   temp <- c(2, 5, 8, 11, 14)
   days <- c(194, 87, 54, 35, 28)
   df <- tibble::tibble(temp, days)
-  m <- fit_model(temp = temp, days = days, species = "sockeye", dev.type = "hatch")
+  m <- fit_model(temp = temp, days = days, species = "sockeye", development_type = "hatch")
 
   expect_type(m, "list")
 
-  expect_s3_class(m$func, "data.frame")
-  # expect_length(m$func, 1)
+  expect_s3_class(m$expression, "data.frame")
+  # expect_length(m$expression, 1)
 
   expect_type(m$log_a , "double")
   expect_type(m$b, "double")

### 2024-12-09 tests/testthat/test-predict_phenology.R

@@ -2,14 +2,14 @@ test_that("predict phenology works", {
   m <- model_select(
     author = "Beacham and Murray 1990",
     species = "sockeye",
-    model = 2,
-    dev.type = "hatch"
+    model_id = 2,
+    development_type = "hatch"
     )
   p <- suppressWarnings(predict_phenology(
     data = woody_island,dates = date,
     temperature = temp_c,
     spawn.date = "1990-08-18",
-    model = m$func
+    model = m$expression
     )
     )
 

### 2024-12-09 R/fit_model.R
remove lifecycle badge

@@ -1,7 +1,6 @@
 #' Fit B&M model 2 to new data using `stats::nls()`
 #'
 #' @description
-#' `r lifecycle::badge("experimental")`
 #' Generate your own custom parameterized models for predicting hatching and emergence phenology.
 #'
 #' @details

### 2024-12-09 R/model_select.R

@@ -1,7 +1,6 @@
 #' Select a development model structure
 #'
 #' @description
-#' `r lifecycle::badge("experimental")`
 #' The function calls a model table with the parameterizations for
 #' different species from different studies built in. Refer to the
 #' table (`model_table`) before using function to find inputs for

### 2024-12-09 R/plot_phenology.R

@@ -1,7 +1,6 @@
 #' Visualize the phenology of fish development
 #'
 #' @description
-#' `r lifecycle::badge("experimental")`
 #' The function takes the output from `predict_phenology()` and creates a ggplot2 object.
 #'
 #' @param plot A list containing the output from `predict_phenology()`

### 2024-12-09 R/predict_phenology.R

@@ -1,7 +1,6 @@
 #' Predict phenology of fish
 #'
 #' @description
-#' `r lifecycle::badge("experimental")`
 #' Predict the phenology of fish using the effective value framework.
 #'
 #' @param data Dataframe with dates and temperature.

### 2024-12-09 R/model_select.R
add return

@@ -39,4 +39,5 @@ model_select <- function(author,
         model_id == {{ model_id }} &
         development_type == {{ development_type }}
     )
+  mod
 }

### 2024-12-09 R/plot_phenology.R
rename variables

@@ -24,7 +24,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   if (labels == TRUE) {
     all_label <- ggplot2::labs(
       x = "Date", y = "Mean daily temperature",
-      title = paste(dat$days2done, "days to develop"),
+      title = paste(dat$days_to_develop, "days to develop"),
       subtitle = paste(
         "Fish spawned",
         dat$dev.period$start,
@@ -35,7 +35,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
 
     ef_cumsum_label <- ggplot2::labs(
       x = "Date", y = "Mean daily temperature",
-      title = paste(dat$days2done, "days to develop"),
+      title = paste(dat$days_to_develop, "days to develop"),
       subtitle = paste(
         "Fish spawned",
         dat$dev.period$start,
@@ -46,7 +46,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
 
     ef_daily_label <- ggplot2::labs(
       x = "Date", y = "Mean daily temperature",
-      title = paste(dat$days2done, "days to develop"),
+      title = paste(dat$days_to_develop, "days to develop"),
       subtitle = paste(
         "Fish spawned",
         dat$dev.period$start,
@@ -63,7 +63,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   }
 
   if (style == "all") {
-    p <- dat$ef.tibble |>
+    p <- dat$ef_table |>
       ggplot2::ggplot(ggplot2::aes(x = .data$dates, y = .data$temperature)) +
       ggplot2::geom_line(color = "darkblue") +
       ggplot2::geom_point(color = "darkblue", size = 0.5) +
@@ -76,7 +76,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   }
 
   if (style == "ef_cumsum") {
-    p <- dat$ef.tibble |>
+    p <- dat$ef_table |>
       ggplot2::ggplot(ggplot2::aes(x = .data$dates, y = .data$temperature)) +
       ggplot2::geom_line(color = "darkblue") +
       ggplot2::geom_point(color = "darkblue", size = 0.5) +
@@ -87,7 +87,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   }
 
   if (style == "ef_daily") {
-    p <- dat$ef.tibble |>
[…]

### 2024-12-09 R/predict_phenology.R

@@ -17,15 +17,15 @@
 #'
 #' @return
 #' A list with the following elements:
-#' * `days2done`: A numeric vector of length 1; number of predicted
+#' * `days_to_develop`: A numeric vector of length 1; number of predicted
 #'    days to hatch or emerge.
-#' * `ef.tibble`: An n x 4 tibble (n = number of days to hatch or emerge) with
+#' * `ef_table`: An n x 4 tibble (n = number of days to hatch or emerge) with
 #'    the dates, temperature, effective values, and cumulative sum of the
 #'    effective values.
 #' * `dev.period`: a 1x2 dataframe with the dates corresponding to when your
 #'    fish's parent spawned (input with `predict_phenology(spawn.date = ...)`)
 #'    and the date when the fish is predicted to hatch or emerge.
-#' * `model.specs`: A data.frame with the model specifications.
+#' * `model_specs`: A data.frame with the model specifications.
 #'
 #' @export
 #'
@@ -144,10 +144,10 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     # dev.period$stop <- lubridate::as_date(NA)
 
     ef.results <- list(
-      days2done = Ef.days,
+      days_to_develop = Ef.days,
       dev.period = dev.period,
-      ef.tibble = dat_out_sub,
-      model.specs = model
+      ef_table = dat_out_sub,
+      model_specs = model
     )
 
     # get dates with NaN values
@@ -173,10 +173,10 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     dev.period$stop <- lubridate::as_date(NA)
 
     ef.results <- list(
-      days2done = Ef.days,
+      days_to_develop = Ef.days,
       dev.period = dev.period,
-      ef.tibble = dat_out_sub,
-      model.specs = model
+      ef_table = dat_out_sub,
+      model_specs = model
     )
 
     # get dates with NaN values
@@ -203,10 +203,10 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     dev.period$stop <- lubridate::as_date(NA)
 
     ef.results <- list(
-      days2done = as.numeric(NA),
+      days_to_develop = as.numeric(NA),
       dev.period = dev.period,
-      ef.tibble = dat_out_sub,
-      model.specs = model
+      ef_table = dat_out_sub,
+      model_specs = model
     )
 
     cli::cli_inform(c(
[…]

### 2024-12-09 tests/testthat/test-predict_phenology.R

@@ -16,17 +16,17 @@ test_that("predict phenology works", {
   expect_type(p, "list")
   expect_length(p, 4)
 
-  # expect_type(p$days2done, "integer")
+  # expect_type(p$days_to_develop, "integer")
 
   # expect_type(p$ef.vals, "double")
 
-  expect_s3_class(p$ef.tibble, "data.frame")
+  expect_s3_class(p$ef_table, "data.frame")
 
   expect_s3_class(p$dev.period, "data.frame")
   expect_equal(ncol(p$dev.period), 2)
 
-  # expect_s3_class(p$model.specs, "data.frame")
-  # expect_equal(ncol(p$model.specs), 5)
+  # expect_s3_class(p$model_specs, "data.frame")
+  # expect_equal(ncol(p$model_specs), 5)
 
 
 })

### 2024-12-10 R/plot_phenology.R
overhaul and re document()

@@ -1,7 +1,8 @@
-#' Visualize the phenology of fish development
+#' Visualize fish phenology
 #'
 #' @description
-#' The function takes the output from `predict_phenology()` and creates a ggplot2 object.
+#' The function takes the output from `predict_phenology()` and creates a
+#' basic ggplot2 plot object to visualize the predicted phenology.
 #'
 #' @param plot A list containing the output from `predict_phenology()`
 #' @param style The style of the plot. A vector with possible values "all",
@@ -9,92 +10,121 @@
 #' @param labels Logical. If TRUE (default), labels are added to the plot.
 #'
 #' @return
-#' A object of class "gg" and "ggplot" that can be printed to the console or saved as an image.
+#' A object of class "gg" and "ggplot".
 #'
 #' @export
 #'
 #' @examples
 #' library(hatchR)
-#' #plot_phenology(WI_hatch, style = "ef_cumsum") # shows a plot with just the ef cumulative sum values
-#' #plot_phenology(WI_hatch, style = "ef_daily") # shows a plot with just the ef daily values
-#' #plot_phenology(WI_hatch, labels = FALSE) # turns off the labeling for a cleaner figure
+#' # get model parameterization
+#' sockeye_hatch_mod <- model_select(
+#'   author = "Beacham and Murray 1990",
+#'   species = "sockeye",
+#'   model = 2,
+#'   development_type = "hatch"
+#' )
+#' # predict phenology
+#' sockeye_hatch <- predict_phenology(
+#'   data = woody_island,
+#'   dates = date,
+#'   temperature = temp_c,
+#'   spawn.date = "1990-08-18",
+#'   model = sockeye_hatch_mod
+#' )
+#' plot_phenology(sockeye_hatch)
+#' plot_phenology(sockeye_hatch, style = "ef_cumsum")
+#' plot_phenology(sockeye_hatch, style = "ef_daily")
+#' plot_phenology(sockeye_hatch, labels = FALSE)
 plot_phenology <- function(plot, style = "all", labels = TRUE) {
   dat <- plot
+  x_lab <- "Date"
+  y_lab <- "Mean daily temperature"
+  cols <- c("#1f78b4", "#d95f02", "#1b9e77")
 
   if (labels == TRUE) {
     all_label <- ggplot2::labs(
-      x = "Date", y = "Mean daily temperature",
+      x = x_lab, y = y_lab,
[…]

### 2024-12-10 README.Rmd
rename model_tabel arguments
update README

@@
 # see days to hatch
-WI_hatch$days2done
+WI_hatch$days_to_develop
 # and development period

### 2024-12-10 inst/manuscript/paper.Rmd
add case study 2 section

--- 2024-12-09 inst/manuscript/paper.Rmd
@@
 library(hatchR)
+library(ggridges)
 ```
@@
 However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development.
-Despite these improvements, their application remains limited due to few parameterizations for different species, being largely limited to salmonids.
-Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
-**hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
-**hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
+Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
+Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
+`hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
+`hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
 To demonstrate potential uses, we present two case studies as likely applications for this software.
-**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
+`hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.
 
@@
 
-To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**.
-Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`.
+Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
 
-To widen the user application of these methods, we distribute two user-interfaces for **hatchR**.
-The first is a R package distributed via CRAN that allows users the most customizable application of these methods.
+To widen the user application of these methods, we distribute two user-interfaces for `hatchR`.
+The first is a R package distributed via CRAN that allows users the most customizable application for these methods.
 The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
-These variable approaches are outlined in the package documentation on **hatchR**'s website.
-Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
-The Shiny form trades-off some of the automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
+These variable approaches are outlined in the package documentation on `hatchR`'s website.
+Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface.
+The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package.
 Below, we present the basic overview of the software and multiple case studies of how it may be applied.
@@
 ### 
-library(tidyverse); library(hatchR)
+library(tidyverse); library(hatchR); library(ggdist)
 ```
@@
 
-**hatchR** is meant to primarily be a tool for predicting phenology.
+`hatchR` is meant to primarily be a tool for predicting phenology.
 In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help.
@@
 For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
-Finally, we provide brief summaries of general applications of **hatchR** below and include a visual representation of the full package workflow (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables and making more complex plots.
+Finally, we provide brief summaries of general applications of `hatchR` below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
 
-```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap = "Generalized workflow for hatchR.
-Data actions are depicted with beige circles while functions are represented with white rectangles."}
+```{r echo = FALSE, out.height = '50%', fig.align = 'center'}
 knitr::include_graphics("flowchart.png")
@@
 
-The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`.
-Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
+The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
+Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
 
@@
 
-Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade).
+Fundamentally, `hatchR` assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade).
 Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required.
@@
 
-**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
-**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
+`hatchR` assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
+`hatchR` can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
 In these cases, we suggest users consider how much of that data type is reasonable with their data.
 
-For users choosing to implement **hatchR** in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
+For users choosing to implement `hatchR` in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
 For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.
@@
 
-**hatchR** is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.
+`hatchR` is built assuming data will be analyzed as daily average temperatures.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
 
@@
 # take temps from a random normal dist with mean 10 sd 3 
-# for every date time combo in dates and append to column (temp) 
-# in year_sim
+# for every date time combo in dates and append to column (temp) in year_sim
 year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
@@
 
-```{r, fig.cap = "Graphical output from plot_check_temp(), a visual data check tool in the package. Note that the upper (red) and lower (blue) bound lines  can be custom set by the user."}
-
+```{r}
 plot_check_temp(data = year_sim,
@@
 Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
-We provide parameterizations for three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-The parameterizations are for commonly cultured sport- and commercial fishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
+We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
 
-We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure 3).
+We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure XXX).
 
@@
 
-Note the *R^2^* fit from the models below. You can see they generally all preform well and are in line with values from model 2 of @beacham1990.
+Note the model the *R^2^* fit from the models below. You can see the generally all preform well and are in line with values from model 2 of @beacham1990.
 
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Predicted days to hatch for Channel Catfish (blue), Lake Sturgeon (black), and Smallmouth Bass (green) using custom paramerizations from hatchR's fit_model() function.
-A simulated thermal regime with an ascending thermograph and a mean temperature of 16 °C is used to demonstate the different parameterizations.
-All fish spawn the same day, July 1."}
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 ### predict_phenology
@@
   )
+
+sockeye_emerge_mod <- model_select(
+  author = "Beacham and Murray 1990", 
+  species = "sockeye", 
+  model = 2, 
+  dev.type = "emerge"
+  )
+
 ```
@@
 
-`WI_hatch$days2done` outputs the predicted days to develop.
+`WI_hatch$days2done` outputs the predicted days to hatch or emerge.
 
-`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted todevelop.
+`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to hatch or emerge.
 
@@
 
-**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results (Figure 4).
+**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results.
 The plot visualizes three specific components: 1.
@@
 
-```{r, fig.cap = "Output from hatchR's function, plot_phenology().
-The output is meant to provide a basic summary of your model using a single spawn date.
-Users are encouraged to visit the packages help page where articles demonstrate more advanced, custom plotting applications."}
+```{r}
 plot_phenology(WI_hatch)
@@
 
+For the second example, we will again use bull trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**.
+In this scenario we will use data from @isaak2018 (included `idaho` dataset), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho.
+For this approach we winnow putative bull trout spawning sites by filtering for sites with mean August temperature \</= 13 °C in accordance with thresholds from @isaak2015.
+For the resulting 139 sites we will demonstrate predicting hatch timing in these putative Bull Trout spawning habitats.
+
+```{r, echo = FALSE}
+
+bull_trout_sites <- idaho |> 
+  mutate(month = month(date)) |> #make a month column (numeric)
+  filter(month == 8) |> # filter out Aug.
+  group_by(site) |> # apply grouping by site
+  summarise(mean_aug_temp = mean(temp_c)) |> 
+  filter(mean_aug_temp <= 13) |> # keep only sites 13 C or cooler
+  pull(site) |> 
+  unique()
+
+# filter to bull trout sites
+idaho_bt <- idaho |> 
+  filter(site %in% bull_trout_sites)
+
+# nest data and summarize
+isaak_summ_bt <- idaho_bt |> 
+  group_by(site) |> 
+  nest() |> 
+  mutate(
+    summ_obj = map(
+      data, 
+      summarize_temp,
+      temperature = temp_c,
+      dates = date
+      )
+    ) |> 
+  select(site, summ_obj)
+
+# spawn dates
+spawn_dates <- map(
+  c(2011:2014), # year vector to map for custom function
+  function(year) { # custom paste function
+    c(
+      paste0(year, "-09-01"),
+      paste0(year, "-09-15"),
+      paste0(year, "-09-30")
+      )
+    }
+  ) |> 
+  unlist()
+
+# bull trout hatch model
+bt_hatch <- model_select(
+  development_type = "hatch",
+  author = "Austin et al. 2017",
+  species = "bull trout",
+  model = "MM"
+)
+
+```
+
+We need to setup our models and data for this analysis, which we don't show those steps here for the sake of concision in this manuscript, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**.
+After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure XXX.
+
+```{r}
+hatch_res <- isaak_summ_bt |> 
+  mutate(
+    dev_period = map2(
+      summ_obj, spawn_dates, # map across our site object and spawn dates
+      predict_phenology,
+      temperature = daily_temp,
+      model = bt_hatch,
+      dates = date
+      ) |> 
+      map_df("dev.period") |> # pull out just dev.period results
+      list()
+    ) |> 
+  select(site, dev_period) |> # just select the columns we want
+  unnest(cols = c(dev_period)) |> # unnest everything
+  mutate(days_to_hatch = stop - start) # make a new column of days to hatch
+
+```
+
+```{r, message=FALSE}
+cut_ints <- date(
+  c(
+    "2011-08-31", "2012-08-31",
+    "2013-08-31", "2014-08-31",
+    "2015-08-31"
+  )
+)
+
+# add year and spawn time factors for plotting
+hatch_factors <- hatch_res |> 
+  mutate(year = cut(start,
+    breaks = cut_ints,
+    labels = c(2011:2014)
+  )) |>
+   # cut coerces our labels to factors, this changes them back to numbers
+  mutate(year = as.numeric(as.character(year))) |> 
+  mutate(day = day(start)) |> 
+  mutate(spawn_time = case_when(
+    day == 1 ~ "Early",
+    day == 15 ~ "Peak",
+    day == 30 ~ "Late"
+  )) |>
+  mutate(spawn_time = factor(
+    spawn_time, levels = c("Late", "Peak", "Early"), 
+    ordered = TRUE)
+    )
+
+hatch_factors_sum <- hatch_factors |> 
+  group_by(year, spawn_time, start) |> 
+  summarise(mean_hatch = ceiling(mean(days_to_hatch)),
+            hatch_5 = ceiling(quantile(days_to_hatch, probs = 0.05)),
+            hatch_95 = ceiling(quantile(days_to_hatch, probs = 0.95)))
+
+p1 <- ggplot(
+  data =  hatch_factors,
+    aes(
+      x = days_to_hatch,
+      y = spawn_time,
+      fill = spawn_time,
+      color = spawn_time
+    )) +
+  geom_density_ridges(alpha = 0.9) +
+    facet_wrap( ~ year, ncol = 1) +
+    scale_fill_brewer(palette = "Blues", direction = 1) +
+    scale_color_brewer(palette = "Blues", direction = 1) +
+    labs(x = "Days to hatch", y = "Spawn time") +
+    theme_classic() +
+    theme(legend.position = "none")
+
+```
+
 # Discussion
@@
 
-talk about how temperatures may not always reflect redd temperture
-
-talk about how other factors like oxygen or predator cues may change phenology
-
-talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns
-
-talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.
-
 # Acknowledgements
 
-We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.
 

### 2024-12-10 inst/manuscript/paper.Rmd
render paper.Rmd

@@
 Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
-Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
-`hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
-`hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
+Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
+**hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
+**hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
 To demonstrate potential uses, we present two case studies as likely applications for this software.
-`hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.
+**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
 
@@
 
-To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`.
-Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**.
+Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
 
-To widen the user application of these methods, we distribute two user-interfaces for `hatchR`.
+To widen the user application of these methods, we distribute two user-interfaces for **hatchR**.
 The first is a R package distributed via CRAN that allows users the most customizable application for these methods.
 The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
-These variable approaches are outlined in the package documentation on `hatchR`'s website.
-Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface.
-The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package.
+These variable approaches are outlined in the package documentation on **hatchR**'s website.
+Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
+The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
 Below, we present the basic overview of the software and multiple case studies of how it may be applied.
 
-```{r echo=FALSE}
+```{r echo=FALSE, message=FALSE}
 ### 
@@
 For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
-Finally, we provide brief summaries of general applications of `hatchR` below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
+Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
 
@@
 
-The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
+The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
 Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
@@
 
-`hatchR` is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
+**hatchR** is built assuming data will be analyzed as daily average temperatures.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
 
@@
   model = 2, 
-  dev.type = "emerge"
+  development_type = "emerge"
   )
@@
 
-We demonstrate this first case study using the graphical user interface portion of the `hatchR` ecosystem found at <https://elifelts.shinyapps.io/hatchR/>.
+We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>.
 Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
@@
 Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on `hatchR`'s website <https://bmait101.github.io/hatchR/>.
+The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on **hatchR**'s website <https://bmait101.github.io/hatchR/>.
 
@@
 
-```{r, message=FALSE}
+```{r, message=FALSE, echo=FALSE}
 cut_ints <- date(
@@
 
-p1 <- ggplot(
+ggplot(
   data =  hatch_factors,

### 2024-12-11 inst/manuscript/paper.Rmd
clean up some code chunks to make things shorter
add fig.caps

rerender Rmd to PDF

@@
 
-```{r echo = FALSE, out.height = '50%', fig.align = 'center'}
+```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap= "Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles."}
 knitr::include_graphics("flowchart.png")
@@
 
-```{r}
+```{r, fig.cap= "Full output of the plot_check_temp() visual data check function from hatchR.
+Note that the red and blue lines are customizable by the user to create outlier thresholds for their visual check."}
 plot_check_temp(data = year_sim,
@@
 
-```{r, eval=FALSE}
-# it is useful to plot again to check your summarized data
-plot_check_temp(data = year_sim_summ,
-                dates = date,
-                temperature = daily_temp,
-                temp_min = 0, # temp_min and max lines are 
-                              # user customizable
-                temp_max = 15)
-```
-
 ## Model Selection
@@
 
-fit_model for three non-salmonid specieslib
-
 #### Fitting models for other fishes
@@
 We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
 
@@
                      development_type = "hatch")
+
+```
+
+```{r, echo=FALSE}
 
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap= "Days to hatch as predicted using custom parameterizations for Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass (green) as generated by the fit_model() function in hathcR.
+Models are all run using a spawn date starting June 1 for a randomly generated ascending thermal regime with mean temperature 16 °C"}
 ### predict_phenology
@@
 
-sockeye_emerge_mod <- model_select(
-  author = "Beacham and Murray 1990", 
-  species = "sockeye", 
-  model = 2, 
-  development_type = "emerge"
-  )
-
 ```
@@
 WI_hatch$dev.period
-```
-
-We can also do the same with emergence:
-
-```{r, warning=FALSE}
-WI_emerge <- predict_phenology(
-  data = woody_island,
-  dates = date,
-  temperature = temp_c,
-  spawn.date = "1990-08-18",
-  model = sockeye_emerge_mod   # notice we're using emerge model expression here
-  )
-
-# see days to hatch and development period
-WI_emerge$days2done
-WI_emerge$dev.period
 ```
@@
 
-```{r}
+```{r, fig.cap= "Output of the plot_phenology() function from hatchR.
+The function plots the output of the predict_phenology() function and displays time to develop, as well as the ambient thermal regime (blue), scaled daily effective value (green), and the scaled cumulative sum of effective values (orange)."}
 plot_phenology(WI_hatch)
@@
 
-```{r, message=FALSE, echo=FALSE}
+```{r, message=FALSE, echo=FALSE, fig.cap= "Densities of predicted days to hatch for Bull Trout for 139 putative spawning locations across Idaho for four years of shared temperature data.
+Early spawning considers if a parent spawned September 1, Peak-September 15, and Late-September 30."}
 cut_ints <- date(

### 2024-12-11 inst/misc/hexlogos/hatchR_logo.R
add hatchR_logo in inst/misc to hold script and images for input/ouput for hatchR logo

@@ -1,0 +1,12 @@
+library(hexSticker)
+
+# current logo
+sticker(here::here("inst/misc/hexlogos/hatchR_FINAL.tif"), package="hatchR", p_size=20, s_x=1, s_y=.75, s_width=.6,
+        h_fill="white", h_color = "black", p_color = "black",
+        filename=here::here("inst/misc/hexlogos/hatchR_logo.png"))
+
+
+# FS colors
+sticker("~/Downloads/hatchR_FINAL.tif", package="hatchR", p_size=20, s_x=1, s_y=.75, s_width=.6,
+        h_fill="#01502f", h_color = "#ffd330", p_color = "#ffd330",
+        filename="~/Downloads/hatchR_hexsticker.png")

### 2024-12-11 README.Rmd

@@
 <!-- badges: start -->
-
 [![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
-
 <!-- badges: end -->

### 2024-12-11 inst/manuscript/paper.Rmd

@@
 
-We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure XXX).
+We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure 3).
 
@@
   )
-
 ```
@@
 
-**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results.
-The plot visualizes three specific components: 1.
-)the temperature regime over which you are predicting, 2.) the cumulative sum of effective values, and 3.) the effective value for each day in your prediction span.
+**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results (Figure 4).
+The plot visualizes three specific components: 1.) the temperature regime over which you are predicting, 2.) the cumulative sum of effective values, and 3.) the effective value for each day in your prediction span.
 The function allows you to output various figures based on your interests, but defaults to a figure with all information and the corresponding labels.
@@
 )
-
 ```
@@
   mutate(days_to_hatch = stop - start) # make a new column of days to hatch
-
 ```
@@
     theme(legend.position = "none")
-
 ```
@@
 talk about how these models represent point estimates and that emergence and hatch will take the form of a distribution
+
+talk about how temperatures may not always reflect redd temperture
+
+talk about how other factors like oxygen or predator cues may change phenology
+
+talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns
+
+talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.
 

### 2024-12-16 README.Rmd
update logo
fix typo

@@
 
-# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" alt="hatchR website" align="right" height="139"/></a>
+# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
 
@@
 
-If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running
+If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running:
 

### 2025-01-07 inst/manuscript/paper.Rmd
writing discussion

@@
 
-talk about how these models also represent local adaptation and heritable plasticity
+With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations.
+To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times.
+Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for salmonids or create their own species- or population specific parameterization.
+Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of programmatic platform.
+
+In the application of using effective value models **hatchR** and the user make some key assumptions.
+Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment.
+These include water quality like dissolved oxygen, pH, or salnity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024).
+Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.
+
+To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019).
+However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species @west-eberhard2003.
+For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes.
+Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks.
+The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation @conover2009, @sparks2022.
+In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.
 

### 2025-01-07 inst/manuscript/paper.Rmd
discussion

@@
 Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.
+Another factor that could bias estimates from these models is that temperatures used from sensors don't reflect the geomoprhic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.
 
@@
 
-talk about how these models represent point estimates and that emergence and hatch will take the form of a distribution
+**BRYAN check the below**
 
-talk about how temperatures may not always reflect redd temperture
-
-talk about how other factors like oxygen or predator cues may change phenology
-
-talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns
-
-talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.
+The models presented in **hatchR** can be further customized beyond the use-cases provided above.
+For instance, while our models as presented are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or embryological developmental stages @velsen1980 or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**.
+Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided, they could pass a different expression into `predict_phenology()` for additional customization.
+Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in an exponential decay relationship and these developmental models could be extended to other organisms beyond fishes.
 

### 2025-01-08 inst/manuscript/paper.Rmd
add to model selection section
reknit

--- 2025-01-07 inst/manuscript/paper.Rmd
@@
 To widen the user application of these methods, we distribute two user-interfaces for **hatchR**.
-The first is a R package distributed via CRAN that allows users the most customizable application for these methods.
+The first is a R package (@r:ther) distributed via CRAN that allows users the most customizable application for these methods.
 The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
 These variable approaches are outlined in the package documentation on **hatchR**'s website.
-Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
+Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
 The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
@@
 **hatchR** is meant to primarily be a tool for predicting phenology.
-In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help.
+In this sense, we limit functionality to these applications and provide minimal data checking and plotting help.
 This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks.
@@
 We do provide two basic data check functions discussed in the Checking Data section.
-Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R.
+Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R.
 For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
@@
 The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
-Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
+Specifically, `model_table` includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
 
@@
 
-For users choosing to implement **hatchR** in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
+For users choosing to implement **hatchR** in R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
 For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.
@@
 **hatchR** is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
+Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable (only in R, it must be summarized for the Shiny app), as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
 
@@
 
-talk about using nls for fit_model
+Users can either use Salmonid models from `model_table` included in the package and the Shiny app.
+As discussed, these models are included because parameterizations already existed in the literature.
+However, these parameterizations are limited to Pacific Salmon and Bull Trout.
+To widen the application of the effective value approach, we include a `fit_model()` function and is species agnostic (as long as develop generally follows the power law.
 
-model_table and fit_model
+The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs and estimates parameter coeficients for *log~e~a* and *b* using `stats::nls()`
+
+**BRYAN TO FINISH THIS PARAGRAPH OFF HERE**
 
@@
 We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
 We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
@@
 Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment.
-These include water quality like dissolved oxygen, pH, or salnity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024).
+These include water quality like dissolved oxygen, pH, or salinity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024).
 Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.
+Another factor that could bias estimates from these models is that temperatures used from sensors don't reflect the geomoprhic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.
 
 To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019).
-However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species @west-eberhard2003.
+However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species (@west-eberhard2003).
 For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes.
 Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks.
-The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation @conover2009, @sparks2022.
+The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation (@conover2009, @sparks2022).
 In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.
 
-talk about how these models represent point estimates and that emergence and hatch will take the form of a distribution
+**BRYAN check the below**
 
-talk about how temperatures may not always reflect redd temperture
-
-talk about how other factors like oxygen or predator cues may change phenology
-
-talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns
-
-talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.
+The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above.
+For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**.
+Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization.
+Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.
 

### 2025-01-23 inst/manuscript/paper_oxford/paper_oxford.Rmd
get oxford style working for Fisheries

--- 2025-01-08 inst/manuscript/paper.Rmd
@@
 ---
-title: 'hatchR: A toolset to predict hatch and emergence phenology in wild fishes'
-tags:
-  - R
-  - fishes
-  - developmental phenology
-authors:
+title: Template for Oxford University Press papers
+author:
   - name: Morgan M. Sparks
+    email: Morgan.Sparks@usda.gov
     orcid: 0000-0003-0787-2218
-    affiliation: 1 # (Multiple affiliations must be quoted)
+    corresponding_author: TRUE # Automatic footnote using corresponding_author
+    affiliation: A
   - name: Eli Felts
     orcid: 0000-0002-5888-9266
-    affiliation: 2
+    affiliation: B
   - name: Bryan M. Maitland
     orcid: 0000-0002-4491-5064
-    affiliation: 1
-affiliations:
- - name: US Forest Service, Rocky Mountain Research Station, Boise, ID, USA
-   index: 1
- - name: US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
-   index: 2
-citation_author: Sparks et al.
-date: 07 July 2025
-year: 2025
-output: rticles::joss_article
-csl: apa.csl
-journal: JOSS
-editor_options:
-  chunk_output_type: console
-bibliography: ms_refs.bib
----
-
-```{r, include=FALSE}
-library(tidyverse)
-library(hatchR)
-library(ggridges)
-```
-
-# Abstract
-
-Understanding the timing of key life history events is necessary for managing and conserving populations.
+    affiliation: A
+    footnote:
+address:
+  - code: A
+    department: Rocky Mountain Research Station
+    organization: U.S. Forest Service
+    #postcode: 
+    city: Boise
+    state: Idaho # Mix and match elements to get proper address
+    country: United States
+  - code: B
+    department: Idaho Fish and Wildlife Conservation Office
+    organization: U.S. Fish and Wildlife Service
+    city: Orofino
+    state: Idaho # Mix and match elements to get proper address
+    country: United States
+abstract:
+  - "Understanding the timing of key life history events is necessary for managing and conserving populations.
 Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models.
@@
 **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
+"
+keywords:
+  - developmental phenology
+  - effective value model
+  - R
+  - Shiny
+#boxedlist: # Not fully compatible with all document formats
+#  - First point
+#  - Second point
+competing_interests: |
+  There are no competing interest.
+author_contributions: |
+  MMS and BMM designed and wrote the package. EF wrote the Shiny app. All authors wrote and edited the manuscript.
+acknowledgments: |
+  We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+
+  The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.
+Any mention of trade names, products, or services does not imply an endorsement by the U.S.
+government, USFS, or USFWS.
+USFS and USFWS do not endorse any commercial products, services or enterprises.
+
+bibliography: ms_refs.bib # Includes refs in OUP example template
+## When using `citation_package = "default"`, a CSL file can be used
+#csl: https://www.zotero.org/styles/oxford-university-press-note
+output:
+  rticles::oup_article:
+    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
+    journal: "Fisheries"
+    document_style: "contemporary" # Can be contemporary, modern, traditional
+    papersize: "large" # Can be large, medium, small
+    citation_package: "default" # Uncomment when using a CSL; default "natbib"
+    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
+    #number_sections: FALSE # Uncomment to not number sections; default TRUE
+    #number_lines: TRUE # Use `lineno` package - Default FALSE
+    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
+    #onecolumn: TRUE # Uncomment for one column format; default FALSE
+    extra_dependencies:
+      - booktabs # to use with knitr::kable() example below
+
+## Example of pandoc's variable
+#urlcolor: orange
+#linkcolor: green
+#citecolor: red
+header-includes:
+  #- \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
+  - \theoremstyle{thmstyleone} # Theorem stuff from OUP template
+  - \newtheorem{theorem}{Theorem} #  meant for continuous numbers.
+%%\newtheorem{theorem}{Theorem}[section] # meant for sectionwise numbers.
+optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+  - \newtheorem{proposition}[theorem]{Proposition} # %%\newtheorem{proposition}{Proposition}" # to get separate numbers for theorem and proposition etc.
+  - \theoremstyle{thmstyletwo}
+  - \newtheorem{example}{Example}
+  - \newtheorem{remark}{Remark}
+  - \theoremstyle{thmstylethree}
+  - \newtheorem{definition}{Definition}
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE, # By default, hide code; set to TRUE to see code
+  fig.pos = 'th', # Places figures at top or here
+  out.width = '100%', dpi = 300, # Figure resolution and size
+  fig.env="figure"
+) # Latex figure environment
+
+options(knitr.table.format = "latex") # For kable tables to write LaTeX table directly
+```
+
+```{r, include=FALSE}
+library(tidyverse)
+library(hatchR)
+library(ggridges)
+```
 
@@
 
-```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap= "Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles."}
+```{r, echo = FALSE, out.height = '75%', fig.align = 'center'}
 knitr::include_graphics("flowchart.png")
@@
 
-| date       | temperature |
-|------------|-------------|
-| 2000-01-01 | 2.51        |
-| ...        | ...         |
-| 2000-07-01 | 16.32       |
-| ...        | ...         |
-| 2000-12-31 | 3.13        |
+```{r kabletable, echo=FALSE}
+df = data.frame(date = c("2000-01-01",
+                         "...",
+                         "2000-07-01",
+                         "...",
+                         "2000-12-31"),
+                temperature= c(2.51,
+                        "...",
+                        16.32,
+                        "...",
+                        3.13
+                        ))
+
+# kable can alse be used for creating tables
+knitr::kable(df,
+             booktabs=TRUE)
+```
 
@@
 
-```{r}
+```{r, echo = TRUE}
 # create date object for a year with 30 min reading intervals
@@
 
-```{r, fig.cap= "Full output of the plot_check_temp() visual data check function from hatchR.
-Note that the red and blue lines are customizable by the user to create outlier thresholds for their visual check."}
+```{r, echo = TRUE}
 plot_check_temp(data = year_sim,
@@
 
-```{r}
+```{r, echo = TRUE}
 # summarize
@@
 However, these parameterizations are limited to Pacific Salmon and Bull Trout.
-To widen the application of the effective value approach, we include a `fit_model()` function and is species agnostic (as long as develop generally follows the power law.
+To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law).
 
-The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs and estimates parameter coeficients for *log~e~a* and *b* using `stats::nls()`
+The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs.
+The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`
 
@@
 We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
-We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
+These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
+We provide parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
 
@@
 
-```{r}
+```{r, echo = TRUE}
 ###  make temp regime
@@
 
-```{r}
+```{r, echo = TRUE}
 #model fits
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap= "Days to hatch as predicted using custom parameterizations for Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass (green) as generated by the fit_model() function in hathcR.
-Models are all run using a spawn date starting June 1 for a randomly generated ascending thermal regime with mean temperature 16 °C"}
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 ### predict_phenology
@@
 
-```{r}
+```{r, echo = TRUE}
 sockeye_hatch_mod <- model_select(
@@
 
-```{r, warning=FALSE}
+```{r, warning=FALSE, echo = TRUE}
 WI_hatch <- predict_phenology(
@@
 
-```{r}
+```{r, echo = TRUE}
 WI_hatch$days2done
@@
 
-```{r, eval=FALSE}
+```{r, eval=FALSE, echo = TRUE}
 str(WI_hatch)
@@
 
-```{r, fig.cap= "Output of the plot_phenology() function from hatchR.
-The function plots the output of the predict_phenology() function and displays time to develop, as well as the ambient thermal regime (blue), scaled daily effective value (green), and the scaled cumulative sum of effective values (orange)."}
-plot_phenology(WI_hatch)
-```
+`{r"} plot_phenology(WI_hatch)`
 
@@
 A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action.
-For instance, will have fish have emerged from redds when a stream section has been opened to grazing or bridge decommissioning will commence?
+For instance, will fish have emerged from redds when a stream section has been opened to grazing or road work?
 
-In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed.
-The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st.
+In this scenario, we will consider the road work example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the Crooked River, a key Bull Trout population in the Boise River watershed.
+In this hypothetical scenario, the Forest Fisheries Biologist wants to know if fish will likely be out of the gravel and free-swimming by June 1st as Bull Trout are particularly sensitive to sediment.
 In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.
@@
 Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
+Data used in this case study are from the `crooked_river` data set from R and can either be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
 Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check.
@@
 Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on **hatchR**'s website <https://bmait101.github.io/hatchR/>.
+The process is demonstrated in full in Figure XXX, but the interface is described more completely on **hatchR**'s Shiny website.
 
@@
 
-```{r}
+```{r, echo = TRUE}
 hatch_res <- isaak_summ_bt |> 
@@
 
-```{r, message=FALSE, echo=FALSE, fig.cap= "Densities of predicted days to hatch for Bull Trout for 139 putative spawning locations across Idaho for four years of shared temperature data.
-Early spawning considers if a parent spawned September 1, Peak-September 15, and Late-September 30."}
+```{r, message=FALSE, echo=FALSE}
 cut_ints <- date(
@@
 
-# Acknowledgements
-
-We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
-
-The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.
-Any mention of trade names, products, or services does not imply an endorsement by the U.S.
-government, USFS, or USFWS.
-USFS and USFWS do not endorse any commercial products, services or enterprises.
-
 # References

### 2025-01-23 inst/manuscript/paper_oxford/paper_oxford.tex
get oxford style working for Fisheries

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}

% Set document class options
\documentclass[webpdf,large,contemporary,namedate]{oup-authoring-template}

% one column

%\usepackage{showframe}

% line numbers

% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

% From Pandoc template for its feature
\usepackage{xcolor}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Template for Oxford University Press papers},
  pdfkeywords={developmental phenology, effective value
model, R, Shiny},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
%From Pandoc 3.1.8
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Counters for addresses and footnotes
\newcounter{correspcnt} % For author footnotes
\renewcommand*{\thecorrespcnt}{\fnsymbol{correspcnt}}
\newcounter{addrcnt} % For author addresses

% Macros for dealing with affiliations, footnotes, etc.
\makeatletter

\def\MyNewLabel#1#2#3{\expandafter\gdef\csname #1@#2\endcsname{#3}}

\def\MyRef#1#2{\@ifundefined{#1@#2}{???}{\csname #1@#2\endcsname}}

\newcommand*\ifcounter[1]{%
  \ifcsname c@#1\endcsname
    \expandafter\@firstoftwo
  \else
    \expandafter\@secondoftwo
  \fi
}

\newcommand*\addrlblbycode[1]{%
  \ifcounter{ADDRLBL@#1}
    {}
    {\refstepcounter{addrcnt}\newcounter{ADDRLBL@#1}\setcounter{ADDRLBL@#1}{\value{addrcnt}}}%
    \arabic{ADDRLBL@#1}%
}

\newcommand*\addrbycode[1]{%
  \ifcounter{ADDR@#1}
    {}
    {\newcounter{ADDR@#1}%
     \address[\addrlblbycode{#1}]{\MyRef{ADDRTXT}{#1}}}%
}

\newcommand*\corresplblbycode[1]{%
  \ifcounter{CORRESPLBL@#1}
    {}
    {\refstepcounter{correspcnt}\newcounter{CORRESPLBL@#1}\setcounter{CORRESPLBL@#1}{\value{correspcnt}}}%
    \fnsymbol{CORRESPLBL@#1}%
}

\newcommand*\correspbycode[1]{%
  \ifcounter{CORRESP@#1}
    {}
    {\newcounter{CORRESP@#1}%
     \corresp[\corresplblbycode{#1}]{\MyRef{CORRESPTXT}{#1}}}%
}

\makeatother

% Add missing \city command mentioned in documentation but absent from cls
\providecommand\city[1]{#1}

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{A}{%
  %
  \orgdiv{Rocky Mountain Research Station}, %
  \orgname{U.S. Forest Service}, %
  \orgaddress{%
   %
  %
  %
  \city{Boise}, %
  \state{Idaho}, %
  \country{United States}%
  }%
   %
 %
 }
   \MyNewLabel{ADDRTXT}{B}{%
  %
  \orgdiv{Idaho Fish and Wildlife Conservation Office}, %
  \orgname{U.S. Fish and Wildlife Service}, %
  \orgaddress{%
   %
  %
  %
  \city{Orofino}, %
  \state{Idaho}, %
  \country{United States}%
  }%
   %
 %
 }

% Create labels for Footnotes if they are given in Elsevier format

% Pandoc header-include feature
\theoremstyle{thmstyleone}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{thmstyletwo}
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\theoremstyle{thmstylethree}
\newtheorem{definition}{Definition}
% Pandoc header-include feature
\usepackage{booktabs}

\begin{document}

\journaltitle{Fisheries}
\DOI{DOI HERE}
\copyrightyear{YYYY}
\pubyear{YYYY}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

\title[]{Template for Oxford University Press papers}

\newcounter{thisauthcorresp} % For storage if author is corresponding author
\newcounter{thisauththanks} % For storage if author has thanks

\author[%
\addrlblbycode{A}%
,\refstepcounter{correspcnt}\setcounter{thisauthcorresp}{\value{correspcnt}}\fnsymbol{thisauthcorresp}%
%
%
]{Morgan M. Sparks}

\addrbycode{A}

\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:Morgan.Sparks@usda.gov}{\nolinkurl{Morgan.Sparks@usda.gov}}}

\author[%
\addrlblbycode{B}%
%
%
%
]{Eli Felts}

\addrbycode{B}

\author[%
\addrlblbycode{A}%
%
%
%
]{Bryan M. Maitland}

\addrbycode{A}

\correspbycode{}

% Add author mark
\authormark{Morgan M. Sparks et al.}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

\abstract{
Understanding the timing of key life history events is necessary for
managing and conserving populations. Historically, models to predict
hatch and emergence timing for fishes were difficult to employ in wild
settings because average incubation temperature was needed as the
primary parameter in predictive models. However, recent improvements to
these techniques reworked models such that they could be applied in wild
environments as long as users had data for when adult fish spawned and a
record of average daily temperature over the course of development.
Despite these improvements, their application remains limited due to few
parameterizations for varying species, being largely limited to
salmonids. Here we present \textbf{hatchR}, a software ecosystem that
allows users to predict hatch and emergence timing for wild fishes, as
well as additional tools to aid in those analyses. \textbf{hatchR}
allows users to leverage popular historic parameterizations for
phenological models or to easily implement custom parameterizations
using data not included in the package. \textbf{hatchR} is also
distributed in two forms---an open source R package for maximum
customization, as well as an HTML graphical-user-interface web
application for individuals not familiar with scripting languages. To
demonstrate potential uses, we present two case studies as likely
applications for this software. \textbf{hatchR} promises to open many
exciting avenues in research and management of fishes during their early
life history.}

\keywords{developmental phenology; effective value model; R; Shiny}

\maketitle

\section{Introduction}\label{introduction}

As primarily poikilothermic organisms, the development and growth of
fishes is tightly linked with the temperature of their ambient
environment. This close relationship has allowed researchers to generate
statistical models that allow the prediction of developmental phenology
with high accuracy and precision. These models were typically developed
in aquaculture settings and their initial formulations were not
applicable to wild populations because they assumed a constant
temperature over the course of development (Beacham and Murray 1990;
McPhail and Murray 1979; Alderdice and Velsen 1978). However, Sparks et
al. (2019) reformulated this approach as an ``Effective Value model'',
in which the input was daily average temperature after a parent spawned
and fish would either hatch or emerge when effective values cumulatively
summed to one.

The resulting effective value approach has now been widely applied in
Salmonids for which parameterizations from aquaculture were readily
available---for example Pacific Salmon (\emph{Oncorhynchus spp.}) models
developed by Beacham and Murray (1990) have been applied to various
species and populations (Adelfio et al. 2024, 2019; Kaylor et al. 2021)
while models developed for Bull Trout (\emph{Salvelinus confluentus}) by
McPhail and Murray (1979) were extended by Austin, Essington, and Quinn
(2019). Despite growing popularity, applications have been largely
limited within Salmonids, presumably because parameterizations for such
models already existed due to their wide use in aquaculture and their
general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value
model applications within individual studies and the lack of
parameterization for other species, we developed the software ecosystem,
\textbf{hatchR}. Specifically, \textbf{hatchR} allows users to input
standard raw or summarized temperature datasets that are commonly
collected in wild settings, run basic checks on those data, use built-in
parameterizations like those from Beacham and Murray (1990) or Sparks et
al. (2017), develop custom models from their own or published
temperature and phenological data, and predict hatch and emergence
timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two
user-interfaces for \textbf{hatchR}. The first is a R package ({``R: The
r Project for Statistical Computing''} (n.d.)) distributed via CRAN that
allows users the most customizable application for these methods. The R
package is especially powerful as it allows users to automate their
analyses over multiple variables such as phenology type, multiple spawn
dates, or different habitats with varying thermal regimes. These
variable approaches are outlined in the package documentation on
\textbf{hatchR}'s website. Alternatively, we also distribute a Shiny
application (Chang et al. (2024)) in the form of an HTML-based web tool
to interact with many of \textbf{hatchR}'s functions in a
graphical-user-interface. The Shiny form trades-off some of automative
power for user simplicity, while still allowing users to leverage much
of the functionality of \textbf{hatchR}'s R package. Below, we present
the basic overview of the software and multiple case studies of how it
may be applied.

\section{Package Overview}\label{package-overview}

\textbf{hatchR} is meant to primarily be a tool for predicting
phenology. In this sense, we limit functionality to these applications
and provide minimal data checking and plotting help. This decision is in
part driven by the diversity of data types that users may import and the
difficulty in addressing all those data types with respect to various
data checks. In other words, we expect users to know their data better
than we do and to check it accordingly. We do provide two basic data
check functions discussed in the Checking Data section. Similarly, we
provide limited functionality for plotting results, but provide examples
of how to build custom visualizations from output, specifically in R.
For the Shiny application, we provide a base output plot, but the
ability to download your results for custom plotting in programs of the
user's choice. Finally, we provide brief summaries of general
applications of \textbf{hatchR} below, but encourage users to visit
articles hosted on the software webpage that extensively outline primary
functions and applications, especially automating the application of
predicting phenology across multiple variables.

\begin{center}\includegraphics[width=1\linewidth,height=0.75\textheight]{flowchart} \end{center}

\subsection{Effective value models}\label{effective-value-models}

Effective value models were created by Sparks et al. (2019) to implement
developmental models in wild environments for Sockeye Salmon (\emph{O.
nerka}). The need for their development arose because historic models,
specifically those in Beacham and Murray (1990), only considered the
average incubation temperature during development and, for wild fishes,
average incubation temperature was impossible to estimate because it was
unknown when fish hatched even if adult spawn timing was known. To
address this, Sparks et al. (2019) used the reciprocal of the
formulation of model 2 from Beacham and Murray (1990) and assigned an
effective value for every day of development using the daily average
temperature.

The model follows the general format of:

\[
Effective Value_i = 1/exp(log_ea - log_e(Temperature_i - b))
\]

Where \emph{i} is the daily value and a fish hatches or emerges when the
cumulative sum of effective values reaches one:
\[\sum_{i =1}^nEffectiveValue_i = 1\]

The effective value model framework is the basis for the phenological
models in \textbf{hatchR}, both in the included \texttt{model\_table} in
the package (though \texttt{model\_table} includes more complex models
developed by Beacham and Murray (1990)), as well as for custom models
users can fit with \texttt{fit\_model()}. Specifically,
\texttt{model\_table} includes parameterizations from Beacham and Murray
(1990), Sparks et al. (2017), and Austin, Essington, and Quinn (2019)
(who extended McPhail and Murray (1979)).

\subsection{Data format}\label{data-format}

Water temperature datasets collected for wild environments are often
either 1.) already summarized by day (\emph{i.e.}, mean daily
temperature) or, 2.) in a raw format from something like a HOBO TidbiT
where readings are taken multiple times per day, which can be summarized
into a mean daily temperatures. Alternatively, new statistical models
like that of Siegel et al. (2023) could be similarly implemented.

Fundamentally, \textbf{hatchR} assumes you have input data with two
columns: a date column, giving the date (and often time) of a
temperature measurement, and a temperature column, giving the associated
temperature measurement (in centigrade). Other columns are okay to
include, but these two columns (with any column name---just
\emph{without} spaces) are required. We expect your data to look
something like this:

\begin{tabular}{ll}
\toprule
date & temperature\\
\midrule
2000-01-01 & 2.51\\
... & ...\\
2000-07-01 & 16.32\\
... & ...\\
2000-12-31 & 3.13\\
\bottomrule
\end{tabular}

\textbf{hatchR} assumes you've checked for missing records or errors in
your data as it \emph{will function with gaps}, so it's important to go
through the data checks discussed below, as well as your own validity
checks. \textbf{hatchR} can use values down to freezing (e.g, 0 °C),
which returns extremely small effective values, and time to hatch or
emerge may be \textgreater{} 1 year. In these cases, we suggest users
consider how much of that data type is reasonable with their data.

For users choosing to implement \textbf{hatchR} in R, data can be
imported from any format the user chooses, as long as users can
eventually coerce their data into a \texttt{dataframe} or
\texttt{tibble} format, in which each row represents a single record.
For the Shiny application, users must have their data stored as a .csv
(comma separated values) file for upload, which can easily be exported
using datasheet software like Microsoft Excel or Google Sheets.

\subsection{Checking Data}\label{checking-data}

\textbf{hatchR} is built assuming data will be analyzed as daily average
temperatures. Despite that assumption, raw data (\emph{e.g.}, as
outputted by HOBO loggers) can be used and \textbf{hatchR} includes
functionality to summarize those data into a format that is usable (only
in R, it must be summarized for the Shiny app), as well as provides
functions for basic visual and programmatic data checks to make sure
outliers or missing data are at least brought to user's attention.

We demonstrate the utility of the summary and check functions
\texttt{summarize\_temp()}, \texttt{plot\_check\_temp()}, and
\texttt{check\_continuous()} using a simulated year-long dataset with
temperature readings every thirty minutes.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# create date object for a year with 30 min reading intervals}
\NormalTok{dates }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\AttributeTok{from =} \FunctionTok{ymd\_hms}\NormalTok{(}\StringTok{"2000{-}07{-}01 00:00:00"}\NormalTok{),}
             \AttributeTok{to =} \FunctionTok{ymd\_hms}\NormalTok{(}\StringTok{"2001{-}06{-}30 23:59:59"}\NormalTok{), }\AttributeTok{length.out =} \DecValTok{17568}\NormalTok{) }

\CommentTok{\# create empty dataframe}
\NormalTok{year\_sim }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{nrow =} \FunctionTok{length}\NormalTok{(dates), }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{)) }

\CommentTok{\# date column}
\FunctionTok{colnames}\NormalTok{(year\_sim)[}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}} \StringTok{"date"}

\CommentTok{\# add dates vector to date column}
\NormalTok{year\_sim[}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ dates }

\CommentTok{\#random seed}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}

\CommentTok{\# take temps from a random normal dist with mean 10 sd 3 }
\CommentTok{\# for every date time combo in dates and append to column (temp) }
\CommentTok{\# in year\_sim}
\NormalTok{year\_sim}\SpecialCharTok{$}\NormalTok{temp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =} \FunctionTok{length}\NormalTok{(dates), }\AttributeTok{mean =} \DecValTok{10}\NormalTok{, }\AttributeTok{sd =} \DecValTok{3}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{abs}\NormalTok{() }

\FunctionTok{dim}\NormalTok{(year\_sim)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 17568     2
\end{verbatim}

First, we recommend checking imported data for any outliers or strange
inputs using \texttt{plot\_check\_temp()}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot\_check\_temp}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim,}
                \AttributeTok{dates =}\NormalTok{ date,}
                \AttributeTok{temperature =}\NormalTok{ temp,}
                \AttributeTok{temp\_min =} \DecValTok{0}\NormalTok{, }\CommentTok{\# temp\_min and max lines are }
                              \CommentTok{\# user customizable}
                \AttributeTok{temp\_max =} \DecValTok{25}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics[width=1\linewidth]{paper_oxford_files/figure-latex/unnamed-chunk-5-1}

There are no obvious outliers but since each day has 48 records, we need
to summarize it to daily mean temperature with
\texttt{summarize\_temp()} and then check for missing days with
\texttt{check\_continuous()}. We also recommend using
\texttt{plot\_check\_temp()} again on the summarized data (though leave
out the resulting plot for space efficiency in this manuscript).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# summarize}
\NormalTok{year\_sim\_summ }\OtherTok{\textless{}{-}} \FunctionTok{summarize\_temp}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim,}
                                \AttributeTok{dates =}\NormalTok{ date,}
                                \AttributeTok{temperature =}\NormalTok{ temp)}

\CommentTok{\# now a year\textquotesingle{}s worth of single{-}day data}
\FunctionTok{dim}\NormalTok{(year\_sim\_summ)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 365   2
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# check continuous (no errors)}
\FunctionTok{check\_continuous}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim\_summ,}
                 \AttributeTok{dates =}\NormalTok{ date)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## i No breaks were found. All clear!
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# we can demonstrate an error by removing Oct. 8 (100th day)}
\FunctionTok{check\_continuous}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim\_summ[}\SpecialCharTok{{-}}\DecValTok{100}\NormalTok{,],}
                 \AttributeTok{dates =}\NormalTok{ date)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: ! Data not continuous
## i Breaks found at rows:
## i 100
\end{verbatim}

\subsection{Model Selection}\label{model-selection}

Users can either use Salmonid models from \texttt{model\_table} included
in the package and the Shiny app. As discussed, these models are
included because parameterizations already existed in the literature.
However, these parameterizations are limited to Pacific Salmon and Bull
Trout. To widen the application of the effective value approach, we
include a \texttt{fit\_model()} function, which is species agnostic (as
long as development generally follows the power law).

The function \texttt{fit\_model()} uses data in which average incubation
temperature (°C) and days to phenological event (as two vectors) are the
inputs. The function estimates parameter coefficients for
\emph{log\textsubscript{e}a} and \emph{b} using \texttt{stats::nls()}

\textbf{BRYAN TO FINISH THIS PARAGRAPH OFF HERE}

\paragraph{Fitting models for other
fishes}\label{fitting-models-for-other-fishes}

Below, we demonstrate how the \texttt{fit\_model()} function may be used
to create custom parameterizations for species beyond the Salmonids in
the \texttt{model\_table} included in the package. We include
parameterizations from three warm-water species to demonstrate the
\texttt{fit\_model()} utility for fishes beyond the scope of the
original effective value approach. These parameterizations are for
commonly cultured sportfishes including Smallmouth Bass
(\emph{Micropterus dolomieu}) from Webster (1948), Channel Catfish
(\emph{Ictalurus punctatus}) from Small and Bates (2001), and Lake
Sturgeon (\emph{Acipenser fulvescens}) from Smith and King (2005). We
provide parameterization for just Smallmouth Bass here for concision,
but the full code set for all species is available at
\url{https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd}.

We demonstrate the utility of this approach by creating a random thermal
regime with an ascending thermograph with a mean temperature of 16 °C,
parameterizing models for each species, and demonstrating days to hatch
and developmental period for each species with the random thermal regime
(Figure 3).

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\#\#  make temp regime}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\CommentTok{\# create random temps and corresponding dates}
\NormalTok{temps\_sim }\OtherTok{\textless{}{-}} \FunctionTok{sort}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\DecValTok{30}\NormalTok{, }\AttributeTok{mean =} \DecValTok{16}\NormalTok{, }\AttributeTok{sd =} \DecValTok{1}\NormalTok{), }\AttributeTok{decreasing =} \ConstantTok{FALSE}\NormalTok{)}
\NormalTok{dates\_sim }\OtherTok{\textless{}{-}}  \FunctionTok{seq}\NormalTok{(}\AttributeTok{from =} \FunctionTok{ymd}\NormalTok{(}\StringTok{"2000{-}07{-}01"}\NormalTok{),}
             \AttributeTok{to =} \FunctionTok{ymd}\NormalTok{(}\StringTok{"2000{-}07{-}31"}\NormalTok{), }\AttributeTok{length.out =} \DecValTok{30}\NormalTok{)}

\NormalTok{data\_sim }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{data.frame}\NormalTok{()}
\NormalTok{data\_sim[,}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ temps\_sim}
\NormalTok{data\_sim[,}\DecValTok{2}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ dates\_sim}

\CommentTok{\# change names so they aren\textquotesingle{}t the same as the vector objects}
\FunctionTok{colnames}\NormalTok{(data\_sim) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"temp\_sim"}\NormalTok{, }\StringTok{"date\_sim"}\NormalTok{)}

\DocumentationTok{\#\#\# smallmouth mod}
\NormalTok{smallmouth }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{data.frame}\NormalTok{()}
\FunctionTok{colnames}\NormalTok{(smallmouth) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"hours"}\NormalTok{, }\StringTok{"temp\_F"}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{hours }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{52}\NormalTok{, }\DecValTok{54}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{78}\NormalTok{, }\DecValTok{90}\NormalTok{, }\DecValTok{98}\NormalTok{, }\DecValTok{150}\NormalTok{, }\DecValTok{167}\NormalTok{, }\DecValTok{238}\NormalTok{, }\DecValTok{234}\NormalTok{)}
\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_F }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{77}\NormalTok{, }\DecValTok{75}\NormalTok{, }\DecValTok{71}\NormalTok{, }\DecValTok{70}\NormalTok{, }\DecValTok{67}\NormalTok{, }\DecValTok{65}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{59}\NormalTok{, }\DecValTok{55}\NormalTok{, }\DecValTok{55}\NormalTok{)}

\CommentTok{\# change F to C and hours to days}
\NormalTok{smallmouth }\OtherTok{\textless{}{-}}\NormalTok{ smallmouth }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{days =} \FunctionTok{ceiling}\NormalTok{(hours}\SpecialCharTok{/}\DecValTok{24}\NormalTok{),}
         \AttributeTok{temp\_C =}\NormalTok{ (temp\_F }\SpecialCharTok{{-}}\DecValTok{32}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ (}\DecValTok{5}\SpecialCharTok{/}\DecValTok{9}\NormalTok{))}

\NormalTok{smb\_mod }\OtherTok{\textless{}{-}} \FunctionTok{fit\_model}\NormalTok{(}\AttributeTok{temp =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C,}
                     \AttributeTok{days =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{days,}
                     \AttributeTok{species =} \StringTok{"smb"}\NormalTok{,}
                     \AttributeTok{development\_type =} \StringTok{"hatch"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Note the \emph{R\textsuperscript{2}} fit from the models below. You can
see they generally all preform well and are in line with values from
model 2 of Beacham and Murray (1990).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#model fits}
\NormalTok{smb\_mod}\SpecialCharTok{$}\NormalTok{r\_squared; cat\_mod}\SpecialCharTok{$}\NormalTok{r\_squared; sturgeon\_mod}\SpecialCharTok{$}\NormalTok{r\_squared}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.9868067
\end{verbatim}

\begin{verbatim}
## [1] 0.9433598
\end{verbatim}

\begin{verbatim}
## [1] 0.9217358
\end{verbatim}

\includegraphics[width=1\linewidth]{paper_oxford_files/figure-latex/unnamed-chunk-10-1}

\subsection{Predicting Phenology and
Output}\label{predicting-phenology-and-output}

To illustrate model selection and phenology prediction we will recreate
a small portion of the analysis done by Sparks et al. (2019) using the
\texttt{woody\_island} dataset included in this package . We will
predict both hatch and emergence timing, so we will obtain a model
expression for each using \texttt{model\_select()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sockeye\_hatch\_mod }\OtherTok{\textless{}{-}} \FunctionTok{model\_select}\NormalTok{(}
  \AttributeTok{author =} \StringTok{"Beacham and Murray 1990"}\NormalTok{, }
  \AttributeTok{species =} \StringTok{"sockeye"}\NormalTok{, }
  \AttributeTok{model =} \DecValTok{2}\NormalTok{, }
  \AttributeTok{development\_type =} \StringTok{"hatch"}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

We can now use our model expressions to predict when sockeye would hatch
and emerge at Woody Island in 1990. First we predict hatch timing using
\texttt{predict\_phenology()}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{WI\_hatch }\OtherTok{\textless{}{-}} \FunctionTok{predict\_phenology}\NormalTok{(}
  \AttributeTok{data =}\NormalTok{ woody\_island,}
  \AttributeTok{dates =}\NormalTok{ date,}
  \AttributeTok{temperature =}\NormalTok{ temp\_c,}
  \AttributeTok{spawn.date =} \StringTok{"1990{-}08{-}18"}\NormalTok{,}
  \AttributeTok{model =}\NormalTok{ sockeye\_hatch\_mod}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

And then look inside the returned object to see days to hatch and
development period:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{WI\_hatch}\SpecialCharTok{$}\NormalTok{days2done}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## NULL
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{WI\_hatch}\SpecialCharTok{$}\NormalTok{dev.period}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        start       stop
## 1 1990-08-18 1990-10-30
\end{verbatim}

\paragraph{Understanding your results}\label{understanding-your-results}

The output from \texttt{predict\_phenology()} includes a lot of
information. If we look at our \texttt{WI\_hatch} object we see there
are multiple elements stored in a list which can be accessed using the
\texttt{\$} operator.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{str}\NormalTok{(WI\_hatch)}
\end{Highlighting}
\end{Shaded}

\texttt{WI\_hatch\$days2done} outputs the predicted days to develop.

\texttt{WI\_hatch\$dev.period} is a 1x2 dataframe with the dates
corresponding to when your fish's parent spawned (which you input with
\texttt{predict\_phenology(spawn.date\ =\ ...)}) and the date when the
fish is predicted todevelop.

\texttt{WI\_hatch\$ef.vals} is a vector of each day's effective value as
evaluated using whatever model is chosen.

\texttt{WI\_hatch\$ef.tibble} is a \emph{n} x 4 tibble (\emph{n} =
number of days to hatch or emerge) and the columns are the date, each
day's temperature and effective value, and the cumulative sum of the
effective values. The \texttt{ef.tibble} object is meant to serve as the
basis for users to make custom figures for their data beyond the
functionality we discuss below.

\paragraph{Plotting phenology}\label{plotting-phenology}

\textbf{hatchR} has a built in function, \texttt{plot\_phenology()},
that allows users to visualize their phenology results (Figure 4). The
plot visualizes three specific components: 1.) the temperature regime
over which you are predicting, 2.) the cumulative sum of effective
values, and 3.) the effective value for each day in your prediction
span. The function allows you to output various figures based on your
interests, but defaults to a figure with all information and the
corresponding labels.

\texttt{\{r"\}\ plot\_phenology(WI\_hatch)}

\section{Case Study 1}\label{case-study-1}

A common management scenario where developmental phenology might be
useful would be trying to understand if fish might be free-moving before
some management action. For instance, will fish have emerged from redds
when a stream section has been opened to grazing or road work?

In this scenario, we will consider the road work example and Bull Trout,
a threatened fish in the United States under the Endangered Species Act
(Nolfi et al. 2024), and the Crooked River, a key Bull Trout population
in the Boise River watershed. In this hypothetical scenario, the Forest
Fisheries Biologist wants to know if fish will likely be out of the
gravel and free-swimming by June 1st as Bull Trout are particularly
sensitive to sediment. In this system, it is expected that Bull Trout
will be done spawning by the end of September, so we'll consider the
last possible spawn date as September 30th.

We demonstrate this first case study using the graphical user interface
portion of the \textbf{hatchR} ecosystem found at
\url{https://elifelts.shinyapps.io/hatchR/}. Users will first upload
their data with the \texttt{Import\ Data} window, which requires them to
select their file on their personal computer, provide the program with
the columns corresponding for dates and temperatures, and then provide
the format in which dates are coded (\emph{e.g.}, year-month-day or
day-month-year). Data used in this case study are from the
\texttt{crooked\_river} data set from R and can either be written out
from the package or accessed at
\url{https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/}.
Once data is uploaded the program automatically plots the user's data
using \texttt{plot\_check\_temp()} in the background and provides them
the outputted graphical check. After uploading and checking data, the
user switches to the \texttt{Model\ Phenology} window. In this
circumstance, we use the preloaded parameterization for bull trout from
Austin, Essington, and Quinn (2019) with the \texttt{Existing} button
for model selection, which the user selects with the various drop down
options in the menu. After the model is selected, the user can choose
multiple spawn dates from the interactive calendar provided. We show
results for spawning for September 30th as indicated in the example
above. Once dates are chosen, a table entry for each spawn date is
outputted in the \texttt{Phenology\ Summaries} tab and corresponding
plot with data from each spawn date in the \texttt{Timeline\ Plot} tab.
Output from predicting phenology and the resulting figure are
downloadable from their respective tabs. The process is demonstrated in
full in Figure XXX, but the interface is described more completely on
\textbf{hatchR}'s Shiny website.

In this example we expect the last fish out of the gravel well before
the June 1st date and the manager could allow grazing in this area
without worrying about direct mechanical disturbance to fish developing
in the gravel.

\section{Case Study 2}\label{case-study-2}

For the second example, we will again use bull trout, but demonstrate a
much more complex application for the purpose of showing the full
flexibility of the programmatic application of \textbf{hatchR}. In this
scenario we will use data from Isaak et al. (2018) (included
\texttt{idaho} dataset), which includes temperature data from 226 sites
across the major upper Columbia River headwater watersheds in Idaho. For
this approach we winnow putative bull trout spawning sites by filtering
for sites with mean August temperature \textless/= 13 °C in accordance
with thresholds from Isaak et al. (2015). For the resulting 139 sites we
will demonstrate predicting hatch timing in these putative Bull Trout
spawning habitats.

We need to setup our models and data for this analysis, which we don't
show those steps here for the sake of concision in this manuscript,
however they are demonstrated in \texttt{paper.Rmd} included in the
GitHub repository for \textbf{hatchR}. After the setup, we can easily
map \texttt{predict\_phenology()} across all putative spawning sites and
three spawn dates (September 1-Early Spawning, September 15-Peak
Spawning, and September 31-Late Spawning), the results of which are
presented in Figure XXX.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hatch\_res }\OtherTok{\textless{}{-}}\NormalTok{ isaak\_summ\_bt }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{dev\_period =} \FunctionTok{map2}\NormalTok{(}
\NormalTok{      summ\_obj, spawn\_dates, }\CommentTok{\# map across our site object and spawn dates}
\NormalTok{      predict\_phenology,}
      \AttributeTok{temperature =}\NormalTok{ daily\_temp,}
      \AttributeTok{model =}\NormalTok{ bt\_hatch,}
      \AttributeTok{dates =}\NormalTok{ date}
\NormalTok{      ) }\SpecialCharTok{|\textgreater{}} 
      \FunctionTok{map\_df}\NormalTok{(}\StringTok{"dev.period"}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \CommentTok{\# pull out just dev.period results}
      \FunctionTok{list}\NormalTok{()}
\NormalTok{    ) }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{select}\NormalTok{(site, dev\_period) }\SpecialCharTok{|\textgreater{}} \CommentTok{\# just select the columns we want}
  \FunctionTok{unnest}\NormalTok{(}\AttributeTok{cols =} \FunctionTok{c}\NormalTok{(dev\_period)) }\SpecialCharTok{|\textgreater{}} \CommentTok{\# unnest everything}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{days\_to\_hatch =}\NormalTok{ stop }\SpecialCharTok{{-}}\NormalTok{ start) }\CommentTok{\# make a new column of days to hatch}
\end{Highlighting}
\end{Shaded}

\includegraphics[width=1\linewidth]{paper_oxford_files/figure-latex/unnamed-chunk-18-1}

\section{Discussion}\label{discussion}

With \textbf{hatchR} we present a software ecosystem that bridges the
analytical gap of predicting developmental phenology for wild fishes and
develops a formal framework for applying effective value models from
user-provided parameterizations. To do so, the software is bundled in
two forms, a fully customizable R package, especially useful for running
repetitive analyses and a graphical-user-interface designed for ease of
use and tasks that may only run once or a handful of times. Both
applications allows users to import their data, run basic data checks,
and apply historic model parameterizations for salmonids or create their
own species- or population specific parameterization. Additionally, we
provide substantial documentation on the \textbf{hatchR} website
(\url{https://bmait101.github.io/hatchR/}) to walk users through basic
to advanced applications of programmatic platform.

In the application of using effective value models \textbf{hatchR} and
the user make some key assumptions. Particularly, numerous studies have
indicated that stressful environmental conditions can cue fish to
prematurely hatch or emerge from their environment. These include water
quality like dissolved oxygen, pH, or salinity, pathogens, and even
mechanical agitation (reviewed in Quinn (2018) and Cowan et al. (2024)).
Moreover, while the \textbf{hatchR} provides point estimates for
developmental phenology, spawning and developmental within populations
of fishes both occur as distributions as opposed single events Mason
(1976) and we encourage users to either predict phenology using early,
peak, and late frameworks (\emph{e.g.}, 5th, 50th, and 95th quantiles)
or with real or modeled distributions. Another factor that could bias
estimates from these models is that temperatures used from sensors don't
reflect the geomoprhic properties in the environment where eggs are
developing such that they may be too cold or warm Geist et al. (2002).

To date, the application of the effective value framework has largely
focused using species-specific models to predict phenology in wild
environments (see Adelfio et al. (2024), Austin, Essington, and Quinn
(2019)). However, we emphasize that the statistical relationships at the
basis of these models are fundamentally reaction norms, meaning that
both the intercept and slope of the responses of fishes to temperature
are reflective of family- or population-level genetic variation, genetic
x environment responses, or may be indicative of phylogenetic
differences among species (West-Eberhard (2003)). For instance, while
Sparks et al. (2017) did not find differences in the rates of
development between the focal populations of their study, they did
observe family-level genetic x environment interactions across different
thermal regimes. Similarly, when they reparameterized their models using
fish from western Alaska, they found differences in both the slope and
intercept of the model relative to the original parameterization from
Beacham and Murray (1990), notably derived from multiple Canadian
stocks. The western Alaskan fish generally developed more slowly than
their southerly counterparts, consistent with cogradient variation
(Conover, Duffy, and Hice (2009), Sparks et al. (2022)). In this sense,
we encourage users not to only think about the end points of these
models (days to development) but also how the statistical relationships
they are premised on inform our understanding of micro- and and
macro-evolutionary processes.

\textbf{BRYAN check the below}

The models presented in \textbf{hatchR} can be further customized in
multiple ways beyond the use-cases provided above. For instance, while
our models are fit to predict hatch or emergence timing, they could be
used to predict other developmental stages prior to the initation of
external feeding like eye-up or other embryological developmental stages
(Velsen (1980)) or \textbf{Some example from a non-salmonid, maybe
initiation of pelagic-larval or riverine current dispersal}.
Additionally, while \texttt{fit\_model()} relies on non-linear modeling
platform, \texttt{predict\_phenology()} only requires users to pass a
model expression, so if users preferred a different model fit than the
non-linear option provided (as long as it integrates daily temperature),
they could pass a different expression into
\texttt{predict\_phenology()} for additional customization. Finally,
while \textbf{hatchR} was designed specifically for fishes, we expect
other poikilothermic organisms such as reptiles, amphibians, or
invertebrates would develop in accordance with the power law and these
developmental models could be extended to other organisms beyond fishes.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-adelfio2019}
Adelfio, Luca A., Steven M. Wondzell, Nathan J. Mantua, and Gordon H.
Reeves. 2019. {``Warm Winters Reduce Landscape-Scale Variability in the
Duration of Egg Incubation for Coho Salmon (Oncorhynchus Kisutch) on the
Copper River Delta, Alaska.''} \emph{Canadian Journal of Fisheries and
Aquatic Sciences} 76 (8): 1362--75.
\url{https://doi.org/10.1139/cjfas-2018-0152}.

\bibitem[\citeproctext]{ref-adelfio2024}
---------. 2024. {``Expanded, Compressed, or Equal? Interactions Between
Spawning Window and Stream Thermal Regime Generate Three Responses in
Modeled Juvenile Emergence for Pacific Salmon.''} \emph{Canadian Journal
of Fisheries and Aquatic Sciences}, February.
\url{https://doi.org/10.1139/cjfas-2023-0238}.

\bibitem[\citeproctext]{ref-alderdice1978}
Alderdice, D. F., and F. P. J. Velsen. 1978. {``Relation Between
Temperature and Incubation Time for Eggs of Chinook Salmon (Oncorhynchus
Tshawytscha).''} \emph{Journal of the Fisheries Research Board of
Canada} 35 (1): 69--75. \url{https://doi.org/10.1139/f78-010}.

\bibitem[\citeproctext]{ref-austin2019}
Austin, Catherine S., Timothy E. Essington, and Thomas P. Quinn. 2019.
{``Spawning and Emergence Phenology of Bull Trout Salvelinus Confluentus
Under Differing Thermal Regimes.''} \emph{Journal of Fish Biology} 94
(1): 191--95. \url{https://doi.org/10.1111/jfb.13864}.

\bibitem[\citeproctext]{ref-beacham1990}
Beacham, Terry D., and Clyde B. Murray. 1990. {``Temperature, Egg Size,
and Development of Embryos and Alevins of Five Species of Pacific
Salmon: A Comparative Analysis.''} \emph{Transactions of the American
Fisheries Society} 119 (6): 927--45.
\url{https://doi.org/10.1577/1548-8659(1990)119\%3C0927:TESADO\%3E2.3.CO;2}.

\bibitem[\citeproctext]{ref-chang2024}
Chang, Winston, Joe Cheng, J. J. Allaire, Carson Sievert, Barret
Schloerke, Yihui Xie, Jeff Allen, et al. 2024. \emph{Shiny: Web
Application Framework for r}.
\url{https://cran.r-project.org/web/packages/shiny/index.html}.

\bibitem[\citeproctext]{ref-conover2009}
Conover, David O., Tara A. Duffy, and Lyndie A. Hice. 2009. {``The
Covariance Between Genetic and Environmental Influences Across
Ecological Gradients.''} \emph{Annals of the New York Academy of
Sciences} 1168 (1): 100--129.
\url{https://doi.org/10.1111/j.1749-6632.2009.04575.x}.

\bibitem[\citeproctext]{ref-cowan2024}
Cowan, Zara-Louise, Leon Green, Timothy D. Clark, Tamzin A. Blewett,
Jeremy De Bonville, Thomas Gagnon, Elizabeth Hoots, et al. 2024.
{``Global Change and Premature Hatching of Aquatic Embryos.''}
\emph{Global Change Biology} 30 (9): e17488.
\url{https://doi.org/10.1111/gcb.17488}.

\bibitem[\citeproctext]{ref-geist2002}
Geist, David R., Timothy P. Hanrahan, Evan V. Arntzen, Geoffrey A.
McMichael, Christopher J. Murray, and Yi-Ju Chien. 2002.
{``Physicochemical Characteristics of the Hyporheic Zone Affect Redd
Site Selection by Chum Salmon and Fall Chinook Salmon in the Columbia
River.''} \emph{North American Journal of Fisheries Management} 22 (4):
1077--85.
\url{https://doi.org/10.1577/1548-8675(2002)022\%3C1077:PCOTHZ\%3E2.0.CO;2}.

\bibitem[\citeproctext]{ref-isaak2018}
Isaak, Daniel J., Charles H. Luce, Gwynne L. Chandler, Dona L. Horan,
and Sherry P. Wollrab. 2018. {``Principal Components of Thermal Regimes
in Mountain River Networks.''} \emph{Hydrology and Earth System
Sciences} 22 (12): 6225--40.
\url{https://doi.org/10.5194/hess-22-6225-2018}.

\bibitem[\citeproctext]{ref-isaak2015}
Isaak, Daniel J., Michael K. Young, David E. Nagel, Dona L. Horan, and
Matthew C. Groce. 2015. {``The Cold-Water Climate Shield: Delineating
Refugia for Preserving Salmonid Fishes Through the 21st Century.''}
\emph{Global Change Biology} 21 (7): 2540--53.
\url{https://doi.org/10.1111/gcb.12879}.

\bibitem[\citeproctext]{ref-kaylor2021}
Kaylor, Matthew J., Casey Justice, Jonathan B. Armstrong, Benjamin A.
Staton, Lauren A. Burns, Edwin Sedell, and Seth M. White. 2021.
{``Temperature, Emergence Phenology and Consumption Drive Seasonal
Shifts in Fish Growth and Production Across Riverscapes.''}
\emph{Journal of Animal Ecology} 90 (7): 1727--41.
\url{https://doi.org/10.1111/1365-2656.13491}.

\bibitem[\citeproctext]{ref-mason1976}
Mason, JC. 1976. {``Some Features of Coho Salmon, Oncorhynchus Kisutch,
Fry Emerging from Simulated Redds and Concurrent Changes in
Photobehavior.''} \emph{Fish. Bull} 74 (1): 167--75.

\bibitem[\citeproctext]{ref-mcphail1979}
McPhail, JD, and CB Murray. 1979. {``The Early Life-History and Ecology
of Dolly Varden (Salvelimus Malma) in the Upper Arrow Lakes.''} Nelson,
BC, Canada.

\bibitem[\citeproctext]{ref-nolfi2024}
Nolfi, Dan, Tracy Melbihess, Sandi Fisher, and Lisa Ellis. 2024.
{``5-Year Statuse Review Coterminous United States Population of Bull
Trout (Salvelinus Confluentus).''}

\bibitem[\citeproctext]{ref-quinn2018}
Quinn, Thomas P. 2018. \emph{The Behavior and Ecology of Pacific Salmon
and Trout}. University of Washington Press.

\bibitem[\citeproctext]{ref-r:ther}
{``R: The r Project for Statistical Computing.''} n.d.
\url{https://www.r-project.org/}.

\bibitem[\citeproctext]{ref-siegel2023}
Siegel, Jared E., Aimee H. Fullerton, Alyssa M. FitzGerald, Damon
Holzer, and Chris E. Jordan. 2023. {``Daily Stream Temperature
Predictions for Free-Flowing Streams in the Pacific Northwest, USA.''}
\emph{PLOS Water} 2 (8): e0000119.
\url{https://doi.org/10.1371/journal.pwat.0000119}.

\bibitem[\citeproctext]{ref-small2001}
Small, Brian C., and Terry D. Bates. 2001. {``Effect of Low-Temperature
Incubation of Channel Catfish Ictalurus Punctatus Eggs on Development,
Survival, and Growth.''} \emph{Journal of the World Aquaculture Society}
32 (2): 189--94.
\url{https://doi.org/10.1111/j.1749-7345.2001.tb01094.x}.

\bibitem[\citeproctext]{ref-smith2005}
Smith, K. M., and D. K. King. 2005. {``Dynamics and Extent of Larval
Lake Sturgeon Acipenser Fulvescens Drift in the Upper Black River,
Michigan.''} \emph{Journal of Applied Ichthyology} 21 (3): 161--68.
\url{https://doi.org/10.1111/j.1439-0426.2005.00623.x}.

\bibitem[\citeproctext]{ref-sparks2019}
Sparks, Morgan M., Jeffrey A. Falke, Thomas P. Quinn, Milo D. Adkison,
Daniel E. Schindler, Krista Bartz, Dan Young, and Peter A. H. Westley.
2019. {``Influences of Spawning Timing, Water Temperature, and Climatic
Warming on Early Life History Phenology in Western Alaska Sockeye
Salmon.''} \emph{Canadian Journal of Fisheries and Aquatic Sciences} 76
(1): 123--35. \url{https://doi.org/10.1139/cjfas-2017-0468}.

\bibitem[\citeproctext]{ref-sparks2022}
Sparks, Morgan M., Joshua C. Kraft, Kliffi M. S. Blackstone, Gordon G.
McNickle, and Mark R. Christie. 2022. {``Large Genetic Divergence
Underpins Cryptic Local Adaptation Across Ecological and Evolutionary
Gradients.''} \emph{Proceedings of the Royal Society B: Biological
Sciences} 289 (1984): 20221472.
\url{https://doi.org/10.1098/rspb.2022.1472}.

\bibitem[\citeproctext]{ref-sparks2017}
Sparks, Morgan M., Peter A. H. Westley, Jeffrey A. Falke, and Thomas P.
Quinn. 2017. {``Thermal Adaptation and Phenotypic Plasticity in a
Warming World: Insights from Common Garden Experiments on Alaskan
Sockeye Salmon.''} \emph{Global Change Biology} 23 (12): 5203--17.
\url{https://doi.org/10.1111/gcb.13782}.

\bibitem[\citeproctext]{ref-velsen1980}
Velsen, F. P. J. 1980. {``Embryonic Development in Eggs of Sockeye
Salmon Oncorhynchus Nerka.''} \emph{Canadian Special Publication of
Fisheries and Aquatic Sciences}, no. 49.

\bibitem[\citeproctext]{ref-webster1948}
Webster, Dwight A. 1948. {``Relation of Temperature to Survival and
Incubation of the Eggs of Smallmouth Bass (Micropterus Dolomieu).''}
\emph{Transactions of the American Fisheries Society} 75 (1): 43--47.
\url{https://doi.org/10.1577/1548-8659(1945)75\%5B43:ROTTSA\%5D2.0.CO;2}.

\bibitem[\citeproctext]{ref-west-eberhard2003}
West-Eberhard, Mary Jane. 2003. \emph{Developmental Plasticity and
Evolution}. Oxford University Press.

\end{CSLReferences}

\section{Competing interests}

There are no competing interest.

\section{Author contributions statement}

MMS and BMM designed and wrote the package. EF wrote the Shiny app. All
authors wrote and edited the manuscript.

\section{Acknowledgments}

We thank Laura Koller for her help designing the \textbf{hatchR} logo.
Dan Isaak provided useful discussion about model development and
temperature datasets.

The views expressed in this manuscript are those of the authors and do
not necessarily represent the views or policies of USFS or USFWS. Any
mention of trade names, products, or services does not imply an
endorsement by the U.S. government, USFS, or USFWS. USFS and USFWS do
not endorse any commercial products, services or enterprises.

%% Author bio-pics with images

\end{document}

### 2025-02-03 R/model_table.R
chnage from Austin et al. 2017 to 2019 - fixes #11

@@ -2,7 +2,7 @@
 #'
 #' Table providing the model parameterizations for the phenology models.
 #'
-#' @source Beacham & Murray (1990), Sparks et al. (2019), Austin et al. (2019).
+#' @source Beacham & Murray (1990), Sparks et al. (2017), Austin et al. (2019).
 #' @format Data frame with 5 columns:
 #' \describe{
 #' \item{author}{Author-year short citation}
@@ -20,10 +20,9 @@
 #'   \emph{Canadian Journal of Zoology},
 #'   \bold{68}, 1931--1940.
 #'
-#' Sparks, M.M., Falke, J.A., Quinn, T.A., Adkinson, M.D.,
-#' Schindler, D.E. (2019). Influences of spawning timing, water temperature,
-#' and climatic warming on early life history phenology in western
-#' Alaska sockeye salmon.
+#' Sparks, M.M., Westley, P.A.H., Falke, J.A., Quinn, T.A. (2017). Thermal
+#' adaptation and phenotypic plasticity in a warming world: Insights from
+#' common garden experiments on Alaskan sockeye salmon.
 #'   \emph{Canadian Journal of Fisheries and Aquatic Sciences},
 #'   \bold{76(1)}, 123--135.
 #'

### 2025-02-14 inst/manuscript/oxford/oxford.Rmd

---
title: Template for Oxford University Press papers
author:
  - name: Alice Anonymous
    email: alice@example.com
    corresponding_author: TRUE # Automatic footnote using corresponding_author
    affiliation:
      - ABC
      - DEF
    biography: |
      A promising young researcher working on her thesis project.
#    biography_pic: "/PATH/TO/image.png"
#    biography_pic_width: "1in"
  - name: Bob Security
    email: bob@example.com
    affiliation: DEF
    footnote: 1 # Footnote using reference
  - name: Cat Memes
    email: cat@anotheru.edu
    affiliation: Another University
    footnote:
        - 1
        - 2
address:
  - code: ABC
    department: Department of Biology
    organization: University of Idaho
    street: 123 Main Street
    #postcode: 
    city: Boise
    state: Idaho 07458 # Mix and match elements to get proper address
    country: United States
  - code: DEF
    address: Address in one element deprecated, but works
  - code: Another University
    address: Address for Another University
footnote:
  - code: 1
    text: "Equal contribution"
  - code: 2
    text: "Current email address: [cat@example.com](mailto:cat@example.com)"
abstract:
  - "This is the first paragraph of the abstract. It has several sentences making it go over several lines. For this it needs to have a lot of text." 
  - "This is the second paragraph."
  - "**Motivation:** You can also have some paragraphs start with bold face."
keywords:
  - keyword1
  - keyword2
  - keyword3
#boxedlist: # Not fully compatible with all document formats
#  - First point
#  - Second point
competing_interests: |
  There are no competing interest.
author_contributions: |
  AA did all the work. The others are just freeloaders.
acknowledgments: |
  This is an acknowledgement.

  It consists of two paragraphs.

bibliography: mybibfile.bib # Includes refs in OUP example template
## When using `citation_package = "default"`, a CSL file can be used
#csl: https://www.zotero.org/styles/oxford-university-press-note
output:
  rticles::oup_article:
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    journal: "Journal Name"
    document_style: "contemporary" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    #number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    #onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below

## Example of pandoc's variable
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
  #- \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \theoremstyle{thmstyleone} # Theorem stuff from OUP template
  - \newtheorem{theorem}{Theorem} #  meant for continuous numbers. %%\newtheorem{theorem}{Theorem}[section] # meant for sectionwise numbers. optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  - \newtheorem{proposition}[theorem]{Proposition} # %%\newtheorem{proposition}{Proposition}" # to get separate numbers for theorem and proposition etc.
  - \theoremstyle{thmstyletwo}
  - \newtheorem{example}{Example}
  - \newtheorem{remark}{Remark}
  - \theoremstyle{thmstylethree}
  - \newtheorem{definition}{Definition}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, # By default, hide code; set to TRUE to see code
  fig.pos = 'th', # Places figures at top or here
  out.width = '100%', dpi = 300, # Figure resolution and size
  fig.env="figure"
) # Latex figure environment

options(knitr.table.format = "latex") # For kable tables to write LaTeX table directly
```

# Introduction

This template is based on the generic OUP authoring template available on CTAN under [oup-authoring-template](https://www.ctan.org/pkg/oup-authoring-template). The CTAN template includes LaTeX documentation and a sample LaTeX document that provide far more details regarding the full functionality of the format. Here, only the basic functioning of the Rmarkdown adaptation of the format is demonstrated.

## A subsection

A numbered list:

1) First point
2) Second point
    - Subpoint

A bullet list:

* First point
* Second point

## Notes

* Extra white space in document will tend to disappear as text is filled in.
* Code blocks tend to generate lots of empty white space when `echo=TRUE` for some reason.

# Literature citations

By default, citations are handled by `natbib` using a numeric citation format. To use name-date citations, sets `namedate: TRUE` in the YAML header.

Here are two sample references:

* **author (year) example:** @horvath2018dna showed some really cool things. Only seems to work properly if `namedate: TRUE`. 
* **(author year) example:** This is a well known result [@ji20123d].

The bibliography will appear at the end of the document.

Though not normally available in the OUP LaTeX format, [CSL style files](https://www.zotero.org/styles) can also be used with the Rmarkdown adaptation by setting in the YAML header `citation_package: "default"` and defining the `csl` element to be the path towards the style file.

# Equations

An equation without a label for cross-referencing:

$$
E=mc^2
$$

An inline equation: $y=ax+b$

An equation with a label for cross-referencing:

\begin{equation}\label{eq:eq1}
\int^{r_2}_0 F(r,\varphi){\rm d}r\,{\rm d}\varphi = 1
\end{equation}

This equation can be referenced as follows: Eq. \ref{eq:eq1}

# Inserting R figures

The code below creates a figure. The code is included in the output because `echo=TRUE`.

```{r fig1, fig.cap="This is the first figure.",echo=TRUE}
plot(1:10,main="Some data",xlab="Distance (cm)",
     ylab="Time (hours)")
```

You can reference this figure as follows: Fig. \ref{fig:fig1}.

## Figures spanning two-columns

Figures can span two columns be setting `fig.env="figure*"`.

```{r fig2, fig.cap="This is a wide figure.",fig.env="figure*"}
plot(1:5,pch=19,main="Some wide data",
     xlab="Distance (cm)",ylab="Time (hours)")
```

Reference to second figure: Fig. \ref{fig:fig2}

# Tables

## Generate a table using `xtable`

```{r xtabletable, results="asis", echo=TRUE}
df = data.frame(ID=1:3,code=letters[1:3])

# Creates tables that follow OUP guidelines 
# using xtable
library(xtable) 
print(xtable(df,caption="This is a xtable table.",
             label="tab:tab1"),
      comment=FALSE,caption.placement="top")
```

You can reference this table as follows: Table \ref{tab:tab1}.

## Generate a table using `kable`

```{r kabletable, echo=TRUE}
df = data.frame(ID=1:3,code=letters[1:3])

# kable can alse be used for creating tables
knitr::kable(df,caption="This is a kable table.",
             booktabs=TRUE,label="tab2")
```

You can reference this table as follows: Table \ref{tab:tab2}.

## Table spanning two columns

Tables can span two columns be setting `table.envir = "table*"` in `knitr::kable`.

```{r widetable, echo=TRUE}
df = data.frame(ID=1:3,code1=letters[1:3],
                code2=letters[4:6],
                code3=letters[7:9],
                code4=letters[10:12],
                code5=letters[13:15])

# kable can alse be used for creating tables
knitr::kable(df,caption="This is a wide kable table.",
             #format="latex",
             table.envir="table*",
             booktabs=TRUE,label="tab3")
```

# Cross-referencing sections

You can cross-reference sections and subsections as follows: Section \ref{literature-citations} and Section \ref{a-subsection}.

***Note:*** the last section in the document will be used as the section title for the bibliography.

For more portable and flexible referencing of sections, equations, figures and tables, use [`bookdown::pdf_document2`](https://github.com/rstudio/bookdown) with YAML header option `base_format: rticles::oup_article`.

# Appendices {-}

::: {.appendices latex=true}

# Section title of first appendix

blabla

## Subsection title of first appendix

and so on....
:::

# References

### 2025-02-14 inst/manuscript/paper.tex

@@
 \renewcommand{\headrulewidth}{0pt}
-\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{C:/Users/BryanMaitland/AppData/Local/R/win-library/4.4/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
+\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{/Library/Frameworks/R.framework/Versions/4.4-arm64/Resources/library/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
 \fancyhead[C]{}
@@
 
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
 % Pandoc citation processing
@@
 Despite these improvements, their application remains limited due to few
-parameterizations for different species, being largely limited to
-salmonids. Here we present \texttt{hatchR}, a software ecosystem that
+parameterizations for varying species, being largely limited to
+salmonids. Here we present \textbf{hatchR}, a software ecosystem that
 allows users to predict hatch and emergence timing for wild fishes, as
-well as additional tools to aid in those analyses. \texttt{hatchR}
+well as additional tools to aid in those analyses. \textbf{hatchR}
 allows users to leverage popular historic parameterizations for
 phenological models or to easily implement custom parameterizations
-using data not included in the package. \texttt{hatchR} is also
+using data not included in the package. \textbf{hatchR} is also
 distributed in two forms---an open source R package for maximum
@@
 demonstrate potential uses, we present two case studies as likely
-applications for this software. \texttt{hatchR} promises to open many
+applications for this software. \textbf{hatchR} promises to open many
 exciting avenues in research and management of fishes during their early
@@
 applicable to wild populations because they assumed a constant
-temperature over the course of development Beacham \& Murray (1990)
-(\textbf{add more!}) . However, Sparks et al. (2019) reformulated this
-approach as an ``Effective Value model'', in which the input was daily
-average temperature after a parent spawned and fish would either hatch
-or emerge when effective values cumulatively summed to one.
+temperature over the course of development (Alderdice \& Velsen, 1978;
+Beacham \& Murray, 1990; McPhail \& Murray, 1979). However, Sparks et
+al. (2019) reformulated this approach as an ``Effective Value model'',
+in which the input was daily average temperature after a parent spawned
+and fish would either hatch or emerge when effective values cumulatively
+summed to one.
 
 The resulting effective value approach has now been widely applied in
-salmonids for which parameterizations from aquaculture were readily
+Salmonids for which parameterizations from aquaculture were readily
 available---for example Pacific Salmon (\emph{Oncorhynchus spp.}) models
-developed by Beacham \& Murray (1990) were have been applied to various
-species and populaions (Adelfio, Wondzell, Mantua, \& Reeves, 2019,
+developed by Beacham \& Murray (1990) have been applied to various
+species and populations (Adelfio, Wondzell, Mantua, \& Reeves, 2019,
 2024; Kaylor et al., 2021) while models developed for Bull Trout
@@
 parameterization for other species, we developed the software ecosystem,
-\texttt{hatchR}. Specifically, \texttt{hatchR} allows users to input
+\textbf{hatchR}. Specifically, \textbf{hatchR} allows users to input
 standard raw or summarized temperature datasets that are commonly
@@
 parameterizations like those from Beacham \& Murray (1990) or Sparks,
-Westley, Falke, \& Quinn (2017), develop custom models from their own
-temperature and phenological data, and predict hatch and emergence
-timing using these models in the effective value framework.
+Westley, Falke, \& Quinn (2017), develop custom models from their own or
+published temperature and phenological data, and predict hatch and
+emergence timing using these models in the effective value framework.
 
 To widen the user application of these methods, we distribute two
-user-interfaces for \texttt{hatchR}. The first is a R package
-distributed via CRAN that allows users the most customizable application
-of these methods. The R package is especially powerful as it allows
-users to automate their analyses over multiple variables such as
-phenology type, multiple spawn dates, or different habitats with varying
-thermal regimes. These variable approaches are outlined in the package
-documentation on \texttt{hatchR}'s website. Alternatively, we also
-distribute a Shiny application in the form of an HTML-based web tool to
-interact with many of \texttt{hatchR}'s functions in a
-graphical-user-interface. The Shiny form trades-off some of automative
-power for user simplicity, while still allowing users to leverage much
-of the functionality of \texttt{hatchR}'s R package. Below, we present
-the basic overview of the software and multiple case studies of how it
-may be applied.
+user-interfaces for \textbf{hatchR}. The first is a R package ({``R''}
+(n.d.)) distributed via CRAN that allows users the most customizable
+application for these methods. The R package is especially powerful as
+it allows users to automate their analyses over multiple variables such
+as phenology type, multiple spawn dates, or different habitats with
+varying thermal regimes. These variable approaches are outlined in the
+package documentation on \textbf{hatchR}'s website. Alternatively, we
+also distribute a Shiny application (Chang et al. (2024)) in the form of
+an HTML-based web tool to interact with many of \textbf{hatchR}'s
+functions in a graphical-user-interface. The Shiny form trades-off some
+of automative power for user simplicity, while still allowing users to
+leverage much of the functionality of \textbf{hatchR}'s R package.
+Below, we present the basic overview of the software and multiple case
+studies of how it may be applied.
 
@@
 
-\texttt{hatchR} is meant to primarily be a tool for predicting
-phenology. In this sense, we mostly limit functionality to these
-applications and provide minimal data checking and plotting help. This
-decision is in part driven by the diversity of data types that users may
-import and the difficulty in addressing all those data types with
-respect to various data checks. In other words, we expect users to know
-their data better than we do and to check it accordingly. We do provide
-two basic data check functions discussed in the Checking Data section.
-Similarly, we provide limited functionality for plotting results, but
-provide examples of how to build custom visualization from output,
-specifically in R. For the Shiny application, we provide a base output
-plot, but the ability to download your results for custom plotting in
-programs of the user's choice.
+\textbf{hatchR} is meant to primarily be a tool for predicting
+phenology. In this sense, we limit functionality to these applications
+and provide minimal data checking and plotting help. This decision is in
+part driven by the diversity of data types that users may import and the
+difficulty in addressing all those data types with respect to various
+data checks. In other words, we expect users to know their data better
+than we do and to check it accordingly. We do provide two basic data
+check functions discussed in the Checking Data section. Similarly, we
+provide limited functionality for plotting results, but provide examples
+of how to build custom visualizations from output, specifically in R.
+For the Shiny application, we provide a base output plot, but the
+ability to download your results for custom plotting in programs of the
+user's choice. Finally, we provide brief summaries of general
+applications of \textbf{hatchR} below, but encourage users to visit
+articles hosted on the software webpage that extensively outline primary
+functions and applications, especially automating the application of
+predicting phenology across multiple variables.
 
-\includegraphics{flowchart.png}
+\begin{figure}
+
+{\centering \includegraphics[width=5.68in,height=0.5\textheight]{flowchart}
+
+}
+
+\caption{Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles.}\label{fig:unnamed-chunk-3}
+\end{figure}
 
@@
 developmental models in wild environments for Sockeye Salmon (\emph{O.
-nerka}). The need for change arose because historic models, specifically
-those in Beacham \& Murray (1990), only considered the average
-incubation temperature during development and, for wild fishes, average
-incubation temperature was impossible to estimate because it was unknown
-when fish hatched even if adult spawn timing was known. To address this,
-Sparks et al. (2019) used the reciprocal of the formulation of model 2
-from Beacham \& Murray (1990) and assigned an effective value for every
-day of development using the daily average temperature.
+nerka}). The need for their development arose because historic models,
+specifically those in Beacham \& Murray (1990), only considered the
+average incubation temperature during development and, for wild fishes,
+average incubation temperature was impossible to estimate because it was
+unknown when fish hatched even if adult spawn timing was known. To
+address this, Sparks et al. (2019) used the reciprocal of the
+formulation of model 2 from Beacham \& Murray (1990) and assigned an
+effective value for every day of development using the daily average
+temperature.
 
@@
 Where \emph{i} is the daily value and a fish hatches or emerges when the
-cumulative sum or \[\sum_{i =1}^nEffectiveValue_i = 1\]
+cumulative sum of effective values reaches one:
+\[\sum_{i =1}^nEffectiveValue_i = 1\]
 
 The effective value model framework is the basis for the phenological
-models in \texttt{hatchR}, both in the included \texttt{model\_table} in
-the package, as well as for custom models users can fit with
-\texttt{fit\_model()}. Specifically, \texttt{model\_table} has been
-extended to include more parameterizations from Beacham \& Murray
+models in \textbf{hatchR}, both in the included \texttt{model\_table} in
+the package (though \texttt{model\_table} includes more complex models
+developed by Beacham \& Murray (1990)), as well as for custom models
+users can fit with \texttt{fit\_model()}. Specifically,
+\texttt{model\_table} includes parameterizations from Beacham \& Murray
 (1990), Sparks et al. (2017), and Austin et al. (2019) (who extended
@@
 
+Water temperature datasets collected for wild environments are often
+either 1.) already summarized by day (\emph{i.e.}, mean daily
+temperature) or, 2.) in a raw format from something like a HOBO TidbiT
+where readings are taken multiple times per day, which can be summarized
+into a mean daily temperatures. Alternatively, new statistical models
+like that of Siegel, Fullerton, FitzGerald, Holzer, \& Jordan (2023)
+could be similarly implemented.
+
+Fundamentally, \textbf{hatchR} assumes you have input data with two
+columns: a date column, giving the date (and often time) of a
+temperature measurement, and a temperature column, giving the associated
+temperature measurement (in centigrade). Other columns are okay to
+include, but these two columns (with any column name---just
+\emph{without} spaces) are required. We expect your data to look
+something like this:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule\noalign{}
+date & temperature \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+2000-01-01 & 2.51 \\
+\ldots{} & \ldots{} \\
+2000-07-01 & 16.32 \\
+\ldots{} & \ldots{} \\
+2000-12-31 & 3.13 \\
+\end{longtable}
+
+\textbf{hatchR} assumes you've checked for missing records or errors in
+your data as it \emph{will function with gaps}, so it's important to go
+through the data checks discussed below, as well as your own validity
+checks. \textbf{hatchR} can use values down to freezing (e.g, 0 °C),
+which returns extremely small effective values, and time to hatch or
+emerge may be \textgreater{} 1 year. In these cases, we suggest users
+consider how much of that data type is reasonable with their data.
+
+For users choosing to implement \textbf{hatchR} in R, data can be
+imported from any format the user chooses, as long as users can
+eventually coerce their data into a \texttt{dataframe} or
+\texttt{tibble} format, in which each row represents a single record.
+For the Shiny application, users must have their data stored as a .csv
+(comma separated values) file for upload, which can easily be exported
+using datasheet software like Microsoft Excel or Google Sheets.
+
 \subsection{Checking Data}\label{checking-data}
 
-\texttt{hatchR} is built assuming data will be analyzed as daily average
+\textbf{hatchR} is built assuming data will be analyzed as daily average
 temperatures. Despite that assumption, raw data (\emph{e.g.}, as
-outputted by HOBO loggers) can be used and \texttt{hatchR} includes
-functionality to summarize those data into a format that is usable, as
-well as provides functions for basic visual and programmatic data checks
-to make sure outliers or missing data are at least brought to users'
-attention.
+outputted by HOBO loggers) can be used and \textbf{hatchR} includes
+functionality to summarize those data into a format that is usable (only
+in R, it must be summarized for the Shiny app), as well as provides
+functions for basic visual and programmatic data checks to make sure
+outliers or missing data are at least brought to user's attention.
 
-summarize\_temp, check\_continuous and plot\_check\_temp
+We demonstrate the utility of the summary and check functions
+\texttt{summarize\_temp()}, \texttt{plot\_check\_temp()}, and
+\texttt{check\_continuous()} using a simulated year-long dataset with
+temperature readings every thirty minutes.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# create date object for a year with 30 min reading intervals}
+\NormalTok{dates }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\AttributeTok{from =} \FunctionTok{ymd\_hms}\NormalTok{(}\StringTok{"2000{-}07{-}01 00:00:00"}\NormalTok{),}
+             \AttributeTok{to =} \FunctionTok{ymd\_hms}\NormalTok{(}\StringTok{"2001{-}06{-}30 23:59:59"}\NormalTok{), }\AttributeTok{length.out =} \DecValTok{17568}\NormalTok{) }
+
+\CommentTok{\# create empty dataframe}
+\NormalTok{year\_sim }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{nrow =} \FunctionTok{length}\NormalTok{(dates), }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{)) }
+
+\CommentTok{\# date column}
+\FunctionTok{colnames}\NormalTok{(year\_sim)[}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}} \StringTok{"date"}
+
+\CommentTok{\# add dates vector to date column}
+\NormalTok{year\_sim[}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ dates }
+
+\CommentTok{\#random seed}
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+
+\CommentTok{\# take temps from a random normal dist with mean 10 sd 3 }
+\CommentTok{\# for every date time combo in dates and append to column (temp) }
+\CommentTok{\# in year\_sim}
+\NormalTok{year\_sim}\SpecialCharTok{$}\NormalTok{temp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =} \FunctionTok{length}\NormalTok{(dates), }\AttributeTok{mean =} \DecValTok{10}\NormalTok{, }\AttributeTok{sd =} \DecValTok{3}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{abs}\NormalTok{() }
+
+\FunctionTok{dim}\NormalTok{(year\_sim)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 17568     2
+\end{verbatim}
+
+First, we recommend checking imported data for any outliers or strange
+inputs using \texttt{plot\_check\_temp()}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{plot\_check\_temp}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim,}
+                \AttributeTok{dates =}\NormalTok{ date,}
+                \AttributeTok{temperature =}\NormalTok{ temp,}
+                \AttributeTok{temp\_min =} \DecValTok{0}\NormalTok{, }\CommentTok{\# temp\_min and max lines are }
+                              \CommentTok{\# user customizable}
+                \AttributeTok{temp\_max =} \DecValTok{25}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1.pdf}
+\caption{Full output of the plot\_check\_temp() visual data check
+function from hatchR. Note that the red and blue lines are customizable
+by the user to create outlier thresholds for their visual check.}
+\end{figure}
+
+There are no obvious outliers but since each day has 48 records, we need
+to summarize it to daily mean temperature with
+\texttt{summarize\_temp()} and then check for missing days with
+\texttt{check\_continuous()}. We also recommend using
+\texttt{plot\_check\_temp()} again on the summarized data (though leave
+out the resulting plot for space efficiency in this manuscript).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# summarize}
+\NormalTok{year\_sim\_summ }\OtherTok{\textless{}{-}} \FunctionTok{summarize\_temp}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim,}
+                                \AttributeTok{dates =}\NormalTok{ date,}
+                                \AttributeTok{temperature =}\NormalTok{ temp)}
+
+\CommentTok{\# now a year\textquotesingle{}s worth of single{-}day data}
+\FunctionTok{dim}\NormalTok{(year\_sim\_summ)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 365   2
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# check continuous (no errors)}
+\FunctionTok{check\_continuous}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim\_summ,}
+                 \AttributeTok{dates =}\NormalTok{ date)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## i No breaks were found. All clear!
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# we can demonstrate an error by removing Oct. 8 (100th day)}
+\FunctionTok{check\_continuous}\NormalTok{(}\AttributeTok{data =}\NormalTok{ year\_sim\_summ[}\SpecialCharTok{{-}}\DecValTok{100}\NormalTok{,],}
+                 \AttributeTok{dates =}\NormalTok{ date)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: ! Data not continuous
+## i Breaks found at rows:
+## i 100
+\end{verbatim}
 
@@
 
-talk about using nls for fit\_model
+Users can either use Salmonid models from \texttt{model\_table} included
+in the package and the Shiny app. As discussed, these models are
+included because parameterizations already existed in the literature.
+However, these parameterizations are limited to Pacific Salmon and Bull
+Trout. To widen the application of the effective value approach, we
+include a \texttt{fit\_model()} function, which is species agnostic (as
+long as development generally follows the power law).
 
-model\_table and fit\_model
+The function \texttt{fit\_model()} uses data in which average incubation
+temperature (°C) and days to phenological event (as two vectors) are the
+inputs. The function estimates parameter coefficients for
+\emph{log\textsubscript{e}a} and \emph{b} using \texttt{stats::nls()}
 
-fit\_model for three non-salmonid species
+\textbf{BRYAN TO FINISH THIS PARAGRAPH OFF HERE}
+
+\paragraph{Fitting models for other
+fishes}\label{fitting-models-for-other-fishes}
+
+Below, we demonstrate how the \texttt{fit\_model()} function may be used
+to create custom parameterizations for species beyond the Salmonids in
+the \texttt{model\_table} included in the package. We include
+parameterizations from three warm-water species to demonstrate the
+\texttt{fit\_model()} utility for fishes beyond the scope of the
+original effective value approach. These parameterizations are for
+commonly cultured sportfishes including Smallmouth Bass
+(\emph{Micropterus dolomieu}) from Webster (1948), Channel Catfish
+(\emph{Ictalurus punctatus}) from Small \& Bates (2001), and Lake
+Sturgeon (\emph{Acipenser fulvescens}) from Smith \& King (2005). We
+provide parameterization for just Smallmouth Bass here for concision,
+but the full code set for all species is available at
+\url{https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd}.
+
+We demonstrate the utility of this approach by creating a random thermal
+regime with an ascending thermograph with a mean temperature of 16 °C,
+parameterizing models for each species, and demonstrating days to hatch
+and developmental period for each species with the random thermal regime
+(Figure 3).
 
@@
 \begin{Highlighting}[]
-\NormalTok{smallmouth }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{tibble}\NormalTok{()}
+\DocumentationTok{\#\#\#  make temp regime}
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\CommentTok{\# create random temps and corresponding dates}
+\NormalTok{temps\_sim }\OtherTok{\textless{}{-}} \FunctionTok{sort}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\DecValTok{30}\NormalTok{, }\AttributeTok{mean =} \DecValTok{16}\NormalTok{, }\AttributeTok{sd =} \DecValTok{1}\NormalTok{), }\AttributeTok{decreasing =} \ConstantTok{FALSE}\NormalTok{)}
+\NormalTok{dates\_sim }\OtherTok{\textless{}{-}}  \FunctionTok{seq}\NormalTok{(}\AttributeTok{from =} \FunctionTok{ymd}\NormalTok{(}\StringTok{"2000{-}07{-}01"}\NormalTok{),}
+             \AttributeTok{to =} \FunctionTok{ymd}\NormalTok{(}\StringTok{"2000{-}07{-}31"}\NormalTok{), }\AttributeTok{length.out =} \DecValTok{30}\NormalTok{)}
+
+\NormalTok{data\_sim }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{data.frame}\NormalTok{()}
+\NormalTok{data\_sim[,}\DecValTok{1}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ temps\_sim}
+\NormalTok{data\_sim[,}\DecValTok{2}\NormalTok{] }\OtherTok{\textless{}{-}}\NormalTok{ dates\_sim}
+
+\CommentTok{\# change names so they aren\textquotesingle{}t the same as the vector objects}
+\FunctionTok{colnames}\NormalTok{(data\_sim) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"temp\_sim"}\NormalTok{, }\StringTok{"date\_sim"}\NormalTok{)}
+
+\DocumentationTok{\#\#\# smallmouth mod}
+\NormalTok{smallmouth }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{2}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{data.frame}\NormalTok{()}
 \FunctionTok{colnames}\NormalTok{(smallmouth) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"hours"}\NormalTok{, }\StringTok{"temp\_F"}\NormalTok{)}
@@
 
+\CommentTok{\# change F to C and hours to days}
 \NormalTok{smallmouth }\OtherTok{\textless{}{-}}\NormalTok{ smallmouth }\SpecialCharTok{|\textgreater{}} 
@@
 
-\FunctionTok{plot}\NormalTok{(smallmouth}\SpecialCharTok{$}\NormalTok{days}\SpecialCharTok{\textasciitilde{}}\NormalTok{smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C)}
-
-\NormalTok{smb\_mod }\OtherTok{\textless{}{-}} \FunctionTok{fit\_model}\NormalTok{(smallmouth,}
-                     \AttributeTok{temp =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C,}
-                     \AttributeTok{days =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{hours)}
+\NormalTok{smb\_mod }\OtherTok{\textless{}{-}} \FunctionTok{fit\_model}\NormalTok{(}\AttributeTok{temp =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{temp\_C,}
+                     \AttributeTok{days =}\NormalTok{ smallmouth}\SpecialCharTok{$}\NormalTok{days,}
+                     \AttributeTok{species =} \StringTok{"smb"}\NormalTok{,}
+                     \AttributeTok{development\_type =} \StringTok{"hatch"}\NormalTok{)}
 \end{Highlighting}
 \end{Shaded}
+
+Note the \emph{R\textsuperscript{2}} fit from the models below. You can
+see they generally all preform well and are in line with values from
+model 2 of Beacham \& Murray (1990).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#model fits}
+\NormalTok{smb\_mod}\SpecialCharTok{$}\NormalTok{r\_squared; cat\_mod}\SpecialCharTok{$}\NormalTok{r\_squared; sturgeon\_mod}\SpecialCharTok{$}\NormalTok{r\_squared}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 0.9868067
+\end{verbatim}
+
+\begin{verbatim}
+## [1] 0.9433598
+\end{verbatim}
+
+\begin{verbatim}
+## [1] 0.9217358
+\end{verbatim}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/unnamed-chunk-10-1.pdf}
+\caption{Days to hatch as predicted using custom parameterizations for
+Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass
+(green) as generated by the fit\_model() function in hathcR. Models are
+all run using a spawn date starting June 1 for a randomly generated
+ascending thermal regime with mean temperature 16 °C}
+\end{figure}
 
@@
 
-Use woody \_example from website
+To illustrate model selection and phenology prediction we will recreate
+a small portion of the analysis done by Sparks et al. (2019) using the
+\texttt{woody\_island} dataset included in this package . We will
+predict both hatch and emergence timing, so we will obtain a model
+expression for each using \texttt{model\_select()}.
 
-show predict\_phenology output slots
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{sockeye\_hatch\_mod }\OtherTok{\textless{}{-}} \FunctionTok{model\_select}\NormalTok{(}
+  \AttributeTok{author =} \StringTok{"Beacham and Murray 1990"}\NormalTok{, }
+  \AttributeTok{species =} \StringTok{"sockeye"}\NormalTok{, }
+  \AttributeTok{model =} \DecValTok{2}\NormalTok{, }
+  \AttributeTok{development\_type =} \StringTok{"hatch"}
+\NormalTok{  )}
+\end{Highlighting}
+\end{Shaded}
 
-show plot\_phenology
+We can now use our model expressions to predict when sockeye would hatch
+and emerge at Woody Island in 1990. First we predict hatch timing using
+\texttt{predict\_phenology()}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{WI\_hatch }\OtherTok{\textless{}{-}} \FunctionTok{predict\_phenology}\NormalTok{(}
+  \AttributeTok{data =}\NormalTok{ woody\_island,}
+  \AttributeTok{dates =}\NormalTok{ date,}
+  \AttributeTok{temperature =}\NormalTok{ temp\_c,}
+  \AttributeTok{spawn.date =} \StringTok{"1990{-}08{-}18"}\NormalTok{,}
+  \AttributeTok{model =}\NormalTok{ sockeye\_hatch\_mod}
+\NormalTok{  )}
+\end{Highlighting}
+\end{Shaded}
+
+And then look inside the returned object to see days to hatch and
+development period:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{WI\_hatch}\SpecialCharTok{$}\NormalTok{days2done}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## NULL
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{WI\_hatch}\SpecialCharTok{$}\NormalTok{dev.period}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##        start       stop
+## 1 1990-08-18 1990-10-30
+\end{verbatim}
+
+\paragraph{Understanding your results}\label{understanding-your-results}
+
+The output from \texttt{predict\_phenology()} includes a lot of
+information. If we look at our \texttt{WI\_hatch} object we see there
+are multiple elements stored in a list which can be accessed using the
+\texttt{\$} operator.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{str}\NormalTok{(WI\_hatch)}
+\end{Highlighting}
+\end{Shaded}
+
+\texttt{WI\_hatch\$days2done} outputs the predicted days to develop.
+
+\texttt{WI\_hatch\$dev.period} is a 1x2 dataframe with the dates
+corresponding to when your fish's parent spawned (which you input with
+\texttt{predict\_phenology(spawn.date\ =\ ...)}) and the date when the
+fish is predicted todevelop.
+
+\texttt{WI\_hatch\$ef.vals} is a vector of each day's effective value as
+evaluated using whatever model is chosen.
+
+\texttt{WI\_hatch\$ef.tibble} is a \emph{n} x 4 tibble (\emph{n} =
+number of days to hatch or emerge) and the columns are the date, each
+day's temperature and effective value, and the cumulative sum of the
+effective values. The \texttt{ef.tibble} object is meant to serve as the
+basis for users to make custom figures for their data beyond the
+functionality we discuss below.
+
+\paragraph{Plotting phenology}\label{plotting-phenology}
+
+\textbf{hatchR} has a built in function, \texttt{plot\_phenology()},
+that allows users to visualize their phenology results (Figure 4). The
+plot visualizes three specific components: 1.) the temperature regime
+over which you are predicting, 2.) the cumulative sum of effective
+values, and 3.) the effective value for each day in your prediction
+span. The function allows you to output various figures based on your
+interests, but defaults to a figure with all information and the
+corresponding labels.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{plot\_phenology}\NormalTok{(WI\_hatch)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/unnamed-chunk-15-1.pdf}
+\caption{Output of the plot\_phenology() function from hatchR. The
+function plots the output of the predict\_phenology() function and
+displays time to develop, as well as the ambient thermal regime (blue),
+scaled daily effective value (green), and the scaled cumulative sum of
+effective values (orange).}
+\end{figure}
 
@@
 useful would be trying to understand if fish might be free-moving before
-some management action. For instance, will have fish have emerged from
-redds when a stream section has been opened to grazing or bridge
-decommissioning will commence?
+some management action. For instance, will fish have emerged from redds
+when a stream section has been opened to grazing or road work?
 
-In this scenario, we will consider the grazing example and Bull Trout, a
-threatened fish in the united states under the Endangered Species Act
-(Nolfi, Melbihess, Fisher, \& Ellis, 2024), and the East Fork Salmon
-River, a key Bull Trout population in the upper Salmon River watershed.
-The fisheries manager there wants to know if fish will likely be out of
-the gravel and free-swimming by June 1st. In this system, it is expected
+In this scenario, we will consider the road work example and Bull Trout,
+a threatened fish in the United States under the Endangered Species Act
+(Nolfi, Melbihess, Fisher, \& Ellis, 2024), and the Crooked River, a key
+Bull Trout population in the Boise River watershed. In this hypothetical
+scenario, the Forest Fisheries Biologist wants to know if fish will
+likely be out of the gravel and free-swimming by June 1st as Bull Trout
+are particularly sensitive to sediment. In this system, it is expected
 that Bull Trout will be done spawning by the end of September, so we'll
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\#\# read in EFS data}
-
-\NormalTok{EFS\_data }\OtherTok{\textless{}{-}}\NormalTok{ crooked\_river}
-
-\DocumentationTok{\#\#\# view bull trout models and select model}
-
-\NormalTok{model\_table }\SpecialCharTok{|\textgreater{}} 
-  \FunctionTok{filter}\NormalTok{(species }\SpecialCharTok{==} \StringTok{"bull trout"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## # A tibble: 2 x 5
-##   author             species    model dev.type func                       
-##   <chr>              <chr>      <chr> <chr>    <chr>                      
-## 1 Austin et al. 2017 bull trout MM    hatch    1/exp(5.086 - (x * 0.131)) 
-## 2 Austin et al. 2017 bull trout MM    emerge   1/exp(5.590 - (x  * 0.126))
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# we care about when fish are out of gravel, so select emerge mod}
-\NormalTok{bt\_emerge\_mod }\OtherTok{\textless{}{-}} \FunctionTok{model\_select}\NormalTok{(}\AttributeTok{author =} \StringTok{"Austin et al. 2017"}\NormalTok{,}
-                            \AttributeTok{species =} \StringTok{"bull trout"}\NormalTok{,}
-                            \AttributeTok{model =} \StringTok{"MM"}\NormalTok{,}
-                            \AttributeTok{dev.type =} \StringTok{"emerge"}\NormalTok{)}
-
-\CommentTok{\# predict emergence time using Sept. 30 as the spawn date }
-\NormalTok{bt\_emergence }\OtherTok{\textless{}{-}} \FunctionTok{predict\_phenology}\NormalTok{(}\AttributeTok{data =}\NormalTok{ EFS\_data,}
-                                  \AttributeTok{dates =}\NormalTok{ date,}
-                                  \AttributeTok{temperature =}\NormalTok{ temp\_c,}
-                                  \AttributeTok{spawn.date =} \StringTok{"2011{-}09{-}30"}\NormalTok{, }
-                                  \AttributeTok{model =}\NormalTok{ bt\_emerge\_mod)}
-
-\CommentTok{\# fish emerge May 1, before June 1}
-\NormalTok{bt\_emergence}\SpecialCharTok{$}\NormalTok{dev.period}\SpecialCharTok{$}\NormalTok{stop}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] "2012-05-01 UTC"
-\end{verbatim}
+We demonstrate this first case study using the graphical user interface
+portion of the \textbf{hatchR} ecosystem found at
+\url{https://elifelts.shinyapps.io/hatchR/}. Users will first upload
+their data with the \texttt{Import\ Data} window, which requires them to
+select their file on their personal computer, provide the program with
+the columns corresponding for dates and temperatures, and then provide
+the format in which dates are coded (\emph{e.g.}, year-month-day or
+day-month-year). Data used in this case study are from the
+\texttt{crooked\_river} data set from R and can either be written out
+from the package or accessed at
+\url{https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/}.
+Once data is uploaded the program automatically plots the user's data
+using \texttt{plot\_check\_temp()} in the background and provides them
+the outputted graphical check. After uploading and checking data, the
+user switches to the \texttt{Model\ Phenology} window. In this
+circumstance, we use the preloaded parameterization for bull trout from
+Austin et al. (2019) with the \texttt{Existing} button for model
+selection, which the user selects with the various drop down options in
+the menu. After the model is selected, the user can choose multiple
+spawn dates from the interactive calendar provided. We show results for
+spawning for September 30th as indicated in the example above. Once
+dates are chosen, a table entry for each spawn date is outputted in the
+\texttt{Phenology\ Summaries} tab and corresponding plot with data from
+each spawn date in the \texttt{Timeline\ Plot} tab. Output from
+predicting phenology and the resulting figure are downloadable from
+their respective tabs. The process is demonstrated in full in Figure
+XXX, but the interface is described more completely on \textbf{hatchR}'s
+Shiny website.
 
@@
 
+For the second example, we will again use bull trout, but demonstrate a
+much more complex application for the purpose of showing the full
+flexibility of the programmatic application of \textbf{hatchR}. In this
+scenario we will use data from Isaak, Luce, Chandler, Horan, \& Wollrab
+(2018) (included \texttt{idaho} dataset), which includes temperature
+data from 226 sites across the major upper Columbia River headwater
+watersheds in Idaho. For this approach we winnow putative bull trout
+spawning sites by filtering for sites with mean August temperature
+\textless/= 13 °C in accordance with thresholds from Isaak, Young,
+Nagel, Horan, \& Groce (2015). For the resulting 139 sites we will
+demonstrate predicting hatch timing in these putative Bull Trout
+spawning habitats.
+
+We need to setup our models and data for this analysis, which we don't
+show those steps here for the sake of concision in this manuscript,
+however they are demonstrated in \texttt{paper.Rmd} included in the
+GitHub repository for \textbf{hatchR}. After the setup, we can easily
+map \texttt{predict\_phenology()} across all putative spawning sites and
+three spawn dates (September 1-Early Spawning, September 15-Peak
+Spawning, and September 31-Late Spawning), the results of which are
+presented in Figure XXX.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{hatch\_res }\OtherTok{\textless{}{-}}\NormalTok{ isaak\_summ\_bt }\SpecialCharTok{|\textgreater{}} 
+  \FunctionTok{mutate}\NormalTok{(}
+    \AttributeTok{dev\_period =} \FunctionTok{map2}\NormalTok{(}
+\NormalTok{      summ\_obj, spawn\_dates, }\CommentTok{\# map across our site object and spawn dates}
+\NormalTok{      predict\_phenology,}
+      \AttributeTok{temperature =}\NormalTok{ daily\_temp,}
+      \AttributeTok{model =}\NormalTok{ bt\_hatch,}
+      \AttributeTok{dates =}\NormalTok{ date}
+\NormalTok{      ) }\SpecialCharTok{|\textgreater{}} 
+      \FunctionTok{map\_df}\NormalTok{(}\StringTok{"dev.period"}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \CommentTok{\# pull out just dev.period results}
+      \FunctionTok{list}\NormalTok{()}
+\NormalTok{    ) }\SpecialCharTok{|\textgreater{}} 
+  \FunctionTok{select}\NormalTok{(site, dev\_period) }\SpecialCharTok{|\textgreater{}} \CommentTok{\# just select the columns we want}
+  \FunctionTok{unnest}\NormalTok{(}\AttributeTok{cols =} \FunctionTok{c}\NormalTok{(dev\_period)) }\SpecialCharTok{|\textgreater{}} \CommentTok{\# unnest everything}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{days\_to\_hatch =}\NormalTok{ stop }\SpecialCharTok{{-}}\NormalTok{ start) }\CommentTok{\# make a new column of days to hatch}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/unnamed-chunk-19-1.pdf}
+\caption{Densities of predicted days to hatch for Bull Trout for 139
+putative spawning locations across Idaho for four years of shared
+temperature data. Early spawning considers if a parent spawned September
+1, Peak-September 15, and Late-September 30.}
+\end{figure}
+
 \section{Discussion}\label{discussion}
 
-talk about how these models also represent local adaptation and
-heritable plasticity
+With \textbf{hatchR} we present a software ecosystem that bridges the
+analytical gap of predicting developmental phenology for wild fishes and
+develops a formal framework for applying effective value models from
+user-provided parameterizations. To do so, the software is bundled in
+two forms, a fully customizable R package, especially useful for running
+repetitive analyses and a graphical-user-interface designed for ease of
+use and tasks that may only run once or a handful of times. Both
+applications allows users to import their data, run basic data checks,
+and apply historic model parameterizations for salmonids or create their
+own species- or population specific parameterization. Additionally, we
+provide substantial documentation on the \textbf{hatchR} website
+(\url{https://bmait101.github.io/hatchR/}) to walk users through basic
+to advanced applications of programmatic platform.
 
-talk about how these models represent point estimates and that emergence
-and hatch will take the form of a distribution
+In the application of using effective value models \textbf{hatchR} and
+the user make some key assumptions. Particularly, numerous studies have
+indicated that stressful environmental conditions can cue fish to
+prematurely hatch or emerge from their environment. These include water
+quality like dissolved oxygen, pH, or salinity, pathogens, and even
+mechanical agitation (reviewed in Quinn (2018) and Cowan et al. (2024)).
+Moreover, while the \textbf{hatchR} provides point estimates for
+developmental phenology, spawning and developmental within populations
+of fishes both occur as distributions as opposed single events Mason
+(1976) and we encourage users to either predict phenology using early,
+peak, and late frameworks (\emph{e.g.}, 5th, 50th, and 95th quantiles)
+or with real or modeled distributions. Another factor that could bias
+estimates from these models is that temperatures used from sensors don't
+reflect the geomoprhic properties in the environment where eggs are
+developing such that they may be too cold or warm Geist et al. (2002).
+
+To date, the application of the effective value framework has largely
+focused using species-specific models to predict phenology in wild
+environments (see Adelfio et al. (2024), Austin et al. (2019)). However,
+we emphasize that the statistical relationships at the basis of these
+models are fundamentally reaction norms, meaning that both the intercept
+and slope of the responses of fishes to temperature are reflective of
+family- or population-level genetic variation, genetic x environment
+responses, or may be indicative of phylogenetic differences among
+species (West-Eberhard (2003)). For instance, while Sparks et al. (2017)
+did not find differences in the rates of development between the focal
+populations of their study, they did observe family-level genetic x
+environment interactions across different thermal regimes. Similarly,
+when they reparameterized their models using fish from western Alaska,
+they found differences in both the slope and intercept of the model
+relative to the original parameterization from Beacham \& Murray (1990),
+notably derived from multiple Canadian stocks. The western Alaskan fish
+generally developed more slowly than their southerly counterparts,
+consistent with cogradient variation (Conover, Duffy, \& Hice (2009),
+Sparks, Kraft, Blackstone, McNickle, \& Christie (2022)). In this sense,
+we encourage users not to only think about the end points of these
+models (days to development) but also how the statistical relationships
+they are premised on inform our understanding of micro- and and
+macro-evolutionary processes.
+
+\textbf{BRYAN check the below}
+
+The models presented in \textbf{hatchR} can be further customized in
+multiple ways beyond the use-cases provided above. For instance, while
+our models are fit to predict hatch or emergence timing, they could be
+used to predict other developmental stages prior to the initation of
+external feeding like eye-up or other embryological developmental stages
+(Velsen (1980)) or \textbf{Some example from a non-salmonid, maybe
+initiation of pelagic-larval or riverine current dispersal}.
+Additionally, while \texttt{fit\_model()} relies on non-linear modeling
+platform, \texttt{predict\_phenology()} only requires users to pass a
+model expression, so if users preferred a different model fit than the
+non-linear option provided (as long as it integrates daily temperature),
+they could pass a different expression into
+\texttt{predict\_phenology()} for additional customization. Finally,
+while \textbf{hatchR} was designed specifically for fishes, we expect
+other poikilothermic organisms such as reptiles, amphibians, or
+invertebrates would develop in accordance with the power law and these
+developmental models could be extended to other organisms beyond fishes.
 
@@
 
-We would like to sincerely thank the editor and reviewers for all of
-their helpful feedback which greatly improved both the software and the
-manuscript.
+We thank Laura Koller for her help designing the \textbf{hatchR} logo.
+Dan Isaak provided useful discussion about model development and
+temperature datasets.
 
@@
 
+\bibitem[\citeproctext]{ref-alderdice1978}
+Alderdice, D. F., \& Velsen, F. P. J. (1978). Relation between
+temperature and incubation time for eggs of chinook salmon (oncorhynchus
+tshawytscha). \emph{Journal of the Fisheries Research Board of Canada},
+\emph{35}(1), 69--75.
+doi:\href{https://doi.org/10.1139/f78-010}{10.1139/f78-010}
+
 \bibitem[\citeproctext]{ref-austin2019}
@@
 
+\bibitem[\citeproctext]{ref-chang2024}
+Chang, W., Cheng, J., Allaire, J. J., Sievert, C., Schloerke, B., Xie,
+Y., Allen, J., et al. (2024). \emph{Shiny: Web application framework for
+r}. Retrieved from
+\url{https://cran.r-project.org/web/packages/shiny/index.html}
+
+\bibitem[\citeproctext]{ref-conover2009}
+Conover, D. O., Duffy, T. A., \& Hice, L. A. (2009). The Covariance
+between Genetic and Environmental Influences across Ecological
+Gradients. \emph{Annals of the New York Academy of Sciences},
+\emph{1168}(1), 100--129.
+doi:\href{https://doi.org/10.1111/j.1749-6632.2009.04575.x}{10.1111/j.1749-6632.2009.04575.x}
+
+\bibitem[\citeproctext]{ref-cowan2024}
+Cowan, Z.-L., Green, L., Clark, T. D., Blewett, T. A., De Bonville, J.,
+Gagnon, T., Hoots, E., et al. (2024). Global change and premature
+hatching of aquatic embryos. \emph{Global Change Biology}, \emph{30}(9),
+e17488. doi:\href{https://doi.org/10.1111/gcb.17488}{10.1111/gcb.17488}
+
+\bibitem[\citeproctext]{ref-geist2002}
+Geist, D. R., Hanrahan, T. P., Arntzen, E. V., McMichael, G. A., Murray,
+C. J., \& Chien, Y.-J. (2002). Physicochemical Characteristics of the
+Hyporheic Zone Affect Redd Site Selection by Chum Salmon and Fall
+Chinook Salmon in the Columbia River. \emph{North American Journal of
+Fisheries Management}, \emph{22}(4), 1077--1085.
+doi:\href{https://doi.org/10.1577/1548-8675(2002)022\%3C1077:PCOTHZ\%3E2.0.CO;2}{10.1577/1548-8675(2002)022\textless1077:PCOTHZ\textgreater2.0.CO;2}
+
+\bibitem[\citeproctext]{ref-isaak2018}
+Isaak, D. J., Luce, C. H., Chandler, G. L., Horan, D. L., \& Wollrab, S.
+P. (2018). Principal components of thermal regimes in mountain river
+networks. \emph{Hydrology and Earth System Sciences}, \emph{22}(12),
+6225--6240.
+doi:\href{https://doi.org/10.5194/hess-22-6225-2018}{10.5194/hess-22-6225-2018}
+
+\bibitem[\citeproctext]{ref-isaak2015}
+Isaak, D. J., Young, M. K., Nagel, D. E., Horan, D. L., \& Groce, M. C.
+(2015). The cold-water climate shield: delineating refugia for
+preserving salmonid fishes through the 21st century. \emph{Global Change
+Biology}, \emph{21}(7), 2540--2553.
+doi:\href{https://doi.org/10.1111/gcb.12879}{10.1111/gcb.12879}
+
 \bibitem[\citeproctext]{ref-kaylor2021}
@@
 doi:\href{https://doi.org/10.1111/1365-2656.13491}{10.1111/1365-2656.13491}
+
+\bibitem[\citeproctext]{ref-mason1976}
+Mason, J. (1976). Some features of coho salmon, oncorhynchus kisutch,
+fry emerging from simulated redds and concurrent changes in
+photobehavior. \emph{Fish. Bull}, \emph{74}(1), 167--175.
 
@@
 
+\bibitem[\citeproctext]{ref-quinn2018}
+Quinn, T. P. (2018). \emph{The Behavior and Ecology of Pacific Salmon
+and Trout}. University of Washington Press.
+
+\bibitem[\citeproctext]{ref-r:ther}
+R: The r project for statistical computing. (n.d.). Retrieved from
+\url{https://www.r-project.org/}
+
+\bibitem[\citeproctext]{ref-siegel2023}
+Siegel, J. E., Fullerton, A. H., FitzGerald, A. M., Holzer, D., \&
+Jordan, C. E. (2023). Daily stream temperature predictions for
+free-flowing streams in the Pacific Northwest, USA. \emph{PLOS Water},
+\emph{2}(8), e0000119.
+doi:\href{https://doi.org/10.1371/journal.pwat.0000119}{10.1371/journal.pwat.0000119}
+
+\bibitem[\citeproctext]{ref-small2001}
+Small, B. C., \& Bates, T. D. (2001). Effect of Low-Temperature
+Incubation of Channel Catfish Ictalurus punctatus Eggs on Development,
+Survival, and Growth. \emph{Journal of the World Aquaculture Society},
+\emph{32}(2), 189--194.
+doi:\href{https://doi.org/10.1111/j.1749-7345.2001.tb01094.x}{10.1111/j.1749-7345.2001.tb01094.x}
+
+\bibitem[\citeproctext]{ref-smith2005}
+Smith, K. M., \& King, D. K. (2005). Dynamics and extent of larval lake
+sturgeon Acipenser fulvescens drift in the Upper Black River, Michigan.
+\emph{Journal of Applied Ichthyology}, \emph{21}(3), 161--168.
+doi:\href{https://doi.org/10.1111/j.1439-0426.2005.00623.x}{10.1111/j.1439-0426.2005.00623.x}
+
 \bibitem[\citeproctext]{ref-sparks2019}
@@
 
+\bibitem[\citeproctext]{ref-sparks2022}
+Sparks, M. M., Kraft, J. C., Blackstone, K. M. S., McNickle, G. G., \&
+Christie, M. R. (2022). Large genetic divergence underpins cryptic local
+adaptation across ecological and evolutionary gradients.
+\emph{Proceedings of the Royal Society B: Biological Sciences},
+\emph{289}(1984), 20221472.
+doi:\href{https://doi.org/10.1098/rspb.2022.1472}{10.1098/rspb.2022.1472}
+
 \bibitem[\citeproctext]{ref-sparks2017}
@@
 
+\bibitem[\citeproctext]{ref-velsen1980}
+Velsen, F. P. J. (1980). Embryonic Development in Eggs of Sockeye Salmon
+Oncorhynchus nerka. \emph{Canadian Special Publication of Fisheries and
+Aquatic Sciences}, (49).
+
+\bibitem[\citeproctext]{ref-webster1948}
+Webster, D. A. (1948). Relation of temperature to survival and
+incubation of the eggs of smallmouth bass (micropterus dolomieu).
+\emph{Transactions of the American Fisheries Society}, \emph{75}(1),
+43--47.
+doi:\href{https://doi.org/10.1577/1548-8659(1945)75\%5B43:ROTTSA\%5D2.0.CO;2}{10.1577/1548-8659(1945)75{[}43:ROTTSA{]}2.0.CO;2}
+
+\bibitem[\citeproctext]{ref-west-eberhard2003}
+West-Eberhard, M. J. (2003). \emph{Developmental Plasticity and
+Evolution}. Oxford University Press.
+
 \end{CSLReferences}

### 2025-02-26 inst/poster/hatchR_posterdown/hatchR_posterdown.Rmd
add poster folder

---
title: "hatchR: A toolset to predict when fish hatch and emerge"
author:
  - name: Morgan Sparks
    affil: 1
  - name: Eli Felts
    affil: 2
  - name: Allison Swartz
    affil: 3
  - name: Bryan Maitland
    affil: 1
affiliation:
  - num: 1
    address: Rocky Mountain Research Station, US Forest Service
  - num: 2
    address: Idaho Fish and Wildlife Conservation Oﬀice, US Fish and Wildlife Service
  - num: 3
    address: College of Forestry, Oregon State University
column_numbers: 4
logoright_name: "./hatchR_posterdown_files/figure-html/agency_logos.png"
logoleft_name: "./hatchR_posterdown_files/figure-html/logo.png"
poster_height: "36in"
poster_width: "48in"
primary_colour: "	#011f4b"
sectitle_bgcol: "#005b96"
sectitle_bordercol: "#005b96"
column_margins: "0.33in"
output: 
  posterdown::posterdown_html:
    self_contained: false
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(hatchR); library(tidyverse); library(ggridges)
```

# hatchR

**hatchR** is a software ecosystem for predicting fish developmental phenology using statistical models. It offers three primary applications:

1.  Basic summarization, plotting, and data checks for water temperature data.
2.  Access published parameterizations for salmonid developmental models or build your own custom parameterizations for any species with user provided data.
3.  Predict hatch and/or emergence timing across ambient temperatures from common data sources such as HOBO loggers.

**hatchR** is deployed in two formats. First, it can be downloaded as a R package. Secondly, for users not familiar with R, it can also be accessed as a web app built in Shiny as a point-and-click tool.

To predict phenology, you need [**daily average temperatures over incubation (or data that can be summarized as such**]{.underline}**)** and [**spawn timing**]{.underline}.

# hatchR's workflow

```{r, out.width='36%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}

knitr::include_graphics("./hatchR_posterdown_files/figure-html/workflow.png")
```

# Included & custom models

**hatchR** includes numerous parameterizations for salmonids because the original models were built from aquaculture studies with salmon. Those models are included in `model_table` and include all North American Pacific Salmon (*Oncorhynchus spp.*) and Bull Trout (*Salvelinus confluentus*).

```{r mytable, out.width='80%'}

knitr::kable(model_table[c(1,9),c(1,2,5)], caption = 'Two example rows (of 51) from `model_table`.',align = 'c',"html")
```

Alternatively, you can build custom models with experimental data. Here we demonstrate with Channel Catfish (*Ictalurus punctatus*).

```{r, echo=TRUE}
# make catfish data object for function
catfish <- matrix(NA, 3, 2) |> data.frame() 
colnames(catfish) <- c("days", "temp_C") 
catfish$days <- c(16,21,26) # days to hatch
catfish$temp_C <- c(22,10,7) # exp. temperatures

# fit model
cat_mod <- fit_model(
  temp = catfish$temp_C, 
  days = catfish$days,
  species = "catfish",
  development_type = "hatch")
```

To further demonstrate this approach we parameterize models (not shown) for Smallmouth Bass (*Micropterus dolomieu*) and Lake Sturgeon (*Acipenser fulvescens*) and apply the models to a temperature regime with mean temperature 16 °C.

```{r, out.width='75%', fig.align='center', fig.cap= "Predicted days to hatch using custom parameterizations for three warmwater species with mean 16 °C thermal regime."}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

data_sim <- matrix(NA, 30, 2) |> data.frame()
data_sim[,1] <- temps_sim
data_sim[,2] <- dates_sim

# change names so they aren't the same as the vector objects
colnames(data_sim) <- c("temp_sim", "date_sim")

### smallmouth mod
smallmouth <- matrix(NA, 10, 2) |> data.frame()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

# change F to C and hours to days
smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

smb_mod <- fit_model(temp = smallmouth$temp_C,
                     days = smallmouth$days,
                     species = "smb",
                     development_type = "hatch")

### lake sturgeon mod
sturgeon <-  matrix(NA, 7, 2) |> data.frame()
colnames(sturgeon) <- c("days", "CTU")
sturgeon$days <- c(7,5,6,6,5,11,7)
sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)

sturgeon <- sturgeon |> 
  mutate(temp_C = CTU/days) # change CTUs to average temp and add column

sturgeon_mod <- fit_model(days = sturgeon$days,
                          temp = sturgeon$temp_C,
                          species = "sturgeon",
                          development_type = "hatch")
### predict_phenology

#smallmouth bass
smb_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = smb_mod$expression)

# catfish
catfish_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = cat_mod$expression)

# sturgeon
# note that 16 C is pretty far out of range of temps for model fit, not best practice
sturgeon_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = sturgeon_mod$expression)

# summary for all species
all_res <- data.frame(matrix(NA, 3, 2))
colnames(all_res) <- c("start", "stop")

all_res$start <- c(catfish_res$dev.period$start, 
                   smb_res$dev.period$start, 
                   sturgeon_res$dev.period$start)

all_res$stop <- c(catfish_res$dev.period$stop,
                  smb_res$dev.period$stop, 
                  sturgeon_res$dev.period$stop)

all_res <- all_res |> 
  mutate(days = ceiling(stop-start),
         index = c(17,16.5,16))

all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")

ggplot() +
  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim )) + 
  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim )) +
  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax =index-.35, ymin = index-.5, fill = Species)) +
  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index -0.425), label = days)) +
  labs(x = "Date", y = "Temperature (°C)", fill = "") +
  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
  theme_classic(base_size = 28) +
  theme(legend.position = c(0.75, 0.25), 
        legend.background = element_rect(fill=alpha('blue', 0)))
```

# Basic prediction

As a case study for a basic application we'll use our included data set of `crooked_river` and an example with Bull Trout.

Let's assume the forest manager wants to do some roadwork upstream of a Bull Trout spawning reach in Crooked River starting **July 1**. They engage the forest biologist who knows Bull Trout, which are sensitive to sediment, will be finished spawning there by **September 30**. They can then predict if fish are likely to have emerged from the gravel by July (using the 2014-2015 spawn year).

```{r, echo = TRUE}
# get model from model_table
bull_trout_emerge <- model_select(
  author = "Austin et al. 2017",
  species = "bull trout",
  model = "MM",
  development_type = "emerge")

# predict phenology assuming Sept. 30th as spawn
crooked_river_emerge <- predict_phenology(
  data = crooked_river, 
  dates = date, 
  temperature = temp_c, 
  spawn.date = "2014-09-30", 
  model = bull_trout_emerge)

#predicted days to emerge
crooked_river_emerge$days_to_develop

```

Our model predicts Bull Trout will emerge 204 days after September 30th, which is April 22 or well before the July 1 date of road construction.

# Advanced prediction

The power of **hatchR** is fully realized when we combine multiple variables for iterating the **hatchR** function, `predict_phenology()`. We'll use 139 putative Bull Trout populations spread across Idaho to demonstrate combining predictions across sites, years, and spawn dates. We will use the included `idaho` data set, with which we will do some filtering (not shown) and then predict hatch timing assuming fish spawn September 1, 15, and 30 (Early, Peak, Late) at each site in each of the four continuous years of temperature data (1668 parameter combinations).

We present an abridged demonstration automating across our parameter set:

```{r}

bull_trout_sites <- idaho |> 
  mutate(month = month(date)) |> #make a month column (numeric)
  filter(month == 8) |> # filter out Aug.
  group_by(site) |> # apply grouping by site
  summarise(mean_aug_temp = mean(temp_c)) |> 
  filter(mean_aug_temp <= 13) |> # keep only sites 13 C or cooler
  pull(site) |> 
  unique()

# filter to bull trout sites
idaho_bt <- idaho |> 
  filter(site %in% bull_trout_sites)

# nest data and summarize
isaak_summ_bt <- idaho_bt |> 
  group_by(site) |> 
  nest() |> 
  mutate(
    summ_obj = map(
      data, 
      summarize_temp,
      temperature = temp_c,
      dates = date
      )
    ) |> 
  select(site, summ_obj)

# spawn dates
spawn_dates <- map(
  c(2011:2014), # year vector to map for custom function
  function(year) { # custom paste function
    c(
      paste0(year, "-09-01"),
      paste0(year, "-09-15"),
      paste0(year, "-09-30")
      )
    }
  ) |> 
  unlist()

# bull trout hatch model
bt_hatch <- model_select(
  development_type = "hatch",
  author = "Austin et al. 2017",
  species = "bull trout",
  model = "MM"
)
```

```{r, echo=TRUE}
# predict hatch time across variable combinations
hatch_res <- isaak_summ_bt |> 
  mutate(dev_period = map2(
      summ_obj, spawn_dates, 
      predict_phenology,
      temperature = daily_temp,
      model = bt_hatch,
      dates = date) |> 
      map_df("dev.period") |> 
      list()) |> 
  select(site, dev_period) |> 
  unnest(cols = c(dev_period)) |> 
  mutate(days_to_hatch = stop - start) 
```

We can then plot those results (code not shown) to visualize phenology across our variable set.

```{r, fig.height = 8.75, fig.align='center', fig.cap= "Predicted hatch timing for 139 putatuve Bull Trout sites for three spawning periods."}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

# add year and spawn time factors for plotting
hatch_factors <- hatch_res |> 
  mutate(year = cut(start,
    breaks = cut_ints,
    labels = c(2011:2014)
  )) |>
   # cut coerces our labels to factors, this changes them back to numbers
  mutate(year = as.numeric(as.character(year))) |> 
  mutate(day = day(start)) |> 
  mutate(spawn_time = case_when(
    day == 1 ~ "Early",
    day == 15 ~ "Peak",
    day == 30 ~ "Late"
  )) |>
  mutate(spawn_time = factor(
    spawn_time, levels = c("Late", "Peak", "Early"), 
    ordered = TRUE)
    )

hatch_factors_sum <- hatch_factors |> 
  group_by(year, spawn_time, start) |> 
  summarise(mean_hatch = ceiling(mean(days_to_hatch)),
            hatch_5 = ceiling(quantile(days_to_hatch, probs = 0.05)),
            hatch_95 = ceiling(quantile(days_to_hatch, probs = 0.95)))

ggplot(
  data =  hatch_factors,
    aes(
      x = days_to_hatch,
      y = spawn_time,
      fill = spawn_time,
      color = spawn_time
    )) +
  geom_density_ridges(alpha = 0.9) +
    facet_wrap( ~ year, ncol = 1) +
    scale_fill_brewer(palette = "Blues", direction = 1) +
    scale_color_brewer(palette = "Blues", direction = 1) +
    labs(x = "Days to hatch", y = "Spawn time") + 
  theme_classic(base_size = 28) +
    theme(legend.position = "none")
```

# Conclusion

**hatchR** includes extensive documentation for these and other examples. To fully explore **hatchR** or try out the **Basic prediction** example in Shiny (*scroll pages down on mobile devices*), we recommend visiting the Shiny and package websites.

```{r, out.width="11%",out.height="11%",fig.show='hold', fig.align='center', fig.cap= "Shiny App (left), Website (Right)", fig.width= 0.5}

knitr::include_graphics(c("./hatchR_posterdown_files/figure-html/hatchR_shiny.png", "./hatchR_posterdown_files/figure-html/hatchR_website.png"))
```

### 2025-02-26 inst/manuscript/paper_pdf/paper.Rmd
add the new paper_pdf folder with the newest version of the paper and associated files

--- 2025-01-23 inst/manuscript/paper_oxford/paper_oxford.Rmd
@@
 ---
-title: Template for Oxford University Press papers
-author:
-  - name: Morgan M. Sparks
-    email: Morgan.Sparks@usda.gov
-    orcid: 0000-0003-0787-2218
-    corresponding_author: TRUE # Automatic footnote using corresponding_author
-    affiliation: A
-  - name: Eli Felts
-    orcid: 0000-0002-5888-9266
-    affiliation: B
-  - name: Bryan M. Maitland
-    orcid: 0000-0002-4491-5064
-    affiliation: A
-    footnote:
-address:
-  - code: A
-    department: Rocky Mountain Research Station
-    organization: U.S. Forest Service
-    #postcode: 
-    city: Boise
-    state: Idaho # Mix and match elements to get proper address
-    country: United States
-  - code: B
-    department: Idaho Fish and Wildlife Conservation Office
-    organization: U.S. Fish and Wildlife Service
-    city: Orofino
-    state: Idaho # Mix and match elements to get proper address
-    country: United States
-abstract:
-  - "Understanding the timing of key life history events is necessary for managing and conserving populations.
+title: ""
+output: pdf_document
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
+bibliography: ms_refs.bib
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
+## hatchR: A toolset to predict when fish hatch and emerge
+
+Morgan Sparks^1^, Eli Felts^2^, Allison Swartz^3^, Bryan Maitland^1^
+
+^1^Rocky Mountain Research Station, U.S. Forest Service, Boise, ID, USA
+
+^2^US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
+
+^3^College of Forestry, Oregon State University, Corvallis, OR, USA
+
+keywords: developmental phenology, effective value model, R, Shiny
+
+\newpage
+
+# Abstract
+
+Understanding the timing of key life history events is necessary for managing and conserving populations.
 Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models.
@@
 To demonstrate potential uses, we present two case studies as likely applications for this software.
-**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
-"
-keywords:
-  - developmental phenology
-  - effective value model
-  - R
-  - Shiny
-#boxedlist: # Not fully compatible with all document formats
-#  - First point
-#  - Second point
-competing_interests: |
-  There are no competing interest.
-author_contributions: |
-  MMS and BMM designed and wrote the package. EF wrote the Shiny app. All authors wrote and edited the manuscript.
-acknowledgments: |
-  We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history
 
-  The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.
-Any mention of trade names, products, or services does not imply an endorsement by the U.S.
-government, USFS, or USFWS.
-USFS and USFWS do not endorse any commercial products, services or enterprises.
-
-bibliography: ms_refs.bib # Includes refs in OUP example template
-## When using `citation_package = "default"`, a CSL file can be used
-#csl: https://www.zotero.org/styles/oxford-university-press-note
-output:
-  rticles::oup_article:
-    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
-    journal: "Fisheries"
-    document_style: "contemporary" # Can be contemporary, modern, traditional
-    papersize: "large" # Can be large, medium, small
-    citation_package: "default" # Uncomment when using a CSL; default "natbib"
-    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
-    #number_sections: FALSE # Uncomment to not number sections; default TRUE
-    #number_lines: TRUE # Use `lineno` package - Default FALSE
-    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
-    #onecolumn: TRUE # Uncomment for one column format; default FALSE
-    extra_dependencies:
-      - booktabs # to use with knitr::kable() example below
-
-## Example of pandoc's variable
-#urlcolor: orange
-#linkcolor: green
-#citecolor: red
-header-includes:
-  #- \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
-  - \theoremstyle{thmstyleone} # Theorem stuff from OUP template
-  - \newtheorem{theorem}{Theorem} #  meant for continuous numbers.
-%%\newtheorem{theorem}{Theorem}[section] # meant for sectionwise numbers.
-optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-  - \newtheorem{proposition}[theorem]{Proposition} # %%\newtheorem{proposition}{Proposition}" # to get separate numbers for theorem and proposition etc.
-  - \theoremstyle{thmstyletwo}
-  - \newtheorem{example}{Example}
-  - \newtheorem{remark}{Remark}
-  - \theoremstyle{thmstylethree}
-  - \newtheorem{definition}{Definition}
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE, # By default, hide code; set to TRUE to see code
-  fig.pos = 'th', # Places figures at top or here
-  out.width = '100%', dpi = 300, # Figure resolution and size
-  fig.env="figure"
-) # Latex figure environment
-
-options(knitr.table.format = "latex") # For kable tables to write LaTeX table directly
-```
-
-```{r, include=FALSE}
-library(tidyverse)
-library(hatchR)
-library(ggridges)
-```
+\newpage
 
@@
 ### 
-library(tidyverse); library(hatchR); library(ggdist)
+library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
 ```
@@
 
-```{r, echo = FALSE, out.height = '75%', fig.align = 'center'}
-knitr::include_graphics("flowchart.png")
+```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow.
+Data inputs are represented by folders, processes by filled circles, functions as plain text rectangles, and decisions as italicized text."}
+knitr::include_graphics("./workflow.png")
 ```
@@
 Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required.
-We expect your data to look something like this:
+Data is expected to be formatted similar to Table 1.
 
-```{r kabletable, echo=FALSE}
+```{r kabletable, echo=FALSE }
 df = data.frame(date = c("2000-01-01",
@@
 knitr::kable(df,
-             booktabs=TRUE)
+             booktabs=TRUE,
+             caption = "Example temperature data for use in hatchR.")
 ```
@@
 
-```{r, echo = TRUE}
+```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
 plot_check_temp(data = year_sim,
@@
 
-#### Fitting models for other fishes
+### Fitting models for other fishes
 
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom built parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
 ### predict_phenology
@@
 
-#### Understanding your results
+### Understanding your results
 
@@
 
-#### Plotting phenology
+### Plotting phenology
 
@@
 
-```{r, message=FALSE, echo=FALSE}
+```{r, message=FALSE, echo=FALSE,  out.width='85%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
 cut_ints <- date(

### 2025-02-26 inst/manuscript/paper_pdf/paper.Rmd
added section to paper about why we don't add parameterized models to model_table

@@
 
-^2^US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
+^2^Idaho Fish and Wildlife Conservation Office, U.S. Fish and Wildlife Service, Orofino, ID, USA
 
@@
 
-We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.
+We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset (`year_sim`, first 6 rows shown) with temperature readings every thirty minutes.
 
-```{r, echo = TRUE}
+```{r, echo = FALSE}
 # create date object for a year with 30 min reading intervals
@@
 
-dim(year_sim)
+head(year_sim)
 
@@
 As discussed, these models are included because parameterizations already existed in the literature.
-However, these parameterizations are limited to Pacific Salmon and Bull Trout.
+However, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg.
+183 for numerous other Salmonids) .
 To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law).
+Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C.
+Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to think hard about whether the custom models they are parameterizing fit the data over which they are predicting.
 

### 2025-02-26 inst/manuscript/paper_pdf/paper.Rmd
add additional sections at end (Conflict of Interest, funding, etc.)
Switch to APA reference style, add APA.csl

@@
 bibliography: ms_refs.bib
+csl: apa.csl
 ---
@@
 
-Morgan Sparks^1^, Eli Felts^2^, Allison Swartz^3^, Bryan Maitland^1^
+Morgan Sparks^1,4^, Eli Felts^2^, Allison Swartz^3^, Bryan Maitland^1^
 
@@
 ^3^College of Forestry, Oregon State University, Corvallis, OR, USA
+
+^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@USDA.gov
 
@@
 
+# Acknowledgements
+
+We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature data sets.
+
+The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.
+Any mention of trade names, products, or services does not imply an endorsement by the U.S.
+government, USFS, or USFWS.
+USFS and USFWS do not endorse any commercial products, services or enterprises.
+
+# Conflicts of Interest
+
+The authors declare no known conflicts of interest.
+
+# Data Availability
+
+**hathcR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.
+
+# Ethics Statement
+
+All data was derived from pre-published sources or created synthetically.
+
+# Funding
+
+The authors declare no funding sources.
+
 # References

### 2025-02-27 inst/manuscript/paper_pdf/paper.Rmd
Mess with final figure sizing

@@
 
-```{r, message=FALSE, echo=FALSE,  out.width='85%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
+```{r, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
 cut_ints <- date(

### 2025-02-27 inst/manuscript/paper_pdf/paper.Rmd
add Morgan's edits to manuscript and re-render with edits

@@
 
-Morgan Sparks^1,4^, Eli Felts^2^, Allison Swartz^3^, Bryan Maitland^1^
+Morgan M. Sparks^1,4^, Eli A. Felts^2^, Allison G. Swartz^3^, Bryan M. Maitland^1^
 
@@
 **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
-**hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
+**hatchR** is distributed in two forms—an open source R package for maximum customization and as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
 To demonstrate potential uses, we present two case studies as likely applications for this software.
@@
 
-The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
+The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example, Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
 Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.
@@
 To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**.
-Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+Specifically, **hatchR** allows users to input standard raw or summarized temperature data sets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
 
@@
 Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
-The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
+The Shiny app trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
 Below, we present the basic overview of the software and multiple case studies of how it may be applied.
@@
 Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R.
-For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
-Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
+For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in software of the user's choice.
+Finally, we provide brief summaries of general applications of **hatchR** below (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
 
 ```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow.
-Data inputs are represented by folders, processes by filled circles, functions as plain text rectangles, and decisions as italicized text."}
+Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decisions choices as italicized text."}
 knitr::include_graphics("./workflow.png")
@@
 nerka*).
-The need for their development arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
+The need for their creation arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
 To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.
@@
 
-Water temperature datasets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures.
+Water temperature data sets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures.
 Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.
@@
 **hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
-**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
+**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge could extend to a year or more.
 In these cases, we suggest users consider how much of that data type is reasonable with their data.
@@
 
-We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset (`year_sim`, first 6 rows shown) with temperature readings every thirty minutes.
+We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long data set (`year_sim`, the dimensions of the data and first 6 rows are shown) with temperature readings every thirty minutes.
 
@@
 
-head(year_sim)
-
 ```
 
-First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()`
+```{r}
+#year_sim data dimensions
+dim(year_sim)
+#fist 6 rows of year_sim
+head(year_sim)
+```
 
-```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
+First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()` (Figure 2).
+
+```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set.
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
 plot_check_temp(data = year_sim,
@@
 
-Users can either use Salmonid models from `model_table` included in the package and the Shiny app.
-As discussed, these models are included because parameterizations already existed in the literature.
-However, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg.
-183 for numerous other Salmonids) .
+Users can either use Salmonid models from `model_table` or generate custom models suing `fit_model()` in both R and Shiny deployments of **hatchR**.
+As discussed, the models in `model_table` are included because parameterizations already existed in the literature, however, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg.
+183 for numerous other Salmonids).
+Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C.
+Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to consider whether the custom models they are parameterizing fit the data over which they are predicting.
+
 To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law).
-Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C.
-Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to think hard about whether the custom models they are parameterizing fit the data over which they are predicting.
-
 The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs.
@@
 
-Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
+We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
 We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
-We provide parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
+These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
+We show parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) in the project GitHub.
 
@@
 
-```{r, echo = TRUE}
+```{r, echo=FALSE}
 ###  make temp regime
@@
 colnames(data_sim) <- c("temp_sim", "date_sim")
+```
+
+```{r, echo = TRUE}
 
@@
 
-# change F to C and hours to days
+# change °F to °C and hours to days
 smallmouth <- smallmouth |> 
@@
 
+# model object for smallmouth bass 
 smb_mod <- fit_model(temp = smallmouth$temp_C,
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom built parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
 ### predict_phenology
@@
 
-To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` dataset included in this package .
-We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`.
+To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` data set included in this package.
+We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`, which calls built-in models from `model_table`.
 
@@
 
-And then look inside the returned object to see days to hatch and development period:
+And then look at the returned object to see days to hatch and development period:
 
 ```{r, echo = TRUE}
-WI_hatch$days2done
+WI_hatch$days_to_develop
 WI_hatch$dev.period
@@
 
-`WI_hatch$days2done` outputs the predicted days to develop.
+`WI_hatch$days_to_develop` outputs the predicted days to develop.
 
-`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted todevelop.
+`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to develop.
 
-`WI_hatch$ef.vals` is a vector of each day's effective value as evaluated using whatever model is chosen.
+`WI_hatch$ef_table` is a *n* x 5 tibble (*n* = number of days to hatch or emerge) and the columns are a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
+The `ef_table` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.
 
-`WI_hatch$ef.tibble` is a *n* x 4 tibble (*n* = number of days to hatch or emerge) and the columns are the date, each day's temperature and effective value, and the cumulative sum of the effective values.
-The `ef.tibble` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.
+`WI_hatch$model_specs` provides details about the model used for predicting phenology, which varies depending on if the model was generated from `model_select()` or `fit_model()`, but most importantly contains the model expression (*e.g.*, the formula) used for prediction.
 
@@
 
-`{r"} plot_phenology(WI_hatch)`
+```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` data set."}
+plot_phenology(WI_hatch)
+```
 
@@
 
-We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>.
+We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}.
 Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
-Data used in this case study are from the `crooked_river` data set from R and can either be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
+Data used in this case study are from the `crooked_river` data set.
+It is included as a demo data set in Shiny, but can also be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
 Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check.
@@
 Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in full in Figure XXX, but the interface is described more completely on **hatchR**'s Shiny website.
+The process is demonstrated in the included ***supplementary video file???***, but the interface is described more completely on **hatchR**'s Shiny website.
 
@@
 
-In this example we expect the last fish out of the gravel well before the June 1st date and the manager could allow grazing in this area without worrying about direct mechanical disturbance to fish developing in the gravel.
+In this example we expect the last fish out of the gravel well before the June 1st date and the manager could allow road work in this area without worrying about sediment disturbance to fish developing in the gravel.
 
@@
 
-For the second example, we will again use bull trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**.
-In this scenario we will use data from @isaak2018 (included `idaho` dataset), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho.
+For the second example, we will again use Bull Trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**.
+In this scenario we will use data from @isaak2018 (included `idaho` data set), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho.
 For this approach we winnow putative bull trout spawning sites by filtering for sites with mean August temperature \</= 13 °C in accordance with thresholds from @isaak2015.
@@
 
-We need to setup our models and data for this analysis, which we don't show those steps here for the sake of concision in this manuscript, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**.
-After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure XXX.
+We need to setup our models and data for this analysis, which we don't demonstrate here for the sake of concision, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**.
+After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure 5.
 
@@
 With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations.
-To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times.
-Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for salmonids or create their own species- or population specific parameterization.
-Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of programmatic platform.
+To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive or complex analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times.
+Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for Salmonids or create their own species- or population specific parameterizations.
+Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of the two platforms.
 
@@
 Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.
-Another factor that could bias estimates from these models is that temperatures used from sensors don't reflect the geomoprhic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.
+Another factor that could bias estimates from these models is that temperatures used from sensors do not reflect the geomorphic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.
 
@@
 The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above.
-For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**.
+For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initiation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or ***Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal***.
 Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization.
 Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.
+
+# Conclusion
+
+**hatchR** bridges the analytical gap for the prediction of developmental phenology for fishes in wild environments.
+It provides basic data checks and summarization, options for existing or custom model parameterization, and simple to complex phenology prediction frameworks.
+Importantly, it extends the effective value framework developed by @sparks2019 to a generalizable tool that can be applied to any fish species or population as long as the appropriate source data are available, which can easily be collected in aquaculture settings.
+We present basic applications for how **hatchR** may be used, with additional and more detailed examples provided on the software's [website](https://bmait101.github.io/hatchR/).
+**hatchR** is designed for applied and basic applications in which users can engage with the resource in a programmatic environment or a point-and-click Shiny app.
+We expect we have only demonstrated a small suite of the applications **hatchR** may be used for and encourage the user community to creatively build on the framework developed here.
 

### 2025-02-28 inst/manuscript/paper_pdf/paper.Rmd
update with newest package version and fix corresponding error
revisions to abstract

@@
 
-^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@USDA.gov
+^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov
 
@@
 
-Understanding the timing of key life history events is necessary for managing and conserving populations.
-Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models.
-However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development.
-Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
-Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
-**hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
-**hatchR** is distributed in two forms—an open source R package for maximum customization and as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
-To demonstrate potential uses, we present two case studies as likely applications for this software.
-**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history
+Understanding the timing of key life history events is essential for effective fish sonservation and management.
+Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on avergae incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings.
+Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records.
+However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids.
+Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes.
+**hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
+The sofware is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for thos unfamiliar with scripting.
+To illustrate its utility, we present two case studies demonstrating its application in research and management.
+By expanding assess to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.
 
@@
   development_type = "hatch",
-  author = "Austin et al. 2017",
+  author = "Austin et al. 2019",
   species = "bull trout",

### 2025-02-28 inst/manuscript/paper_pdf/paper.Rmd

--- 2025-02-27 inst/manuscript/paper_pdf/paper.Rmd
@@
 However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development.
-Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
+Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to Salmonids.
 Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
@@
 To demonstrate potential uses, we present two case studies as likely applications for this software.
-**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history
+**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
 

### 2025-02-28 inst/manuscript/paper_pdf/paper.Rmd

--- 2025-02-28 inst/manuscript/paper_pdf/paper.Rmd
@@
 
-Understanding the timing of key life history events is essential for effective fish sonservation and management.
-Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on avergae incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings.
+Understanding the timing of key life history events is essential for effective fish conservation and management.
+Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings.
 Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records.
@@
 **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
-The sofware is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for thos unfamiliar with scripting.
+The software is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for those unfamiliar with scripting.
 To illustrate its utility, we present two case studies demonstrating its application in research and management.

### 2025-02-28 inst/manuscript/JOSS_yaml.R
cleaning ms folder

@@ -1,31 +1,0 @@
----
-  title: 'hatchR: A toolset to predict hatch and emergence phenology in wild fishes'
-tags:
-  - R
-- fishes
-- developmental phenology
-authors:
-  - name: Morgan M. Sparks
-orcid: 0000-0003-0787-2218
-affiliation: 1 # (Multiple affiliations must be quoted)
-- name: Eli Felts
-orcid: 0000-0002-5888-9266
-affiliation: 2
-- name: Bryan M. Maitland
-orcid: 0000-0002-4491-5064
-affiliation: 1
-affiliations:
-  - name: US Forest Service, Rocky Mountain Research Station, Boise, ID, USA
-index: 1
-- name: US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
-index: 2
-citation_author: Sparks et al.
-date: 07 July 2025
-year: 2025
-output: rticles::joss_article
-csl: apa.csl
-journal: JOSS
-editor_options:
-  chunk_output_type: console
-bibliography: ms_refs.bib
----

### 2025-02-28 inst/manuscript/flowchart-render.R

@@ -1,49 +1,0 @@
-
-# path <- here::here("manuscript", "flowchart.gv")
-#
-# DiagrammeR::grViz(g)
-
-g <- "digraph {
-
-      # a 'graph' statement
-      graph [overlap = false, label = 'hatchR workflow', labelloc = t, fontsize = 30]
-
-      # label nodes
-      node [fontsize = 20]
-      import [label = 'import data', shape = circle, fillcolor = Beige, style=filled]
-      dates [label = 'check dates', shape = circle, fillcolor = Beige, style=filled]
-      plot1 [label = 'plot_check_data()' shape = rectangle]
-      format [label = 'is data as\ndaily average?', shape = circle, fillcolor = Beige, style=filled]
-      plot2 [label = 'plot_check_data()' shape = rectangle]
-      summarize [label = 'sumamrize_temp()',  shape = rectangle]
-      continuous [label = 'check_continuous()', shape = rectangle]
-      model [label = 'model type:\ncustom or included?', shape = circle, fillcolor = Beige, style=filled]
-      select [label = 'model_select()', shape = rectangle]
-      fit [label = 'fit_model()', shape = rectangle]
-      phenology [label = 'predict_phenology()', shape = rectangle]
-      plot3 [label = 'plot_phenology()', shape = rectangle]
-
-      # edge definitions with the node IDs
-      edge [fontsize = 20]
-      import -> dates
-      dates -> dates [label = 'not dttm\nuse lubridate']
-      dates -> plot1 [label = 'is dttm']
-      plot1 -> format
-      format -> summarize [label = 'multiple measures\nw/in a day']
-      format -> model [label = '  mean daily average']
-      summarize -> continuous -> plot2 -> model
-      model -> select [label = 'hatchR built-\nin model']
-      model -> fit [label = ' user custom data']
-      {select, fit} -> phenology
-      phenology -> plot3
-
-      }
-
-"
-DiagrammeR::grViz(g)
-
-DiagrammeR::grViz(g) |>
-  DiagrammeRsvg::export_svg() |>
-  charToRaw() |>
-  rsvg::rsvg()  |>
-  png::writePNG("inst/manuscript/flowchart.png")

### 2025-02-28 inst/misc/interactive_checking.R
purge old files

@@ -1,42 +1,0 @@
-load_all()
-
-# get model parameterization
-sockeye_hatch_mod <- model_select(
-  author = "Beacham and Murray 1990",
-  species = "sockeye",
-  model = 2,
-  development_type = "hatch"
-)
-# predict phenology
-sockeye_hatch <- predict_phenology(
-  data = woody_island,
-  dates = date,
-  temperature = temp_c,
-  spawn.date = "1990-08-18",
-  model = sockeye_hatch_mod
-)
-
-# vector of temperatures
-temp <- c(2, 5, 8, 11, 14)
-# vector of days to hatch
-days <- c(194, 87, 54, 35, 28)
-hatch <- tibble::tibble(temp, days)
-hatch <- fit_model(df = hatch, temp = temp, days = days, species = "fish")
-hatch$expression
-
-crooked_river$date <- as.character(crooked_river$date)
-crooked_river$date <- as.Date(crooked_river$date)
-
-plot_check_temp(
-  data = crooked_river,
-  dates = date,
-  temperature = temp_c
-)
-
-# remove a 20 random rows from crooked_river
-tmp <- crooked_river[-sample(1:nrow(crooked_river), 20),]
-
-check_continuous(tmp, date)
-
-chk <- check_continuous(tmp, date)
-chk

### 2025-02-28 inst/misc/to-do-list.R

@@ -1,4 +1,0 @@
-# update lifecycle bages
-
-# add better error and warning messages
-

### 2025-02-28 inst/manuscript/paper.Rmd
introduction edits

@@
 ---
-title: 'hatchR: A toolset to predict hatch and emergence phenology in wild fishes'
-tags:
-  - R
-  - fishes
-  - developmental phenology
-authors:
-  - name: Morgan M. Sparks
-    orcid: 0000-0003-0787-2218
-    affiliation: 1 # (Multiple affiliations must be quoted)
-  - name: Eli Felts
-    orcid: 0000-0002-5888-9266
-    affiliation: 2
-  - name: Bryan M. Maitland
-    orcid: 0000-0002-4491-5064
-    affiliation: 1
-affiliations:
- - name: US Forest Service, Rocky Mountain Research Station, Boise, ID, USA
-   index: 1
- - name: US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
-   index: 2
-citation_author: Sparks et al.
-date: 07 July 2025
-year: 2025
-output: rticles::joss_article
+title: ""
+output: pdf_document
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
+bibliography: ms_refs.bib
 csl: apa.csl
-journal: JOSS
-editor_options:
-  chunk_output_type: console
-bibliography: ms_refs.bib
 ---
 
-```{r, include=FALSE}
-library(tidyverse)
-library(hatchR)
-library(ggridges)
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
+
+## hatchR: A toolset to predict when fish hatch and emerge
+
+Morgan M. Sparks^1,4^, Eli A. Felts^2^, Allison G. Swartz^3^, Bryan M. Maitland^1^
+
+^1^Rocky Mountain Research Station, U.S. Forest Service, Boise, ID, USA
+
+^2^Idaho Fish and Wildlife Conservation Office, U.S. Fish and Wildlife Service, Orofino, ID, USA
+
+^3^College of Forestry, Oregon State University, Corvallis, OR, USA
+
+^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov
+
+keywords: developmental phenology, effective value model, R, Shiny
+
+\newpage
 
@@
 
-Understanding the timing of key life history events is necessary for managing and conserving populations.
-Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models.
-However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development.
-Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids.
-Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses.
-**hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package.
-**hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages.
-To demonstrate potential uses, we present two case studies as likely applications for this software.
-**hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
+Understanding the timing of key life history events is essential for effective fish conservation and management.
+Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings.
+Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records.
+However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids.
+Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes.
+**hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
+The software is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for those unfamiliar with scripting.
+To illustrate its utility, we present two case studies demonstrating its application in research and management.
+By expanding assess to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.
+
+\newpage
 
@@
 
-As primarily poikilothermic organisms, the development and growth of fishes is tightly linked with the temperature of their ambient environment.
-This close relationship has allowed researchers to generate statistical models that allow the prediction of developmental phenology with high accuracy and precision.
-These models were typically developed in aquaculture settings and their initial formulations were not applicable to wild populations because they assumed a constant temperature over the course of development [@beacham1990; @mcphail1979; @alderdice1978].
-However, @sparks2019 reformulated this approach as an "Effective Value model", in which the input was daily average temperature after a parent spawned and fish would either hatch or emerge when effective values cumulatively summed to one.
+As poikilothermic organisms, fish development and growth are closely tied to ambient environment.
+This strong relationship has enabled researchers to generate statistical models that predict developmental phenology with high accuracy.
+Historically, these models were formulated in aquaculture settings under the assumption of constant temperature throughout development [@beacham1990; @mcphail1979; @alderdice1978], limiting their applicability to wild populations.
+However, @sparks2019 reformulated this approach into an "Effective Value model", which instead uses daily average temperature after spawning, predicting hatch or emergence when cumulative effective values reach a threshold of one.
 
-The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019.
-Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.
+This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available.
+For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
+Despite its growing adoption, applications of the effective value model remian largely confined to salmonids, likelu due to the availability of existing parameterizations and the commercial and recreational importance of these species.
 
-To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**.
-Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.
+To extrend these modeling capabilities beyond salminds and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations.
+**hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017.
+Users can also develop custom models using their own or published temperature and phenological data within the effective value framework.
 
-To widen the user application of these methods, we distribute two user-interfaces for **hatchR**.
-The first is a R package (@r:ther) distributed via CRAN that allows users the most customizable application for these methods.
-The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes.
-These variable approaches are outlined in the package documentation on **hatchR**'s website.
-Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface.
-The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package.
-Below, we present the basic overview of the software and multiple case studies of how it may be applied.
+To maximize accessibility, **hatchR** is available in two formats.
+The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing mutiple variables, such as phenology type, spawn timing, or thermal regimes.
+Comprehensize documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
+The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
+Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.
 
@@
 Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R.
-For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.
-Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
+For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in software of the user's choice.
+Finally, we provide brief summaries of general applications of **hatchR** below (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
 
-```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap= "Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles."}
-knitr::include_graphics("flowchart.png")
+```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow.
+Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decisions choices as italicized text."}
+knitr::include_graphics("./workflow.png")
 ```
@@
 Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required.
-We expect your data to look something like this:
+Data is expected to be formatted similar to Table 1.
 
-| date       | temperature |
-|------------|-------------|
-| 2000-01-01 | 2.51        |
-| ...        | ...         |
-| 2000-07-01 | 16.32       |
-| ...        | ...         |
-| 2000-12-31 | 3.13        |
+```{r kabletable, echo=FALSE }
+df = data.frame(date = c("2000-01-01",
+                         "...",
+                         "2000-07-01",
+                         "...",
+                         "2000-12-31"),
+                temperature= c(2.51,
+                        "...",
+                        16.32,
+                        "...",
+                        3.13
+                        ))
+
+# kable can alse be used for creating tables
+knitr::kable(df,
+             booktabs=TRUE,
+             caption = "Example temperature data for use in hatchR.")
+```
 
 **hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
-**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year.
+**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge could extend to a year or more.
 In these cases, we suggest users consider how much of that data type is reasonable with their data.
@@
 
-We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.
+We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long data set (`year_sim`, the dimensions of the data and first 6 rows are shown) with temperature readings every thirty minutes.
 
-```{r}
+```{r, echo = FALSE}
 # create date object for a year with 30 min reading intervals
@@
 
-dim(year_sim)
-
 ```
 
-First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()`
+```{r}
+#year_sim data dimensions
+dim(year_sim)
+#fist 6 rows of year_sim
+head(year_sim)
+```
 
-```{r, fig.cap= "Full output of the plot_check_temp() visual data check function from hatchR.
-Note that the red and blue lines are customizable by the user to create outlier thresholds for their visual check."}
+First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()` (Figure 2).
+
+```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set.
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
 plot_check_temp(data = year_sim,
@@
 
-Users can either use Salmonid models from `model_table` included in the package and the Shiny app.
-As discussed, these models are included because parameterizations already existed in the literature.
-However, these parameterizations are limited to Pacific Salmon and Bull Trout.
-To widen the application of the effective value approach, we include a `fit_model()` function and is species agnostic (as long as develop generally follows the power law.
+Users can either use Salmonid models from `model_table` or generate custom models suing `fit_model()` in both R and Shiny deployments of **hatchR**.
+As discussed, the models in `model_table` are included because parameterizations already existed in the literature, however, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg.
+183 for numerous other Salmonids).
+Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C.
+Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to consider whether the custom models they are parameterizing fit the data over which they are predicting.
 
-The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs and estimates parameter coeficients for *log~e~a* and *b* using `stats::nls()`
+To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law).
+The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs.
+The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`
 
@@
 
-#### Fitting models for other fishes
+### Fitting models for other fishes
 
-Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
+We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
 We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.
-We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
+These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
+We show parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) in the project GitHub.
 
@@
 
-```{r}
+```{r, echo=FALSE}
 ###  make temp regime
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap= "Days to hatch as predicted using custom parameterizations for Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass (green) as generated by the fit_model() function in hathcR.
-Models are all run using a spawn date starting June 1 for a randomly generated ascending thermal regime with mean temperature 16 °C"}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
 ### predict_phenology
@@
 
-To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` dataset included in this package .
-We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`.
+To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` data set included in this package.
+We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`, which calls built-in models from `model_table`.
 
-```{r}
+```{r, echo = TRUE}
 sockeye_hatch_mod <- model_select(
@@
 
-And then look inside the returned object to see days to hatch and development period:
+And then look at the returned object to see days to hatch and development period:
 
-```{r}
-WI_hatch$days2done
+```{r, echo = TRUE}
+WI_hatch$days_to_develop
 WI_hatch$dev.period
@@
 
-`WI_hatch$days2done` outputs the predicted days to develop.
+`WI_hatch$days_to_develop` outputs the predicted days to develop.
 
-`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted todevelop.
+`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to develop.
 
-`WI_hatch$ef.vals` is a vector of each day's effective value as evaluated using whatever model is chosen.
+`WI_hatch$ef_table` is a *n* x 5 tibble (*n* = number of days to hatch or emerge) and the columns are a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
+The `ef_table` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.
 
-`WI_hatch$ef.tibble` is a *n* x 4 tibble (*n* = number of days to hatch or emerge) and the columns are the date, each day's temperature and effective value, and the cumulative sum of the effective values.
-The `ef.tibble` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.
+`WI_hatch$model_specs` provides details about the model used for predicting phenology, which varies depending on if the model was generated from `model_select()` or `fit_model()`, but most importantly contains the model expression (*e.g.*, the formula) used for prediction.
 
-#### Plotting phenology
+### Plotting phenology
 
@@
 
-```{r, fig.cap= "Output of the plot_phenology() function from hatchR.
-The function plots the output of the predict_phenology() function and displays time to develop, as well as the ambient thermal regime (blue), scaled daily effective value (green), and the scaled cumulative sum of effective values (orange)."}
+```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` data set."}
 plot_phenology(WI_hatch)
@@
 A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action.
-For instance, will have fish have emerged from redds when a stream section has been opened to grazing or bridge decommissioning will commence?
+For instance, will fish have emerged from redds when a stream section has been opened to grazing or road work?
 
-In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed.
-The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st.
+In this scenario, we will consider the road work example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the Crooked River, a key Bull Trout population in the Boise River watershed.
+In this hypothetical scenario, the Forest Fisheries Biologist wants to know if fish will likely be out of the gravel and free-swimming by June 1st as Bull Trout are particularly sensitive to sediment.
 In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.
 
-We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>.
+We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}.
 Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
+Data used in this case study are from the `crooked_river` data set.
+It is included as a demo data set in Shiny, but can also be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
 Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check.
@@
 Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on **hatchR**'s website <https://bmait101.github.io/hatchR/>.
+The process is demonstrated in the included ***supplementary video file???***, but the interface is described more completely on **hatchR**'s Shiny website.
 
@@
 
-We need to setup our models and data for this analysis, which we don't show those steps here for the sake of concision in this manuscript, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**.
-After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure XXX.
+We need to setup our models and data for this analysis, which we don't demonstrate here for the sake of concision, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**.
+After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure 5.
 
-```{r}
+```{r, echo = TRUE}
 hatch_res <- isaak_summ_bt |> 
@@
 
-```{r, message=FALSE, echo=FALSE, fig.cap= "Densities of predicted days to hatch for Bull Trout for 139 putative spawning locations across Idaho for four years of shared temperature data.
-Early spawning considers if a parent spawned September 1, Peak-September 15, and Late-September 30."}
+```{r, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
 cut_ints <- date(
@@
 The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above.
-For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**.
+For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initiation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or ***Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal***.
 Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization.
@@
 
+# Conclusion
+
+**hatchR** bridges the analytical gap for the prediction of developmental phenology for fishes in wild environments.
+It provides basic data checks and summarization, options for existing or custom model parameterization, and simple to complex phenology prediction frameworks.
+Importantly, it extends the effective value framework developed by @sparks2019 to a generalizable tool that can be applied to any fish species or population as long as the appropriate source data are available, which can easily be collected in aquaculture settings.
+We present basic applications for how **hatchR** may be used, with additional and more detailed examples provided on the software's [website](https://bmait101.github.io/hatchR/).
+**hatchR** is designed for applied and basic applications in which users can engage with the resource in a programmatic environment or a point-and-click Shiny app.
+We expect we have only demonstrated a small suite of the applications **hatchR** may be used for and encourage the user community to creatively build on the framework developed here.
+
 # Acknowledgements
 
-We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
+We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature data sets.
 
@@
 
+# Conflicts of Interest
+
+The authors declare no known conflicts of interest.
+
+# Data Availability
+
+**hathcR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.
+
+# Ethics Statement
+
+All data was derived from pre-published sources or created synthetically.
+
+# Funding
+
+The authors declare no funding sources.
+
 # References

### 2025-02-28 inst/manuscript/paper.Rmd
edits to overview and ef framework

@@
 
-**hatchR** is meant to primarily be a tool for predicting phenology.
-In this sense, we limit functionality to these applications and provide minimal data checking and plotting help.
-This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks.
-In other words, we expect users to know their data better than we do and to check it accordingly.
-We do provide two basic data check functions discussed in the Checking Data section.
-Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R.
-For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in software of the user's choice.
-Finally, we provide brief summaries of general applications of **hatchR** below (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.
+**hatchR** is designed primarily as a tool for predicting fish phenology.
+To maintain focus on this core function, we provide minimal build-in data validation and vizualization tools, as users are expected to understand and check their own data.
+Given the diversity of potential data types, it is impractical to implement comprehensize validation checks.
+However, we include basic data-checking functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`).
+Intuitive functions are provided for users to apply models (either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function).
+Users can then apply these models to water temperature data (e.g.
+from a HOBO temp logger) to predict when hatching phenology will occur.
+This is accomplished with the `predict_phenology()` function.
+The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external vizualizations.
+For a high-level overview of **hatchR**'s applications, see Figure 1.
+Additional details on key functions and workflows–particulalry for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.
 
@@
 
-Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O.
+Effective value models were introduced by @sparks2019 to predict developmental timing in wild populations, initially for Sockeye Salmon (*O.
 nerka*).
-The need for their creation arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known.
-To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.
+Their development was necessitated by limitations in traditional models, such as those in @beacham1990, which relied on average incubation temperature over the full developmental period.
+In wild settings, estimating this average temperature was impracticable since hatch timing was unknown, even when spawning dates were recorded.
+To overcome this challenge, @sparks2019 reformulated model 2 from @beacham1990 by taking its reciprocal and assigning an *effective value* to each day of development based on the daily average temperature.
+This approach allowed for cumulative tracking of developmental progress, enabling hatch and emergence predictions without requiring prior knowledge of incubation temperatures.
 
@@
 $$
-Effective Value_i = 1/exp(log_ea - log_e(Temperature_i - b))
+\mathrm{E_i} = \frac{1}{\exp(\log_e(a) - \log_e(T_i - b)}
 $$
 
-Where *i* is the daily value and a fish hatches or emerges when the cumulative sum of effective values reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$
+where $E_i$ is the effective value and $T_i$ the temperature for day $i$, and $a$ and $b$ are model parameterization estimates (i.e. species- or model-specific constants).
 
-The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`.
-Specifically, `model_table` includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
+This framework as the foundation for phenological models in **hatchR**.
+The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
+While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function.
+This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.
 

### 2025-02-28 inst/poster/hatchR_posterdown/hatchR_posterdown.Rmd
change title and fix Austin et al. error

@@
 ---
-title: "hatchR: A toolset to predict when fish hatch and emerge"
+title: "hatchR: A tool set to predict when fish hatch and emerge"
 author:
@@
 bull_trout_emerge <- model_select(
-  author = "Austin et al. 2017",
+  author = "Austin et al. 2019",
   species = "bull trout",

### 2025-02-28 R/idaho.R
fix typo

@@ -9,7 +9,7 @@
 #' \describe{
 #' \item{date}{Date measurement was taken.}
 #' \item{site}{Site ID.}
-#' \item{twmp_c}{Temperature in degrees Celsius.}
+#' \item{temp_c}{Temperature in degrees Celsius.}
 #' }
 #' @examples
 #' idaho

### 2025-02-28 inst/manuscript/paper.Rmd
edits to data format and checking

@@
 
-Water temperature data sets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures.
-Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.
+Water temperature datasets collected in the field typically fall into two categories: 1) summarized daily data, where mean daily temperatures are pre-computed, or 2) raw high-frequency data, such as those recorded by HOBO TidbiT loggers, which require summary into mean daily temperatures before use.
+Additionally, new statistical models, such as @siegel2023, could also be implemented into this framework.
 
-Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade).
-Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required.
-Data is expected to be formatted similar to Table 1.
+**hatchR** assumes input data consists if at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
+Other columns may be present, but columns names should not include spaces.
+Data should follow the format outlined in Table 1.
 
@@
 
-**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks.
-**hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge could extend to a year or more.
-In these cases, we suggest users consider how much of that data type is reasonable with their data.
+Since **hatchR** does not automatically handle missing data, users must check for gaps or errors before running analyses.
+The package will function with missing values, but gaps in the dataset may affect predictions.
+**hatchR** supports temperatures as low as 0 °C, though such values yield extremely small effective values, potentially extending hatch or emergence timing to a year or more.
+Users should critically assess whether such data align with biological expectations.
 
-For users choosing to implement **hatchR** in R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record.
-For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.
+For R users, **hatchR** can import data in any format, provided it is converted into a `data.frame` or `tibble`, where each row represents a single temperature record.
+The Shiny application requires data to be uploaded as a .csv (comma separated values) file, which can easily be exported from spreadsheet software such as Microsoft Excel or Google Sheets.
 
@@
 
-**hatchR** is built assuming data will be analyzed as daily average temperatures.
-Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable (only in R, it must be summarized for the Shiny app), as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.
+**hatchR** is designed to analyze daily average temperatures.
+While high-frequency data (e.g., from HOBO loggers) can be used, it must be summarized into daily averages.
+**hatchR** provides built-in functionality for this summarization in R but requires pre-summarized data for use in the Shiny app.
+
+To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programattically.
 
@@
 ```{r, echo = FALSE}
+library(lubridate)
 # create date object for a year with 30 min reading intervals

### 2025-02-28 inst/manuscript/paper.Rmd
edits to checking data

@@
 
-To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programattically.
+To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically.
+These checks ensure data integroty before model application.
 
-We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long data set (`year_sim`, the dimensions of the data and first 6 rows are shown) with temperature readings every thirty minutes.
+We demonstrate the utility three functions: `summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`).
+This dataset contains temperature readings taken every thirty minutes, and its structure (dimentions and first six rows) is shown below.
 
@@
 
-First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()` (Figure 2).
+First, we recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 2).
 
@@
 
-There are no obvious outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`.
-We also recommend using `plot_check_temp()` again on the summarized data (though leave out the resulting plot for space efficiency in this manuscript).
+In this case, no obvious outliers are present, but since each day contains 48 records, the data must be summarized to daily mean temperature using `summarize_temp()`.
+After summarization, `check_continuous()` should be used to identify any missing days.
+We also suggest running `plot_check_temp()` again on the summarized data to verify its integrity, though we omit the resulting plot here for space efficiency.
 

### 2025-02-28 R/idaho.R
add source link

@@ -4,7 +4,7 @@
 #' Water temperature data from Isaak et al. (2018).  Generally covers the
 #' Boise, Payette, Clearwater, and upper Salmon River watersheds.
 #'
-#' @source To come...
+#' @source https://hess.copernicus.org/articles/22/6225/2018/
 #' @format Data frame
 #' \describe{
 #' \item{date}{Date measurement was taken.}

### 2025-02-28 inst/manuscript/paper.Rmd
edits to model selection

@@
 
-Users can either use Salmonid models from `model_table` or generate custom models suing `fit_model()` in both R and Shiny deployments of **hatchR**.
-As discussed, the models in `model_table` are included because parameterizations already existed in the literature, however, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg.
-183 for numerous other Salmonids).
-Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C.
-Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to consider whether the custom models they are parameterizing fit the data over which they are predicting.
+Users can select from existing Salmonid models in `model_table` or generate custom models using `fit_model()` in both the R and Shiny deployments of **hatchR**.
+The models in `model_table` are included because their parameterizations are well-documented in the literature, though they are currently limited to Pacific Salmon and Bull Trout (see @quinn2018 pg.
+183, for additional Salmonid models).
+To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C.
 
-To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law).
-The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs.
-The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`
+Custom models, by contrast, often have narrower parameterization ranges.
+To prevent misapplication, we exclude other model parameterization from `model_table`, requiring users to carefully asses whether their parameterized models are appropriate for the temperature ranges in their datasets.
 
-**BRYAN TO FINISH THIS PARAGRAPH OFF HERE**
+To expand the applicability of the effective value approach beyond Salmonids, **hacthR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature.
+The function takes two input vectors--one for average incubation temperature (°C) and one for the number of days to a given phenological event--and estimates model parameters (*log~e~a* and *b*) using `stats::nls()`.
+
+This approach allows users to generate models tailored to non-Salmonid species, provided they have experimental or field data linking development to temperature.
+However, users should be mindful that custom models may not generalize well beyond the temperature range of the data used to fit the model.
+Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, pending sufficient validation in the literature.
 
@@
 ```{r, echo=FALSE}
+library(dplyr)
 ###  make temp regime

### 2025-02-28 R/plot_check_temp.R
more description for \code section

@@ -4,7 +4,7 @@
 #' The `plot_check_temp` function is used to plot imported data to check
 #' temperature values. The function takes a data frame with dates and
 #' temperature values, and plots the temperature values over time. The
-#' function also allows users to specify the minimum and maximum temperature
+#' function also allows users to specify the visual thresholds for minimum and maximum temperature
 #' values to be plotted.
 #'
 #' @param data A data.frame, or data frame extension (e.g. a tibble).

### 2025-02-28 inst/manuscript/paper.Rmd
edits to fitting custom models

@@
 
-We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package.
-We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach.
-These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
-We show parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) in the project GitHub.
+We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`.
+To showcase its utility, we provide parameterizations for  three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
+These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach.
+For conciseness, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) on the  GitHub project repository.
 
-We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime (Figure 3).
+Next, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C.
+Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3).
+This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid system.
 
@@
 
-Note the *R^2^* fit from the models below. You can see they generally all preform well and are in line with values from model 2 of @beacham1990.
+Note the *R^2^* fit from the models below. You can see they generally preform well and are in line with values from model 2 of @beacham1990.
 
@@
 smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared
-
 ```

### 2025-02-28 inst/manuscript/paper.Rmd
edits to prediction, plotting, and understanding results

@@
 
-To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` data set included in this package.
-We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`, which calls built-in models from `model_table`.
+To illustrate model selection and phenology prediction, we will replicate a portion of the analysis from  @sparks2019 using the `woody_island` dataset included with **hatchR**.
+Speficially, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
+
+First, we obtain model expression for both hatch and emergence using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
 
@@
 
-We can now use our model expressions to predict when sockeye would hatch and emerge at Woody Island in 1990. First we predict hatch timing using `predict_phenology()`:
+These model expressions are then applied in `predict_phenology()` to estimate the days to hatch and development:
 
@@
 
-And then look at the returned object to see days to hatch and development period:
+The returned object provides key outputs, including days to hatch and the full developmental period, allowing us to assess phenological patterns under the recorded thermal conditions:
 
@@
 
-The output from `predict_phenology()` includes a lot of information.
-If we look at our `WI_hatch` object we see there are multiple elements stored in a list which can be accessed using the `$` operator.
+The output from `predict_phenology()` contains mutiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:
 
@@
 
-`WI_hatch$days_to_develop` outputs the predicted days to develop.
+`WI_hatch$days_to_develop` -- Returns the predicted number of days required for development.
 
-`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to develop.
+`WI_hatch$dev.period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
 
-`WI_hatch$ef_table` is a *n* x 5 tibble (*n* = number of days to hatch or emerge) and the columns are a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
-The `ef_table` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.
+`WI_hatch$ef_table` -- An *n* x 5 tibble (*n* = number of days to hatch or emerge), containing a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
+This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discuss below.
 
-`WI_hatch$model_specs` provides details about the model used for predicting phenology, which varies depending on if the model was generated from `model_select()` or `fit_model()`, but most importantly contains the model expression (*e.g.*, the formula) used for prediction.
+`WI_hatch$model_specs` -- Provides details about the model used for prediction, including whether it was retrieved from `model_select()` or generated using `fit_model()`.
+Most importantly, it contains the model expression (i.e., the formula) used for phenology predictions.
 
@@
 
-**hatchR** has a built in function, `plot_phenology()`, that allows users to visualize their phenology results (Figure 4).
-The plot visualizes three specific components: 1.) the temperature regime over which you are predicting, 2.) the cumulative sum of effective values, and 3.) the effective value for each day in your prediction span.
-The function allows you to output various figures based on your interests, but defaults to a figure with all information and the corresponding labels.
+**hatchR** includes a built in function, `plot_phenology()`, for visualizing phenology predictions (Figure 4).
+This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span.
+Be default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles.
+However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results.
+
+This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.
 

### 2025-02-28 inst/manuscript/paper.Rmd
edits to case study 1

@@
 
-# Case Study 1
+# Case Study 1:
 
-A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action.
-For instance, will fish have emerged from redds when a stream section has been opened to grazing or road work?
+**Bull Trout in the Crooked River** OR
+**Predicting Emergence Timing for Management Actions**
 
-In this scenario, we will consider the road work example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the Crooked River, a key Bull Trout population in the Boise River watershed.
-In this hypothetical scenario, the Forest Fisheries Biologist wants to know if fish will likely be out of the gravel and free-swimming by June 1st as Bull Trout are particularly sensitive to sediment.
-In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.
+A common management application of developmental phenology is assessing whether fish will be free-moving before a scheduled management action, such as stream section access for grazing or road work.
+For instance, will fish have emerged from redds before construction begins, reducing potential disturbance?
 
-We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}.
-Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year).
-Data used in this case study are from the `crooked_river` data set.
-It is included as a demo data set in Shiny, but can also be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
-Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check.
-After uploading and checking data, the user switches to the `Model Phenology` window.
-In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu.
-After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided.
-We show results for spawning for September 30th as indicated in the example above.
-Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab.
-Output from predicting phenology and the resulting figure are downloadable from their respective tabs.
-The process is demonstrated in the included ***supplementary video file???***, but the interface is described more completely on **hatchR**'s Shiny website.
+In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, where a key Bull Trout (*Salvelinus confluentus*) population exists.
+Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensative to sediment disturbance.
+The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st.
+In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.
+
+To demonstrate this case study, we use the **hatchR**  graphical user interface available at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}.
+Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year).
+For this example, we use the `crooked_river` dataset, included in the Shiny app as a demo dataset.
+It can also be accessed directly at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
+
+Once uploaded, **hatchR** automatically generates a vizual data check using `plot_check_temp()`.
+After confirming data integrity, users navigate to the `Model Phenology` window.
+For this case study, we use the preloaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus.
+The user then chooses multiple spawn dates using an interactive calendar.
+Here, we focus on September 30th as outlined in our management scenario.
+
+Following model selection, **hatchR** outputs results in two key locations: the `Phenology Summaries` tab, which provides a table with predictions for each spawn date, and the `Timeline Plot` tab, which shows the corresponding visualization of development timing.
+Both the prediction table and plot can be downloaded directly from their respective tabs.
+
+The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walkthrough is available on the **hatchR** Shiny website.
 
@@
 
-In this example we expect the last fish out of the gravel well before the June 1st date and the manager could allow road work in this area without worrying about sediment disturbance to fish developing in the gravel.
+In this example, the model predicts that the last Bull Trout will emerge before the June 1st traget date.
+This suggests that the Fisheries Biologist can confidently approve the road work in the area without concern for sediment disturbance impacting fish developing in the gravel.
+This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with necessary land-use activities.
 

### 2025-02-28 inst/manuscript/paper.Rmd
edits to case study 2

@@
 
-For the second example, we will again use Bull Trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**.
-In this scenario we will use data from @isaak2018 (included `idaho` data set), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho.
-For this approach we winnow putative bull trout spawning sites by filtering for sites with mean August temperature \</= 13 °C in accordance with thresholds from @isaak2015.
-For the resulting 139 sites we will demonstrate predicting hatch timing in these putative Bull Trout spawning habitats.
+**Bull Trout in the Upper Columbia River Headwaters** OR
+**Large-scale Predictions of Bull Trout development timing**
+
+For the second case study, we demonstrate a more complex, large-sclae application of **hatchR**, highlighting its full flexibility when applied programmatically in R.
+This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the upper Columbia River headwaters in Idaho.
+
+We use the `idaho` dataset from @isaak2018, which includes temperature data from these sites.
+To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature \</= 13 °C, a known thermal threshold for Bull Trout spawning suitability.
+The filtering process reduced the dataset to 139 sites potential spawning sites.
 
@@
 
-We need to setup our models and data for this analysis, which we don't demonstrate here for the sake of concision, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**.
-After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure 5.
+To predict hatch timing across these sites, we:
+
+1.
+Set up the necessary moels and data, using Bull Trout parameterization.
+(For conciseness, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository for **hatchR**.)
+
+2. Apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn date:
+
+* Early: September 1st
+* Peak: September 15th
+* Late: September 30th
+
+By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing caries across different spawning habitats.
+The results of this analysis are presented in Figure 5, providing insights into how emergence timing might vary under different thermal regimes and across the species' geographic range.
+This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
 

### 2025-02-28 inst/manuscript/paper.Rmd
edits to discussion

@@
 This is accomplished with the `predict_phenology()` function.
-The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external vizualizations.
+The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external visualizations.
 For a high-level overview of **hatchR**'s applications, see Figure 1.
-Additional details on key functions and workflows–particulalry for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.
+Additional details on key functions and workflows–particularly for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.
 
@@
 
-With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations.
-To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive or complex analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times.
-Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for Salmonids or create their own species- or population specific parameterizations.
-Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of the two platforms.
+With **hatchR**, we present a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes.
+It establishes a formal framework for applying effective value models from user-provided parameterizations.
+The software is available in two formats:
 
-In the application of using effective value models **hatchR** and the user make some key assumptions.
-Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment.
-These include water quality like dissolved oxygen, pH, or salinity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024).
-Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.
-Another factor that could bias estimates from these models is that temperatures used from sensors do not reflect the geomorphic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.
+* A fully customizable R package, ideal for complex and repetitive analyses.
+* A graphical user interface (GUI) for ease of use, designed for tasks that may only need to be run once or a few times.
 
-To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019).
-However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species (@west-eberhard2003).
-For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes.
-Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks.
-The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation (@conover2009, @sparks2022).
-In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.
+Both versions allow users to import data, perform basic data checks, and apply either pre-existing salmonid model parameterizations or generate custom models specific to other species or populations.
+To support users at various levels of expertise, we provide extensive documentation on the hatchR website (https://bmait101.github.io/hatchR/), covering basic and advanced applications.
 
-**BRYAN check the below**
+## Assumptions and considerations in applying effective value models
 
-The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above.
-For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initiation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or ***Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal***.
-Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization.
-Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.
+The application of hatchR and the effective value modeling framework relies on several key assumptions.
+First, environmental stressors may alter developmental timing.
+While effective value models predict developmental timing based on temperature, studies have shown that stressful environmental conditions—such as low dissolved oxygen, altered pH, high salinity, pathogen exposure, or mechanical disturbance—can induce premature hatching or emergence (@quinn2018; @cowan2024).
+Users should consider how such factors may influence their predictions.
+
+Second, developmental timing occurs as a distribution, not a fixed point.
+While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events (@mason1976).
+We encourage users to predict phenology using early, peak, and late thresholds (e.g., 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.
+
+Third, sensor-based temperature data may differ from actual egg temperatures.
+Water temperatures recorded by environmental sensors may not fully reflect thermal conditions in spawning microhabitats, where geomorphic factors influence temperature regimes (@geist2002).
+Users should consider how differences between measured and actual incubation temperatures may affect predictions.
+
+## Evolutionary and population-level considerations
+
+To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations (@adelfio2024; @austin2019).
+However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by genetic variation, gene-environment interactions, and phylogenetic differences (@west-eberhard2003).
+
+For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes.
+Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada (@beacham1990), consistent with cogradient variation (@conover2009; @sparks2022).
+These findings highlight the importance of considering not only developmental time predictions but also how the underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.
+
+## Future directions OR Expanding the utility of **hatchR**
+
+The models within **hatchR** can be customized in multiple ways beyond the examples provided.
+While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones, such as early embryonic stages (e.g., eye-up in salmonids; @velsen1980), initiation of pelagic-larval dispersal in marine fish, where temperature regulates the transition from demersal to pelagic behavior, or current-mediated dispersal in riverine species, where larvae begin downstream drift at temperature-dependent thresholds.
+
+Additionally, while `fit_model()` uses non-linear regression to estimate parameters, `predict_phenology()` only requires users to provide a model expression.
+This means that users can integrate alternative model structures, as long as they incorporate daily temperature, allowing further customization of predictions.
+
+Finally, while hatchR was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power-law relationship with temperature.
+Extending the effective value framework to these taxa could provide valuable insights into their life history timing under variable environmental conditions.
 
@@
 
-**hatchR** bridges the analytical gap for the prediction of developmental phenology for fishes in wild environments.
-It provides basic data checks and summarization, options for existing or custom model parameterization, and simple to complex phenology prediction frameworks.
-Importantly, it extends the effective value framework developed by @sparks2019 to a generalizable tool that can be applied to any fish species or population as long as the appropriate source data are available, which can easily be collected in aquaculture settings.
-We present basic applications for how **hatchR** may be used, with additional and more detailed examples provided on the software's [website](https://bmait101.github.io/hatchR/).
-**hatchR** is designed for applied and basic applications in which users can engage with the resource in a programmatic environment or a point-and-click Shiny app.
-We expect we have only demonstrated a small suite of the applications **hatchR** may be used for and encourage the user community to creatively build on the framework developed here.
+**hatchR** provides a versatile and accessible tool for predicting developmental phenology in wild fish populations.
+It offers basic data checks and summarization tools, pre-existing and customizable model parameterization options, and scalable applications from simple site-level predictions to complex multi-site analyses.
+
+Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings.
+We present foundational applications of **hatchR**, with additional use cases and implementation guides available on the software's website.
+The software is designed for both applied and basic research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app.
+We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.
 

### 2025-02-28 inst/manuscript/paper.Rmd
fix typos, move lib call to setup chunk

@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, warning=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
+library(tidyverse)
+library(hatchR)
+library(ggdist)
+library(ggridges)
 ```
@@
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
-Despite its growing adoption, applications of the effective value model remian largely confined to salmonids, likelu due to the availability of existing parameterizations and the commercial and recreational importance of these species.
+Despite its growing adoption, applications of the effective value model remain largely confined to salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.
 
-To extrend these modeling capabilities beyond salminds and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations.
+To extend these modeling capabilities beyond Salmonids and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations.
 **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017.
@@
 To maximize accessibility, **hatchR** is available in two formats.
-The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing mutiple variables, such as phenology type, spawn timing, or thermal regimes.
-Comprehensize documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
+The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
+Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
 The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
 Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.
-
-```{r echo=FALSE, message=FALSE}
-### 
-library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
-```
 

### 2025-02-28 inst/manuscript/paper.Rmd

@@
 
-```{r setup, include=FALSE, warning=FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
-library(tidyverse)
-library(hatchR)
-library(ggdist)
-library(ggridges)
 ```
@@
 
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
+```
+
 # Package Overview
@@
 
-where $E_i$ is the effective value and $T_i$ the temperature for day $i$, and $a$ and $b$ are model parameterization estimates (i.e. species- or model-specific constants).
+where $E_i$ is the effective value and $T_i$ the temperature for day $i$, and $a$ and $b$ are model parameterization estimates (i.e.
+species- or model-specific constants).
+A fish hatches or emerges when the cumulative sum of effective values reaches one:
+
+$$
+\sum_{i =1}^nE_i = 1
+$$
 

### 2025-03-03 README.Rmd
Fixes for several issues; mostly copy edits (#24)
* Copy edits fix #15

* Revise structure and add new fixes

* Copy eidts to fix #17

* Copy edits; closes #18

* Copy edits; closes #19

* Adding minimum versions for dependencies

* Remove minimum on base R packages; closes #16
update install instructions, title, desc

@@
 
-# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
+# hatchR: Predict Fish Hatch and Emergence Timing <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
 
@@
 
-**hatchR** is an R package for predicting fish developmental phenology using statistical models.
-It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (*e.g.*, Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al.
-(2019).
+**hatchR** is an R package for predicting hatch and emergence timing for a wide range of wild fishes using the effective value framework (Sparks et al., (2019) <DOI:10.1139/cjfas-2017-0468>).
+'hatchR' offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
+The package includes functions for model selection, phenology prediction, and visualization.
+**hatchR** is designed to be user-friendly and accessible to a wide range of users, including researchers, managers, and educators.
 
@@
 
-You can install the development version of **hatchR** from [GitHub](https://github.com/) with:
+Install and load the most recent approved version from CRAN by running:
 
 ```{r}
-# install.packages("pak")
+# install the most recent approved version from CRAN
+install.packages("hatchR")
+```
+
+Install the development version of **hatchR** from [GitHub](https://github.com/) by running:
+
+```{r}
+# install the most recent version from GitHub
 pak::pak("bmait101/hatchR")
 ```
+
+## Interactive Shiny App
+
+You can also use the **hatchR** Shiny app to interactively predict fish developmental phenology. To run the app, visit the [hatchR Shiny app](https://elifelts.shinyapps.io/hatchR_demo/).
 
@@
 ```
+
+**hatchR** requires three primary data inputs:
+
+1.  Water temperature data
+2.  Species-specific model parameters
+3.  Spawning date or date ranges
 

### 2025-03-03 inst/manuscript/paper.Rmd
edit abstract

@@
 Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records.
-However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids.
+However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on Salmonids.
 Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes.
@@
 To illustrate its utility, we present two case studies demonstrating its application in research and management.
-By expanding assess to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.
+By expanding access to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.
 
@@
 
-**Bull Trout in the Crooked River** OR
-**Predicting Emergence Timing for Management Actions**
+**Bull Trout in the Crooked River** OR **Predicting Emergence Timing for Management Actions**
 
@@
 
-**Bull Trout in the Upper Columbia River Headwaters** OR
-**Large-scale Predictions of Bull Trout development timing**
+**Bull Trout in the Upper Columbia River Headwaters** OR **Large-scale Predictions of Bull Trout development timing**
 
@@
 
-* Early: September 1st
-* Peak: September 15th
-* Late: September 30th
+-   Early: September 1st
+-   Peak: September 15th
+-   Late: September 30th
 
@@
 
-* A fully customizable R package, ideal for complex and repetitive analyses.
-* A graphical user interface (GUI) for ease of use, designed for tasks that may only need to be run once or a few times.
+-   A fully customizable R package, ideal for complex and repetitive analyses.
+-   A graphical user interface (GUI) for ease of use, designed for tasks that may only need to be run once or a few times.
 
 Both versions allow users to import data, perform basic data checks, and apply either pre-existing salmonid model parameterizations or generate custom models specific to other species or populations.
-To support users at various levels of expertise, we provide extensive documentation on the hatchR website (https://bmait101.github.io/hatchR/), covering basic and advanced applications.
+To support users at various levels of expertise, we provide extensive documentation on the hatchR website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.
 

### 2025-03-03 inst/manuscript/paper.Rmd
Introduction edits

@@
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
-Despite its growing adoption, applications of the effective value model remain largely confined to salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.
+Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.
 
@@
 To maximize accessibility, **hatchR** is available in two formats.
-The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
+The first is a R package, **hatchR**, distributed via CRAN [@r:ther], providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
 Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
-The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
+The second is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
 Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.

### 2025-03-03 inst/manuscript/paper.Rmd
Package overview edits

@@
 **hatchR** is designed primarily as a tool for predicting fish phenology.
-To maintain focus on this core function, we provide minimal build-in data validation and vizualization tools, as users are expected to understand and check their own data.
+To maintain focus on this core function, we provide minimal built-in data validation and vizualization tools, as users are expected to understand and check their own data.
 Given the diversity of potential data types, it is impractical to implement comprehensize validation checks.
-However, we include basic data-checking functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`).
-Intuitive functions are provided for users to apply models (either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function).
-Users can then apply these models to water temperature data (e.g.
-from a HOBO temp logger) to predict when hatching phenology will occur.
+However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`).
+Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function.
+Users can then apply these models to water temperature data (*e.g.,* from a HOBO temperature logger) to predict when hatching phenology will occur.
 This is accomplished with the `predict_phenology()` function.
@@
 ```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow.
-Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decisions choices as italicized text."}
+Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decision choices as italicized text."}
 knitr::include_graphics("./workflow.png")

### 2025-03-03 inst/manuscript/paper.Rmd
checking data edits

@@
 We demonstrate the utility three functions: `summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`).
-This dataset contains temperature readings taken every thirty minutes, and its structure (dimentions and first six rows) is shown below.
+This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below.
 
@@
 ```{r}
-#year_sim data dimensions
+#year_sim data dimensions (rows x columns)
 dim(year_sim)
@@
 # now a year's worth of single-day data
+# 365 days x two columns
 dim(year_sim_summ)

### 2025-03-03 inst/manuscript/paper.Rmd
Fitting models edits

@@
 
-To expand the applicability of the effective value approach beyond Salmonids, **hacthR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature.
+To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature.
 The function takes two input vectors--one for average incubation temperature (°C) and one for the number of days to a given phenological event--and estimates model parameters (*log~e~a* and *b*) using `stats::nls()`.
@@
 
-Next, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C.
+To demonstrate parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd).
 Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3).
@@
 
-Note the *R^2^* fit from the models below. You can see they generally preform well and are in line with values from model 2 of @beacham1990.
+Note the *R^2^* fit from the models below. You can see they generally preform well and are close to values from model 2 of @beacham1990, which fall between 0.95-0.99 range.
 

### 2025-03-03 inst/manuscript/paper.Rmd
predicting phenology edits 2

@@
 
-^2^Idaho Fish and Wildlife Conservation Office, U.S. Fish and Wildlife Service, Orofino, ID, USA
+^2^Clearwater Solutions, Lenore, ID, USA
 
@@
 ```{r, echo = TRUE}
+#model_select() to get sockeye model
 sockeye_hatch_mod <- model_select(
@@
 ```{r, warning=FALSE, echo = TRUE}
+#predict_phenology() with sockeye model and woody_island temperature data
 WI_hatch <- predict_phenology(
@@
   temperature = temp_c,
-  spawn.date = "1990-08-18",
+  spawn.date = "1990-08-18", #notice the character string for spawn date
   model = sockeye_hatch_mod
@@
 
-```{r, eval=FALSE, echo = TRUE}
-str(WI_hatch)
+```{r, echo = TRUE}
+summary(WI_hatch)
 ```
@@
 `WI_hatch$model_specs` -- Provides details about the model used for prediction, including whether it was retrieved from `model_select()` or generated using `fit_model()`.
-Most importantly, it contains the model expression (i.e., the formula) used for phenology predictions.
+Most importantly, it contains the model expression (*i.e.*, the formula) used for phenology predictions.
 

### 2025-03-03 inst/manuscript/paper.Rmd
Case Study 1 edits

@@
 
-# Case Study 1:
+# Case Study 1
 
-**Bull Trout in the Crooked River** OR **Predicting Emergence Timing for Management Actions**
+**Predicting Emergence Timing for Management Actions**
 
@@
 
-In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, where a key Bull Trout (*Salvelinus confluentus*) population exists.
+In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population.
 Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensative to sediment disturbance.
@@
 After confirming data integrity, users navigate to the `Model Phenology` window.
-For this case study, we use the preloaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus.
+For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus.
 The user then chooses multiple spawn dates using an interactive calendar.
-Here, we focus on September 30th as outlined in our management scenario.
+Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.
 
@@
 
-In this example, the model predicts that the last Bull Trout will emerge before the June 1st traget date.
-This suggests that the Fisheries Biologist can confidently approve the road work in the area without concern for sediment disturbance impacting fish developing in the gravel.
+In this example, the model predicts that the last Bull Trout will emerge before the June 1st road work target date.
+This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel.
 This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with necessary land-use activities.

### 2025-03-03 inst/manuscript/paper.Rmd
case study 2 edits

@@
 
-**Bull Trout in the Upper Columbia River Headwaters** OR **Large-scale Predictions of Bull Trout development timing**
+**Large Scale Predictions of Bull Trout Development Timing**
 
-For the second case study, we demonstrate a more complex, large-sclae application of **hatchR**, highlighting its full flexibility when applied programmatically in R.
+For the second case study, we demonstrate a more complex, large-scale application of **hatchR**, highlighting its full flexibility when applied programmatically in R.
 This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the upper Columbia River headwaters in Idaho.
 
-We use the `idaho` dataset from @isaak2018, which includes temperature data from these sites.
+We use the `idaho` dataset from @isaak2018, which includes temperature data for these sites.
 To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature \</= 13 °C, a known thermal threshold for Bull Trout spawning suitability.
@@
 
-To predict hatch timing across these sites, we:
+To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization.
+(For conciseness, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository for **hatchR**.) We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates (Early: September 1st, Peak: September 15th, Late: September 30th).
 
-1.
-Set up the necessary moels and data, using Bull Trout parameterization.
-(For conciseness, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository for **hatchR**.)
-
-2.  Apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn date:
-
--   Early: September 1st
--   Peak: September 15th
--   Late: September 30th
-
-By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing caries across different spawning habitats.
-The results of this analysis are presented in Figure 5, providing insights into how emergence timing might vary under different thermal regimes and across the species' geographic range.
+By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
+The results of this analysis are presented in Figure 5, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range.
 This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.

### 2025-03-03 inst/manuscript/paper.Rmd
Conclusion edits and knit

@@
 To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature.
-The function takes two input vectors--one for average incubation temperature (°C) and one for the number of days to a given phenological event--and estimates model parameters (*log~e~a* and *b*) using `stats::nls()`.
+The function takes two input vectors—one for average incubation temperature (°C) and one for the number of days to a given phenological event—and estimates model parameters (*log~e~a* and *b*) using `stats::nls()`.
 
@@
 It establishes a formal framework for applying effective value models from user-provided parameterizations.
-The software is available in two formats:
-
--   A fully customizable R package, ideal for complex and repetitive analyses.
--   A graphical user interface (GUI) for ease of use, designed for tasks that may only need to be run once or a few times.
+The software is available in two formats: 1) A fully customizable R package, ideal for complex and repetitive analyses and 2) a graphical user interface for ease of use, designed for tasks that may only need to be run once or a few times.
 
@@
 First, environmental stressors may alter developmental timing.
-While effective value models predict developmental timing based on temperature, studies have shown that stressful environmental conditions—such as low dissolved oxygen, altered pH, high salinity, pathogen exposure, or mechanical disturbance—can induce premature hatching or emergence (@quinn2018; @cowan2024).
+While effective value models predict developmental timing based on temperature, studies have shown that stressful environmental condition such as low dissolved oxygen, altered pH, high salinity, pathogen exposure, or mechanical disturbance can induce premature hatching or emergence [@quinn2018; @cowan2024].
 Users should consider how such factors may influence their predictions.
@@
 Second, developmental timing occurs as a distribution, not a fixed point.
-While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events (@mason1976).
+While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events [@mason1976].
 We encourage users to predict phenology using early, peak, and late thresholds (e.g., 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.
 
-Third, sensor-based temperature data may differ from actual egg temperatures.
-Water temperatures recorded by environmental sensors may not fully reflect thermal conditions in spawning microhabitats, where geomorphic factors influence temperature regimes (@geist2002).
+Third, sensor-based temperature data may differ from actual embryonic ambient temperatures.
+Water temperatures recorded by environmental sensors may not fully reflect thermal conditions in spawning microhabitats, where geomorphic factors influence temperature regimes [@geist2002].
 Users should consider how differences between measured and actual incubation temperatures may affect predictions.
@@
 
-To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations (@adelfio2024; @austin2019).
-However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by genetic variation, gene-environment interactions, and phylogenetic differences (@west-eberhard2003).
+To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations [@adelfio2024; @austin2019].
+However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by genetic variation, gene-environment interactions, and phylogenetic differences [@west-eberhard2003].
 
 For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes.
-Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada (@beacham1990), consistent with cogradient variation (@conover2009; @sparks2022).
+Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022].
 These findings highlight the importance of considering not only developmental time predictions but also how the underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.
 
-## Future directions OR Expanding the utility of **hatchR**
+## Expanding the utility of **hatchR**
 
 The models within **hatchR** can be customized in multiple ways beyond the examples provided.
-While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones, such as early embryonic stages (e.g., eye-up in salmonids; @velsen1980), initiation of pelagic-larval dispersal in marine fish, where temperature regulates the transition from demersal to pelagic behavior, or current-mediated dispersal in riverine species, where larvae begin downstream drift at temperature-dependent thresholds.
+While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones, such as early embryonic stages (*e.g.*, eye-up in salmonids; [@velsen1980]), initiation of pelagic-larval dispersal in marine fish, where temperature regulates the transition from demersal to pelagic behavior, or current-mediated dispersal in riverine species, where larvae begin downstream drift at temperature-dependent thresholds.
 
@@
 
-Finally, while hatchR was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power-law relationship with temperature.
+Finally, while hatchR was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power law relationship with temperature.
 Extending the effective value framework to these taxa could provide valuable insights into their life history timing under variable environmental conditions.
@@
 
-The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS.
+The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS.
 Any mention of trade names, products, or services does not imply an endorsement by the U.S.
-government, USFS, or USFWS.
-USFS and USFWS do not endorse any commercial products, services or enterprises.
+government or USFS.
+USFS does not endorse any commercial products, services or enterprises.
 
@@
 
-**hathcR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.
+**hatchR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.
 

### 2025-03-04 inst/manuscript/paper.Rmd
try to fix hanging HTML link in Case Study 1 (failed)
final edits to manuscript and then render

@@
                 temp_min = 0, # temp_min and max lines are 
-                              # user customizable
-                temp_max = 25)
+                temp_max = 25) # user customizable
 ```
@@
 These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach.
-For conciseness, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) on the GitHub project repository.
+For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository.
 
@@
 Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3).
-This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid system.
+This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
 
@@
 
-These model expressions are then applied in `predict_phenology()` to estimate the days to hatch and development:
+These model expressions are then implemented in `predict_phenology()` to estimate the days to hatch and development:
 
@@
 
-To demonstrate this case study, we use the **hatchR** graphical user interface available at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}.
+To demonstrate this case study, we use the **hatchR** graphical user interface available at  <https://elifelts.shinyapps.io/hatchR_demo/>.
 Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year).
 For this example, we use the `crooked_river` dataset, included in the Shiny app as a demo dataset.
-It can also be accessed directly at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.
+It can also be accessed directly at [https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/](#0){.uri}.
 
@@
 This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel.
-This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with necessary land-use activities.
+This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.
 
@@
 
-To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization.
-(For conciseness, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository for **hatchR**.) We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates (Early: September 1st, Peak: September 15th, Late: September 30th).
+To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization (for concision, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository or in the Predict fish phenology: nested article on **hatchR**'s website) We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates (Early: September 1st, Peak: September 15th, Late: September 30th).
 
@@
 To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations [@adelfio2024; @austin2019].
-However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by genetic variation, gene-environment interactions, and phylogenetic differences [@west-eberhard2003].
+However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by local adaptation, gene-environment interactions, and phylogenetic differences [@west-eberhard2003].
 
@@
 Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022].
-These findings highlight the importance of considering not only developmental time predictions but also how the underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.
+These findings highlight the importance of considering how developmental rates are keyedto specific environments but also how these underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.
 
@@
 The models within **hatchR** can be customized in multiple ways beyond the examples provided.
-While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones, such as early embryonic stages (*e.g.*, eye-up in salmonids; [@velsen1980]), initiation of pelagic-larval dispersal in marine fish, where temperature regulates the transition from demersal to pelagic behavior, or current-mediated dispersal in riverine species, where larvae begin downstream drift at temperature-dependent thresholds.
+While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones not reliant on exogenous feeding, such as early embryonic stages (*e.g.*, eye-up; [@velsen1980]), initiation or cessation of pelagic-larval dispersal, or current-mediated dispersal in riverine species, though not all cases may be as specifically tied to temperature as hatch and emergence.
 
@@
 
-**hatchR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.
+**hatchR** is fully open source and reproducible.
+Source code and data can be found at <https://github.com/bmait101/hatchR/>.
+The Rmardkdown document with all the code to reproduce the examples from this manuscript is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
+The The latest version will be archived upon acceptance of the manuscript.
 

### 2025-03-04 inst/manuscript/paper.Rmd
one last copy edit
expanding fit_model section

@@
 To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature.
-The function takes two input vectors—one for average incubation temperature (°C) and one for the number of days to a given phenological event—and estimates model parameters (*log~e~a* and *b*) using `stats::nls()`.
+The function takes two input vectors: average incubation temperature (°C) and the number of days to a given phenological event.
+A model is then fit to the data using `stats::nls()`, which performs nonlinear least squares regression to estimate the parameters *log~e~a* and *b*.
+Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to fit the nonlinear model.
 
 This approach allows users to generate models tailored to non-Salmonid species, provided they have experimental or field data linking development to temperature.
-However, users should be mindful that custom models may not generalize well beyond the temperature range of the data used to fit the model.
-Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, pending sufficient validation in the literature.
+However, users should be mindful of several factors, such as extrapolations risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
+Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, invertebrates, provided sufficient validation in the literature.
 
@@
 
-Once uploaded, **hatchR** automatically generates a vizual data check using `plot_check_temp()`.
+Once uploaded, **hatchR** automatically generates a visual data check using `plot_check_temp()`.
 After confirming data integrity, users navigate to the `Model Phenology` window.

### 2025-03-04 inst/manuscript/paper.Rmd
final copy edits and render 2

@@
 
-This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available.
+This effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations were readily available.
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
@@
 **hatchR** is designed primarily as a tool for predicting fish phenology.
-To maintain focus on this core function, we provide minimal built-in data validation and vizualization tools, as users are expected to understand and check their own data.
-Given the diversity of potential data types, it is impractical to implement comprehensize validation checks.
+To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data.
+Given the diversity of potential data types, it is impractical to implement comprehensive validation checks.
 However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`).
@@
 While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function.
-This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.
+This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond Salmonids.
 
@@
 To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically.
-These checks ensure data integroty before model application.
+These checks ensure data integrity before model application.
 
@@
 Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3).
-This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
+This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
 
@@
 To illustrate model selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**.
-Speficially, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
+Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
 
@@
 
-The output from `predict_phenology()` contains mutiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:
+The output from `predict_phenology()` contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:
 
@@
 In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population.
-Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensative to sediment disturbance.
+Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensitive to sediment disturbance.
 The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st.
@@
 
-The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walkthrough is available on the **hatchR** Shiny website.
+The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walk through is available on the **hatchR** Shiny website.
 
@@
 Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022].
-These findings highlight the importance of considering how developmental rates are keyedto specific environments but also how these underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.
+These findings highlight the importance of considering how developmental rates are keyed to specific environments but also how these underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.
 
@@
 
-We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature data sets.
+We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided helpful discussion about model development and temperature data sets.
 

### 2025-03-04 inst/manuscript/paper.Rmd
incorporate ms comments by @allisonswartz

@@
 However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on Salmonids.
-Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes.
+Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and stream temperature data.
 **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
@@
 
-As poikilothermic organisms, fish development and growth are closely tied to ambient environment.
+As poikilothermic organisms, fish development and growth are closely tied to ambient environmental conditions.
 This strong relationship has enabled researchers to generate statistical models that predict developmental phenology with high accuracy.
 Historically, these models were formulated in aquaculture settings under the assumption of constant temperature throughout development [@beacham1990; @mcphail1979; @alderdice1978], limiting their applicability to wild populations.
-However, @sparks2019 reformulated this approach into an "Effective Value model", which instead uses daily average temperature after spawning, predicting hatch or emergence when cumulative effective values reach a threshold of one.
+However, @sparks2019 reformulated this approach into an "Effective Value model" that instead uses daily average temperatures after spawning, which predicts hatch or emergence timing when cumulative effective values reach a threshold of one.
 
-This effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations were readily available.
+The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations were readily available.
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
@@
 **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017.
-Users can also develop custom models using their own or published temperature and phenological data within the effective value framework.
+Users can also develop custom models using their own or published temperature and phenological data within the effective value framework for application to unparameterized species or populations.
 
 To maximize accessibility, **hatchR** is available in two formats.
-The first is a R package, **hatchR**, distributed via CRAN [@r:ther], providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
+The first is a R package, **hatchR**, distributed via The Comprehensive R Archive Network (CRAN), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
 Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
@@
 This is accomplished with the `predict_phenology()` function.
-The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external visualizations.
+The R package provides user-friendly example vignettes that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs.
+In contrast, the Shiny application includes a default output plot and an option to download results for external visualizations.
 For a high-level overview of **hatchR**'s applications, see Figure 1.
@@
 
-This framework as the foundation for phenological models in **hatchR**.
+This framework is the foundation for phenological models in **hatchR**.
 The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
@@
 Water temperature datasets collected in the field typically fall into two categories: 1) summarized daily data, where mean daily temperatures are pre-computed, or 2) raw high-frequency data, such as those recorded by HOBO TidbiT loggers, which require summary into mean daily temperatures before use.
-Additionally, new statistical models, such as @siegel2023, could also be implemented into this framework.
+Additionally, new statistical models that predict daily stream temperatures, such as those developed by @siegel2023, could also be implemented into this framework.
 
@@
 
-Since **hatchR** does not automatically handle missing data, users must check for gaps or errors before running analyses.
-The package will function with missing values, but gaps in the dataset may affect predictions.
+**hatchR** will function with missing values, but gaps in the dataset may affect predictions.
+Since the package does not automatically handle missing data, users must check for gaps or errors before running analyses with the provided check functions (see section *Checking Data*).
 **hatchR** supports temperatures as low as 0 °C, though such values yield extremely small effective values, potentially extending hatch or emergence timing to a year or more.
@@
 
-This approach allows users to generate models tailored to non-Salmonid species, provided they have experimental or field data linking development to temperature.
-However, users should be mindful of several factors, such as extrapolations risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
+This approach allows users to generate models tailored to non-Salmonid species, or even different populations within species, provided they have experimental or field data linking development to temperature.
+However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
 Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, invertebrates, provided sufficient validation in the literature.
@@
 
-To demonstrate parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd).
-Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3).
+To demonstrate custom parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd).
+Using this dataset, we apply the custom models for each species to predict hatch timing and total developmental duration (Figure 3).
 This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
@@
 ```{r, echo = TRUE}
-
-### smallmouth mod
+# smallmouth model
 smallmouth <- matrix(NA, 10, 2) |> data.frame()
@@
                      development_type = "hatch")
-
 ```
@@
 ```{r, echo=FALSE}
-
-### catfish mod
+# catfish model
 catfish <- matrix(NA, 3, 2) |> data.frame()
@@
 
-### lake sturgeon mod
+# lake sturgeon model
 sturgeon <-  matrix(NA, 7, 2) |> data.frame()
@@
                           development_type = "hatch")
-
 ```
@@
 ```{r, echo = TRUE}
-#model fits
+# model fits
 smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared
@@
 
-First, we obtain model expression for both hatch and emergence using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
+First, we obtain the model expression for hatching using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
 
 ```{r, echo = TRUE}
-#model_select() to get sockeye model
+# model_select() to get sockeye model
 sockeye_hatch_mod <- model_select(
@@
 ```{r, warning=FALSE, echo = TRUE}
-#predict_phenology() with sockeye model and woody_island temperature data
+# predict_phenology() with sockeye model and woody_island temperature data
 WI_hatch <- predict_phenology(
@@
   temperature = temp_c,
-  spawn.date = "1990-08-18", #notice the character string for spawn date
+  spawn.date = "1990-08-18",  # notice the character string for spawn date
   model = sockeye_hatch_mod
@@
 `WI_hatch$ef_table` -- An *n* x 5 tibble (*n* = number of days to hatch or emerge), containing a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
-This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discuss below.
+This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discussed below.
 
@@
 For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus.
-The user then chooses multiple spawn dates using an interactive calendar.
+The user then may chooses multiple spawn dates using an interactive calendar.
 Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.
@@
 
-We use the `idaho` dataset from @isaak2018, which includes temperature data for these sites.
+We use the `idaho` dataset from @isaak2018, which includes four years of temperature data (2011-2014) for these sites.
 To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature \</= 13 °C, a known thermal threshold for Bull Trout spawning suitability.
@@
 ```{r, echo = FALSE}
-
 bull_trout_sites <- idaho |> 
@@
 
-To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization (for concision, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository or in the Predict fish phenology: nested article on **hatchR**'s website) We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates (Early: September 1st, Peak: September 15th, Late: September 30th).
+To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization (for concision, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository or in the Predict fish phenology: nested article on **hatchR**'s website).
+We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
 
 By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
-The results of this analysis are presented in Figure 5, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range.
+The results of this analysis are presented in Figure 5, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range and across years.
 This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
@@
 
-Both versions allow users to import data, perform basic data checks, and apply either pre-existing salmonid model parameterizations or generate custom models specific to other species or populations.
+Both versions allow users to import data, perform basic data checks, perform basic visualizations, and apply either pre-existing Salmonid model parameterizations or generate custom models specific to other species or populations.
 To support users at various levels of expertise, we provide extensive documentation on the hatchR website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.
@@
 However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by local adaptation, gene-environment interactions, and phylogenetic differences [@west-eberhard2003].
-
 For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes.
@@
 Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings.
-We present foundational applications of **hatchR**, with additional use cases and implementation guides available on the software's website.
+We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website.
 The software is designed for both applied and basic research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app.

### 2025-03-06 inst/manuscript/paper.Rmd
incoporate ms comments by @pfrater

@@
 
-Morgan M. Sparks^1,4^, Eli A. Felts^2^, Allison G. Swartz^3^, Bryan M. Maitland^1^
+Morgan M. Sparks^1,5^, Bryan M. Maitland^1^, Eli A. Felts^2^, Allison G. Swartz^3^, Paul Frater^4^
 
@@
 
-^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov
+^4^Wisconsin Department of Natural Resources, Madison, WI, USA
+
+^5^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov
 
@@
 
-The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations were readily available.
+The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations are readily available.
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
@@
 To maximize accessibility, **hatchR** is available in two formats.
-The first is a R package, **hatchR**, distributed via The Comprehensive R Archive Network (CRAN), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
+The first is an R package, **hatchR**, distributed via The Comprehensive R Archive Network (CRAN), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes.
 Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
-The second is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
-Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.
+The second format is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
+Below, we provide an overview of **hatchR** and present case studies demonstrating its application in research and management.
 
@@
 $$
-\mathrm{E_i} = \frac{1}{\exp(\log_e(a) - \log_e(T_i - b)}
+\mathrm{E_i} = \frac{1}{\exp(\log_e(a) - \log_e(T_i - b))}
 $$
@@
 
-**hatchR** assumes input data consists if at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
+**hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
 Other columns may be present, but columns names should not include spaces.
@@
 
-We demonstrate the utility three functions: `summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`).
+Here, we demonstrate the utility three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`).
 This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below.
@@
 These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach.
-For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository.
+For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).
 
@@
 Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year).
-For this example, we use the `crooked_river` dataset, included in the Shiny app as a demo dataset.
-It can also be accessed directly at [https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/](#0){.uri}.
+For this example, we use the `crooked_river` dataset, which is included in **hatchR**.
 
@@
 Both the prediction table and plot can be downloaded directly from their respective tabs.
-
-The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walk through is available on the **hatchR** Shiny website.
+The full process is demonstrated through a detailed walkthrough on the **hatchR** Shiny website.
 
@@
 We use the `idaho` dataset from @isaak2018, which includes four years of temperature data (2011-2014) for these sites.
-To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature \</= 13 °C, a known thermal threshold for Bull Trout spawning suitability.
-The filtering process reduced the dataset to 139 sites potential spawning sites.
+To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature ≤ 13 °C, a known thermal threshold for Bull Trout spawning suitability.
+The filtering process reduced the dataset to 139 potential spawning sites.
 
@@
 
-With **hatchR**, we present a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes.
+**hatchR** is a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes.
 It establishes a formal framework for applying effective value models from user-provided parameterizations.
@@
 
-Finally, while hatchR was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power law relationship with temperature.
+Finally, while **hatchR** was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power law relationship with temperature.
 Extending the effective value framework to these taxa could provide valuable insights into their life history timing under variable environmental conditions.
+We encourage users to submit fitted models using this framework as a pull request to **hatchR**'s Github repository provided that data used in the models are published.
 
@@
 We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website.
-The software is designed for both applied and basic research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app.
+The software is designed for both applied and fundamental research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app.
 We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.

### 2025-03-06 inst/manuscript/paper.Rmd
author list
remove unicode

@@
 
-Morgan M. Sparks^1,5^, Bryan M. Maitland^1^, Eli A. Felts^2^, Allison G. Swartz^3^, Paul Frater^4^
+Morgan M. Sparks^1,5,6^, Bryan M. Maitland^1,6^, Eli A. Felts^2^, Allison G. Swartz^3^, Paul Frater^4^
 
@@
 ^5^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov
+
+^6^Dual first authors
 
@@
 We use the `idaho` dataset from @isaak2018, which includes four years of temperature data (2011-2014) for these sites.
-To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature ≤ 13 °C, a known thermal threshold for Bull Trout spawning suitability.
+To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature at or less than 13 °C, a known thermal threshold for Bull Trout spawning suitability.
 The filtering process reduced the dataset to 139 potential spawning sites.

### 2025-03-06 inst/manuscript/paper.Rmd
code revisions

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
 ```
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
-```
-
 # Package Overview
@@
 
-```{r kabletable, echo=FALSE }
-df = data.frame(date = c("2000-01-01",
+```{r kabletable, echo=FALSE}
+df <- data.frame(
+  date = c(
+    "2000-01-01",
     "...",
@@
     "...",
-                         "2000-12-31"),
-                temperature= c(2.51,
+    "2000-12-31"
+  ),
+  temperature = c(
+    2.51,
     "...",
@@
     3.13
-                        ))
+  )
+)
 
-# kable can alse be used for creating tables
 knitr::kable(df,
-             booktabs=TRUE,
-             caption = "Example temperature data for use in hatchR.")
+  booktabs = TRUE,
+  caption = "Example temperature data for use in hatchR."
+)
 ```
@@
 Here, we demonstrate the utility three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`).
-This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below.
+This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:
 
-```{r, echo = FALSE}
+```{r fakedata, echo=FALSE}
 library(lubridate)
+
 # create date object for a year with 30 min reading intervals
-dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
-             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)
+dates <- seq(
+  from = ymd_hms("2000-07-01 00:00:00"),
+  to = ymd_hms("2001-06-30 23:59:59"),
+  length.out = 17568
+)
 
@@
 
-#random seed
+# random seed
 set.seed(123)
@@
 # in year_sim
-year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
+year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) |> 
   abs()
-
 ```
 
-```{r}
-#year_sim data dimensions (rows x columns)
+```{r show-fakedata}
+# year_sim object dimensions (rows and columns)
 dim(year_sim)
-#fist 6 rows of year_sim
+# fist 6 rows of year_sim
 head(year_sim)
@@
 
-First, we recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 2).
+We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 2).
+This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures.
+In this example, we set the minimum and maximum thresholds to 0 and 25 °C, respectively.
 
-```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set.
+```{r plot-check-1, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set.
 Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
@@
 
-In this case, no obvious outliers are present, but since each day contains 48 records, the data must be summarized to daily mean temperature using `summarize_temp()`.
+In this case, no obvious outliers are present.
+However, each day contains 48 records, so the data must be summarized to daily mean temperature using `summarize_temp()`.
 After summarization, `check_continuous()` should be used to identify any missing days.
@@
 
-```{r, echo = TRUE}
+```{r sum-temps, echo = TRUE}
 # summarize
-year_sim_summ <- summarize_temp(data = year_sim,
+year_sim_summ <- summarize_temp(
+  data = year_sim,
   dates = date,
-                                temperature = temp)
+  temperature = temp
+)
 
-# now a year's worth of single-day data
-# 365 days x two columns
+# check dimensions of summarized object
 dim(year_sim_summ)
 
-# check continuous (no errors)
-check_continuous(data = year_sim_summ,
-                 dates = date)
+# check if continuous
+check_continuous(
+  data = year_sim_summ,
+  dates = date
+)
 
 # we can demonstrate an error by removing Oct. 8 (100th day)
-check_continuous(data = year_sim_summ[-100,],
-                 dates = date)
-
+check_continuous(
+  data = year_sim_summ[-100, ],
+  dates = date
+)
 ```
@@
 
-```{r, echo=FALSE}
+```{r make-temps, echo=FALSE}
 library(dplyr)
-###  make temp regime
 set.seed(123)
+
 # create random temps and corresponding dates
-temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
-dates_sim <-  seq(from = ymd("2000-07-01"),
-             to = ymd("2000-07-31"), length.out = 30)
+temps_sim <- sort(rnorm(n = 30, mean = 16, sd = 1), decreasing = FALSE)
+dates_sim <- seq(
+  from = ymd("2000-07-01"),
+  to = ymd("2000-07-31"), length.out = 30
+)
 
 data_sim <- matrix(NA, 30, 2) |> data.frame()
-data_sim[,1] <- temps_sim
-data_sim[,2] <- dates_sim
+data_sim[, 1] <- temps_sim
+data_sim[, 2] <- dates_sim
 
@@
 
-```{r, echo = TRUE}
-# smallmouth model
+```{r, echo=TRUE}
+# Smallmouth Bass model
 smallmouth <- matrix(NA, 10, 2) |> data.frame()
@@
 smallmouth <- smallmouth |>
-  mutate(days = ceiling(hours/24),
-         temp_C = (temp_F -32) * (5/9))
+  mutate(
+    days = ceiling(hours / 24),
+    temp_C = (temp_F - 32) * (5 / 9)
+  )
 
-# model object for smallmouth bass 
-smb_mod <- fit_model(temp = smallmouth$temp_C,
+# model object for Smallmouth Bass
+smb_mod <- fit_model(
+  temp = smallmouth$temp_C,
   days = smallmouth$days,
   species = "smb",
-                     development_type = "hatch")
+  development_type = "hatch"
+)
 ```
@@
 ```{r, echo=FALSE}
-# catfish model
+# Catfish model
 catfish <- matrix(NA, 3, 2) |> data.frame()
@@
 
-# lake sturgeon model
+# Lake Sturgeon model
 sturgeon <-  matrix(NA, 7, 2) |> data.frame()
@@
 
-```{r, echo = TRUE}
-# model fits
+```{r, echo=TRUE}
+# R^2 values from model fits
 smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
-### predict_phenology
-
-#smallmouth bass
-smb_res <- predict_phenology(data = data_sim,
+```{r pred-pheno-1, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
+# Smallmouth Bass
+smb_res <- predict_phenology(
+  data = data_sim,
   dates = date_sim,
@@
   spawn.date = "2000-07-01",
-                  model = smb_mod$expression)
+  model = smb_mod$expression
+)
 
-# catfish
-catfish_res <- predict_phenology(data = data_sim,
+# Catfish
+catfish_res <- predict_phenology(
+  data = data_sim,
   dates = date_sim,
@@
   spawn.date = "2000-07-01",
-                  model = cat_mod$expression)
+  model = cat_mod$expression
+)
 
-# sturgeon
+# Sturgeon
 # note that 16 C is pretty far out of range of temps for model fit, not best practice
-sturgeon_res <- predict_phenology(data = data_sim,
+sturgeon_res <- predict_phenology(
+  data = data_sim,
   dates = date_sim,
@@
   spawn.date = "2000-07-01",
-                  model = sturgeon_mod$expression)
+  model = sturgeon_mod$expression
+)
 
-# summary for all species
+# Summary for all species
 all_res <- data.frame(matrix(NA, 3, 2))
@@
 
-all_res$start <- c(catfish_res$dev.period$start, 
+all_res$start <- c(
+  catfish_res$dev.period$start,
   smb_res$dev.period$start,
-                   sturgeon_res$dev.period$start)
+  sturgeon_res$dev.period$start
+)
 
-all_res$stop <- c(catfish_res$dev.period$stop,
+all_res$stop <- c(
+  catfish_res$dev.period$stop,
   smb_res$dev.period$stop,
-                  sturgeon_res$dev.period$stop)
+  sturgeon_res$dev.period$stop
+)
 
 all_res <- all_res |>
-  mutate(days = ceiling(stop-start),
-         index = c(17,16.5,16))
+  mutate(
+    days = ceiling(stop - start),
+    index = c(17, 16.5, 16)
+  )
 
@@
 ggplot() +
-  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim )) + 
-  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim )) +
-  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax =index-.35, ymin = index-.5, fill = Species)) +
-  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index -0.425), label = days)) +
+  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
+  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim)) +
+  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax = index - .35, ymin = index - .5, fill = Species)) +
+  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index - 0.425), label = days)) +
   labs(x = "Date", y = "Temperature (°C)") +
@@
 
-```{r, echo = TRUE}
-# model_select() to get sockeye model
+```{r, echo=TRUE}
+# model_select() to get sockeye hatch model
 sockeye_hatch_mod <- model_select(
@@
 
-```{r, warning=FALSE, echo = TRUE}
-# predict_phenology() with sockeye model and woody_island temperature data
+```{r, warning=FALSE, echo=TRUE}
+# predict_phenology() with sockeye hatch model and woody_island temperature data
 WI_hatch <- predict_phenology(
@@
 
-```{r, echo = TRUE}
+```{r, echo=TRUE}
 WI_hatch$days_to_develop
@@
 
-```{r, echo = TRUE}
+```{r, echo=TRUE}
 summary(WI_hatch)
@@
 ```{r, eval=FALSE, echo=FALSE}
-
-### read in EFS data
-
+# read in EFS data
 EFS_data <- crooked_river
 
-### view bull trout models and select model
-
+# view bull trout models and select model
 model_table |>
@@
 # we care about when fish are out of gravel, so select emerge mod
-bt_emerge_mod <- model_select(author = "Austin et al. 2017",
+bt_emerge_mod <- model_select(
+  author = "Austin et al. 2017",
   species = "bull trout",
   model = "MM",
-                            development_type = "emerge")
+  development_type = "emerge"
+)
 
 # predict emergence time using Sept. 30 as the spawn date
-bt_emergence <- predict_phenology(data = EFS_data,
+bt_emergence <- predict_phenology(
+  data = EFS_data,
   dates = date,
@@
   spawn.date = "2011-09-30",
-                                  model = bt_emerge_mod)
+  model = bt_emerge_mod
+)
 
@@
 bt_emergence$dev.period$stop
-
 ```
@@
 
-```{r, echo = FALSE}
+```{r, echo=FALSE}
 bull_trout_sites <- idaho |>
-  mutate(month = month(date)) |> #make a month column (numeric)
+  mutate(month = month(date)) |>  # make a month column (numeric)
   filter(month == 8) |>  # filter out Aug.
@@
 
-```{r, echo = TRUE}
+```{r, echo=TRUE}
 hatch_res <- isaak_summ_bt |>
@@
   mutate(spawn_time = factor(
-    spawn_time, levels = c("Late", "Peak", "Early"), 
-    ordered = TRUE)
-    )
+    spawn_time,
+    levels = c("Late", "Peak", "Early"),
+    ordered = TRUE
+  ))
 
@@
   group_by(year, spawn_time, start) |>
-  summarise(mean_hatch = ceiling(mean(days_to_hatch)),
+  summarise(
+    mean_hatch = ceiling(mean(days_to_hatch)),
     hatch_5 = ceiling(quantile(days_to_hatch, probs = 0.05)),
-            hatch_95 = ceiling(quantile(days_to_hatch, probs = 0.95)))
+    hatch_95 = ceiling(quantile(days_to_hatch, probs = 0.95))
+  )
 
@@
     color = spawn_time
-    )) +
+  )
+) +
   geom_density_ridges(alpha = 0.9) +
-    facet_wrap( ~ year, ncol = 1) +
+  facet_wrap(~year, ncol = 1) +
   scale_fill_brewer(palette = "Blues", direction = 1) +

### 2025-03-06 inst/manuscript/paper.Rmd
add new figure 2

@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
+library(tidyverse); library(hatchR); library(ggdist); library(ggridges); library(patchwork)
 ```
@@
 $$
+
+To demonstrate this approach, we use European Grayling (*Thymallus thymallus*) that were raised in five experimental average temperatures ranging from 2-14 °C [@quinn2018].
+We fit a developmental model to those experimental data (Figure 2A), reparameterize that model into an effective value model, and predict effective values for daily average temperatures ranging from 2-14 °C (Figure 2B).
+
+```{r, fig.cap= "Effective value model fit and application for European Grayling.
+Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line).
+Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C.
+The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference.",fig.height= 8, fig.width=4,  fig.align = 'center', echo =FALSE}
+
+grayling <- matrix(NA, 5, 2) |> tibble::tibble()
+colnames(grayling) <- c("temp", "days")
+
+grayling$temp <- c(2,5,8,11,14)
+grayling$days <- c(106,51,29,19,13)
+
+grayling_mod <- hatchR::fit_model(temp = grayling$temp,
+                                  days = grayling$days,
+                                  species = "grayling",
+                                  development_type = "hatch")
+
+temps <- c(2:14)
+
+ef_vals <- NULL
+for (x in temps){
+  ef <- eval(parse(text = grayling_mod$expression$ expression))
+  ef_vals <- rbind(ef_vals, ef)
+}
+
+grayling_ef <- matrix(NA, 13, 2) |> tibble::tibble()
+colnames(grayling_ef) <- c("temp", "ef")
+grayling_ef$temp <- temps
+grayling_ef$ef <- ef_vals[,1]
+
+p1 <- grayling_mod$pred_plot +
+  labs(x = "Developmental average temperature (°C)", y = "Days to hatch") +
+  theme_classic()
+
+data_1 <- tibble(t = c(0:14), e = seq(0, 0.14, by =0.01))
+
+p2 <- grayling_ef |>
+  ggplot() +
+  geom_point(aes(x = temp, y = ef)) +
+  geom_line(aes(x = temp, y = ef)) +
+  geom_line(data = data_1, aes(x = t, y = e), linetype = "dashed") +
+  #geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
+  labs(x = "Daily average temperature (°C)", y = "Effective value") +
+  theme_classic()
+
+p1/p2 + plot_annotation(tag_levels = 'A')
+
+```
 
@@
 ```{r plot-check-1, echo = TRUE,  fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set.
-Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height= 4, fig.width=6.5}
 plot_check_temp(data = year_sim,

### 2025-03-06 README.Rmd
update getting help
add CRAN badges

@@
 <!-- badges: start -->
+[![CRAN status](https://www.r-pkg.org/badges/version/hatchR)](https://CRAN.R-project.org/package=hatchR)
+[![cran checks](https://badges.cranchecks.info/worst/hatchR.svg)](https://cran.r-project.org/web/checks/check_results_hatchR.html)
 [![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
+[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
 <!-- badges: end -->
@@
 1.  Open a GitHub issue [link here](https://github.com/bmait101/hatchR/issues).
-2.  Email the **hatchR** support team ([Morgan Sparks](morgan.sparks@usda.gov), [Bryan Maitland](bryan.maitland@usda.gov))
-3.  Post on a support website like Stack Overflow or Cross Validated.
+2.  Post on a support website like Stack Overflow or Cross Validated.
 

### 2025-03-06 inst/manuscript/paper.Rmd
add in new figure 2 and corresponding paragraph

@@
 
-Here, we demonstrate the utility three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`).
+Here, we demonstrate the utility three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long dataset (`year_sim`).
 This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:
@@
 
-```{r plot-check-1, echo = TRUE,  fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set.
+```{r plot-check-1, echo = TRUE,  fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
 Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height= 4, fig.width=6.5}
@@
 
-```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` data set."}
+```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
 plot_phenology(WI_hatch)

### 2025-03-06 README.Rmd
Fix install instructions (eval=FALSE)

--- 2025-03-03 README.Rmd
@@
 [![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
+[![CRAN status](https://www.r-pkg.org/badges/version/hatchR)](https://CRAN.R-project.org/package=hatchR)
+[![cran checks](https://badges.cranchecks.info/worst/hatchR.svg)](https://cran.r-project.org/web/checks/check_results_hatchR.html)
+[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
 <!-- badges: end -->
@@
 
-Install and load the most recent approved version from CRAN by running:
+Install the released version from CRAN:
 
-```{r}
-# install the most recent approved version from CRAN
+```{r, eval = FALSE}
 install.packages("hatchR")
@@
 
-Install the development version of **hatchR** from [GitHub](https://github.com/) by running:
+Install the development version of **hatchR** from GitHub:
 
-```{r}
-# install the most recent version from GitHub
+```{r, eval = FALSE}
+# install.packages("pak")
 pak::pak("bmait101/hatchR")

### 2025-03-06 inst/manuscript/paper.Rmd
WIP

@@
 
-```{r, fig.cap= "Effective value model fit and application for European Grayling.
+```{r, echo = FALSE, fig.height = 8, fig.width = 4,  fig.align = 'center', fig.cap = "Effective value model fit and application for European Grayling.
 Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line).
 Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C.
-The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference.",fig.height= 8, fig.width=4,  fig.align = 'center', echo =FALSE}
-
+The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference."}
 grayling <- matrix(NA, 5, 2) |> tibble::tibble()
@@
 
-grayling$temp <- c(2,5,8,11,14)
-grayling$days <- c(106,51,29,19,13)
+grayling$temp <- c(2, 5, 8, 11, 14)
+grayling$days <- c(106, 51, 29, 19, 13)
 
-grayling_mod <- hatchR::fit_model(temp = grayling$temp,
+grayling_mod <- hatchR::fit_model(
+  temp = grayling$temp,
   days = grayling$days,
   species = "grayling",
-                                  development_type = "hatch")
+  development_type = "hatch"
+)
 
@@
 ef_vals <- NULL
-for (x in temps){
+for (x in temps) {
   ef <- eval(parse(text = grayling_mod$expression$ expression))
@@
 grayling_ef$temp <- temps
-grayling_ef$ef <- ef_vals[,1]
+grayling_ef$ef <- ef_vals[, 1]
 
@@
 
-data_1 <- tibble(t = c(0:14), e = seq(0, 0.14, by =0.01))
+data_1 <- tibble(t = c(0:14), e = seq(0, 0.14, by = 0.01))
 
@@
   geom_line(data = data_1, aes(x = t, y = e), linetype = "dashed") +
-  #geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
+  # geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
   labs(x = "Daily average temperature (°C)", y = "Effective value") +
@@
 
-p1/p2 + plot_annotation(tag_levels = 'A')
-
+p1 / p2 + plot_annotation(tag_levels = "A")
 ```

### 2025-03-06 inst/manuscript/paper.Rmd
Copy edits

@@
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
-Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.
+Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial, recreational, and cultural importance of these species.
 

### 2025-03-07 inst/manuscript/paper.Rmd
Incorporating @efelts-usfws copy edits

@@
 Understanding the timing of key life history events is essential for effective fish conservation and management.
-Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings.
+Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult to obtain in natural settings.
 Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records.
@@
 
-Here, we demonstrate the utility three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long dataset (`year_sim`).
+Here, we demonstrate the utility of three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long dataset (`year_sim`).
 This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:
@@
 
-```{r plot-check-1, echo = TRUE,  fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
-Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height= 4, fig.width=6.5}
+```{r plot-check-1, echo = TRUE,  fig.cap = "Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 4, fig.width = 6.5}
 plot_check_temp(data = year_sim,
@@
                 temperature = temp,
-                temp_min = 0, # temp_min and max lines are 
-                temp_max = 25) # user customizable
+                temp_min = 0,
+                temp_max = 25)
 ```
@@
 
-This approach allows users to generate models tailored to non-Salmonid species, or even different populations within species, provided they have experimental or field data linking development to temperature.
+This approach allows users to generate models tailored to any species of fish, or even different populations within species, provided they have experimental or field data linking development to temperature.
 However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
@@
 
-The returned object provides key outputs, including days to hatch and the full developmental period, allowing us to assess phenological patterns under the recorded thermal conditions:
+The returned object provides key outputs, including days to hatch and the full development period, allowing us to assess phenological patterns under the recorded thermal conditions:
 
@@
 In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population.
-Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensitive to sediment disturbance.
+Bull Trout, a federally threatened species under the Endangered Species Act [@usfws_bt_1999], are particularly sensitive to sediment disturbance [@nolfi2024].
 The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st.
@@
 
-In this example, the model predicts that the last Bull Trout will emerge before the June 1st road work target date.
+In this example, the model predicts that the last Bull Trout will emerge on April 21, before the June 1st road work target date.
 This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel.

### 2025-03-07 inst/manuscript/paper.Rmd
Fix typo in Case Study 1

@@
 
-```{r, eval=FALSE, echo=FALSE}
+```{r bt-emerge-crooked, eval=FALSE, echo=FALSE}
 # read in EFS data
@@
 bt_emerge_mod <- model_select(
-  author = "Austin et al. 2017",
+  author = "Austin et al. 2019",
   species = "bull trout",
@@
 
-In this example, the model predicts that the last Bull Trout will emerge on April 21, before the June 1st road work target date.
+In this example, the model predicts that the last Bull Trout will emerge on May 1, before the road work target date (June 1).
 This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel.

### 2025-03-07 inst/manuscript/paper.Rmd
Reorder ms sections for better flow

@@
 
-## hatchR: A toolset to predict when fish hatch and emerge
+# hatchR: A toolset to predict when fish hatch and emerge
 
@@
 
-# Package Overview
+# **hatchR** Overview
 
-**hatchR** is designed primarily as a tool for predicting fish phenology.
+**hatchR** is designed primarily as a tool for predicting early life history phenology in wild fishes.
 To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data.
@@
 Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function.
-Users can then apply these models to water temperature data (*e.g.,* from a HOBO temperature logger) to predict when hatching phenology will occur.
+Users can then apply these models to water temperature data (*e.g.,* from an in-stream temperature logger) to predict when hatching phenology will occur.
 This is accomplished with the `predict_phenology()` function.
@@
 ```{r show-fakedata}
-# year_sim object dimensions (rows and columns)
+# simulated object (year_sim) dimensions (rows and columns)
 dim(year_sim)
@@
 
-We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 2).
+We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 3).
 This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures.
@@
 Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 4, fig.width = 6.5}
+# visual check of temperature data
 plot_check_temp(data = year_sim,
@@
 ```{r sum-temps, echo = TRUE}
-# summarize
+# summarize temperature data
 year_sim_summ <- summarize_temp(
@@
 
-# check dimensions of summarized object
+# check dimensions of summarized data
 dim(year_sim_summ)
@@
 
-## Model Selection
+# **hatchR** Workflow
 
-Users can select from existing Salmonid models in `model_table` or generate custom models using `fit_model()` in both the R and Shiny deployments of **hatchR**.
+Users can select from published model parameterizations contained in `model_table`, or generate custom model parameterizations using `fit_model()`, in both the R and Shiny deployments of **hatchR**.
 The models in `model_table` are included because their parameterizations are well-documented in the literature, though they are currently limited to Pacific Salmon and Bull Trout (see @quinn2018 pg.
 183, for additional Salmonid models).
-To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C.
+To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C, requiring users to carefully asses whether their parameterized models are appropriate for the temperature ranges in their datasets.
 
-Custom models, by contrast, often have narrower parameterization ranges.
-To prevent misapplication, we exclude other model parameterization from `model_table`, requiring users to carefully asses whether their parameterized models are appropriate for the temperature ranges in their datasets.
+To illustrate parameter selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**.
+Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
+
+First, we obtain the model expression for hatching using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
+
+```{r, echo=TRUE}
+# use model_select() to get sockeye hatch model
+sockeye_hatch_mod <- model_select(
+  author = "Beacham and Murray 1990", 
+  species = "sockeye", 
+  model = 2, 
+  development_type = "hatch"
+  )
+```
+
+The selected model expressions are then implemented using `predict_phenology()` to estimate the days to hatch (or emergence):
+
+```{r, warning=FALSE, echo=TRUE}
+# use predict_phenology() to predict sockeye hatch
+WI_hatch <- predict_phenology(
+  data = woody_island,        # temperature data
+  dates = date,               # date column
+  temperature = temp_c,       # temperature column
+  spawn.date = "1990-08-18",  # spawning date
+  model = sockeye_hatch_mod   # model expression
+  )
+```
+
+The returned object provides several outputs, including days to hatch and the full development period, allowing us to assess phenological patterns under the recorded thermal conditions:
+
+```{r, echo=TRUE}
+WI_hatch$days_to_develop; WI_hatch$dev.period
+```
+
+## Understanding your results
+
+The output from `predict_phenology()` contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:
+
+```{r, echo=TRUE}
+summary(WI_hatch)
+```
+
+`WI_hatch$days_to_develop` -- Returns the predicted number of days required for development.
+
+`WI_hatch$dev.period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
+
+`WI_hatch$ef_table` -- An *n* x 5 tibble (*n* = number of days to hatch or emerge), containing a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
+This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discussed below.
+
+`WI_hatch$model_specs` -- Provides details about the model used for prediction, including whether it was retrieved from `model_select()` or generated using `fit_model()`.
+Most importantly, it contains the model expression (*i.e.*, the formula) used for phenology predictions.
+
+## Plotting phenology
+
+**hatchR** includes a built in function, `plot_phenology()`, for visualizing phenology predictions (Figure 4).
+This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span.
+Be default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles.
+However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results.
+This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.
+
+```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
+plot_phenology(WI_hatch)
+```
+
+## Fitting models for other fishes
 
@@
 However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
-Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, invertebrates, provided sufficient validation in the literature.
+Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.
 
-### Fitting models for other fishes
+To demonstrate how `fit_model()` may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
+These species were selected due to their common use in aquaculture and sport fisheries.
+For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).
 
-We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`.
-To showcase its utility, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
-These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach.
-For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).
-
-To demonstrate custom parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd).
-Using this dataset, we apply the custom models for each species to predict hatch timing and total developmental duration (Figure 3).
-This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
-
-```{r make-temps, echo=FALSE}
-library(dplyr)
-set.seed(123)
-
-# create random temps and corresponding dates
-temps_sim <- sort(rnorm(n = 30, mean = 16, sd = 1), decreasing = FALSE)
-dates_sim <- seq(
-  from = ymd("2000-07-01"),
-  to = ymd("2000-07-31"), length.out = 30
-)
-
-data_sim <- matrix(NA, 30, 2) |> data.frame()
-data_sim[, 1] <- temps_sim
-data_sim[, 2] <- dates_sim
-
-# change names so they aren't the same as the vector objects
-colnames(data_sim) <- c("temp_sim", "date_sim")
-```
-
-```{r, echo=TRUE}
-# Smallmouth Bass model
-smallmouth <- matrix(NA, 10, 2) |> data.frame()
-colnames(smallmouth) <- c("hours", "temp_F")
-smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
-smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)
-
-# change °F to °C and hours to days
-smallmouth <- smallmouth |>
-  mutate(
-    days = ceiling(hours / 24),
-    temp_C = (temp_F - 32) * (5 / 9)
-  )
-
-# model object for Smallmouth Bass
-smb_mod <- fit_model(
-  temp = smallmouth$temp_C,
-  days = smallmouth$days,
-  species = "smb",
-  development_type = "hatch"
-)
-```
-
-```{r, echo=FALSE}
+```{r cat-stg-fit, echo=FALSE}
 # Catfish model
@@
 
-Note the *R^2^* fit from the models below. You can see they generally preform well and are close to values from model 2 of @beacham1990, which fall between 0.95-0.99 range.
+```{r smb-fit, echo=TRUE}
+# Smallmouth Bass data (Webster 1948)
+smallmouth <- matrix(NA, 10, 2) |> data.frame()
+colnames(smallmouth) <- c("hours", "temp_F")
+smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
+smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)
 
-```{r, echo=TRUE}
+# change °F to °C and hours to days
+smallmouth <- smallmouth |>
+  mutate(
+    days = ceiling(hours / 24),
+    temp_C = (temp_F - 32) * (5 / 9)
+  )
+
+# estimate parameters for Smallmouth Bass using fit_model()
+smb_mod <- fit_model(
+  temp = smallmouth$temp_C,
+  days = smallmouth$days,
+  species = "smb",
+  development_type = "hatch"
+)
+```
+
+Note the *R^2^* fit from the models below. You can see they generally preform well and are close to values from model 2 of @beacham1990, which fall between 0.95 and 0.99.
+
+```{r r2-vals, echo=TRUE}
 # R^2 values from model fits
@@
 
-```{r pred-pheno-1, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
+Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (available in `paper.Rmd`).
+Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Base shown below):
+
+```{r make-temps, echo=FALSE, warning=FALSE, message=FALSE}
+library(dplyr)
+set.seed(123)
+
+# create random temps and corresponding dates
+temps_sim <- sort(rnorm(n = 30, mean = 16, sd = 1), decreasing = FALSE)
+dates_sim <- seq(
+  from = ymd("2000-07-01"),
+  to = ymd("2000-07-31"), length.out = 30
+)
+
+data_sim <- matrix(NA, 30, 2) |> data.frame()
+data_sim[, 1] <- temps_sim
+data_sim[, 2] <- dates_sim
+
+# change names so they aren't the same as the vector objects
+colnames(data_sim) <- c("temp_sim", "date_sim")
+```
+
+```{r smb-pred, echo=TRUE, warning=FALSE, message=FALSE}
 # Smallmouth Bass
 smb_res <- predict_phenology(
-  data = data_sim,
-  dates = date_sim,
-  temperature = temp_sim,
-  spawn.date = "2000-07-01",
-  model = smb_mod$expression
+  data = data_sim,             # simulated data
+  dates = date_sim,            # simulated dates
+  temperature = temp_sim,      # simulated temperatures
+  spawn.date = "2000-07-01",   # spawn date
+  model = smb_mod$expression   # model expression from fit_model()
 )
+```
 
+Finally, we summarize the outputs across species to visualize hatch timing and total developmental duration (Figure 5).
+This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
+
+```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
 # Catfish
@@
 
-## Predicting Phenology and Output
-
-To illustrate model selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**.
-Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
-
-First, we obtain the model expression for hatching using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
-
-```{r, echo=TRUE}
-# model_select() to get sockeye hatch model
-sockeye_hatch_mod <- model_select(
-  author = "Beacham and Murray 1990", 
-  species = "sockeye", 
-  model = 2, 
-  development_type = "hatch"
-  )
-```
-
-These model expressions are then implemented in `predict_phenology()` to estimate the days to hatch and development:
-
-```{r, warning=FALSE, echo=TRUE}
-# predict_phenology() with sockeye hatch model and woody_island temperature data
-WI_hatch <- predict_phenology(
-  data = woody_island,
-  dates = date,
-  temperature = temp_c,
-  spawn.date = "1990-08-18",  # notice the character string for spawn date
-  model = sockeye_hatch_mod
-  )
-```
-
-The returned object provides key outputs, including days to hatch and the full development period, allowing us to assess phenological patterns under the recorded thermal conditions:
-
-```{r, echo=TRUE}
-WI_hatch$days_to_develop
-WI_hatch$dev.period
-```
-
-### Understanding your results
-
-The output from `predict_phenology()` contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:
-
-```{r, echo=TRUE}
-summary(WI_hatch)
-```
-
-`WI_hatch$days_to_develop` -- Returns the predicted number of days required for development.
-
-`WI_hatch$dev.period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
-
-`WI_hatch$ef_table` -- An *n* x 5 tibble (*n* = number of days to hatch or emerge), containing a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
-This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discussed below.
-
-`WI_hatch$model_specs` -- Provides details about the model used for prediction, including whether it was retrieved from `model_select()` or generated using `fit_model()`.
-Most importantly, it contains the model expression (*i.e.*, the formula) used for phenology predictions.
-
-### Plotting phenology
-
-**hatchR** includes a built in function, `plot_phenology()`, for visualizing phenology predictions (Figure 4).
-This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span.
-Be default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles.
-However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results.
-
-This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.
-
-```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
-plot_phenology(WI_hatch)
-```
-
 # Case Study 1
@@
 
-```{r bt-emerge-crooked, eval=FALSE, echo=FALSE}
+```{r case-1, eval=FALSE, echo=FALSE}
 # read in EFS data
@@
 
-```{r, echo=FALSE}
+```{r case-2-prep, echo=FALSE}
 bull_trout_sites <- idaho |>
@@
 
-To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization (for concision, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository or in the Predict fish phenology: nested article on **hatchR**'s website).
+To predict hatch timing across these sites, we first set up the necessary models and data, using the built-in Bull Trout parameterization (for concision, we omit this setup here, but full details are available in the `paper.Rmd` file in the GitHub repository, or in the *Predict fish phenology: Nested* article on **hatchR**'s website).
+
 We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
 
-By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
-The results of this analysis are presented in Figure 5, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range and across years.
-This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
-
 ```{r, echo=TRUE}
+# apply predict_phenology() over multiple sites and spawn dates
 hatch_res <- isaak_summ_bt |>
@@
     dev_period = map2(
-      summ_obj, spawn_dates,  # map across our site object and spawn dates
-      predict_phenology,
-      temperature = daily_temp,
-      model = bt_hatch,
-      dates = date
+      summ_obj,                  # summarized temperature data object
+      spawn_dates,               # list of spawn dates
+      predict_phenology,         # function to apply
+      temperature = daily_temp,  # column of temperature data in `summ_obj`
+      dates = date,              # column of date in `summ_obj`
+      model = bt_hatch           # model to use (from model_select())
     ) |>
@@
   select(site, dev_period) |>           # just select the columns we want
-  unnest(cols = c(dev_period)) |>  # unnest everything
+  unnest(cols = c(dev_period)) |>       # un-nest everything
   mutate(days_to_hatch = stop - start)  # make a new column of days to hatch
@@
 
-```{r, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
+By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
+The results of this analysis are presented in Figure 6, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range and across years.
+This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
+
+```{r plot-bt, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
 cut_ints <- date(
@@
 Both versions allow users to import data, perform basic data checks, perform basic visualizations, and apply either pre-existing Salmonid model parameterizations or generate custom models specific to other species or populations.
-To support users at various levels of expertise, we provide extensive documentation on the hatchR website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.
+To support users at various levels of expertise, we provide extensive documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.
 
@@
 
-The application of hatchR and the effective value modeling framework relies on several key assumptions.
+The application of **hatchR** and the effective value modeling framework relies on several key assumptions.
 First, environmental stressors may alter developmental timing.

### 2025-03-07 inst/manuscript/paper.Rmd
commit final copy edits before changin figure 2 orientation

@@
 
-Morgan M. Sparks^1,5,6^, Bryan M. Maitland^1,6^, Eli A. Felts^2^, Allison G. Swartz^3^, Paul Frater^4^
+Morgan M. Sparks^1,5,6^, Bryan M. Maitland^1,6^, Eli A. Felts^2^, Allison G. Swartz^3^, Paul N. Frater^4^
 
@@
 
-As poikilothermic organisms, fish development and growth are closely tied to ambient environmental conditions.
-This strong relationship has enabled researchers to generate statistical models that predict developmental phenology with high accuracy.
-Historically, these models were formulated in aquaculture settings under the assumption of constant temperature throughout development [@beacham1990; @mcphail1979; @alderdice1978], limiting their applicability to wild populations.
-However, @sparks2019 reformulated this approach into an "Effective Value model" that instead uses daily average temperatures after spawning, which predicts hatch or emergence timing when cumulative effective values reach a threshold of one.
+Fish, as poikilothermic organisms, have developmental rates that are strongly influenced by ambient environmental conditions.
+This close relationship has allowed researchers to develop statistical models that predict developmental phenology with high accuracy.
+Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [\@beacham1990; \@mcphail1979; \@alderdice1978].
+As a result, accumulated thermal unit (ATU) thresholds have become widely used to predict development in wild populations.
+However, this approach can be inaccurate, particularly at the cold and warm extremes of temperature ranges, due to the non-linear relationship between average incubation temperature and the thermal units required for development [\@quinn2018].
+
+In colder environments, fish require fewer accumulated thermal units to develop than those in warmer conditions, yet still experience longer overall developmental periods.
+Recognizing this non-linear relationship, \@sparks2019 introduced the Effective Value model, which improves prediction accuracy by accounting for daily average temperatures following spawning.
+This model predicts hatch or emergence timing based on the cumulative effective values reaching a threshold of one, offering a more precise framework for forecasting developmental phenology in wild fish populations.
 
@@
 For a high-level overview of **hatchR**'s applications, see Figure 1.
-Additional details on key functions and workflows–particularly for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.
+Additional details on key functions and workflows, particularly for automating phenology predictions across multiple variables, are available in articles hosted on the software's webpage.
 
-```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow.
+```{r, echo = FALSE, out.height= "72%", fig.align = 'center', fig.cap="hatchR workflow.
 Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decision choices as italicized text."}
@@
 **hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
-Other columns may be present, but columns names should not include spaces.
+Other columns may be present, but column names should not include spaces.
 Data should follow the format outlined in Table 1.
@@
 **hatchR** is designed to analyze daily average temperatures.
-While high-frequency data (e.g., from HOBO loggers) can be used, it must be summarized into daily averages.
+While high-frequency data (*e.g.*, from HOBO loggers) can be used, it must be summarized into daily averages.
 **hatchR** provides built-in functionality for this summarization in R but requires pre-summarized data for use in the Shiny app.
@@
 183, for additional Salmonid models).
-To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C, requiring users to carefully asses whether their parameterized models are appropriate for the temperature ranges in their datasets.
+To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C, instead requiring users to carefully asses whether their own parameterized models are appropriate for the temperature ranges in their datasets.
 
@@
 
-The output from `predict_phenology()` contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:
+The output from `predict_phenology()` (in this example, our named object `WI_hatch`) contains multiple elements in a list, which can be accessed using the `$` operator.
+Each component provides different insight into the predicted phenology:
 
@@
 # R^2 values from model fits
-smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared
+c(smb_mod$r_squared, cat_mod$r_squared, sturgeon_mod$r_squared)
 ```

### 2025-03-07 inst/manuscript/paper.Rmd
fux w/ figure spacing

@@
 
-```{r, echo = FALSE, fig.height = 8, fig.width = 4,  fig.align = 'center', fig.cap = "Effective value model fit and application for European Grayling.
+```{r, echo = FALSE, fig.height=3, fig.align = 'center', fig.cap = "Effective value model fit and application for European Grayling.
 Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line).
@@
 
-p1 / p2 + plot_annotation(tag_levels = "A")
+p1 + p2 + plot_annotation(tag_levels = "A")
 ```
@@
 ```{r plot-check-1, echo = TRUE,  fig.cap = "Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
-Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 4, fig.width = 6.5}
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 3.25}
 # visual check of temperature data
@@
 
-```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
+```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.", fig.height=5}
 # Catfish

### 2025-03-07 inst/manuscript/paper.Rmd
fix refereneces in intro

@@
 This close relationship has allowed researchers to develop statistical models that predict developmental phenology with high accuracy.
-Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [\@beacham1990; \@mcphail1979; \@alderdice1978].
+Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [@beacham1990, @mcphail1979, @alderdice1978].
 As a result, accumulated thermal unit (ATU) thresholds have become widely used to predict development in wild populations.
-However, this approach can be inaccurate, particularly at the cold and warm extremes of temperature ranges, due to the non-linear relationship between average incubation temperature and the thermal units required for development [\@quinn2018].
+However, this approach can be inaccurate, particularly at the cold and warm extremes of temperature ranges, due to the non-linear relationship between average incubation temperature and the thermal units required for development [@quinn2018].
 
 In colder environments, fish require fewer accumulated thermal units to develop than those in warmer conditions, yet still experience longer overall developmental periods.
-Recognizing this non-linear relationship, \@sparks2019 introduced the Effective Value model, which improves prediction accuracy by accounting for daily average temperatures following spawning.
+Recognizing this non-linear relationship, @sparks2019 introduced the Effective Value model, which improves prediction accuracy by accounting for daily average temperatures following spawning.
 This model predicts hatch or emergence timing based on the cumulative effective values reaching a threshold of one, offering a more precise framework for forecasting developmental phenology in wild fish populations.

### 2025-03-07 inst/manuscript/paper.Rmd
final copy edits and render

@@
 In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population.
-Bull Trout, a federally threatened species under the Endangered Species Act [@usfws_bt_1999], are particularly sensitive to sediment disturbance [@nolfi2024].
+Bull Trout, a federally threatened species under the Endangered Species Act [@usfws_bt_1999, @nolfi2024], are particularly sensitive to sediment disturbance.
 The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st.
@@
   temperature = temp_c,
-  spawn.date = "2011-09-30",
+  spawn.date = "2014-09-30",
   model = bt_emerge_mod
@@
 
-In this example, the model predicts that the last Bull Trout will emerge on May 1, before the road work target date (June 1).
+In this example, the model predicts that the last Bull Trout will emerge on April 21, before the road work target date (June 1).
 This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel.
@@
 While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events [@mason1976].
-We encourage users to predict phenology using early, peak, and late thresholds (e.g., 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.
+We encourage users to predict phenology using early, peak, and late thresholds (*e.g.*, 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.
 
@@
 
-Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings.
+Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available.
 We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website.

### 2025-03-10 inst/manuscript/paper.Rmd
fix citations and rerender
Fix in-text citation

@@
 This close relationship has allowed researchers to develop statistical models that predict developmental phenology with high accuracy.
-Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [@beacham1990, @mcphail1979, @alderdice1978].
+Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [@beacham1990; @mcphail1979; @alderdice1978].
 As a result, accumulated thermal unit (ATU) thresholds have become widely used to predict development in wild populations.
@@
 In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population.
-Bull Trout, a federally threatened species under the Endangered Species Act [@usfws_bt_1999, @nolfi2024], are particularly sensitive to sediment disturbance.
+Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensitive to sediment disturbance.
 The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st.

### 2025-03-10 inst/manuscript/paper.Rmd
add supplementary materials line
rerender with double spaces and move some figures for better spacing

@@
    - \linenumbers
+   - \usepackage{setspace}
+   - \doublespacing
 bibliography: ms_refs.bib
@@
 
+\newpage
+
 # Case Study 1
@@
 Both the prediction table and plot can be downloaded directly from their respective tabs.
-The full process is demonstrated through a detailed walkthrough on the **hatchR** Shiny website.
+The full process is demonstrated through a detailed walkthrough in the supplementary video file (Supplementary Materials).
 
@@
 
-We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided helpful discussion about model development and temperature data sets.
+We thank Laura Koller for her help designing the **hatchR** logo.
+Dan Isaak provided helpful discussion about model development and temperature data sets.
+Matt Kaylor provided useful comments on an earlier trial run of the package and shiny app.
 

### 2025-03-13 inst/poster/hatchR_posterdown/hatchR_posterdown.Rmd

@@
 poster_width: "48in"
-primary_colour: "	#011f4b"
+primary_colour: "#011f4b"
+secondary_colour: "#011f4b"
+author_textcol: "white"
+titlebox_borderwidth: "1cm"
+titlebox_bordercol: "#011f4b"
 sectitle_bgcol: "#005b96"
@@
 
-```{r, out.width='36%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}
+```{r, out.width='34%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}
 
@@
 
-```{r, out.width='75%', fig.align='center', fig.cap= "Predicted days to hatch using custom parameterizations for three warmwater species with mean 16 °C thermal regime."}
+```{r, out.width='73%', fig.align='center', fig.cap= "Predicted days to hatch using custom parameterizations for three warmwater species with mean 16 °C thermal regime."}
 ###  make temp regime
@@
 
-Our model predicts Bull Trout will emerge 204 days after September 30th, which is April 22 or well before the July 1 date of road construction.
+Our model predicts Bull Trout will emerge 204 days after September 30, which is April 22 or well before the July 1 date of road construction.
 
@@
 
-```{r, fig.height = 8.75, fig.align='center', fig.cap= "Predicted hatch timing for 139 putatuve Bull Trout sites for three spawning periods."}
+```{r, fig.height = 8.75, fig.align='center', fig.cap= "Predicted hatch timing for 139 putative Bull Trout sites for three spawning periods."}
 cut_ints <- date(

### 2025-03-14 inst/misc/bmm_ppt.qmd
Add qmd ppt stuff

---
title: "hatchR"
format: pptx
editor: visual
execute:
  echo: TRUE
  warning: false
---

## Prep

```{r}
library(hatchR)
library(tidyverse)
library(ggridges) 
```

## Woody Island Dataset

```{r}
print(woody_island, n = 5)
```

## `plot_temp_check()`

```{r}
plot_check_temp(
  data = woody_island,
  dates = date, 
  temperature = temp_c
  ) +
  geom_rect(
    aes(
      xmin = ymd("1990-08-18"),
      xmax = ymd("1991-04-01"),
      ymin = -10,
      ymax = 25),
    fill = "grey",
    alpha = 0.01
    ) 
```

## Select built-in model parameterization

```{r}
model_table
```

------------------------------------------------------------------------

```{r}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  development_type = "hatch"
  )
sockeye_hatch_mod$expression
```

## Predict phenology!

```{r}
WI_hatch <- predict_phenology(
  data = woody_island,
  dates = date,
  temperature = temp_c,
  spawn.date = "1990-08-18",
  model = sockeye_hatch_mod
  )
```

## Checking outputs

```{r}
WI_hatch$days_to_develop
WI_hatch$dev.period
```

## Plot phenology

```{r}
plot_phenology(WI_hatch)
```

------------------------------------------------------------------------

```{r echo=FALSE}
# Smallmouth Bass data (Webster 1948)
smallmouth <- matrix(NA, 10, 2) |> data.frame()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

# change °F to °C and hours to days
smallmouth <- smallmouth |>
  mutate(
    days = ceiling(hours / 24),
    temp_C = (temp_F - 32) * (5 / 9)
  ) |> 
  select(-hours, -temp_F)
```

```{r}
smallmouth
```

------------------------------------------------------------------------

```{r}
# estimate parameters for Smallmouth Bass using fit_model()
smb_mod <- fit_model(
  temp = smallmouth$temp_C,
  days = smallmouth$days,
  species = "smb",
  development_type = "hatch"
)
```

------------------------------------------------------------------------

```{r}
smb_mod$pred_plot
```

------------------------------------------------------------------------

```{r}
smb_mod$r_squared
smb_mod$expression$expression
```

------------------------------------------------------------------------

```{r echo=FALSE, message=FALSE}
set.seed(123)

# create random temps and corresponding dates
temps_sim <- sort(rnorm(n = 30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <- seq(
  from = ymd("2000-07-01"),
  to = ymd("2000-07-31"), length.out = 30
)

data_sim <- matrix(NA, 30, 2) |> data.frame()
data_sim[, 1] <- temps_sim
data_sim[, 2] <- dates_sim

# change names so they aren't the same as the vector objects
colnames(data_sim) <- c("temp_sim", "date_sim")
```

```{r message=FALSE}
smb_hatch <- predict_phenology(
  data = data_sim,             # simulated data
  dates = date_sim,            # simulated dates
  temperature = temp_sim,      # simulated temperatures
  spawn.date = "2000-07-01",   # spawn date
  model = smb_mod$expression   # model expression from fit_model()
)
smb_hatch$days_to_develop
smb_hatch$dev.period
```

------------------------------------------------------------------------

```{r echo=FALSE}
# Summary for all species
all_res <- data.frame(matrix(NA, 1, 2))
colnames(all_res) <- c("start", "stop")

all_res$start <- c(smb_hatch$dev.period$start)

all_res$stop <- c(smb_hatch$dev.period$stop)

all_res <- all_res |>
  mutate(
    days = ceiling(stop - start),
    index = c(16.5)
  )

all_res$Species <- c("Smallmouth Bass")

ggplot() +
  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim)) +
  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax = index - .35, ymin = index - .5, fill = Species)) +
  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index - 0.425), label = days)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
  theme_classic(base_size = 16) +
  theme(legend.position = c(0.75, 0.25))
```

------------------------------------------------------------------------

```{r}
plot_phenology(smb_hatch)
```

------------------------------------------------------------------------

```{r}
idaho
```

------------------------------------------------------------------------

```{r}
bull_trout_sites <- idaho |>
  mutate(month = month(date)) |>                # make a month column (numeric)
  filter(month == 8) |>                         # filter out Aug.
  group_by(site) |>                             # apply grouping by site
  summarise(mean_aug_temp = mean(temp_c)) |>
  filter(mean_aug_temp <= 13) |>                # keep only sites 13 C or cooler
  pull(site) |>
  unique()

# filter to bull trout sites
idaho_bt <- idaho |>
  filter(site %in% bull_trout_sites)
```

------------------------------------------------------------------------

```{r}
# nest data and summarize
isaak_summ_bt <- idaho_bt |>
  group_by(site) |>
  nest() |>
  mutate(
    summ_obj = map(
      data,
      summarize_temp,
      temperature = temp_c,
      dates = date
    )
  ) |>
  select(site, summ_obj)
```

------------------------------------------------------------------------

```{r}
isaak_summ_bt
```

------------------------------------------------------------------------

```{r}
# lets look at a couple individual sites
PIBO_1345 <- idaho_bt |> filter(site == "PIBO_1345")

# looks nice
plot_check_temp(PIBO_1345,
                dates = date,
                temperature = temp_c)
```

------------------------------------------------------------------------

```{r}
spawn_dates <- map(
  c(2011:2014),     # year vector to map for custom function
  function(year) {  # custom paste function
    c(
      paste0(year, "-09-01"),
      paste0(year, "-09-15"),
      paste0(year, "-09-30")
    )
  }
) |>
  unlist()
```

------------------------------------------------------------------------

```{r}
spawn_dates
```

------------------------------------------------------------------------

```{r}
bt_hatch <- model_select(
  development_type = "hatch",
  author = "Austin et al. 2019",
  species = "bull trout",
  model = "MM"
)
```

------------------------------------------------------------------------

```{r}
# apply predict_phenology() over multiple sites and spawn dates
hatch_res <- isaak_summ_bt |>
  mutate(
    dev_period = map2(
      summ_obj,                  # summarized temperature data object
      spawn_dates,               # list of spawn dates
      predict_phenology,         # function to apply
      temperature = daily_temp,  # column of temperature data in `summ_obj`
      dates = date,              # column of date in `summ_obj`
      model = bt_hatch           # model to use (from model_select())
    ) |>
      map_df("dev.period") |>    # pull out just dev.period results
      list()
  ) |>
  select(site, dev_period) |>           # just select the columns we want
  unnest(cols = c(dev_period)) |>       # un-nest everything
  mutate(days_to_hatch = stop - start)  # make a new column of days to hatch
```

------------------------------------------------------------------------

```{r}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

# add year and spawn time factors for plotting
hatch_factors <- hatch_res |>
  mutate(year = cut(start,
    breaks = cut_ints,
    labels = c(2011:2014)
  )) |>
  # cut coerces our labels to factors, this changes them back to numbers
  mutate(year = as.numeric(as.character(year))) |>
  mutate(day = day(start)) |>
  mutate(spawn_time = case_when(
    day == 1 ~ "Early",
    day == 15 ~ "Peak",
    day == 30 ~ "Late"
  )) |>
  mutate(spawn_time = factor(
    spawn_time,
    levels = c("Late", "Peak", "Early"),
    ordered = TRUE
  ))

hatch_factors_sum <- hatch_factors |>
  group_by(year, spawn_time, start) |>
  summarise(
    mean_hatch = ceiling(mean(days_to_hatch)),
    hatch_5 = ceiling(quantile(days_to_hatch, probs = 0.05)),
    hatch_95 = ceiling(quantile(days_to_hatch, probs = 0.95))
  )

ggplot(
  data = hatch_factors,
  aes(
    x = days_to_hatch,
    y = spawn_time,
    fill = spawn_time,
    color = spawn_time
  )
) +
  geom_density_ridges(alpha = 0.9) +
  facet_wrap(~year, ncol = 1) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  scale_color_brewer(palette = "Blues", direction = 1) +
  labs(x = "Days to hatch", y = "Spawn time") +
  theme_classic() +
  theme(legend.position = "none")
```

### 2025-04-07 README.Rmd
add CRAN dl badge

@@
 [![cran checks](https://badges.cranchecks.info/worst/hatchR.svg)](https://cran.r-project.org/web/checks/check_results_hatchR.html)
+[![](https://cranlogs.r-pkg.org/badges/hatchR)](https://cran.r-project.org/package=hatchR)
 [![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

### 2025-04-22 inst/poster/hatchR_betterland/hatchR_betterland.Rmd
create a betterland poster

---
title: hatchR
author:
  - name: Morgan Sparks
    affil: 1
  - name: Bryan Maitland
    affil: 1
  - name: Eli Felts
    affil: 2
  - name: Allison Swartz
    affil: 3
  - name: Paul Frater
    affil: 4
affiliation:
  - num: 1
    address: Rocky Mountain Research Station, US Forest Service
  - num: 2
    address: Idaho Fish and Wildlife Conservation Oﬀice, US Fish and Wildlife Service
  - num: 3
    address: College of Forestry, Oregon State University
  - num: 4
    address: Wisconsin Department of Natural Resources, Madison, WI, USA
main_findings:
  - "A toolset to predict when fish hatch and emerge in the wild"
  - '![](hatchR_betterland_files/figure-html/logo.png){.main_pic}'
logoleft_name: '![](hatchR_betterland_files/figure-html/hatchR_website.png){.main-img-left}'
logoright_name:
main_fontfamily: "Rasa"
output: 
  posterdown::posterdown_betterland:
    self_contained: false
    pandoc_args: --mathjax
    highlight: haddock
    number_sections: false
link-citations: true
bibliography: packages.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(results = 'asis',
                      echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html")

```

# Introduction

**hatchR** is a software ecosystem for predicting fish developmental phenology using statistical models. It offers three primary applications:

1.  Basic summarization, plotting, and data checks for water temperature data.
2.  Access published parameterizations for salmonid developmental models or build your own custom parameterizations for any species with user provided data.
3.  Predict hatch and/or emergence timing across ambient temperatures from common data sources such as HOBO loggers.

**hatchR** is deployed in two formats. First, it can be downloaded as a R package. Secondly, for users not familiar with R, it can also be accessed as a web app built in Shiny as a point-and-click tool.

To predict phenology, you need [**daily average temperatures over incubation (or data that can be summarized as such**]{.underline}**)** and [**spawn timing**]{.underline}.

# Using hatchR

## `hatchR` can be accessed two ways:

### R package

`hatchR` can be downloaded from CRAN and used in the R programming language for full functionality. The `hatchR` website provides numerous Articles on basic to advanced use.

```{r,out.width="50%",out.height="50%",fig.show='hold', fig.align='center'}

knitr::include_graphics("./hatchR_betterland_files/figure-html/hatchR_website.png")
```

### Shiny app

`hatchR` can be accessed in a point-and-click interface via Shiny. The app loads in your browser and provides much of the functionality of the R package but is less automative.

```{r,out.width="50%",out.height="50%",fig.show='hold', fig.align='center'}

  knitr::include_graphics("./hatchR_betterland_files/figure-html/hatchR_shiny.png")
```

# Workflow

```{r, out.width='60%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}

knitr::include_graphics("./hatchR_betterland_files/figure-html/workflow.png")
```

### 2025-04-23 inst/poster/hatchR_betterland/hatchR_betterland.Rmd

@@
   - "A toolset to predict when fish hatch and emerge in the wild"
-  - '![](hatchR_betterland_files/figure-html/logo.png){.main_pic}'
-logoleft_name: '![](hatchR_betterland_files/figure-html/hatchR_website.png){.main-img-left}'
+  - '![](hatchR_betterland_files/figure-html/hatchR_logo_hires.png){.main_pic}'
+logoleft_name: 
 logoright_name:
@@
 
-## `hatchR` can be accessed two ways:
+## **hatchR** can be accessed two ways:
 
@@
 
-`hatchR` can be downloaded from CRAN and used in the R programming language for full functionality. The `hatchR` website provides numerous Articles on basic to advanced use.
+**hatchR** can be downloaded from CRAN and used in the R programming language for full functionality. The **hatchR** website provides numerous Articles on basic to advanced use.
 
@@
 
-`hatchR` can be accessed in a point-and-click interface via Shiny. The app loads in your browser and provides much of the functionality of the R package but is less automative.
+**hatchR** can be accessed in a point-and-click interface via Shiny. The app loads in your browser and provides much of the functionality of the R package but is less automative.
 

### 2025-04-24 inst/poster/hatchR_betterland/hatchR_betterland.Rmd
fix layout of betterland poster

@@
 main_fontfamily: "Rasa"
+column_padding: "10mm"
 output: 
@@
 
+<br>
+
 ### Shiny app
@@
 
-```{r, out.width='60%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}
+```{r, out.width='55%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}
 

### 2025-04-24 inst/poster/hatchR_betterland/hatchR_betterland.Rmd
copy edits to betterland poster

--- 2025-04-23 inst/poster/hatchR_betterland/hatchR_betterland.Rmd
@@
   - num: 2
-    address: Idaho Fish and Wildlife Conservation Oﬀice, US Fish and Wildlife Service
+    address: Idaho Fish and Wildlife Conservation Office, US Fish and Wildlife Service
   - num: 3
@@
 2.  Access published parameterizations for salmonid developmental models or build your own custom parameterizations for any species with user provided data.
-3.  Predict hatch and/or emergence timing across ambient temperatures from common data sources such as HOBO loggers.
+3.  Predict hatch and/or emergence timing in *wild* environments from common data sources such as HOBO loggers.
 
 **hatchR** is deployed in two formats.
-First, it can be downloaded as a R package.
+First, it can be downloaded as an R package.
 Secondly, for users not familiar with R, it can also be accessed as a web app built in Shiny as a point-and-click tool.
 
-To predict phenology, you need [**daily average temperatures over incubation (or data that can be summarized as such**]{.underline}**)** and [**spawn timing**]{.underline}.
+To predict phenology, you need [**daily average temperatures over devlopment (or data that can be summarized as such**]{.underline}**)** and [**spawn timing**]{.underline}.
 
@@
 
+<br>
+
 ### Shiny app
 
-**hatchR** can be accessed in a point-and-click interface via Shiny. The app loads in your browser and provides much of the functionality of the R package but is less automative.
+**hatchR** can be accessed in a point-and-click interface via Shiny. The app loads in your browser and provides much of the functionality of the R package but is less flexible.
 
@@
 
-```{r, out.width='60%', fig.align='center', fig.cap= "hatchR workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}
+```{r, out.width='55%', fig.align='center', fig.cap= "**hatchR** workflow. Processes are presented as filled circles, functions as plain text rectangles, and decisions as italicized text."}
 

### 2025-04-25 R/plot_phenology.R
change dev.period to dev_period; closes #50

@@ -52,7 +52,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
       title = paste(dat$days_to_develop, "days to develop"),
       subtitle = stringr::str_wrap(
         stringr::str_glue(
-          "Fish spawned: {dat$dev.period$start}; fish developed: {dat$dev.period$stop}",
+          "Fish spawned: {dat$dev_period$start}; fish developed: {dat$dev_period$stop}",
           "<br><span style='color:#1f78b4'>Temperature</span>;
           <span style='color:#d95f02'>Cumulative EF value</span>;
           <span style='color:#1b9e77'>Daily EF value (x100)</span>"
@@ -66,7 +66,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
       title = paste(dat$days_to_develop, "days to develop"),
       subtitle = stringr::str_wrap(
         stringr::str_glue(
-          "Fish spawned: {dat$dev.period$start}; fish developed: {dat$dev.period$stop}",
+          "Fish spawned: {dat$dev_period$start}; fish developed: {dat$dev_period$stop}",
           "<br><span style='color:#1f78b4'>Temperature</span>;
           <span style='color:#d95f02'>Cumulative EF value</span>"
         ),
@@ -79,7 +79,7 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
       title = paste(dat$days_to_develop, "days to develop"),
       subtitle = stringr::str_wrap(
         stringr::str_glue(
-          "Fish spawned: {dat$dev.period$start}; fish developed: {dat$dev.period$stop}",
+          "Fish spawned: {dat$dev_period$start}; fish developed: {dat$dev_period$stop}",
           "<br><span style='color:#1f78b4'>Temperature</span>;
           <span style='color:#1b9e77'>Daily EF value (x100)</span>"
         ),

### 2025-04-25 R/predict_phenology.R

@@ -19,7 +19,7 @@
 #' * `ef_table`: An n x 4 tibble (n = number of days to hatch or emerge) with
 #'    the dates, temperature, effective values, and cumulative sum of the
 #'    effective values.
-#' * `dev.period`: a 1x2 dataframe with the dates corresponding to when your
+#' * `dev_period`: a 1x2 dataframe with the dates corresponding to when your
 #'    fish's parent spawned (input with `predict_phenology(spawn.date = ...)`)
 #'    and the date when the fish is predicted to hatch or emerge.
 #' * `model_specs`: A data.frame with the model specifications.
@@ -131,15 +131,15 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     dat_out_sub <- dat_out[1:(1 + (Ef.days - 1)), ]
 
     # get development period
-    dev.period <- data.frame(matrix(NA, nrow = 1, ncol = 2))
-    colnames(dev.period) <- c("start", "stop")
-    dev.period$start <- min(dat_out_sub$dates)
-    dev.period$stop <- max(dat_out_sub$dates)
-    # dev.period$stop <- lubridate::as_date(NA)
+    dev_period <- data.frame(matrix(NA, nrow = 1, ncol = 2))
+    colnames(dev_period) <- c("start", "stop")
+    dev_period$start <- min(dat_out_sub$dates)
+    dev_period$stop <- max(dat_out_sub$dates)
+    # dev_period$stop <- lubridate::as_date(NA)
 
     ef.results <- list(
       days_to_develop = Ef.days,
-      dev.period = dev.period,
+      dev_period = dev_period,
       ef_table = dat_out_sub,
       model_specs = model
     )
@@ -160,15 +160,15 @@ predict_phenology <- function(data, dates, temperature, spawn.date, model) {
     dat_out_sub <- dat_out
 
     # get development period
-    dev.period <- data.frame(matrix(NA, nrow = 1, ncol = 2))
-    colnames(dev.period) <- c("start", "stop")
-    dev.period$start <- min(dat_out_sub$dates)
-    # dev.period$stop <- max(dat_out_sub$dates)
-    dev.period$stop <- lubridate::as_date(NA)
+    dev_period <- data.frame(matrix(NA, nrow = 1, ncol = 2))
+    colnames(dev_period) <- c("start", "stop")
+    dev_period$start <- min(dat_out_sub$dates)
[…]

### 2025-04-25 README.Rmd
change dev.period to dev_period; closes #50

--- 2025-03-06 README.Rmd
@@
 
-# hatchR: Predict Fish Hatch and Emergence Timing <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
+# hatchR <a href="https://bmait101.github.io/hatchR/"><img src="man/figures/logo.png" align="right" height="139" alt="hatchR website" /></a>
 
-<!-- badges: start -->
-[![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml)
-[![CRAN status](https://www.r-pkg.org/badges/version/hatchR)](https://CRAN.R-project.org/package=hatchR)
-[![cran checks](https://badges.cranchecks.info/worst/hatchR.svg)](https://cran.r-project.org/web/checks/check_results_hatchR.html)
-[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
-<!-- badges: end -->
+[![publication](https://img.shields.io/badge/Cite-Unpublished-yellow)](https://github.com/bmait101/hatchR/blob/master/inst/CITATION)
+[![downloads](https://cranlogs.r-pkg.org/badges/hatchR)](https://cran.r-project.org/package=hatchR)
+[![total](https://cranlogs.r-pkg.org/badges/grand-total/hatchR)](https://cranlogs.r-pkg.org/)
+
+***Predict hatch and emergence timing for wild fish***
+
+---
 
 **hatchR** is an R package for predicting hatch and emergence timing for a wide range of wild fishes using the effective value framework (Sparks et al., (2019) <DOI:10.1139/cjfas-2017-0468>).
-'hatchR' offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
+**hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
 The package includes functions for model selection, phenology prediction, and visualization.
@@
 
-## Citation
+## Installation
 
-If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running:
+[![CRAN](https://www.r-pkg.org/badges/version/hatchR)](https://CRAN.R-project.org/package=hatchR)
+[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
+<!-- [![R-CMD-check](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bmait101/hatchR/actions/workflows/R-CMD-check.yaml) -->
+<!-- [![cran checks](https://badges.cranchecks.info/worst/hatchR.svg)](https://cran.r-project.org/web/checks/check_results_hatchR.html) -->
 
-```{r}
-citation(package = "hatchR")
-```
-
-## Installation Instructions
-
-Install the released version from CRAN:
+Install the released version of **hatchR** from CRAN:
 
@@
 
-Install the development version of **hatchR** from GitHub:
+Install the latest development version of **hatchR** from GitHub:
 
@@
 
-## Interactive Shiny App
+## Interactive Shiny App <img src='https://upload.wikimedia.org/wikipedia/commons/b/bf/Shiny_hex_logo.svg' align="right" height="50" />
 
@@
 # and development period
-WI_hatch$dev.period
+WI_hatch$dev_period
 ```

### 2025-05-05 inst/poster/hatchR_betterland/hatchR_betterland.Rmd
more poster copy edits

@@
   - num: 4
-    address: Wisconsin Department of Natural Resources, Madison, WI, USA
+    address: Wisconsin Department of Natural Resources
 main_findings:
@@
 
-### R package
+### 1. R package
 
@@
 
-### Shiny app
+### 2. Shiny app
 

### 2025-06-02 R/plot_phenology.R
hard code label colors

@@ -53,9 +53,9 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
       subtitle = stringr::str_wrap(
         stringr::str_glue(
           "Fish spawned: {dat$dev_period$start}; fish developed: {dat$dev_period$stop}",
-          "<br><span style='color:#1f78b4'>Temperature</span>;
-          <span style='color:#d95f02'>Cumulative EF value</span>;
-          <span style='color:#1b9e77'>Daily EF value (x100)</span>"
+          "<br><span style='color:{cols[1]}'>Temperature</span>;
+          <span style='color:{cols[2]}'>Cumulative EF value</span>;
+          <span style='color:{cols[3]}'>Daily EF value (x100)</span>"
         ),
         width = 30
       )
@@ -67,8 +67,8 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
       subtitle = stringr::str_wrap(
         stringr::str_glue(
           "Fish spawned: {dat$dev_period$start}; fish developed: {dat$dev_period$stop}",
-          "<br><span style='color:#1f78b4'>Temperature</span>;
-          <span style='color:#d95f02'>Cumulative EF value</span>"
+          "<br><span style='color:{cols[1]}'>Temperature</span>;
+          <span style='color:{cols[2]}'>Cumulative EF value</span>"
         ),
         width = 30
       )
@@ -80,8 +80,8 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
       subtitle = stringr::str_wrap(
         stringr::str_glue(
           "Fish spawned: {dat$dev_period$start}; fish developed: {dat$dev_period$stop}",
-          "<br><span style='color:#1f78b4'>Temperature</span>;
-          <span style='color:#1b9e77'>Daily EF value (x100)</span>"
+          "<br><span style='color:{cols[1]}'>Temperature</span>;
+          <span style='color:{cols[3]}'>Daily EF value (x100)</span>"
         ),
         width = 30
       )

### 2025-06-02 R/plot_phenology.R
Hard code pt size and linewidth

@@ -45,6 +45,8 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   x_lab <- "Date"
   y_lab <- "Mean daily temperature"
   cols <- c("#1f78b4", "#d95f02", "#1b9e77")
+  size.pts <- 1.0
+  size.line <- 1.0
 
   if (labels == TRUE) {
     all_label <- ggplot2::labs(
@@ -98,12 +100,12 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   if (style == "all") {
     p <- dat$ef_table |>
       ggplot2::ggplot(ggplot2::aes(x = .data$dates, y = .data$temperature)) +
-      ggplot2::geom_line(color = cols[1]) +
-      ggplot2::geom_point(color = cols[1], size = 0.5) +
-      ggplot2::geom_line(ggplot2::aes(y = .data$ef_cumsum * max(.data$temperature)), color = cols[2]) +
-      ggplot2::geom_point(ggplot2::aes(y = .data$ef_cumsum * max(.data$temperature)), color = cols[2], size = 0.25) +
-      ggplot2::geom_line(ggplot2::aes(y = .data$ef_vals * 100), color = cols[3]) +
-      ggplot2::geom_point(ggplot2::aes(y = .data$ef_vals * 100), color = cols[3], size = 0.25) +
+      ggplot2::geom_line(color = cols[1], linewidth = size.line) +
+      ggplot2::geom_point(color = cols[1], size = size.pts) +
+      ggplot2::geom_line(ggplot2::aes(y = .data$ef_cumsum * max(.data$temperature)), color = cols[2], linewidth = size.line) +
+      ggplot2::geom_point(ggplot2::aes(y = .data$ef_cumsum * max(.data$temperature)), color = cols[2], size = size.pts) +
+      ggplot2::geom_line(ggplot2::aes(y = .data$ef_vals * 100), color = cols[3], linewidth = size.line) +
+      ggplot2::geom_point(ggplot2::aes(y = .data$ef_vals * 100), color = cols[3], size = size.pts) +
       all_label +
       ggplot2::theme_classic() +
       ggplot2::theme(plot.subtitle = ggtext::element_markdown()) +
@@ -115,10 +117,10 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   if (style == "ef_cumsum") {
     p <- dat$ef_table |>
[…]

### 2025-06-02 R/plot_phenology.R
Changed to Okabe-Ito cols

@@ -44,9 +44,9 @@ plot_phenology <- function(plot, style = "all", labels = TRUE) {
   dat <- plot
   x_lab <- "Date"
   y_lab <- "Mean daily temperature"
-  cols <- c("#1f78b4", "#d95f02", "#1b9e77")
+  cols <- c("#0072B2", "#E69F00", "#009E73")
   size.pts <- 1.0
-  size.line <- 1.0
+  size.line <- 0.75
 
   if (labels == TRUE) {
     all_label <- ggplot2::labs(

### 2025-06-17 inst/manuscript/paper.Rmd
Reviewer 1 comments

@@
 Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [@beacham1990; @mcphail1979; @alderdice1978].
-As a result, accumulated thermal unit (ATU) thresholds have become widely used to predict development in wild populations.
+As a result, accumulated thermal unit thresholds (ATU models) have become widely used to predict development in wild populations.
 However, this approach can be inaccurate, particularly at the cold and warm extremes of temperature ranges, due to the non-linear relationship between average incubation temperature and the thermal units required for development [@quinn2018].
@@
 This is accomplished with the `predict_phenology()` function.
-The R package provides user-friendly example vignettes that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs.
+The R package provides user-friendly example articles that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs.
 In contrast, the Shiny application includes a default output plot and an option to download results for external visualizations.

### 2025-06-17 inst/manuscript/paper.Rmd
most of reviewer 2 comments.

@@
 However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on Salmonids.
-Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and stream temperature data.
+Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and wtaer temperature data.
 **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
@@
 Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>).
-The second format is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality.
+The second format is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality (<https://elifelts.shinyapps.io/hatchR_demo/>).
 Below, we provide an overview of **hatchR** and present case studies demonstrating its application in research and management.
@@
 Given the diversity of potential data types, it is impractical to implement comprehensive validation checks.
-However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`).
+However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`), the process for which is highlighted in **Box XXX**.
 Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function.
@@
   labs(x = "Developmental average temperature (°C)", y = "Days to hatch") +
+  lims(x = c(0,15)) +
   theme_classic()
@@
   labs(x = "Daily average temperature (°C)", y = "Effective value") +
+  lims(x = c(0,15)) +
   theme_classic()
@@
 
-**hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
+**hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time in the case of fiedl) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
 Other columns may be present, but column names should not include spaces.
@@
 
-To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature.
+To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which only requires species' development follows a power law relationship with temperature.
 The function takes two input vectors: average incubation temperature (°C) and the number of days to a given phenological event.
 A model is then fit to the data using `stats::nls()`, which performs nonlinear least squares regression to estimate the parameters *log~e~a* and *b*.
-Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to fit the nonlinear model.
+Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to
 
@@
 Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (available in `paper.Rmd`).
-Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Base shown below):
+Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Bass shown below):
 
@@
 For instance, will fish have emerged from redds before construction begins, reducing potential disturbance?
+This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.
 
-In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population.
+In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed in Idaho, USA, home to a key Bull Trout (*S.
+confluentus*) population.
 Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensitive to sediment disturbance.
@@
 
-To demonstrate this case study, we use the **hatchR** graphical user interface available at <https://elifelts.shinyapps.io/hatchR_demo/>.
+To demonstrate this case study, we use the **hatchR** graphical user interface.
 Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year).
@@
 This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel.
-This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.
 
@@
 For the second case study, we demonstrate a more complex, large-scale application of **hatchR**, highlighting its full flexibility when applied programmatically in R.
-This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the upper Columbia River headwaters in Idaho.
+This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the greater Snake River watershed in Idaho, USA.
 
@@
 Dan Isaak provided helpful discussion about model development and temperature data sets.
-Matt Kaylor provided useful comments on an earlier trial run of the package and shiny app.
+Matt Kaylor provided useful comments on an earlier trial run of the package and Shiny app.
 

### 2025-06-17 inst/manuscript/paper.Rmd
rerender with updated apa.csl

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE,
+                      warning = FALSE)
 ```
@@
 ```{r, echo=TRUE}
-WI_hatch$days_to_develop; WI_hatch$dev.period
+WI_hatch$days_to_develop; WI_hatch$dev_period
 ```
@@
 
-`WI_hatch$dev.period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
+`WI_hatch$dev_period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
 
@@
 all_res$start <- c(
-  catfish_res$dev.period$start,
-  smb_res$dev.period$start,
-  sturgeon_res$dev.period$start
+  catfish_res$dev_period$start,
+  smb_res$dev_period$start,
+  sturgeon_res$dev_period$start
 )
@@
 all_res$stop <- c(
-  catfish_res$dev.period$stop,
-  smb_res$dev.period$stop,
-  sturgeon_res$dev.period$stop
+  catfish_res$dev_period$stop,
+  smb_res$dev_period$stop,
+  sturgeon_res$dev_period$stop
 )
@@
 # fish emerge May 1, before June 1
-bt_emergence$dev.period$stop
+bt_emergence$dev_period$stop
 ```
@@
     ) |>
-      map_df("dev.period") |>    # pull out just dev.period results
+      map_df("dev_period") |>    # pull out just dev_period results
       list()

### 2025-06-18 inst/manuscript/paper.Rmd
Editor comments (most)

@@
 Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records.
-However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on Salmonids.
+However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids.
 Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and wtaer temperature data.
@@
 
-The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations are readily available.
+The effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations are readily available.
 For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
@@
 
-To extend these modeling capabilities beyond Salmonids and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations.
+To extend these modeling capabilities beyond salmonids and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations.
 **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017.
@@
 
-**hatchR** is designed primarily as a tool for predicting early life history phenology in wild fishes.
+The **hatchR** softeware is designed primarily as a tool for predicting early life history phenology in wild fishes.
 To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data.
 Given the diversity of potential data types, it is impractical to implement comprehensive validation checks.
-However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`), the process for which is highlighted in **Box XXX**.
-Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function.
-Users can then apply these models to water temperature data (*e.g.,* from an in-stream temperature logger) to predict when hatching phenology will occur.
-This is accomplished with the `predict_phenology()` function.
+However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`), the process for which is highlighted in Box 1.
+Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function (Boxes 2,3) or fitting custom functions from data using the `fit_model()` function (Box 4).
+Users can then apply these models to water temperature data (*e.g.,* from an in-stream temperature logger) to predict when developmental phenology will occur using the `predict_phenology()` function (Box 2).
+Alternatively, if users had known or observed hatch or emergence dates, they could use `predict_spawn()` to backcalculate spawn timing using the appropriate corresponding developmental models.
+Finally, users can either use model output (Box 3) for custom graphics or generate simple plots using the `plot_phenology()` function which can use the output object generated from `predict_phenology()` or `predict_spawn()` (Box 5).
+
 The R package provides user-friendly example articles that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs.
@@
 While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function.
-This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond Salmonids.
+This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.
 
@@
 
-**hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time in the case of fiedl) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
+The **hatchR** software assumes input data consists of at least two required columns: a date column indicating the date (and optionally time in the case of field) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
 Other columns may be present, but column names should not include spaces.
@@
 
-**hatchR** will function with missing values, but gaps in the dataset may affect predictions.
+The **hatchR** software will function with missing values, but gaps in the dataset may affect predictions.
 Since the package does not automatically handle missing data, users must check for gaps or errors before running analyses with the provided check functions (see section *Checking Data*).
-**hatchR** supports temperatures as low as 0 °C, though such values yield extremely small effective values, potentially extending hatch or emergence timing to a year or more.
+The **hatchR** software supports temperatures as low as 0 °C, though such values yield extremely small effective values, potentially extending hatch or emergence timing to a year or more.
 Users should critically assess whether such data align with biological expectations.
@@
 
-**hatchR** is designed to analyze daily average temperatures.
+The **hatchR** software is designed to analyze daily average temperatures.
 While high-frequency data (*e.g.*, from HOBO loggers) can be used, it must be summarized into daily averages.
-**hatchR** provides built-in functionality for this summarization in R but requires pre-summarized data for use in the Shiny app.
+The **hatchR** software provides built-in functionality for this summarization in R but requires pre-summarized data for use in the Shiny app.
 
 To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically.
-These checks ensure data integrity before model application.
+These checks ensure data integrity before model application and are demonstrated in Box 1.
+Additionally, `plot_check_temp()` and `check_continuous()` are automatically run when data is loaded into the Shiny app and presented in the `Import Data` window.
 
-Here, we demonstrate the utility of three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long dataset (`year_sim`).
-This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:
-
-```{r fakedata, echo=FALSE}
-library(lubridate)
-
-# create date object for a year with 30 min reading intervals
-dates <- seq(
-  from = ymd_hms("2000-07-01 00:00:00"),
-  to = ymd_hms("2001-06-30 23:59:59"),
-  length.out = 17568
-)
-
-# create empty dataframe
-year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))
-
-# date column
-colnames(year_sim)[1] <- "date"
-
-# add dates vector to date column
-year_sim[1] <- dates
-
-# random seed
-set.seed(123)
-
-# take temps from a random normal dist with mean 10 sd 3
-# for every date time combo in dates and append to column (temp)
-# in year_sim
-year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) |> 
-  abs()
-```
-
-```{r show-fakedata}
-# simulated object (year_sim) dimensions (rows and columns)
-dim(year_sim)
-# fist 6 rows of year_sim
-head(year_sim)
-```
-
-We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 3).
-This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures.
-In this example, we set the minimum and maximum thresholds to 0 and 25 °C, respectively.
-
-```{r plot-check-1, echo = TRUE,  fig.cap = "Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
-Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 3.25}
-# visual check of temperature data
-plot_check_temp(data = year_sim,
-                dates = date,
-                temperature = temp,
-                temp_min = 0,
-                temp_max = 25)
-```
-
-In this case, no obvious outliers are present.
-However, each day contains 48 records, so the data must be summarized to daily mean temperature using `summarize_temp()`.
-After summarization, `check_continuous()` should be used to identify any missing days.
-We also suggest running `plot_check_temp()` again on the summarized data to verify its integrity, though we omit the resulting plot here for space efficiency.
-
-```{r sum-temps, echo = TRUE}
-# summarize temperature data
-year_sim_summ <- summarize_temp(
-  data = year_sim,
-  dates = date,
-  temperature = temp
-)
-
-# check dimensions of summarized data
-dim(year_sim_summ)
-
-# check if continuous
-check_continuous(
-  data = year_sim_summ,
-  dates = date
-)
-
-# we can demonstrate an error by removing Oct. 8 (100th day)
-check_continuous(
-  data = year_sim_summ[-100, ],
-  dates = date
-)
-```
-
-# **hatchR** Workflow
-
-Users can select from published model parameterizations contained in `model_table`, or generate custom model parameterizations using `fit_model()`, in both the R and Shiny deployments of **hatchR**.
-The models in `model_table` are included because their parameterizations are well-documented in the literature, though they are currently limited to Pacific Salmon and Bull Trout (see @quinn2018 pg.
-183, for additional Salmonid models).
+Users can select from published model parameterizations contained in `model_table` (Box 2), or generate custom model parameterizations using `fit_model()` (Box 4) in both the R and Shiny deployments of **hatchR**.
+The models in `model_table` are included because their parameterizations are well-documented in the literature, though they are currently limited to Pacific salmon and Bull Trout (see @quinn2018 pg.
+183, for additional salmonid models).
 To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C, instead requiring users to carefully asses whether their own parameterized models are appropriate for the temperature ranges in their datasets.
-
-To illustrate parameter selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**.
-Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
-
-First, we obtain the model expression for hatching using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
-
-```{r, echo=TRUE}
-# use model_select() to get sockeye hatch model
-sockeye_hatch_mod <- model_select(
-  author = "Beacham and Murray 1990", 
-  species = "sockeye", 
-  model = 2, 
-  development_type = "hatch"
-  )
-```
-
-The selected model expressions are then implemented using `predict_phenology()` to estimate the days to hatch (or emergence):
-
-```{r, warning=FALSE, echo=TRUE}
-# use predict_phenology() to predict sockeye hatch
-WI_hatch <- predict_phenology(
-  data = woody_island,        # temperature data
-  dates = date,               # date column
-  temperature = temp_c,       # temperature column
-  spawn.date = "1990-08-18",  # spawning date
-  model = sockeye_hatch_mod   # model expression
-  )
-```
-
-The returned object provides several outputs, including days to hatch and the full development period, allowing us to assess phenological patterns under the recorded thermal conditions:
-
-```{r, echo=TRUE}
-WI_hatch$days_to_develop; WI_hatch$dev_period
-```
-
-## Understanding your results
-
-The output from `predict_phenology()` (in this example, our named object `WI_hatch`) contains multiple elements in a list, which can be accessed using the `$` operator.
-Each component provides different insight into the predicted phenology:
-
-```{r, echo=TRUE}
-summary(WI_hatch)
-```
-
-`WI_hatch$days_to_develop` -- Returns the predicted number of days required for development.
-
-`WI_hatch$dev_period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
-
-`WI_hatch$ef_table` -- An *n* x 5 tibble (*n* = number of days to hatch or emerge), containing a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
-This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discussed below.
-
-`WI_hatch$model_specs` -- Provides details about the model used for prediction, including whether it was retrieved from `model_select()` or generated using `fit_model()`.
-Most importantly, it contains the model expression (*i.e.*, the formula) used for phenology predictions.
-
-## Plotting phenology
-
-**hatchR** includes a built in function, `plot_phenology()`, for visualizing phenology predictions (Figure 4).
-This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span.
-Be default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles.
-However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results.
-This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.
-
-```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
-plot_phenology(WI_hatch)
-```
 
@@
 
-To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which only requires species' development follows a power law relationship with temperature.
+To expand the applicability of the effective value approach beyond salmonids, **hatchR** includes a `fit_model()` function, which only requires species' development follows a power law relationship with temperature (Box 4).
 The function takes two input vectors: average incubation temperature (°C) and the number of days to a given phenological event.
@@
 However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
-Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.
-
-To demonstrate how `fit_model()` may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
-These species were selected due to their common use in aquaculture and sport fisheries.
-For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).
-
-```{r cat-stg-fit, echo=FALSE}
-# Catfish model
-catfish <- matrix(NA, 3, 2) |> data.frame()
-colnames(catfish) <- c("days", "temp_C")
-catfish$days <- c(16,21,26)
-catfish$temp_C <- c(22,10,7)
-
-cat_mod <- fit_model(temp = catfish$temp_C,
-                     days = catfish$days,
-                     species = "catfish",
-                     development_type = "hatch")
-
-# Lake Sturgeon model
-sturgeon <-  matrix(NA, 7, 2) |> data.frame()
-colnames(sturgeon) <- c("days", "CTU")
-sturgeon$days <- c(7,5,6,6,5,11,7)
-sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)
-
-sturgeon <- sturgeon |> 
-  mutate(temp_C = CTU/days) # change CTUs to average temp and add column
-
-sturgeon_mod <- fit_model(days = sturgeon$days,
-                          temp = sturgeon$temp_C,
-                          species = "sturgeon",
-                          development_type = "hatch")
-```
-
-```{r smb-fit, echo=TRUE}
-# Smallmouth Bass data (Webster 1948)
-smallmouth <- matrix(NA, 10, 2) |> data.frame()
-colnames(smallmouth) <- c("hours", "temp_F")
-smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
-smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)
-
-# change °F to °C and hours to days
-smallmouth <- smallmouth |>
-  mutate(
-    days = ceiling(hours / 24),
-    temp_C = (temp_F - 32) * (5 / 9)
-  )
-
-# estimate parameters for Smallmouth Bass using fit_model()
-smb_mod <- fit_model(
-  temp = smallmouth$temp_C,
-  days = smallmouth$days,
-  species = "smb",
-  development_type = "hatch"
-)
-```
-
-Note the *R^2^* fit from the models below. You can see they generally preform well and are close to values from model 2 of @beacham1990, which fall between 0.95 and 0.99.
-
-```{r r2-vals, echo=TRUE}
-# R^2 values from model fits
-c(smb_mod$r_squared, cat_mod$r_squared, sturgeon_mod$r_squared)
-```
-
-Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (available in `paper.Rmd`).
-Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Bass shown below):
-
-```{r make-temps, echo=FALSE, warning=FALSE, message=FALSE}
-library(dplyr)
-set.seed(123)
-
-# create random temps and corresponding dates
-temps_sim <- sort(rnorm(n = 30, mean = 16, sd = 1), decreasing = FALSE)
-dates_sim <- seq(
-  from = ymd("2000-07-01"),
-  to = ymd("2000-07-31"), length.out = 30
-)
-
-data_sim <- matrix(NA, 30, 2) |> data.frame()
-data_sim[, 1] <- temps_sim
-data_sim[, 2] <- dates_sim
-
-# change names so they aren't the same as the vector objects
-colnames(data_sim) <- c("temp_sim", "date_sim")
-```
-
-```{r smb-pred, echo=TRUE, warning=FALSE, message=FALSE}
-# Smallmouth Bass
-smb_res <- predict_phenology(
-  data = data_sim,             # simulated data
-  dates = date_sim,            # simulated dates
-  temperature = temp_sim,      # simulated temperatures
-  spawn.date = "2000-07-01",   # spawn date
-  model = smb_mod$expression   # model expression from fit_model()
-)
-```
-
-Finally, we summarize the outputs across species to visualize hatch timing and total developmental duration (Figure 5).
-This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.
-
-```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.", fig.height=5}
-# Catfish
-catfish_res <- predict_phenology(
-  data = data_sim,
-  dates = date_sim,
-  temperature = temp_sim,
-  spawn.date = "2000-07-01",
-  model = cat_mod$expression
-)
-
-# Sturgeon
-# note that 16 C is pretty far out of range of temps for model fit, not best practice
-sturgeon_res <- predict_phenology(
-  data = data_sim,
-  dates = date_sim,
-  temperature = temp_sim,
-  spawn.date = "2000-07-01",
-  model = sturgeon_mod$expression
-)
-
-# Summary for all species
-all_res <- data.frame(matrix(NA, 3, 2))
-colnames(all_res) <- c("start", "stop")
-
-all_res$start <- c(
-  catfish_res$dev_period$start,
-  smb_res$dev_period$start,
-  sturgeon_res$dev_period$start
-)
-
-all_res$stop <- c(
-  catfish_res$dev_period$stop,
-  smb_res$dev_period$stop,
-  sturgeon_res$dev_period$stop
-)
-
-all_res <- all_res |>
-  mutate(
-    days = ceiling(stop - start),
-    index = c(17, 16.5, 16)
-  )
-
-all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")
-
-ggplot() +
-  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
-  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim)) +
-  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax = index - .35, ymin = index - .5, fill = Species)) +
-  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index - 0.425), label = days)) +
-  labs(x = "Date", y = "Temperature (°C)") +
-  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
-  theme_classic() +
-  theme(legend.position = c(0.75, 0.25))
-```
-
-\newpage
+Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.
 
@@
 For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus.
-The user then may chooses multiple spawn dates using an interactive calendar.
+The user then may chooses a single or multiple spawn dates using an interactive calendar.
 Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.
@@
 
-**hatchR** is a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes.
+The **hatchR** software ecosystem bridges the analytical gap in predicting developmental phenology for wild fishes.
 It establishes a formal framework for applying effective value models from user-provided parameterizations.
@@
 
-Both versions allow users to import data, perform basic data checks, perform basic visualizations, and apply either pre-existing Salmonid model parameterizations or generate custom models specific to other species or populations.
+Both versions allow users to import data, perform basic data checks, perform basic visualizations, and apply either pre-existing salmonid model parameterizations or generate custom models specific to other species or populations.
 To support users at various levels of expertise, we provide extensive documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.
@@
 We encourage users to predict phenology using early, peak, and late thresholds (*e.g.*, 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.
+We outline an approach for predicting phenology from a modeled distribution in our "Advanced plotting" article on the **hatchR** website (<https://bmait101.github.io/hatchR/articles/Advanced_plotting.html>).
 
@@
 
+# Box 1: Checking and summarizing your data
+
+To demonstrate the process of checking your data we show the workflow for the functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset (`year_sim`).
+This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:
+
+```{r fakedata, echo=FALSE}
+library(lubridate)
+
+# create date object for a year with 30 min reading intervals
+dates <- seq(
+  from = ymd_hms("2000-07-01 00:00:00"),
+  to = ymd_hms("2001-06-30 23:59:59"),
+  length.out = 17568
+)
+
+# create empty dataframe
+year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))
+
+# date column
+colnames(year_sim)[1] <- "date"
+
+# add dates vector to date column
+year_sim[1] <- dates
+
+# random seed
+set.seed(123)
+
+# take temps from a random normal dist with mean 10 sd 3
+# for every date time combo in dates and append to column (temp)
+# in year_sim
+year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) |> 
+  abs()
+```
+
+```{r show-fakedata}
+# simulated object (year_sim) dimensions (rows and columns)
+dim(year_sim)
+# fist 6 rows of year_sim
+head(year_sim)
+```
+
+We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values.
+This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures.
+In this example, we set the minimum and maximum thresholds to 0 and 25 °C, respectively.
+
+```{r plot-check-1, echo = TRUE,  fig.cap = "Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 3.25}
+# visual check of temperature data
+plot_check_temp(data = year_sim,
+                dates = date,
+                temperature = temp,
+                temp_min = 0,
+                temp_max = 25)
+```
+
+In this case, no obvious outliers are present.
+However, each day contains 48 records, so the data must be summarized to daily mean temperature using `summarize_temp()`.
+After summarization, `check_continuous()` should be used to identify any missing days.
+We also suggest running `plot_check_temp()` again on the summarized data to verify its integrity, though we omit the resulting plot here for space efficiency.
+
+```{r sum-temps, echo = TRUE}
+# summarize temperature data
+year_sim_summ <- summarize_temp(
+  data = year_sim,
+  dates = date,
+  temperature = temp
+)
+
+# check dimensions of summarized data
+dim(year_sim_summ)
+
+# check if continuous
+check_continuous(
+  data = year_sim_summ,
+  dates = date
+)
+
+# we can demonstrate an error by removing Oct. 8 (100th day)
+check_continuous(
+  data = year_sim_summ[-100, ],
+  dates = date
+)
+```
+
+# Box 2: Model selection and prediction
+
+To illustrate parameter selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**.
+Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
+
+First, we obtain the model expression for hatching using `model_select()`, which retrieves the appropriate parameterizations from `model_table`:
+
+```{r, echo=TRUE}
+# use model_select() to get sockeye hatch model
+sockeye_hatch_mod <- model_select(
+  author = "Beacham and Murray 1990", 
+  species = "sockeye", 
+  model = 2, 
+  development_type = "hatch"
+  )
+```
+
+The selected model expressions are then implemented using `predict_phenology()` to estimate the days to hatch (or emergence):
+
+```{r, warning=FALSE, echo=TRUE}
+# use predict_phenology() to predict sockeye hatch
+WI_hatch <- predict_phenology(
+  data = woody_island,        # temperature data
+  dates = date,               # date column
+  temperature = temp_c,       # temperature column
+  spawn.date = "1990-08-18",  # spawning date
+  model = sockeye_hatch_mod   # model expression
+  )
+```
+
+The returned object provides several outputs, including days to hatch and the full development period, allowing us to assess phenological patterns under the recorded thermal conditions:
+
+```{r, echo=TRUE}
+WI_hatch$days_to_develop; WI_hatch$dev_period
+```
+
+# Box 3: Understanding model output
+
+The output from `predict_phenology()` (in this example, our named object `WI_hatch` generated in Box 2) contains multiple elements in a list, which can be accessed using the `$` operator.
+Each component provides different insight into the predicted phenology:
+
+```{r, echo=TRUE}
+summary(WI_hatch)
+```
+
+`WI_hatch$days_to_develop` -- Returns the predicted number of days required for development.
+
+`WI_hatch$dev_period` -- A 1x2 dataframe containing the spawning date (as input via `predict_phenology(spawn.date = ...)`) and predicted development completion date.
+
+`WI_hatch$ef_table` -- An *n* x 5 tibble (*n* = number of days to hatch or emerge), containing a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values.
+This table serves as a foundation for users to create custom visualizations beyond the built-in functionality discussed below.
+
+`WI_hatch$model_specs` -- Provides details about the model used for prediction, including whether it was retrieved from `model_select()` or generated using `fit_model()`.
+Most importantly, it contains the model expression (*i.e.*, the formula) used for phenology predictions.
+
+# Box 4: Generating custom models
+
+To demonstrate how `fit_model()` may be used to create custom parameterizations for species beyond the salmonids included in `model_table`, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005].
+These species were selected due to their common use in aquaculture and sport fisheries.
+For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).
+
+```{r cat-stg-fit, echo=FALSE}
+# Catfish model
+catfish <- matrix(NA, 3, 2) |> data.frame()
+colnames(catfish) <- c("days", "temp_C")
+catfish$days <- c(16,21,26)
+catfish$temp_C <- c(22,10,7)
+
+cat_mod <- fit_model(temp = catfish$temp_C,
+                     days = catfish$days,
+                     species = "catfish",
+                     development_type = "hatch")
+
+# Lake Sturgeon model
+sturgeon <-  matrix(NA, 7, 2) |> data.frame()
+colnames(sturgeon) <- c("days", "CTU")
+sturgeon$days <- c(7,5,6,6,5,11,7)
+sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)
+
+sturgeon <- sturgeon |> 
+  mutate(temp_C = CTU/days) # change CTUs to average temp and add column
+
+sturgeon_mod <- fit_model(days = sturgeon$days,
+                          temp = sturgeon$temp_C,
+                          species = "sturgeon",
+                          development_type = "hatch")
+```
+
+```{r smb-fit, echo=TRUE}
+# Smallmouth Bass data (Webster 1948)
+smallmouth <- matrix(NA, 10, 2) |> data.frame()
+colnames(smallmouth) <- c("hours", "temp_F")
+smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
+smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)
+
+# change °F to °C and hours to days
+smallmouth <- smallmouth |>
+  mutate(
+    days = ceiling(hours / 24),
+    temp_C = (temp_F - 32) * (5 / 9)
+  )
+
+# estimate parameters for Smallmouth Bass using fit_model()
+smb_mod <- fit_model(
+  temp = smallmouth$temp_C,
+  days = smallmouth$days,
+  species = "smb",
+  development_type = "hatch"
+)
+```
+
+Note the *R^2^* fit from the models below. You can see they generally preform well and are close to values from model 2 of @beacham1990, which fall between 0.95 and 0.99.
+
+```{r r2-vals, echo=TRUE}
+# R^2 values from model fits
+c(smb_mod$r_squared, cat_mod$r_squared, sturgeon_mod$r_squared)
+```
+
+Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (available in `paper.Rmd`).
+Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Bass shown below):
+
+```{r make-temps, echo=FALSE, warning=FALSE, message=FALSE}
+library(dplyr)
+set.seed(123)
+
+# create random temps and corresponding dates
+temps_sim <- sort(rnorm(n = 30, mean = 16, sd = 1), decreasing = FALSE)
+dates_sim <- seq(
+  from = ymd("2000-07-01"),
+  to = ymd("2000-07-31"), length.out = 30
+)
+
+data_sim <- matrix(NA, 30, 2) |> data.frame()
+data_sim[, 1] <- temps_sim
+data_sim[, 2] <- dates_sim
+
+# change names so they aren't the same as the vector objects
+colnames(data_sim) <- c("temp_sim", "date_sim")
+```
+
+```{r smb-pred, echo=TRUE, warning=FALSE, message=FALSE}
+# Smallmouth Bass
+smb_res <- predict_phenology(
+  data = data_sim,             # simulated data
+  dates = date_sim,            # simulated dates
+  temperature = temp_sim,      # simulated temperatures
+  spawn.date = "2000-07-01",   # spawn date
+  model = smb_mod$expression   # model expression from fit_model()
+)
+```
+
+Finally, we summarize the outputs across species to visualize hatch timing and total developmental duration.
+This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of salmonid systems.
+
+```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.", fig.height=5}
+# Catfish
+catfish_res <- predict_phenology(
+  data = data_sim,
+  dates = date_sim,
+  temperature = temp_sim,
+  spawn.date = "2000-07-01",
+  model = cat_mod$expression
+)
+
+# Sturgeon
+# note that 16 C is pretty far out of range of temps for model fit, not best practice
+sturgeon_res <- predict_phenology(
+  data = data_sim,
+  dates = date_sim,
+  temperature = temp_sim,
+  spawn.date = "2000-07-01",
+  model = sturgeon_mod$expression
+)
+
+# Summary for all species
+all_res <- data.frame(matrix(NA, 3, 2))
+colnames(all_res) <- c("start", "stop")
+
+all_res$start <- c(
+  catfish_res$dev_period$start,
+  smb_res$dev_period$start,
+  sturgeon_res$dev_period$start
+)
+
+all_res$stop <- c(
+  catfish_res$dev_period$stop,
+  smb_res$dev_period$stop,
+  sturgeon_res$dev_period$stop
+)
+
+all_res <- all_res |>
+  mutate(
+    days = ceiling(stop - start),
+    index = c(17, 16.5, 16)
+  )
+
+all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")
+
+ggplot() +
+  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
+  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim)) +
+  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax = index - .35, ymin = index - .5, fill = Species)) +
+  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index - 0.425), label = days)) +
+  labs(x = "Date", y = "Temperature (°C)") +
+  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
+  theme_classic() +
+  theme(legend.position = c(0.75, 0.25))
+```
+
+# Box 5: Plotting phenology
+
+The **hatchR software** includes a built in function, `plot_phenology()`, for visualizing phenology predictions.
+This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span.
+By default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles.
+However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results.
+This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.
+
+```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
+plot_phenology(WI_hatch)
+
+```
+
 # Acknowledgements

### 2025-06-18 inst/manuscript/paper.Rmd
output figures to separat file,  not inline
add figures and tables at end of manuscript, no longer inline

@@
 
+## Workflow
+
 The **hatchR** softeware is designed primarily as a tool for predicting early life history phenology in wild fishes.
@@
 
-```{r, echo = FALSE, out.height= "72%", fig.align = 'center', fig.cap="hatchR workflow.
-Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decision choices as italicized text."}
+```{r, echo = FALSE, eval=FALSE, out.height= "72%", fig.align = 'center', fig.cap=""}
 knitr::include_graphics("./workflow.png")
@@
 
-```{r, echo = FALSE, fig.height=3, fig.align = 'center', fig.cap = "Effective value model fit and application for European Grayling.
-Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line).
-Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C.
-The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference."}
+```{r, echo = FALSE, fig.height=3, fig.align = 'center', fig.cap = ""}
 grayling <- matrix(NA, 5, 2) |> tibble::tibble()
@@
 
-p1 + p2 + plot_annotation(tag_levels = "A")
+figure_2 <- p1 + p2 + plot_annotation(tag_levels = "A")
+ggsave("./manuscript_figures/Figure_2.pdf", figure_2, width = 8, height = 4, dpi = 1000)
 ```
@@
 Data should follow the format outlined in Table 1.
-
-```{r kabletable, echo=FALSE}
-df <- data.frame(
-  date = c(
-    "2000-01-01",
-    "...",
-    "2000-07-01",
-    "...",
-    "2000-12-31"
-  ),
-  temperature = c(
-    2.51,
-    "...",
-    16.32,
-    "...",
-    3.13
-  )
-)
-
-knitr::kable(df,
-  booktabs = TRUE,
-  caption = "Example temperature data for use in hatchR."
-)
-```
 
@@
 
-```{r, echo=TRUE}
+```{r, echo=FALSE}
 # apply predict_phenology() over multiple sites and spawn dates
@@
 
-```{r plot-bt, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
+```{r plot-bt, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap=""}
 cut_ints <- date(
@@
 
-ggplot(
+figure_3 <- ggplot(
   data = hatch_factors,
@@
   theme(legend.position = "none")
+
+ggsave("./manuscript_figures/Figure_3.pdf", figure_3, width = 3.5, height = 5, dpi = 1000)
 ```
@@
 
-```{r plot-check-1, echo = TRUE,  fig.cap = "Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
-Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 3.25}
+```{r plot-check-1, echo = TRUE, eval=FALSE, fig.cap = "", fig.height = 3.25}
 # visual check of temperature data
@@
                 temp_max = 25)
+
+```
+
+```{r, echo=FALSE}
+box_1 <- plot_check_temp(data = year_sim,
+                dates = date,
+                temperature = temp,
+                temp_min = 0,
+                temp_max = 25)
+
+ggsave("./manuscript_figures/Box_1.pdf", box_1, width = 6, height = 3, dpi = 1000)
 ```
@@
 
-```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.", fig.height=5}
+```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="", fig.height=5}
 # Catfish
@@
 
-ggplot() +
+box_4 <- ggplot() +
   geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
@@
   theme(legend.position = c(0.75, 0.25))
+
+ggsave("./manuscript_figures/Box_4.pdf", box_4, width = 4, height = 4, dpi = 1000)
 ```
@@
 
-The **hatchR software** includes a built in function, `plot_phenology()`, for visualizing phenology predictions.
+The **hatchR software** includes a built in function, `plot_phenology()`, for visualizing phenology predictions from either `predict_phenology()` or `predict_spawn()`.
 This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span.
@@
 
-```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
+```{r, eval = FALSE, fig.cap=""}
 plot_phenology(WI_hatch)
 
+```
+
+```{r, echo=FALSE}
+box_5 <- plot_phenology(WI_hatch)
+ggsave("./manuscript_figures/Box_5.pdf", box_5, width = 6, height = 6, dpi = 1000)
 ```
@@
 # References
+
+::: {#refs}
+:::
+
+\newpage
+
+# Table Captions
+
+**Table 1.** Example temperature data for use in hatchR.
+
+\newpage
+
+# Figure Captions
+
+**Figure 1.** hatchR workflow. Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decision choices as italicized text.
+
+**Figure 2.** Effective value model fit and application for European Grayling.
+Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line).
+Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C.
+The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference.
+
+**Figure 3.** Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data.
+
+**Box 1.** Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset.
+Users can set custom thresholds for minimum and maximum temperatures (dashed lines).
+
+**Box 4.** Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.
+
+**Box 5.** Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset.
+
+\newpage
+
+# Tables
+
+```{r kabletable, echo=FALSE}
+df <- data.frame(
+  date = c(
+    "2000-01-01",
+    "...",
+    "2000-07-01",
+    "...",
+    "2000-12-31"
+  ),
+  temperature = c(
+    2.51,
+    "...",
+    16.32,
+    "...",
+    3.13
+  )
+)
+
+knitr::kable(df,
+  booktabs = TRUE
+)
+```

### 2025-06-18 inst/manuscript/paper.Rmd
finsih up with figures and tables to end, render as word doc

@@
 title: ""
-output: pdf_document
+output: word_document
 header-includes:
@@
 
-Fish, as poikilothermic organisms, have developmental rates that are strongly influenced by ambient environmental conditions.
+As poikilothermic organisms, fish have developmental rates that are strongly influenced by ambient environmental conditions.
 This close relationship has allowed researchers to develop statistical models that predict developmental phenology with high accuracy.
@@
 To predict hatch timing across these sites, we first set up the necessary models and data, using the built-in Bull Trout parameterization (for concision, we omit this setup here, but full details are available in the `paper.Rmd` file in the GitHub repository, or in the *Predict fish phenology: Nested* article on **hatchR**'s website).
-
-We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
+We then map `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
 
@@
 By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
-The results of this analysis are presented in Figure 6, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range and across years.
+The results of this analysis are presented in Figure 3, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range and across years.
 This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
@@
 
+\newpage
+
 # Box 1: Checking and summarizing your data
@@
 
+\newpage
+
 # Box 2: Model selection and prediction
@@
 
+\newpage
+
 # Box 3: Understanding model output
@@
 Most importantly, it contains the model expression (*i.e.*, the formula) used for phenology predictions.
+
+\newpage
 
@@
 
+\newpage
+
 # Box 5: Plotting phenology
@@
 ```
+
+\newpage
 

### 2025-06-18 inst/manuscript/paper.Rmd
copy edits and rerender as pdf and word docs

@@
 The effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations are readily available.
-For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
-Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial, recreational, and cultural importance of these species.
+For example, Pacific salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
+Despite its growing adoption, applications of the effective value model remain largely confined to salmonids, likely due to the availability of existing parameterizations and the commercial, recreational, and cultural importance of these species.
 
@@
 
-The **hatchR** softeware is designed primarily as a tool for predicting early life history phenology in wild fishes.
+The **hatchR** software is designed primarily as a tool for predicting early life history phenology in wild fishes.
 To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data.
@@
 
-The R package provides user-friendly example articles that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs.
+The software website provides user-friendly example articles that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow, as well as customization for plotting model outputs.
 In contrast, the Shiny application includes a default output plot and an option to download results for external visualizations.
 For a high-level overview of **hatchR**'s applications, see Figure 1.
-Additional details on key functions and workflows, particularly for automating phenology predictions across multiple variables, are available in articles hosted on the software's webpage.
+Additional details on key functions and workflows, particularly for automating phenology predictions across multiple variables, are available on the software's webpage.
 
@@
 
-Effective value models were introduced by @sparks2019 to predict developmental timing in wild populations, initially for Sockeye Salmon (*O.
+Effective value models were introduced by @sparks2019 to predict developmental timing in wild fish populations, initially for Sockeye Salmon (*O.
 nerka*).
@@
 where $E_i$ is the effective value and $T_i$ the temperature for day $i$, and $a$ and $b$ are model parameterization estimates (i.e.
-species- or model-specific constants).
+species/population- or model-specific constants).
 A fish hatches or emerges when the cumulative sum of effective values reaches one:
@@
 The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).
-While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function.
+While `model_table` incorporates more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function.
 This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.
@@
 
-The **hatchR** software assumes input data consists of at least two required columns: a date column indicating the date (and optionally time in the case of field) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
+The **hatchR** software assumes input data consists of at least two required columns: a date column indicating the date (and optionally time, as is commonly the case for field loggers) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C).
 Other columns may be present, but column names should not include spaces.
@@
 A model is then fit to the data using `stats::nls()`, which performs nonlinear least squares regression to estimate the parameters *log~e~a* and *b*.
-Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to
+Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to fit the nonlinear model.
 
@@
 However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
+If users design experiments for custom model fits, we recommend choosing a minumum of four treatments and choosing the coolest and warmest temperature treatments to match near thermal thresholds for the species or population for best fits.
 Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.
@@
 Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year).
-For this example, we use the `crooked_river` dataset, which is included in **hatchR**.
+For this example, we use the `crooked_river` dataset, which is included in **hatchR**, which can be automatically loaded in the Shiny app's `Import Data` window.
 
-Once uploaded, **hatchR** automatically generates a visual data check using `plot_check_temp()`.
+Once uploaded, **hatchR** automatically generates a visual data check using `plot_check_temp()` and a continuous data check with `check_continuous()`.
 After confirming data integrity, users navigate to the `Model Phenology` window.
@@
 For the second case study, we demonstrate a more complex, large-scale application of **hatchR**, highlighting its full flexibility when applied programmatically in R.
-This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the greater Snake River watershed in Idaho, USA.
+This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the greater Snake River basin in Idaho, USA.
 
@@
 The filtering process reduced the dataset to 139 potential spawning sites.
+For concision, we omit the example workflow, but full details are available in the `paper.Rmd` file in the GitHub repository, or in the *Predict fish phenology: Nested* article on **hatchR**'s website.
 
@@
 
-To predict hatch timing across these sites, we first set up the necessary models and data, using the built-in Bull Trout parameterization (for concision, we omit this setup here, but full details are available in the `paper.Rmd` file in the GitHub repository, or in the *Predict fish phenology: Nested* article on **hatchR**'s website).
+To predict hatch timing across these sites, we first set up the necessary models and data, using the built-in Bull Trout parameterization.
 We then map `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
@@
 We encourage users to predict phenology using early, peak, and late thresholds (*e.g.*, 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.
-We outline an approach for predicting phenology from a modeled distribution in our "Advanced plotting" article on the **hatchR** website (<https://bmait101.github.io/hatchR/articles/Advanced_plotting.html>).
+We outline an approach for predicting phenology from a modeled distribution in our *Advanced plotting* article on the **hatchR** website (<https://bmait101.github.io/hatchR/articles/Advanced_plotting.html>).
 
@@
 
-Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available.
+Importantly, **hatchR** extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available.
 We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website.
@@
 We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values.
-This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures.
+This function plots the temperature data against date and time, allowing users to set custom thresholds for minimum and maximum temperatures.
 In this example, we set the minimum and maximum thresholds to 0 and 25 °C, respectively.
@@
 
-```{r sum-temps, echo = TRUE}
+```{r sum-temps, echo = TRUE, warning=TRUE}
 # summarize temperature data
@@
 
-# Box 2: Model selection and prediction
+# Box 2: Model selection from existing parameterizations and prediction
 
 To illustrate parameter selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**.
-Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.
+Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island, Alaska, USA in 1990.
 
@@
 
-The output from `predict_phenology()` (in this example, our named object `WI_hatch` generated in Box 2) contains multiple elements in a list, which can be accessed using the `$` operator.
+The output from `predict_phenology()` and `predict_spawn()` (in this example, our named object `WI_hatch` generated in Box 2) contains multiple elements in a list, which can be accessed using the `$` operator.
 Each component provides different insight into the predicted phenology:
@@
 These species were selected due to their common use in aquaculture and sport fisheries.
-For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).
+For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR> and the *Parameterize hatchR Models* article on the software's website.
 
@@
 
-Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (available in `paper.Rmd`).
+Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C and standard deviation 1 °C (available in `paper.Rmd`, chosen to represent a reasonable putative temperature regime).
 Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Bass shown below):
@@
 Matt Kaylor provided useful comments on an earlier trial run of the package and Shiny app.
+Two anonymous reviewers and Michael Donaldson added comments that greatly improved the manuscript.
 

### 2025-06-24 inst/manuscript/paper.Rmd
copy edit and rerender

@@
 title: ""
-output: word_document
+output: pdf_document
 header-includes:
@@
 The Rmardkdown document with all the code to reproduce the examples from this manuscript is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
-The The latest version will be archived upon acceptance of the manuscript.
+The latest version will be archived upon acceptance of the manuscript.
 
@@
 
-**Box 5.** Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset.
+**Box 5.** Output of `plot_phenology()` function using predicted hatch time from `woody_island` dataset.
 

### 2025-06-25 inst/manuscript/paper.Rmd
Add i.e. to parenthetical
Remove Table 1

@@
 Historically, these models were created in aquaculture settings under the assumption of constant temperatures throughout development [@beacham1990; @mcphail1979; @alderdice1978].
-As a result, accumulated thermal unit thresholds (ATU models) have become widely used to predict development in wild populations.
+As a result, accumulated thermal unit thresholds (*i.e.*, ATU models) have become widely used to predict development in wild populations.
 However, this approach can be inaccurate, particularly at the cold and warm extremes of temperature ranges, due to the non-linear relationship between average incubation temperature and the thermal units required for development [@quinn2018].
@@
 Other columns may be present, but column names should not include spaces.
-Data should follow the format outlined in Table 1.
 
@@
 
-# Table Captions
-
-**Table 1.** Example temperature data for use in hatchR.
-
-\newpage
-
 # Figure Captions

### 2025-06-25 inst/manuscript/paper.Rmd
Typo
Remove the Table call

@@
 **Figure 2.** Effective value model fit and application for European Grayling.
-Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line).
+Panel A shows developmental rates (days to hatch) at five different average rearing temperatures and the corresponding model fit (blue line).
 Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C.
@@
 **Box 4.** Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.
+Note:
 
 **Box 5.** Output of `plot_phenology()` function using predicted hatch time from `woody_island` dataset.
-
-\newpage
-
-# Tables
-
-```{r kabletable, echo=FALSE}
-df <- data.frame(
-  date = c(
-    "2000-01-01",
-    "...",
-    "2000-07-01",
-    "...",
-    "2000-12-31"
-  ),
-  temperature = c(
-    2.51,
-    "...",
-    16.32,
-    "...",
-    3.13
-  )
-)
-
-knitr::kable(df,
-  booktabs = TRUE
-)
-```

### 2025-06-25 inst/manuscript/paper.Rmd
Box 4 clarification

@@
 
-Next, we generated a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C and standard deviation 1 °C (available in `paper.Rmd`, chosen to represent a reasonable putative temperature regime).
-Using this dataset, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Bass shown below):
+Next, to emulate a wild environment, we simulated a thermal regime featuring an ascending thermograph with a mean temperature of 16 °C and standard deviation 1 °C (available in `paper.Rmd`, chosen to represent a reasonable putative temperature regime).
+To highlight how different species develop at different rates under identical thermal conditions, we apply the custom models for each species using `predict_phenology()` (only Smallmouth Bass shown below):
 

### 2025-06-25 inst/manuscript/paper.Rmd

@@
 The software is designed for both applied and fundamental research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app.
-We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.
+We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs in the face of interacting stressors in aquatic ecosystems [@craig2017].
 

### 2025-06-26 inst/manuscript/paper.Rmd
added usfws to acknowledgements and note about reviewers

@@
 Matt Kaylor provided useful comments on an earlier trial run of the package and Shiny app.
-Two anonymous reviewers and Michael Donaldson added comments that greatly improved the manuscript.
+Three anonymous reviewers added thoghtful comments that greatly improved the manuscript.
 
-The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS.
+The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS and USFWS.
 Any mention of trade names, products, or services does not imply an endorsement by the U.S.
-government or USFS.
-USFS does not endorse any commercial products, services or enterprises.
+government or USFS or USFWS.
+USFS and USFWS does not endorse any commercial products, services or enterprises.
 

### 2025-06-26 inst/manuscript/paper.Rmd
change eli's affiliations
rerender

@@
 
-^2^Clearwater Solutions, Lenore, ID, USA
+^2^Idaho Fish and Wildlife Conservation Office, U.S. Fish and Wildlife Service, Orofino, ID, USA
 
@@
 **Box 4.** Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.
-Note:
 

### 2025-06-30 inst/manuscript/paper.Rmd
Fix typo and add culvert/sediment ref

@@
 However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids.
-Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and wtaer temperature data.
+Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and water temperature data.
 **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
@@
 
-A common management application of developmental phenology is assessing whether fish will be free-moving before a scheduled management action, such as stream section access for grazing or road work.
-For instance, will fish have emerged from redds before construction begins, reducing potential disturbance?
+A common management application of developmental phenology is assessing whether fish will be free-moving before a scheduled management action, such as stream section access for grazing, or road work for culvert replacements.
+Such infrasturcture and activities can cause sedimentation and alter fish habitat and community structure [@maitland2016].
+Will fish have emerged from redds before construction begins, reducing potential disturbance?
 This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.

### 2025-07-02 inst/manuscript/paper.Rmd
final copy edits and rerender before revision submission

@@
 To illustrate its utility, we present two case studies demonstrating its application in research and management.
-By expanding access to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.
+By expanding access to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation and management efforts across diverse species.
 
@@
 The effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations are readily available.
-For example, Pacific salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
+For example, Pacific salmon (*Oncorhynchus* *spp.*) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019.
 Despite its growing adoption, applications of the effective value model remain largely confined to salmonids, likely due to the availability of existing parameterizations and the commercial, recreational, and cultural importance of these species.
@@
 This approach allows users to generate models tailored to any species of fish, or even different populations within species, provided they have experimental or field data linking development to temperature.
-However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses).
-If users design experiments for custom model fits, we recommend choosing a minumum of four treatments and choosing the coolest and warmest temperature treatments to match near thermal thresholds for the species or population for best fits.
+However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and population-specific variation (genetic differences among populations may affect developmental responses).
+If users design experiments for custom model fits, we recommend choosing a minimum of four treatments and choosing the coolest and warmest temperature treatments to match near thermal thresholds for the species or population for best fits.
 Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.
@@
 
-A common management application of developmental phenology is assessing whether fish will be free-moving before a scheduled management action, such as stream section access for grazing, or road work for culvert replacements.
-Such infrasturcture and activities can cause sedimentation and alter fish habitat and community structure [@maitland2016].
-Will fish have emerged from redds before construction begins, reducing potential disturbance?
+A common management application of developmental phenology is assessing whether fish will be free-moving before a scheduled management action, such as stream section access for grazing or road work for culvert replacements.
+Such infrastructure and activities can cause sedimentation and alter fish habitat and community structure [@maitland2016] and mangers need to understand if fish will have emerged from redds before construction begins to minimize impact.
 This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.
@@
 Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensitive to sediment disturbance.
-The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st.
-In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.
+The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st, when work on the project will begin.
+In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date as September 30th.
 
@@
 We use the `idaho` dataset from @isaak2018, which includes four years of temperature data (2011-2014) for these sites.
-To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature at or less than 13 °C, a known thermal threshold for Bull Trout spawning suitability.
+To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature at or less than 13 °C, a known thermal threshold for Bull Trout habitat suitability.
 The filtering process reduced the dataset to 139 potential spawning sites.
@@
 These species were selected due to their common use in aquaculture and sport fisheries.
-For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR> and the *Parameterize hatchR Models* article on the software's website.
+For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>) and the *Parameterize hatchR Models* article on the software's website.
 
@@
 smallmouth <- smallmouth |>
-  mutate(
-    days = ceiling(hours / 24),
+  mutate(days = ceiling(hours / 24),
          temp_C = (temp_F - 32) * (5 / 9)

### 2025-07-02 inst/manuscript/paper.Rmd
final rerender for submission of revisions

@@
 Matt Kaylor provided useful comments on an earlier trial run of the package and Shiny app.
-Three anonymous reviewers added thoghtful comments that greatly improved the manuscript.
+Three anonymous reviewers added thoughtful comments that greatly improved the manuscript.
 

### 2025-07-28 inst/manuscript/paper.Rmd
manuscript proofs edits

@@
 
-^3^College of Forestry, Oregon State University, Corvallis, OR, USA
+^3^Forest Ecosystems and Society, Oregon State University, Corvallis, OR, USA
 
@@
   labs(x = "Developmental average temperature (°C)", y = "Days to hatch") +
-  lims(x = c(0,15)) +
+  lims(x = c(0,15), y = c(0,115)) +
   theme_classic()
@@
   labs(x = "Daily average temperature (°C)", y = "Effective value") +
-  lims(x = c(0,15)) +
+  lims(x = c(0,15), y = c(0,.17)) +
   theme_classic()
@@
 To predict hatch timing across these sites, we first set up the necessary models and data, using the built-in Bull Trout parameterization.
-We then map `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
+We then iterate `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates each year (Early: September 1st, Peak: September 15th, Late: September 30th).
 
@@
 
-By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
+By iterating predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing varies across different spawning habitats.
 The results of this analysis are presented in Figure 3, providing insights into how hatch timing might vary under different thermal regimes and across the species' geographic range and across years.
 This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
+
+-   <div>
 
@@
     ) +
-  geom_density_ridges(alpha = 0.9) +
+      geom_density_ridges(alpha = 0.9, scale = .99,
+                           jittered_points = TRUE,
+        position = position_points_jitter(width = 0.05, height = 0),
+        point_shape = '|', point_size = 1.5, point_alpha = 1, alpha = 0.7,
+      ) +
       facet_wrap(~year, ncol = 1) +
+      expand_limits(y = c(1, 4)) +
       scale_fill_brewer(palette = "Blues", direction = 1) +
@@
     ```
+
+    </div>
 
@@
 The Rmardkdown document with all the code to reproduce the examples from this manuscript is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.
-The latest version will be archived upon acceptance of the manuscript.
+The software and manuscript versions associated with the accepted manuscript are archived on the Releases page of the software GitHub (<https://github.com/bmait101/hatchR/releases>).
 

### 2025-07-28 README.Rmd
Fix URL

@@
 
-**hatchR** is an R package for predicting hatch and emergence timing for a wide range of wild fishes using the effective value framework (Sparks et al., (2019) <DOI:10.1139/cjfas-2017-0468>).
+**hatchR** is an R package for predicting hatch and emergence timing for a wide range of wild fishes using the effective value framework (Sparks et al., 2019, *Canadian Journal of Fisheries and Aquatic Sciences*, 76(1): 123-135).
 **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets.
@@
 **hatchR** is designed to be user-friendly and accessible to a wide range of users, including researchers, managers, and educators.
+
+A manuscript accompanying this package is *in press* at *Fisheries*, forthcoming this Fall.
 
@@
 
-## Interactive Shiny App <img src='https://upload.wikimedia.org/wikipedia/commons/b/bf/Shiny_hex_logo.svg' align="right" height="50" />
+## Interactive Shiny App
 

### 2025-09-18 README.Rmd
Update citation and readme for publiication

@@
 
-[![publication](https://img.shields.io/badge/Cite-Unpublished-yellow)](https://github.com/bmait101/hatchR/blob/master/inst/CITATION)
+[![publication](https://img.shields.io/badge/Cite-Fisheries_2025-green)](https://doi.org/10.1093/fshmag/vuaf078)
 [![downloads](https://cranlogs.r-pkg.org/badges/hatchR)](https://cran.r-project.org/package=hatchR)
@@
 **hatchR** is designed to be user-friendly and accessible to a wide range of users, including researchers, managers, and educators.
-
-A manuscript accompanying this package is *in press* at *Fisheries*, forthcoming this Fall.
 
